### 2017-11-16 sensors.tex
First draft

%  LaTeX support: latex@mdpi.com 
%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).

% You need to save the "mdpi.cls" and "mdpi.bst" files into the same folder as this template file.

%=================================================================
\documentclass[sensors,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}

%
%--------------------
% Class Options:
%--------------------
% journal
%----------
% Choose between the following MDPI journals:
% actuators, admsci, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, atmosphere, atoms, axioms, batteries, behavsci, beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci, buildings, carbon, cancers, catalysts, cells, challenges, chemosensors, children, chromatography, climate, coatings, computation, computers, condensedmatter, cosmetics, cryptography, crystals, data, dentistry, designs, diagnostics, diseases, diversity, econometrics, economies, education, electronics, energies, entropy, environments, epigenomes, fermentation, fibers, fishes, fluids, foods, forests, futureinternet, galaxies, games, gels, genealogy, genes, geosciences, geriatrics, healthcare, horticulturae, humanities, hydrology, informatics, information, infrastructures, inorganics, insects, instruments, ijerph, ijfs, ijms, ijgi, ijtpp, inventions, jcdd, jcm, jdb, jfb, jfmk, jimaging, jof, jintelligence, jlpea, jmse, jpm, jrfm, jsan, land, languages, laws, life, literature, lubricants, machines, magnetochemistry, marinedrugs, materials, mathematics, mca, mti, medsci, medicines, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, molbank, molecules, mps, nanomaterials, ncrna, neonatalscreening, nutrients, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, plants, polymers, processes, proteomes, publications, recycling, religions, remotesensing, resources, risks, robotics, safety, sensors, separations, sexes, sinusitis, socsci, societies, soils, sports, standards, sustainability, symmetry, systems, technologies, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vetsci, viruses, water
%---------
% article
%---------
% The default type of manuscript is article, but can be replaced by: 
% addendum, article, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, correction, conferenceproceedings, conferencereport, expressionofconcern, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimage, letter, newbookreceived, opinion, obituary, projectreport, reply, retraction, review, preprints, shortnote, supfile, technicalnote
% supfile = supplementary materials
%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g. the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figure are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother 
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2017}
\copyrightyear{2017}
\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Remark, Definition
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{EEG Waveform Analysis with applications to Brain Computer Interfaces}

% Authors, for the paper (add full first names)
\Author{Rodrigo Ramele $^{1,\dagger}$, Ana Julia Villar $^{1}$ and Juan Miguel Santos  $^{1}$*}
% Authors, for metadata in PDF
\AuthorNames{Rodrigo Ramele, Ana Julia Villar and Juan Miguel Santos}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
$^{1}$ \quad Computer Engineering Department, Instituto Tecnológico de Buenos Aires (ITBA); info@itba.edu.ar}

% Contact information of the corresponding author
\corres{Correspondence: rramele@itba.edu.ar; Tel.: +54-9-11-4193-9382}

% Current address and/or shared authorship
\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina}

% Simple summary
%\simplesumm{}

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The Electroencephalography is not just a mere clinical tool anymore.  It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time and translating or decoding brain signals, detecting information that can be used to diagnose disease or implement Human Computer Interaction devices.  The traditional automatic approach which is based heavily on using digital tools to detect the cloaked information  in the signal, outshining the research done by the EEG community which was based heavily on EEG waveforms: the structure of signal plots.  The purpose of this work is to provide a bridge by doing a survey and description of the automatic methods that has been used to detect patterns in the waveforms, and to perform a benchmarking approach to determine different characteristics of those methods aiming to detect  specific waveforms mimicking what technician has been done since the inception of this fruitful technology. }

% Keywords
\keyword{electroencephalography (EEG); ERP,VEP,waveform, signal structure}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}
%\AMS{}

% If this is an expanded version of a conference paper, please cite it here: enter the full citation of your conference paper, and add $^\S$ in the end of the title of this article.
%\conference{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data:

%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately. If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal. In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}

%\datasetlicense{license under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% For Conference Proceedings Papers: add the conference title here
%\conferencetitle{}

%\setcounter{secnumdepth}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only for the journal Gels: Please place the Experimental Section after the Conclusions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{-1} %% Remove this when starting to work on the template.

\section{Introduction}

Current society is demanding technology to once and for all provide the means to realize the utopia of social inclusion for people with disabilities.  At the same time, a worldwide aging population is also asking for means to extend active lifestyles throughout all life, including the extending years that medicine is allowing us to have.
Finally, our digital gadgets and the digital revolution have modified the way we interact with all our devices.  A new emerging digital society is consolidating: smart wearable health sensors are required to implement automatic detection of situatedness or detecting health activities, in an active or passive way.
All our communication with devices is based on movement (introduction gauger state of the art 7), but all these trends are precisely pushing this boundary beyond the limitation of our body, beyond the limitation of our own movement.  A new form of Human Machine Interaction was needed and this is how Brain Machine Interfaces were born.  Moreover, as long as we keep putting computer inside every machine (REF IOT), this will be not other thing that Brain Computer Interfaces (BCI).
In the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, computer power, wireless protocols and advanced electronics: the Electroencephalogram (EEG).

EEG sensors are wearable (Sensors EEG Review, TOward mobile EEG), non-invasive, portable and mobile, with excellent temporal resolution, and acceptable spatial resolution (EEG Patterns).  This humble diagnosis device has been transformed into  currently the best approach that we have to precisely detect, out-of-the lab, information from the Brain, directly from the source of volition (footnote), and to use that information to drive our cars, steer our drones, write our emails, and control our wheelchairs (references of everything)

The traditional approach to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph (Atlas of EEG patterns): analog and reference signals were extracted and plotted over a piece of paper, continuously. Electroencephalographers or Electroencephalography technician have decoded and detected patterns in the signals by visually inspecting the signals.  Moreover, clinically, nowadays EEG remains a visually interpreted test (ATLAS EEG)

With the development of automatic electronics, first analog electronic devices and later computarized digital processing methods, mathematically and algorithmically complex procedures were established to decode the information with incredible success.

However, the traditional and knowledgeable approach was mainly overshadow, and the waveform of the EEG was replaced by a black box approach.

The aim of this study is threefold: first  to review current literature of EEG processing techniques which are based on analysis of the waveform.  The second is to evaluate and study the methods by analyzing its performance against a pseudo-real dataset. We aim to provide a better understanding of the different approaches.  We believe that the importance of the waveform analysis methods, as described here, is that by using these kind of approaches colaboration is fostered because there is a clear description and characterization of the signal and the extensive literature which explores EEG can be reviewed from the same point of view.

Enhancing this bridge will be beneficial to both ends: connections between BCI community and other communities (ref Nijboer, BCI a Review), and at the same time an automatic approach to empower what technician regularly do in EEG diagnosis.

BCI REsearch is 71.2 percent based on EEG (recent advancent summary 2016)

We believe that studying specific componentes may lead to better understanding of the neuro component.

%Novel sensor design and smart embodiment for pervasive healthcare;
%New platforms for self-tracking or measurement of physical activity, sleep quality, emotion, gait, kinematics, kinetics and other factors related to personal well-being;
%Information processing for localization and tracking for healthcare applications
%Resource constrained network information processing for healthcare applications
%Sensor data analytics, information integration, pattern mining/recognition, behaviour profiling, data visualisation and user feedback related to personal well-being;
%
%Mobile/wearable sensor-based/remote health monitoring systems, with illustrative case studies on use of innovative information processing approaches.
%Disease focused exemplars and case studies (e.g. neurological disorders, cardiovascular diseases, diabetes and obesity, neuro-rehabilitation);

\section{Electroencephalography Waveform Analysis}

distinguishing the pertinent signal characteristics from extraneous content and representing them in a compact or menaingful form, amenable to interpretation by a human or computer (Wolpaw and Wolpaw)

Shape or waveform analysis methods are considered as nonparametric methods (in opposition to statistical or dynamical models).  They can explore the amplitude, energy, teager energy operand, or more complex like the L-Z Lempel-Ziv complexity measurement (thakor qEEG).

The Electroencephalogram is one of the most widespread used device to capture brain signals.  It was initially developed by Hans Berger and has been extensively used for decades to diagnose neural diseases and other medical conditions.

The first thing that he saw was the visual cortical alpha wave characterization, the Berger Rythm.  Its amplitude and power determine that this feature can be coherently assosiated to a cognitive phenoma (eyes closed).  We should ask ourselves if the EEG should be described years later if this particular event wasn't so evident.

The EEG signal is not stationary, highly complex.  It can be characterized as a linear stochastic process with great similarities to noise (Thakor EEG).  They are measured in microvolts.
they have many artifacts.

The most important part to consider is that of montage.  It could be bipolar or referencial.  In general EEG can be rereferenced offline.The traditional convention, somehow maintained in neuro research, downward polarity was considered negative and upward deflection for negative (Knott, 1985)  It is of utmost importance to remark that  EEG waveforms repre- sent the differential voltage between a given electrode and the recording reference. It is therefore clear that the choice of reference completely determines EEG wave- forms (Lehmann, 1987; Dien, 1998), an important method- ological consideration that all too often is still not recog- nized in the EEG literature. For example, recording with a vertex (Cz) reference would lead to small EEG deflections in the proximity of Cz due to potential synchronization of firing activities within closely spaced brain regions and vol- ume propagation of the EEG signal. Similarly, recording with mastoid or linked ears montage would lead to rather small waveforms at electrodes positioned over temporal brain regions (Pivik et al., 1993).

The automatic processing of EEG signals can be called qEEG is precisely the opposite of what we are doing here, quantite analysis (OJO)

The conventional clinical method of observing the waveform is thought to be subjective and laborius because the results depend on the technicians' experience and expertise.  This call for objective measurements pushed the adoption of qEEG methods and the need for automation (Thakor qEEG)

Some initial works on EEG explored the idea to extend human capacities analyzing EEG waveforms  (automatic detection of k complexes), (A Waveform Analyzer Applied to the Human EEG) where a feature from amplitude and frequency of its signal and its derivative in time-domain is used.  Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.  It is likeyly that the Matched Filter approach (linear) gave poor results and nonlinear methods based on neural network were being explored.It has been the traditional approach for technician, but the advent of elecrical processing, more sophisticated methods.

EEG has been used extensively on Sleep Research, by performing Polysomnographic recordings (PSG)  (reference of the german paper of Sleep Research) where scoring of EEG to determine different sleep stages were performed by visually marking waveforms or graphoelements looking for patterns based on standardized guidelines.

Visual characterization includes the identification or classification of componentes, or transient events, based on phase, monophasic or biphasic  (i.e. positive and negative component) where different components are measured, relation between them is stablished and index are created (e.g. sleep K-Complex is well characterized based on relation rates between positvie vs negative amplitude) (REFERENCE).  Relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,
sleep spindles, polysplindles, K-complexes, vertex sharp waves (VSW), and sawtooth waves (REM Sleep)

Transient phenomena allows also to record occurrence and temporal sequence (mimicking spike analysis in neuro reserach)

The traditional approach do not consider waveform even though the brain oscilations are in general nonsinusoidal (reference)

This approach is relative common in chemical analysis (i.e. chemometrics Skoog, D.; West, D.; Holler, F. Analtyical Chemistry, An Introduction; Saunders: Philadelphia, 1994.), geology (sismic analysis), and quantitive financial analysis.  EKG, or Electrocardiogram, on the other hand, has been extensively processed and analyzed by waveform methods.

Slight variations in whether slopes or other features are scored may be important to certain applications.

waveform, phase, amplitude and frequency

waveform amplitud plotted against time

characteristic shape, or waveform.
rising phase
falling phase
pronounced plateau
ripples, wiggles,

P,Q,R,S,T peaks in EKG analysis (REFERENCE)

aEEG, amplitude integrated electroencephalography or cerebral function monitoring (REFERENCE)

Artefacts: endogeneous exogeneous
Non-Stationarity: this means that this and that.  What does it mean in terms of the signal.
DC drift and trending
Basal EEG activity

Inter-subject and intra-subject variability

Non-exhaustive list of transient events

Determination of transient events, and particularly amplitude of different subcomponents, latency or even phase, has proved very importan concequences in terms of the different cognitive approach.

Rhythms in neural activity are observed across various temporal and spatial scales and are often
referred to as oscillations (see Glossary) [1]. Traditionally, neural oscillations have been
clustered into canonical frequency bands, including delta (1–4 Hz), theta (4–8 Hz), alpha (8–
12 Hz), beta (15–30 Hz), gamma (30–90 Hz), and high gamma (>50 Hz). These bands roughly
correspond to frequency ranges commonly observed in human electroencephalography (EEG)
studies. Although they have been observed for nearly a century, recent theories suggest that
these oscillations play an active role in neural communication

 Hippocampal theta oscillations, for example, are among
the best-studied rhythms in the local field potential (LFP); they have a stereotyped sawtooth
shape

Standard signal processing characterization (chemical, time series reference, sismic reference) can also be applied.

EEG is a multidimensional signal non-stationary

Amplitude
Arch
Frequency
Phase
Nonsinusoidal sinusoidal
Oscilation
Sawtooth: motor cortical beta oscillations
Sharpness
Spike-wave discharge
Transient event

inverse problem is mathematically intractable Voytek 2009

Brainwaves may be missleading because what is actually been harnessed is electrical potentials over the scalp

Reference to the good paper about how EEG actually works

(Paper de Spinelli)

CFC, Cross Frequency Coupling, sharper, arch comb or wicket shape, sawtooth, rectangular, spike-wave like, decay phase, voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace, peak and trough sharpness ratio, symmetry between rise and decay phase, slope ratio (steepness of the rise period to that of the adjacent decay period.

Central trough is sharper and more negative that the adjacent troughs.

Phase-Amplitude Coupling, Phase-Phase Coupling.

Dimensionality reduction which is truly applied when the waveform is actually formed

Oscillatory activity can also have their differente or distinctive waveforms.  Slow oscillations, which are assosiated with REM, sawtooth-shaped.
Sleep spindles can also be considered oscilations and they have a distintinc form assosiated with stage 2 dream.
Visual Cortical alpha and rolandic central mu waves arch-like structure (similar to the greek letter mu).  Slope Ratio.  Trough voltage remains contstant while peak voltage fluctuates.   Steep slopes,   Amplitude asymmetry 
ponto-geniculo-occipital
(PGO) waves

In sleep research sleep 2 stage the background of KComplex and sleep spindles is theta waves

Another approach which is very related to EEG waveform analysis is the detection of latency, amplitude and phase components in ERP.  It is known that these characteristics can be mapped to cognitive or disease situations.  We hypothesize that that including more approaches can lead to a better understanding or opening of a new field.

by application to Brain Computer Interfaces or EEG diagnosis (BCI a review, BCI Vidal 1973)

Seizures captured in their entirety typically show progression from low-voltage, high-frequency spikes to high-voltage, low-fre- quency spike-and-slow wave activity, before stopping abruptly and being replaced by background slowing or suppression (Fig. 29.9). Usual morphologic features include typical rhythmic, gen- eralized, symmetric spike-and-waves or polyspikes and waves at 2 to 3.5 Hz; atypical spike and wave with lower frequency and less symmetry; multiple spike-and-wave (repetitive complexes of two or more spikes followed by a slow wave); and high-voltage, repetitive, rhythmic, focal or generalized delta activity with inter- mixed spikes, sharp waves, or sharp components (Fig. 29.24) (45). Diagnosis is more difficult when the seizure (or SE) pre- cedes the beginning of the tracing and continues beyond its end. In such cases, rhythmic sharp features, typically faster than 1 Hz, may be seen, often with variability.

(e.g. FIRDA - Frontal Intermittent Rhythmic Delta) and posteriorly in children e.g. OIRDA - Occipital Intermittent Rhythmic Delta).

alpha dissapears when alerting by any mechanism (thinking, calculating)

Rhythmic. EEG activity consisting in waves of approximately constant frequency.
Arrhythmic. EEG activity in which no stable rhythms are present.
Dysrhythmic. Rhythms and/or patterns of EEG activity that characteristically appear in patient groups or rarely or seen in healthy subjects.

Attenuation (synonyms: suppression, depression). Reduction of amplitude of EEG activity resulting from decreased voltage. When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".

Hypersynchrony. Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range. The term implies an increase in the number of neural elements contributing to the rhythm. (Note: term is used in interpretative sense but as a descriptor of change in the EEG).
Paroxysmal. Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity. Though the term does not directly imply abnormality, much abnormal activity is paroxysmal.

Monomorphic. Distinct EEG activity appearing to be composed of one dominant activity

Polymorphic. distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.

Sinusoidal. Waves resembling sine waves. Monomorphic activity usually is sinusoidal.

Transient. An isolated wave or pattern that is distinctly different from background activity.

a) Spike: a transient with a pointed peak and a duration from 20 to under 70 msec.
b) Sharp wave: a transient with a pointed peak and duration of 70-200 msec.

generalized, focal or lateralized.

Artifact detection and averaging.

\section{Materials and Methods}

%To verify the validity of the proposed framework and method, the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by  IRCCS Fondazione Santa Lucia, was used to perform a binary classification task on the provided signals.  The algorithm was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).

In order to determine which  processing method to be considered as a  "waveform template matching" we restrict ourselves to the following criteria:

\begin{itemize}
\item The pattern can be identified and verified by visual inspection.
\item The pattern matching is performed in time-domain.
\item The analysis take account the shape of the plot of the signal.
\item They can be used as templates.
\item Transient events ?
\item Single Channel
\end{itemize}

Although the term morphology has been used to identify this approach, We specifically excluded morphological based methods
\cite{Yamaguchi2009}.  On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm. (cita a esa parte del paper).

As described in (Mixed Domain Signal Analysis) the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the pattern.

\subsection{Peak Picking}

The name of many of the EEG features reference indeed peaks like P300 or P3a P3b or N100.   This leads to a natural way to classify them visually by selecting appropiate peaks and matching their positions and amplitudes in an orderly manner.

Although of limited usability, peak picking is the first automatic method of detection and has been used to determine latency of transient events in EEG.  Straighforward in its implementation, it consist in selecting a component, particularly a simple component based on the expected location of its more prominent deflection.

Ouyang2017,Zhang2011

difference of areas (trabajo Mexicanos)

 integrated activity (area measure) (area measure)

 It was used in Farwell and Donchin work on P300 (Wolpaw and Wolpaw).

\subsection{Woody's Template Matching}

Cross Covariance with Template or Cross Correlation

Applying a FIR filter with templates

Krusienski et al 2007
Serby et al 2005

extended to wavelent analysis.

\subsection{Matching Pursuit}

Pursuit algorithms refer, in their many variants, as blind source separation techniques that assume that the EEG signal is a linear combination of different sources that comes from template dictionaries.

We include them here because they are traditional in terms for the traditional view of Morphological Analysis or Shape domain analysis, though they do not restrict to our own definition of "shape domain processing".

However in terms of

\subsection{Dynamic Time Warping}

Dynamic Time Warping is a technique

\subsection{Wrapping Method}

\subsection{Permutation Entropy}

Bond and Pompe PE method.

\subsection{ACF: Autocorrelation Function}

\subsection{Waveform Complexity}

Lempel-Ziv method L-Z complexity

\subsection{Slope Horizontal Chain Code}

contour representation based on an adapted version of the Slope Chain Code (SCC)

\subsection{Warp Averaging}

%\subsection{Morphological Signal Analysis}
%
%Several approaches and methods have been applied to decode them.  Of particular interests are morphological signal analysis 
%CITE (Paper del chabon del ITBA que detalla bien los diferentes metodos) \citep{Alvarado-Gonzalez2016,Yamaguchi2009}.
%
%Morphological or 
%
%template matching
%
%spike
%sharp wave
%spike wave
%polyspike 
%slow wave complex
%
%shape domain
%
%\subsection{P300}
%
%Uhhhh I can talk a lot about P300 \citep{Knuth2006}.
%
%P300 single trial is a golden grail of Brain computer interfaces.  It is studied and analyzed in this way, here in this other way, in this other way is here.

\subsection{EEG based on Histogram of Gradients}

\subsection{Experimental Protocol}

In order to verify each one of the approaches we will inject into a known dataset spurious p300 signal that we may be able to control.  By implementing this pseudo-real approach, we can effectively control null-signals of this evoked potentials following the same approach of other similar works ( ref   Ouyang2017, el trabajo de estudio de templates y el trabajo de p3).

At the same time adding a mixed signal with varying degrees of SNR will allow us to determine robustness of the different methods.

The second part, we will specifically find the P300 signal from a dataset.

amplitude
latency

in a single trial approach.  The template was obtained from the point to point average of a standard p300 experiment (published elsewhere).

The original signal-to-noise ratio was calculated as Hue 2010 as

\begin{itemize}
\item Peak Finding
\item Woody Template Matching
\item Permutation Entropy
\item SHCC (los mexicanos me dijeron que me dan el codigo para matlab)
\item BCI SIFT
\end{itemize}

\begin{itemize}
\item Inyectar una señal en un tren de EEG simulando el P300
\item Probar con diferentes metodos la capacidad de detectar si un segmento contiene o no el patron
\item Probar con diferentes metodos la capacidad de identificar correctamente la localizacion del patron.
\item Probar variando la señal relacion ruido usando la formula de Huo y con eso ver como cada metodo permite mantenerse.
\end{itemize}

%The central part of this method is the feature generation: the histogram of image gradients.  To do so, the first step is the transformation of the signal into a temporary binary image.
%
%The signal is first scaled and centered at zero by 
%
%\begin{equation}
%\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot ( x(t,c) - \bar{x}(t,c)  )}\right \rfloor
%\end{equation}
%
%\noindent where $\gamma$ is the image scale, $t$ is time and $ x(t,c) $ is the EEG matrix defined for each $t$ and for a particular channel $c$.
%
%Then the image is constructed by
%
%\begin{equation}
%I(z_1,z_2) = \left\{ \begin{array}{rl}
%255 & z_1 = \gamma \cdot t; z_2 = \tilde{x}(t,c) + z(c) \\
%0   & \mbox{otherwise}
%\end{array}\right.
%\label{eq:images}
%\end{equation}

BLABLABLABLA \ref{fig:sampledescriptor}.

\begin{figure}[H]
\centering
\includegraphics[width=6cm]{sampledescriptor.png}
\caption{The local patch is located around a sample point plotted on the temporary image. All the sample points are interpolated using the Bresenham algorithm. The 4x4 block grid can be seen on the patch as well as the green arrows showing the dominant gradient on each block.}
\label{fig:sampledescriptor}
\end{figure}

\subsubsection{Classification}

In order to classify we will use the same classifier for every method.  By doing this we will avoid

\subsubsection{Parameters}

BLABLALBALBALBAL

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{section:results}
sections

BLABLABLABLALBA

\begin{table}[H]
\caption{Accuracy levels obtained by a 3-fold cross validation. The values reported by the dataset publishers for Cz are reproduced here for comparison. Additionally, BCI accuracies for channel Cz can be seen as well as performance levels and their standard deviation obtained for the BPC, the best performing channel for each subject.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Participant}	& \textbf{Original Cz}	& \textbf{ACC at Cz}	& \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &   0.84 &   0.79 & Cz  &   0.79 $\pm$ 0.01 \\
2     &   0.86 &   0.81 & PO7 &   0.93 $\pm$ 0.01 \\
3     &   0.87 &   0.78 & Cz  &   0.78 $\pm$ 0.03 \\
4     &   0.86 &   0.68 & PO8 &   0.90 $\pm$ 0.01 \\
5     &   0.86 &   0.80 & PO7 &   0.93 $\pm$ 0.01 \\
6     &   0.89 &   0.96 & Cz  &   0.96 $\pm$ 0.01 \\
7     &   0.89 &   0.78 & PO7 &   0.93 $\pm$ 0.01 \\
8     &   0.92 &   0.91 & PO7 &   1.00 $\pm$ 0.00 \\
1     &   0.84 &   0.81& Fz&   0.82 $\pm$ 0.0017 \\
2     &   0.86 &   0.82& Fz&   0.85 $\pm$ 0.0011 \\
3     &   0.87 &   0.82& Cz&   0.82 $\pm$ 0.0011 \\
4     &   0.86 &   0.81& PO7&   0.82 $\pm$ 0.0012 \\
5     &   0.86 &   0.82& Fz&   0.82 $\pm$ 0.0021 \\
6     &   0.89 &   0.81& Fz&   0.82 $\pm$ 0.0016 \\
7     &   0.89 &   0.83& Fz&   0.84 $\pm$ 0.0011 \\
8     &   0.92 &   0.82& PO7&   0.84 $\pm$ 0.0020 \\
\bottomrule
\end{tabular}
\label{tab:results}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

BLABLABLABLABLALBA

This method, different from other methods which is based on the nonlinearity of the gradient of histograms which can be used to detect

is also based on how the image look like.

SNR of p300 and how to detect it

Check if you can use this to detect any kind of transient signal.

Compare if it is possible with the descriptors from one subject, discriminate the others.

Channal identification based on the metric distance between the bags

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Subsection}
%
%\subsubsection{Subsubsection}
%
%Bulleted lists look like this:
%\begin{itemize}[leftmargin=*,labelsep=4mm]
%\item	First bullet
%\item	Second bullet
%\item	Third bullet
%\end{itemize}
%
%Numbered lists can be added as follows:
%\begin{enumerate}[leftmargin=*,labelsep=3mm]
%\item	First item
%\item	Second item
%\item	Third item
%\end{enumerate}
%
%The text continues here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following are available online at www.mdpi.com/link, Figure S1: title, Table S1: title, Video S1: title.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{This project was supported by the ITBACyT-15 funding program issued by ITBA University.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{This projects is part of a the first author's PhD Thesis which is directed by Juan Miguel Santos and codirected by Ana Julia Villar.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\conflictofinterests{The authors declare no conflict of interest.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this manuscript:\\

\noindent EEG: electroencephalography\\
BCI: Brain Computer Interfaces\\
SNR: Signal to Noise Ratio\\
CNS: Central Nervous System\\
ALS: Amyotrophic Lateral Sclerosis\\
ERP: Event-Related Potential\\
P300: Positive deflection of an Event-Related Potential which occurs 300 ms after onset of stimulus\\
ITR: Information Transfer Rate\\
BTR: Bit Transfer Rate\\
SIFT: Scale Invariant Feature Transform\\
HOG: Histogram Of Gradients}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\appendixtitles{no} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
%\appendixsections{multiple} %Leave argument "multiple" if there are multiple sections. Then a counter is printed ("Appendix A?). If there is only one appendix section then change the argument to ?one? and no counter is printed (?Appendix?).
%\appendix
%\section{}
%The appendix is an optional section that can contain details and data supplemental to the main text. For example, explanations of experimental details that would disrupt the flow of the main text, but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data is shown in the main text can be added here if brief, or as Supplementary data. Mathemtaical proofs of results not central to the paper can be added as an appendix.
%
%\section{}
%All appendix sections must be cited in the main text. In the appendixes, Figures, Tables, etc. should be labeled starting with `A', e.g., Figure A1, Figure A2, etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here. 
\bibliographystyle{mdpi}

%=====================================
% References, variant A: internal bibliography
%=====================================
%\renewcommand\bibname{References}
%\begin{thebibliography}{999}
% Reference 1
%\bibitem{ref-journal}
%Lastname, F.; Author, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142-149.
% Reference 2
%\bibitem{ref-book}
%Lastname, F.F.; Author, T. The title of the cited contribution. In {\em The Book Title}; Editor, F., Meditor, A., Eds.; Publishing House: City, Country, 2007; pp. 32-58.
%\end{thebibliography}

%=====================================
% References, variant B: external bibliography
%=====================================
\bibliography{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\sampleavailability{Samples of the compounds ...... are available from the authors.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2018-04-04 sensors.tex
More robust version

@@
 % Abstract (Do not use inserted blank lines, i.e. \\) 
-\abstract{The Electroencephalography is not just a mere clinical tool anymore.
-It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time and translating or decoding brain signals, detecting information that can be used to diagnose disease or implement Human Computer Interaction devices.
-The traditional automatic approach which is based heavily on using digital tools to detect the cloaked information  in the signal, outshining the research done by the EEG community which was based heavily on EEG waveforms: the structure of signal plots.
-The purpose of this work is to provide a bridge by doing a survey and description of the automatic methods that has been used to detect patterns in the waveforms, and to perform a benchmarking approach to determine different characteristics of those methods aiming to detect  specific waveforms mimicking what technician has been done since the inception of this fruitful technology.
+\abstract{The Electroencephalography (EEG) is not just a mere clinical tool anymore.
+It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time and translating or decoding brain signals that can be used to diagnose disease or implement Human Computer Interaction devices.
+The automatic approach which is based on using digital tools to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community which was based intensively on EEG waveforms and the structure of signal plots.
+The purpose of this work is to help to start to fill this gap by doing a review and description of the automatic methods that have been used to detect patterns in the waveforms, and to perform a benchmarking analysis to determine different characteristics of those methods that aim to detect specific waveforms mimicking what physicians has been doing since the inception of this fruitful technology.
 }
@@
 % Keywords
-\keyword{electroencephalography (EEG); ERP,VEP,waveform, signal structure}
+\keyword{electroencephalography (EEG); ERP,BCI,waveform, signal structure}
 
@@
 Current society is demanding technology to once and for all provide the means to realize the utopia of social inclusion for people with disabilities.
-At the same time, a worldwide aging population is also asking for means to extend active lifestyles throughout all life, including the extending years that medicine is allowing us to have.
-Finally, our digital gadgets and the digital revolution have modified the way we interact with all our devices.
-A new emerging digital society is consolidating: smart wearable health sensors are required to implement automatic detection of situatedness or detecting health activities, in an active or passive way.
-All our communication with devices is based on movement (introduction gauger state of the art 7), but all these trends are precisely pushing this boundary beyond the limitation of our body, beyond the limitation of our own movement.
-A new form of Human Machine Interaction was needed and this is how Brain Machine Interfaces were born.
-Moreover, as long as we keep putting computer inside every machine (REF IOT), this will be not other thing that Brain Computer Interfaces (BCI).
-In the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, computer power, wireless protocols and advanced electronics: the Electroencephalogram (EEG).
+Additionally, a worldwide aging population is also claiming for solutions to extend active lifestyles throughout all life, including additional years that medicine is allowing humanity to have \citep{Lutz2008}.
+At the same time,  the digital gadgets and the digital revolution have modified the way people interact with all their devices.
+A new emerging digital society is consolidating and smart wearable sensors are starting to be used routinely to obtain bio-signals, in an active or passive way (Wolpaw Communication and control)
+Although this communication is based on muscular movement \citep{Guger2017}, all these trends are precisely pushing this boundary beyond the confines of the body and beyond the limitation of human movement.
+A new form of Human Machine communication which directly connects the Central Nervous System to a machine is currently being under development: Brain Machine Interfaces.
+Moreover, as long as computers are being used inside every machine that humanity posses \citep{Domingo2012} this discipline is specializing as Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).
+On the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG) \citep{Schomer2010}.
 
-EEG sensors are wearable (Sensors EEG Review, TOward mobile EEG), non-invasive, portable and mobile, with excellent temporal resolution, and acceptable spatial resolution (EEG Patterns).
-This humble diagnosis device has been transformed into  currently the best approach that we have to precisely detect, out-of-the lab, information from the Brain, directly from the source of volition (footnote), and to use that information to drive our cars, steer our drones, write our emails, and control our wheelchairs (references of everything)
+EEG sensors are wearable \citep{Puce2017} non-invasive, portable and mobile \citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution \citep{Hartman2005}.
+This humble diagnosis device has been transformed into currently the best approach that we posses to precisely detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive our cars, steer our drones, write our emails, and control our wheelchairs \citep{Yuste2017}.
 
-The traditional approach to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph (Atlas of EEG patterns): analog and reference signals were extracted and plotted over a piece of paper, continuously.
-Electroencephalographers or Electroencephalography technician have decoded and detected patterns in the signals by visually inspecting the signals.
-Moreover, clinically, nowadays EEG remains a visually interpreted test (ATLAS EEG)
+The clinical and historical approach to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph (Atlas of EEG patterns): multichannel signals were extracted and plotted over a piece of paper, continuously.
+Electroencephalographers or Electroencephalography technician have decoded and detected patterns along the signals by visually inspecting them \citep{Schomer2010}.
+Even nowadays clinical EEG remains a visually interpreted test \citep{Hartman2005}.
 
-With the development of automatic electronics, first analog electronic devices and later computarized digital processing methods, mathematically and algorithmically complex procedures were established to decode the information with incredible success.
+Automatic processing, or quantitave EEG, was based first on analog electronic devices and later on computerized digital processing methods \citep{Jansen1991}.
+They implemented mathematically and algorithmically complex procedures to decode the information with outstanding success \citep{Yuste2017}.
+The best materialization of the automatic processing of EEG signals lays in the BCI discipline, where $71.2\%$ is based on EEG \citep{Guger2017}.
 
-However, the traditional and knowledgeable approach was mainly overshadow, and the waveform of the EEG was replaced by a black box approach.
+%rich clinical literature
 
-The aim of this study is threefold: first  to review current literature of EEG processing techniques which are based on analysis of the waveform.
-The second is to evaluate and study the methods by analyzing its performance against a pseudo-real dataset.
-We aim to provide a better understanding of the different approaches.
-We believe that the importance of the waveform analysis methods, as described here, is that by using these kind of approaches colaboration is fostered because there is a clear description and characterization of the signal and the extensive literature which explores EEG can be reviewed from the same point of view.
+Hence, the traditional and knowledgeable approach was mainly overshadow particularly in BCI Research, and the waveform of the EEG was replaced by pragmatically sound procedures that were difficult to link to existing clinical EEG knowledge.
+We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and that can help to decode them in an automated way.
 
-Enhancing this bridge will be beneficial to both ends: connections between BCI community and other communities (ref Nijboer, BCI a Review), and at the same time an automatic approach to empower what technician regularly do in EEG diagnosis.
+The aim of this study is twofold: first to review current literature of EEG processing techniques which are based on analysis of the waveform.
+The second is to evaluate and study these methods by analyzing its classification performance against a pseudo-real dataset.
+We aim to outline a better understanding of the different approaches.
+We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective \citep{Nijboer2009,Wei2017}.
 
-BCI REsearch is 71.2 percent based on EEG (recent advancent summary 2016)
+This article unfolds as follows: Section 1 will provide a brief introduction to EEG characterization and the particularities of the EEG waveform characterization.
+Section 3 will explain the algorithms based on the waveform that will be used to perform the benchmark.
+Section 4 the experimentation procedure will be explained.
+Results will be presented in Section 5 and finally Discussion and Conclusions will be established in the final sections.
 
-We believe that studying specific componentes may lead to better understanding of the neuro component.
-
-%Novel sensor design and smart embodiment for pervasive healthcare;
-%New platforms for self-tracking or measurement of physical activity, sleep quality, emotion, gait, kinematics, kinetics and other factors related to personal well-being;
-%Information processing for localization and tracking for healthcare applications
-%Resource constrained network information processing for healthcare applications
-%Sensor data analytics, information integration, pattern mining/recognition, behaviour profiling, data visualisation and user feedback related to personal well-being;
-%
-%Mobile/wearable sensor-based/remote health monitoring systems, with illustrative case studies on use of innovative information processing approaches.
-%Disease focused exemplars and case studies (e.g. neurological disorders, cardiovascular diseases, diabetes and obesity, neuro-rehabilitation);
-
-\section{Electroencephalography Waveform Analysis}
-
-distinguishing the pertinent signal characteristics from extraneous content and representing them in a compact or menaingful form, amenable to interpretation by a human or computer (Wolpaw and Wolpaw)
-
-Shape or waveform analysis methods are considered as nonparametric methods (in opposition to statistical or dynamical models).
-They can explore the amplitude, energy, teager energy operand, or more complex like the L-Z Lempel-Ziv complexity measurement (thakor qEEG).
+\section{Electroencephalography}
 
 The Electroencephalogram is one of the most widespread used device to capture brain signals.
-It was initially developed by Hans Berger and has been extensively used for decades to diagnose neural diseases and other medical conditions.
+It was initially developed by Hans Berger in 1924 and has been extensively used for decades to diagnose neural diseases and other medical conditions.
 
-The first thing that he saw was the visual cortical alpha wave characterization, the Berger Rythm.
-Its amplitude and power determine that this feature can be coherently assosiated to a cognitive phenoma (eyes closed).
-We should ask ourselves if the EEG should be described years later if this particular event wasn't so evident.
+The first characterization that Dr.
+Berger detected was the visual cortical alpha wave, the \textit{Berger Rythm} \citep{Jansen1991}.
+He understood that the amplitude and shape of this rhythm was coherently associated to a cognitive action (eyes closing).
+We should ask ourselves if the EEG research explosion that came after that discovery would have happened, if it weren't so evident that the shape alteration was due to a very simple and verifiable cognitive process.
 
-The EEG signal is not stationary, highly complex.  It can be characterized as a linear stochastic process with great similarities to noise (Thakor EEG).  They are measured in microvolts.
-they have many artifacts.
+The EEG signal is highly complex.
+It can be modeled as a linear stochastic process with great similarities to noise \citep{Thakor2004}.
+They are measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.
 
-The most important part to consider is that of montage.
-It could be bipolar or referencial.
-In general EEG can be rereferenced offline.The traditional convention, somehow maintained in neuro research, downward polarity was considered negative and upward deflection for negative (Knott, 1985)  It is of utmost importance to remark that  EEG waveforms repre- sent the differential voltage between a given electrode and the recording reference.
-It is therefore clear that the choice of reference completely determines EEG wave- forms (Lehmann, 1987; Dien, 1998), an important method- ological consideration that all too often is still not recog- nized in the EEG literature.
-For example, recording with a vertex (Cz) reference would lead to small EEG deflections in the proximity of Cz due to potential synchronization of firing activities within closely spaced brain regions and vol- ume propagation of the EEG signal.
-Similarly, recording with mastoid or linked ears montage would lead to rather small waveforms at electrodes positioned over temporal brain regions (Pivik et al., 1993).
+\begin{figure}[H]
+\centering
+\includegraphics[width=6cm]{eegsample.png}
+\caption{Sample EEG signal captured and plotted on a 2D image using the popular and commercial EPOC Emotiv using EEGLAB.}
+\label{fig:sampleeeg}
+\end{figure}
 
-The automatic processing of EEG signals can be called qEEG is precisely the opposite of what we are doing here, quantite analysis (OJO)
+The device that captures these small variations in the current potentials over the scalp is called the Electroencephalograph.
+An  explanation on the particularities and modelling of EEG can be obtained from \citep{Jackson2014}, as well as in the description of the electrophysiological aspects of the method \citep{Haberman2012}.
 
-The conventional clinical method of observing the waveform is thought to be subjective and laborius because the results depend on the technicians' experience and expertise.
-This call for objective measurements pushed the adoption of qEEG methods and the need for automation (Thakor qEEG)
+EEG signals have the following general components:
 
-Some initial works on EEG explored the idea to extend human capacities analyzing EEG waveforms  (automatic detection of k complexes), (A Waveform Analyzer Applied to the Human EEG) where a feature from amplitude and frequency of its signal and its derivative in time-domain is used.
-Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.
-It is likeyly that the Matched Filter approach (linear) gave poor results and nonlinear methods based on neural network were being explored.It has been the traditional approach for technician, but the advent of elecrical processing, more sophisticated methods.
+\begin{itemize}
+\item Artefacts: endogeneous exogeneous as well as physiological or non-physiological (EEG MEG REview).
+\item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e.
+its mean and variance, and any other higher-order moments are not time-invariant \citep{Jansen1991}.
+\item DC drift and trending
+\item Basal EEG activity
+\item Inter-subject and intra-subject variability:It is known that EEG can be affected by the person behavior like sleep hygiene, caffeine intake, smoking habit, alcohol intake before EEG \citep{Farzan2017}.
+\end{itemize}
 
-EEG has been used extensively on Sleep Research, by performing Polysomnographic recordings (PSG)  (reference of the german paper of Sleep Research) where scoring of EEG to determine different sleep stages were performed by visually marking waveforms or graphoelements looking for patterns based on standardized guidelines.
+Overall, EEG can be characterized by their phase, amplitude,  frequency and \textit{waveform}.
 
-Visual characterization includes the identification or classification of componentes, or transient events, based on phase, monophasic or biphasic  (i.e.
-positive and negative component) where different components are measured, relation between them is stablished and index are created (e.g.
-sleep K-Complex is well characterized based on relation rates between positvie vs negative amplitude) (REFERENCE).
-Relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,
-sleep spindles, polysplindles, K-complexes, vertex sharp waves (VSW), and sawtooth waves (REM Sleep)
+\begin{itemize}
+\item Spontaneous: generally treated as noise or basal EEG.
+\item Evoked: activity that is time-locked to an incoming stimulus or an executed motor action.  In contrast it is often called Induced activty
+\end{itemize}
 
-Transient phenomena allows also to record occurrence and temporal sequence (mimicking spike analysis in neuro reserach)
+Understood according to the belongings or not to a repeated rhythm.
 
-The traditional approach do not consider waveform even though the brain oscilations are in general nonsinusoidal (reference)
+\begin{itemize}
+\item Rhythmic. EEG activity consisting in waves of approximately constant frequency.  It is often abbreviated RA (regular rythmic activity).
+\item Arrhythmic. EEG activity in which no stable rhythms are present.
+\item Dysrhythmic. Rhythms and/or patterns of EEG activity that characteristically appear in patient groups or rarely or seen in healthy subjects.
+\end{itemize}
 
-This approach is relative common in chemical analysis (i.e.
-chemometrics Skoog, D.; West, D.; Holler, F.
-Analtyical Chemistry, An Introduction; Saunders: Philadelphia, 1994.), geology (sismic analysis), and quantitive financial analysis.
-EKG, or Electrocardiogram, on the other hand, has been extensively processed and analyzed by waveform methods.
+Additionally, the number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: waveforms can be generalized, focal or lateralized (EEG).
 
-Slight variations in whether slopes or other features are scored may be important to certain applications.
+Finally, indexes can be derived as  CFC, Cross Frequency Coupling,  Phase-Amplitude Coupling, Phase-Phase Coupling.
 
-waveform, phase, amplitude and frequency
+The traditional clinical approach consisted in analyzing the paper strip that was generated by the plot of the signal obtained from the device.
+Expert technician and physicians analyzed visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.
+Atlases and guidelines were created in order to help in the recognition of these complex patterns.
+Even Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .
+The EEG research has focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements} \citep{Schomer2010}.
 
-waveform amplitud plotted against time
+Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)  \citep{Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines.
+Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g.
+positive or negative peak polarity) or the location within the strip.
+It is regular to establish an amplitude difference between different waveforms from which a relation between them is established and structured index are created (e.g.
+sleep K-Complex is well characterized based on rates between positive vs negative amplitude) (REFERENCE).
+Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, K-complexes, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).
 
-characteristic shape, or waveform.
-rising phase
-falling phase
-pronounced plateau
-ripples, wiggles,
+Moreover, in Epilepsy EEG data acquisition is crucial during the assessment of patients with focal epilepsies for potential seizure surgery, where the origin of the seizure activity must be reliably detected.
+The determination of the Epileptic Seizure onset is defined as the first electrical change seen in the EEG rhythm as compared to baseline or any clinical sign indicating seizure onset (Clinical EEG).
+Most epilepsy patients also show characteristic interictal (or between-seizure) epileptiform discharges (IEDs) termed spike (<70 microsec duration), spike and wave, or sharp-wave (70–200 microsec duration) discharges.
+(Electroencephalography Introduction Canadian Epileptic Assosiation).
+All these characterization are strictly based on the waveform shape.
 
-P,Q,R,S,T peaks in EKG analysis (REFERENCE)
-
-aEEG, amplitude integrated electroencephalography or cerebral function monitoring (REFERENCE)
-
-Artefacts: endogeneous exogeneous
-Non-Stationarity: this means that this and that.  What does it mean in terms of the signal.
-DC drift and trending
-Basal EEG activity
-
-Inter-subject and intra-subject variability
-
-Non-exhaustive list of transient events
-
-Determination of transient events, and particularly amplitude of different subcomponents, latency or even phase, has proved very importan concequences in terms of the different cognitive approach.
-
-Rhythms in neural activity are observed across various temporal and spatial scales and are often
-referred to as oscillations (see Glossary) [1]. Traditionally, neural oscillations have been
-clustered into canonical frequency bands, including delta (1–4 Hz), theta (4–8 Hz), alpha (8–
-12 Hz), beta (15–30 Hz), gamma (30–90 Hz), and high gamma (>50 Hz). These bands roughly
-correspond to frequency ranges commonly observed in human electroencephalography (EEG)
-studies. Although they have been observed for nearly a century, recent theories suggest that
-these oscillations play an active role in neural communication
-
- Hippocampal theta oscillations, for example, are among
-the best-studied rhythms in the local field potential (LFP); they have a stereotyped sawtooth
-shape
-
-Standard signal processing characterization (chemical, time series reference, sismic reference) can also be applied.
-
-EEG is a multidimensional signal non-stationary
-
-Amplitude
-Arch
-Frequency
-Phase
-Nonsinusoidal sinusoidal
-Oscilation
-Sawtooth: motor cortical beta oscillations
-Sharpness
-Spike-wave discharge
-Transient event
-
-inverse problem is mathematically intractable Voytek 2009
-
-Brainwaves may be missleading because what is actually been harnessed is electrical potentials over the scalp
-
-Reference to the good paper about how EEG actually works
-
-(Paper de Spinelli)
-
-CFC, Cross Frequency Coupling, sharper, arch comb or wicket shape, sawtooth, rectangular, spike-wave like, decay phase, voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace, peak and trough sharpness ratio, symmetry between rise and decay phase, slope ratio (steepness of the rise period to that of the adjacent decay period.
-
-Central trough is sharper and more negative that the adjacent troughs.
-
-Phase-Amplitude Coupling, Phase-Phase Coupling.
-
-Dimensionality reduction which is truly applied when the waveform is actually formed
-
-Oscillatory activity can also have their differente or distinctive waveforms.  Slow oscillations, which are assosiated with REM, sawtooth-shaped.
-Sleep spindles can also be considered oscilations and they have a distintinc form assosiated with stage 2 dream.
-Visual Cortical alpha and rolandic central mu waves arch-like structure (similar to the greek letter mu).
-Slope Ratio.
-Trough voltage remains contstant while peak voltage fluctuates.
-Steep slopes,   Amplitude asymmetry
-ponto-geniculo-occipital
-(PGO) waves
-
-In sleep research sleep 2 stage the background of KComplex and sleep spindles is theta waves
-
-Another approach which is very related to EEG waveform analysis is the detection of latency, amplitude and phase components in ERP.
-It is known that these characteristics can be mapped to cognitive or disease situations.
-We hypothesize that that including more approaches can lead to a better understanding or opening of a new field.
-
-by application to Brain Computer Interfaces or EEG diagnosis (BCI a review, BCI Vidal 1973)
-
-Seizures captured in their entirety typically show progression from low-voltage, high-frequency spikes to high-voltage, low-fre- quency spike-and-slow wave activity, before stopping abruptly and being replaced by background slowing or suppression (Fig.
+%Seizures captured in their entirety typically show progression from low-voltage, high-frequency spikes to high-voltage, low-fre- quency spike-and-slow wave activity, before stopping abruptly and being replaced by background slowing or suppression (Fig.
 29.9).
@@
 
-(e.g. FIRDA - Frontal Intermittent Rhythmic Delta) and posteriorly in children e.g. OIRDA - Occipital Intermittent Rhythmic Delta).
+Finally, there are specific EEG patterns that can be used to determine Depth of anestisia and aEEG, amplitude integrated electroencephalography or cerebral function monitoring (REFERENCE) is used to determine coma levels based on shapes obtained from shapes observed in the EEG.
 
-alpha dissapears when alerting by any mechanism (thinking, calculating)
+%Evoked Potentials (EPs) and Event Related Potentials (ERPs) are in general characterized by the waveforms.
 
-Rhythmic. EEG activity consisting in waves of approximately constant frequency.
-Arrhythmic. EEG activity in which no stable rhythms are present.
-Dysrhythmic. Rhythms and/or patterns of EEG activity that characteristically appear in patient groups or rarely or seen in healthy subjects.
+%The waveform of the EEG depends on the settings of the capturing device.
+The most important part to consider is that of montage.
+It could be bipolar or referencial.
+The traditional convention, somehow maintained in neuro research, downward polarity was considered negative and upward deflection for negative (Knott, 1985)  It is of utmost importance to remark that  EEG waveforms represent the differential voltage between a given electrode and the recording reference.
+It is therefore clear that the choice of reference completely determines EEG waveforms (Lehmann, 1987; Dien, 1998), an important methodological consideration that all too often is still not recognized in the EEG literature.
+For example, recording with a vertex (Cz) reference would lead to small EEG deflections in the proximity of Cz due to potential synchronization of firing activities within closely spaced brain regions and volume propagation of the EEG signal.
+Similarly, recording with mastoid or linked ears montage would lead to rather small waveforms at electrodes positioned over temporal brain regions (Pivik et al., 1993).
+Digital EEG can be rereferenced offline.
 
-Attenuation (synonyms: suppression, depression).
+%
+%BCI --> distinguishing the pertinent signal characteristics from extraneous content and representing them in a compact or menaingful form, amenable to interpretation by a human or computer (Wolpaw and Wolpaw)
+
+%Transient phenomena allows also to record occurrence and temporal sequence (mimicking spike analysis in neuro reserach)
+
+%The traditional approach do not consider waveform even though the brain oscilations are in general nonsinusoidal (reference)
+% Waveform characterization, on the other hand, has been extensively used in artifact detection and averaging.
+
+\subsection{EEG Waveform Characterization}
+
+The shape of the signal, the waveform, can be defined as the graphed line that represents the signal's amplitude plotted against time.
+It can  also called EEG biomarker,  EEG pattern, signal shape, signal form and a morphological signal (CITAS DONDE SE LAS LLAMA ASI).
+
+Waveform characterization pays importance to context, both in a spatial and in a temporal sense (Jensen).
+On the basis of contextual consideration, some specific waveform can be considered as noise while in other context is precisely the element which has a cognitive functional implication.
+
+%aforegoing
+%hitherto
+
+%phenomenological, the signal is treated as a black box.
+
+Shape or waveform analysis methods are considered as nonparametric methods (in opposition to statistical or dynamical models).
+They can explore the amplitude, energy or more complex like the teager energy operand and L-Z Lempel-Ziv complexity measurement (thakor qEEG).
+
+%Signal Morphology is not precisely defined in the literature but may refer to
+%
+%\begin{itemize}
+%\item RA (regular rythmic activity) 
+%\item low-voltage rapid activity 
+%\item sharp waves
+%\item longitudinal-bipolar and transverse- bipolar montages (Clinical EEG)
+%\end{itemize}
+
+Slight variations in weather slopes or other features are scored may be important to certain applications.
+
+It has
+\begin{itemize}
+\item characteristic shape, or waveform.
+\item rising phase
+\item falling phase
+\item pronounced plateau
+\item ripples and wiggles. 
+\end{itemize}
+
+\begin{itemize}
+\item Amplitude
+\item Arch
+\item Frequency
+\item Phase
+\item Nonsinusoidal sinusoidal
+\item Oscilation
+\item Sawtooth: motor cortical beta oscillations
+\item Sharpness
+\item Spike-wave discharge
+\item Transient event
+\end{itemize}
+
+Other characterization include, subjective definitions of sharper, arch comb or wicket shape, sawtooth, rectangular, spike-wave like, decay phase, voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace, peak and trough sharpness ratio, symmetry between rise and decay phase, slope ratio (steepness of the rise period to that of the adjacent decay period.
+For instance,  descriptions like "Central trough is sharper and more negative that the adjacent troughs" are common in the literature.
+
+\begin{itemize}
+\item Attenuation (synonyms: suppression, depression).
 Reduction of amplitude of EEG activity resulting from decreased voltage.
 When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".
-
-Hypersynchrony.
+\item Hypersynchrony.
 Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range.
@@
 (Note: term is used in interpretative sense but as a descriptor of change in the EEG).
-Paroxysmal.
+\item Paroxysmal.
 Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity.
 Though the term does not directly imply abnormality, much abnormal activity is paroxysmal.
+\end{itemize}
 
-Monomorphic. Distinct EEG activity appearing to be composed of one dominant activity
+\begin{itemize}
+\item Monomorphic: Distinct EEG activity appearing to be composed of one dominant activity
+\item Polymorphic: distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
+\item Sinusoidal. Waves resembling sine waves. Monomorphic activity usually is sinusoidal.
+\item Transient. An isolated wave or pattern that is distinctly different from background activity.
+\end{itemize}
 
-Polymorphic. distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
-
-Sinusoidal. Waves resembling sine waves. Monomorphic activity usually is sinusoidal.
-
-Transient. An isolated wave or pattern that is distinctly different from background activity.
-
-a) Spike: a transient with a pointed peak and a duration from 20 to under 70 msec.
-b) Sharp wave: a transient with a pointed peak and duration of 70-200 msec.
-
-generalized, focal or lateralized.
-
-Artifact detection and averaging.
+%\subsection{EEG Patterns}
+%
+%%The clinical EEG patterns are identified by the shape of waveform amplitude plotted against time.
+%
+%Non-exhaustive list of transient events
+%Waveform characterization is of quite importance in terms of Event Related Potentials.  
+%Determination of transient events, and particularly amplitude of different subcomponents, latency or even phase, has proved very importan concequences in terms of the different cognitive approach.
+%
+%%inverse problem is mathematically intractable Voytek 2009
+%
+%Oscillatory activity can also have their differente or distinctive waveforms.  Slow oscillations, which are assosiated with REM, sawtooth-shaped.
+%Sleep spindles can also be considered oscilations and they have a distintinc form assosiated with stage 2 dream.
+%Visual Cortical alpha and rolandic central mu waves arch-like structure (similar to the greek letter mu).
+Slope Ratio.
+Trough voltage remains contstant while peak voltage fluctuates.
+Steep slopes,   Amplitude asymmetry
+%ponto-geniculo-occipital
+%(PGO) waves.
+%Rhythms in neural activity are observed across various temporal and spatial scales and are often
+%referred to as oscillations (see Glossary) [1]. Traditionally, neural oscillations have been
+%clustered into canonical frequency bands, including delta (1–4 Hz), theta (4–8 Hz), alpha (8–
+%12 Hz), beta (15–30 Hz), gamma (30–90 Hz), and high gamma (>50 Hz). These bands roughly
+%correspond to frequency ranges commonly observed in human electroencephalography (EEG)
+%studies. Although they have been observed for nearly a century, recent theories suggest that
+%these oscillations play an active role in neural communication.
+Hippocampal theta oscillations, for example, are among the best-studied rhythms in the local field potential (LFP); they have a stereotyped sawtooth shape.
+In sleep research sleep 2 stage the background of KComplex and sleep spindles is theta waves
+%
+%%by application to Brain Computer Interfaces or EEG diagnosis (BCI a review, BCI Vidal 1973)
+%
+%
+%
+%%(e.g. FIRDA - Frontal Intermittent Rhythmic Delta) and posteriorly in children e.g. OIRDA - Occipital Intermittent Rhythmic Delta).
+%
+%%alpha dissapears when alerting by any mechanism (thinking, calculating)
+%
+%a) Spike: a transient with a pointed peak and a duration from 20 to under 70 msec.
+%b) Sharp wave: a transient with a pointed peak and duration of 70-200 msec.
+%
+%Some initial works on EEG explored the idea to extend human capacities analyzing EEG waveforms  (automatic detection of k complexes), (A Waveform Analyzer Applied to the Human EEG) where a feature from amplitude and frequency of its signal and its derivative in time-domain is used.
+Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.
+%
+%ACA VA EL RESTO DE LOS METODOS (fujimori, PAA, etc).  Uchida 1999
 
@@
 
-\begin{itemize}
+\begin{enumerate}
 \item The pattern can be identified and verified by visual inspection.
@@
 \item The analysis take account the shape of the plot of the signal.
-\item They can be used as templates.
-\item Transient events ?
-\item Single Channel
+\item The feature extraction procedure allows to create a Template database.
+\item A Single Channel Processing approach.
+\end{enumerate}
+
+As described in (Mixed Domain Signal Analysis) the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the pattern.
+
+\subsection{Selected Algorithms}
+
+According to the defined criteria, the algorithms that will be evaluated are as follows:
+
+\begin{itemize}
+\item Matching Pursuit
+\item Permutation Entropy
+\item Slope Horizontal Chain Code
+\item Histogram of Gradient Orientations
+\item Merging of Increasing and Decreasing Sequences
 \end{itemize}
 
-Although the term morphology has been used to identify this approach, We specifically excluded morphological based methods
-\cite{Yamaguchi2009}.
+Although the term morphology has been used to identify this approach, We specifically excluded Mathematical Morphological based methods \cite{Yamaguchi2009}.
 On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm.
@@
 
-As described in (Mixed Domain Signal Analysis) the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the pattern.
+\subsection{Excluded Methods}
 
-\subsection{Peak Picking}
+\subsubsection{Waveform features}
 
-The name of many of the EEG features reference indeed peaks like P300 or P3a P3b or N100.
-This leads to a natural way to classify them visually by selecting appropiate peaks and matching their positions and amplitudes in an orderly manner.
+One of the earliest approach to automatically process EEG data is the Peak Picking method.
+Although of limited usability, peak picking has been used to determine latency of transient events in EEG \citep{Jaskowski2000,Zhang2011}.
+Straightforward in its implementation, it consists in selecting a component, particularly a simple component based on the expected location of its more prominent deflection \citep{Ouyang2017}.
+Particularly common in ERP Research, the name of many of the EEG features directly reference a peak within the component, e.g.
+P300 or P3a P3b or N100.
+This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.
+The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).
 
-Although of limited usability, peak picking is the first automatic method of detection and has been used to determine latency of transient events in EEG.
-Straighforward in its implementation, it consist in selecting a component, particularly a simple component based on the expected location of its more prominent deflection.
+A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.
+This was even used in the seminal work of Farwell and Donchin on P300 \citep{Farwell1988,WolpawJonathanR2012}.
+Additionally, a logarithmic graph of the peak-to-peak amplitude or aEEG \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.
 
-Ouyang2017,Zhang2011
+Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms  (automatic detection of k complexes), (A Waveform Analyzer Applied to the Human EEG) where a feature from amplitude and frequency of its signal and its derivative in time-domain is used.
 
-difference of areas (trabajo Mexicanos)
+Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.
 
- integrated activity (area measure) (area measure)
+ (fujimori, PAA, etc).  Uchida 1999
 
- It was used in Farwell and Donchin work on P300 (Wolpaw and Wolpaw).
+\subsubsection{Averaging Methods}
 
-\subsection{Woody's Template Matching}
+The methods that allow to identified waveforms are used to determine different alignments while averaging epochs.
 
-Cross Covariance with Template or Cross Correlation
+This long-standing problem has been tackled from different perspectives.  Woody's Template Matching is perhaps the best effort as well as Pham, and Tuan and others EML.
+
+Dictionary, template based method.  Many do not consider the waveform, albeit they do use a set of templates obtained from a dictionary.
+
+Cross Covariance with Template or Cross Correlation or ACF: Autocorrelation Function: PHAM METHOD
 
@@
 extended to wavelent analysis.
+
+Dynamic Time Warping and other Warp Averaging methods were also used.
+
+Wrapping Analysis
+
+Warp averaging
 
@@
 
-\subsection{Dynamic Time Warping}
-
-Dynamic Time Warping is a technique
-
-\subsection{Wrapping Method}
-
 \subsection{Permutation Entropy}
@@
 
-\subsection{ACF: Autocorrelation Function}
+This method is the best representative of Waveform Complexity: Lempel-Ziv method L-Z complexity are other variants which use a different definition.
 
-\subsection{Waveform Complexity}
-
-Lempel-Ziv method L-Z complexity
+Permutation Entropy a new feature for brain computer interfaces.
 
@@
 
-\subsection{Warp Averaging}
+This method can be encompass as a syntactic analysis technique, where the EEG segment is represented as a series of elementary patterns, similar to tokens.
 
-%\subsection{Morphological Signal Analysis}
-%
-%Several approaches and methods have been applied to decode them.  Of particular interests are morphological signal analysis 
-%CITE (Paper del chabon del ITBA que detalla bien los diferentes metodos) \citep{Alvarado-Gonzalez2016,Yamaguchi2009}.
-%
-%Morphological or 
-%
-%template matching
-%
-%spike
-%sharp wave
-%spike wave
-%polyspike 
-%slow wave complex
-%
-%shape domain
-%
-%\subsection{P300}
-%
-%Uhhhh I can talk a lot about P300 \citep{Knuth2006}.
-%
-%P300 single trial is a golden grail of Brain computer interfaces.  It is studied and analyzed in this way, here in this other way, in this other way is here.
+\subsection{Histogram of Gradients}
 
-\subsection{EEG based on Histogram of Gradients}
+\subsection{Merger of Increasing and Decreasing Sequences}
+
+Stepwise downsampling
 
@@
 In order to verify each one of the approaches we will inject into a known dataset spurious p300 signal that we may be able to control.
-By implementing this pseudo-real approach, we can effectively control null-signals of this evoked potentials following the same approach of other similar works ( ref   Ouyang2017, el trabajo de estudio de templates y el trabajo de p3).
+By implementing this pseudo-real approach, we can effectively control null-signals of this evoked potentials following the same approach of other similar works ( ref   Ouyang2017, el trabajo de estudio de templates y el trabajo de p3 y la tesis).
+
+Jaskowski and Verleger (2000)
 
@@
 
-The second part, we will specifically find the P300 signal from a dataset.
+We will extend this results to offline processing a public dataset of real P300 patterns.
 
-amplitude
-latency
-
-in a single trial approach.  The template was obtained from the point to point average of a standard p300 experiment (published elsewhere).
-
-The original signal-to-noise ratio was calculated as Hue 2010 as
-
-\begin{itemize}
-\item Peak Finding
-\item Woody Template Matching
-\item Permutation Entropy
-\item SHCC (los mexicanos me dijeron que me dan el codigo para matlab)
-\item BCI SIFT
-\end{itemize}
-
-\begin{itemize}
-\item Inyectar una señal en un tren de EEG simulando el P300
-\item Probar con diferentes metodos la capacidad de detectar si un segmento contiene o no el patron
-\item Probar con diferentes metodos la capacidad de identificar correctamente la localizacion del patron.
-\item Probar variando la señal relacion ruido usando la formula de Huo y con eso ver como cada metodo permite mantenerse.
-\end{itemize}
-
-%The central part of this method is the feature generation: the histogram of image gradients.  To do so, the first step is the transformation of the signal into a temporary binary image.
-%
-%The signal is first scaled and centered at zero by 
-%
-%\begin{equation}
-%\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot ( x(t,c) - \bar{x}(t,c)  )}\right \rfloor
-%\end{equation}
-%
-%\noindent where $\gamma$ is the image scale, $t$ is time and $ x(t,c) $ is the EEG matrix defined for each $t$ and for a particular channel $c$.
-%
-%Then the image is constructed by
-%
-%\begin{equation}
-%I(z_1,z_2) = \left\{ \begin{array}{rl}
-%255 & z_1 = \gamma \cdot t; z_2 = \tilde{x}(t,c) + z(c) \\
-%0   & \mbox{otherwise}
-%\end{array}\right.
-%\label{eq:images}
-%\end{equation}
-
-BLABLABLABLA \ref{fig:sampledescriptor}.
+The original signal-to-noise ratio was calculated as Hue 2010.
 
@@
 \centering
-\includegraphics[width=6cm]{sampledescriptor.png}
-\caption{The local patch is located around a sample point plotted on the temporary image.
-All the sample points are interpolated using the Bresenham algorithm.
-The 4x4 block grid can be seen on the patch as well as the green arrows showing the dominant gradient on each block.}
-\label{fig:sampledescriptor}
+\includegraphics[width=6cm]{images/erptemplate1.eps}
+\caption{ERP Template obtained from the point-to-point averaging of a P300 component of a subject obtained from the public dataset.}
+\label{fig:erptemplate1}
 \end{figure}
@@
 
-BLABLALBALBALBAL
+Each method has its own set of parameters which are optimized.
+We will try to use an approach used in (matching pursuit paper for bci) where the best parameters are those which provide the best performance.
+
+Hence, in order to obtain the right set of parameters for each method, an optimization problem was used fixing the injected signal to a gain of 3, and optimizing the parameters.
 
@@
 \label{section:results}
-sections
 
-BLABLABLABLALBA
+Results are shown in Table 1 and in Figure 1.
+Table 1 shows the performance to identify each letter of the standard P300 Speller Matrix.
+Figure 2 shows how the performance increase when averaging is implemented.
 
@@
 %% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccccc}
+\begin{tabular}{cc}
 \toprule
-\textbf{Participant}	& \textbf{Original Cz}	& \textbf{ACC at Cz}	& \textbf{BPC}	& \textbf{Performance}\\
+\textbf{Method}	& \textbf{Single Trial Performance}	\\
 \midrule
-1     &   0.84 &   0.79 & Cz  &   0.79 $\pm$ 0.01 \\
-2     &   0.86 &   0.81 & PO7 &   0.93 $\pm$ 0.01 \\
-3     &   0.87 &   0.78 & Cz  &   0.78 $\pm$ 0.03 \\
-4     &   0.86 &   0.68 & PO8 &   0.90 $\pm$ 0.01 \\
-5     &   0.86 &   0.80 & PO7 &   0.93 $\pm$ 0.01 \\
-6     &   0.89 &   0.96 & Cz  &   0.96 $\pm$ 0.01 \\
-7     &   0.89 &   0.78 & PO7 &   0.93 $\pm$ 0.01 \\
-8     &   0.92 &   0.91 & PO7 &   1.00 $\pm$ 0.00 \\
-1     &   0.84 &   0.81& Fz&   0.82 $\pm$ 0.0017 \\
-2     &   0.86 &   0.82& Fz&   0.85 $\pm$ 0.0011 \\
-3     &   0.87 &   0.82& Cz&   0.82 $\pm$ 0.0011 \\
-4     &   0.86 &   0.81& PO7&   0.82 $\pm$ 0.0012 \\
-5     &   0.86 &   0.82& Fz&   0.82 $\pm$ 0.0021 \\
-6     &   0.89 &   0.81& Fz&   0.82 $\pm$ 0.0016 \\
-7     &   0.89 &   0.83& Fz&   0.84 $\pm$ 0.0011 \\
-8     &   0.92 &   0.82& PO7&   0.84 $\pm$ 0.0020 \\
+MP & 90 \\
+PE & 89 \\
+SHCC & 20 \\
 \bottomrule
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=6cm]{images/performance.eps}
+\caption{Performance obtained from the different methods.}
+\label{fig:performance}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=6cm]{images/P300Performance.eps}
+\caption{Performance obtained from the different methods while averaging different trials.}
+\label{fig:p300performance}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-BLABLABLABLABLALBA
+In spite of all this, the conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.
+The need for more objective measurements pushed the adoption of more automated means of decoding the signals and demanded a need for replication \citep{Thakor2004}.
+This lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread.
 
-This method, different from other methods which is based on the nonlinearity of the gradient of histograms which can be used to detect
+This approach is relative common in chemical analysis (i.e.
+chemometrics Skoog, D.; West, D.; Holler, F.
+Analtyical Chemistry, An Introduction; Saunders: Philadelphia, 1994.), geology (sismic analysis), and quantitive financial analysis.
+EKG, or Electrocardiogram, on the other hand, has been extensively processed and analyzed by waveform methods.
+P,Q,R,S,T peaks in EKG analysis (REFERENCE)
 
-is also based on how the image look like.
+Dimensionality reduction which is truly applied when the waveform is actually formed.
 
-SNR of p300 and how to detect it
-
-Check if you can use this to detect any kind of transient signal.
-
-Compare if it is possible with the descriptors from one subject, discriminate the others.
-
-Channal identification based on the metric distance between the bags
+We left out the segmetation issue which is itself a huge problem.  Where should I look from within the EEG trace ?
 
@@
 \section{Conclusion}
+
+The purpose of this work was twofold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to verify and compare those methods.
+
+Invasive methods remains an issue (referencia del paper de nature que muestra como causan deficits los implantes)
+
+BCI as Assistive technology and BCI supplement natural motor output
+The value of clinical focus, 
+BCI reliability (WOkpaw and wolpaw).
+Clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can lead to more robust implementation of BCI devices.
+
+Let us now turn to the question posed in the title of this paper.
+
+Successful approaches in Computer Vision or pattern recognition in other areas use a set of different features to compount ensemble classifiers.  All these methods had the advantage 
+that they can truly map a visual component with a clinical meaning, assessed by a physician, and provide an automatic identification which can be constructed by the set of methods.
+We believe this approach can be used to provide tools which are more meaningful to the clinical set.
+
+Particularly considering BCI, this approach, in line with the general idea, can be used to foster BCI solutions that can help clinical settings.  Precisely for the same reasons.
+
+More work has to be conducted in order to extend this methods to other patterns of waves.
+
+Berger H: tiber das Elektrenkephalogramm des Menschen, Archiv Psychiatr, 87 (1929) pp. 527-570.
 
@@
 %=====================================
-\bibliography{article}
+\bibliography{sensors}
 

### 2018-04-30 sensors.tex
Changes up

@@
 It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time and translating or decoding brain signals that can be used to diagnose disease or implement Human Computer Interaction devices.
-The automatic approach which is based on using digital tools to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community which was based intensively on EEG waveforms and the structure of signal plots.
-The purpose of this work is to help to start to fill this gap by doing a review and description of the automatic methods that have been used to detect patterns in the waveforms, and to perform a benchmarking analysis to determine different characteristics of those methods that aim to detect specific waveforms mimicking what physicians has been doing since the inception of this fruitful technology.
+The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community which was based intensively on EEG waveforms and the structure of signal plots.
+Hence, the purpose of this work is to help to start to fill this gap by doing a review and description of the procedures that have been used to detect patterns in the waveforms, and to perform a benchmarking analysis of them.
 }
+
+% We aim to detect specific waveforms mimicking what physicians has been doing since the inception of this fruitful technology.
 
@@
 
-Current society is demanding technology to once and for all provide the means to realize the utopia of social inclusion for people with disabilities.
-Additionally, a worldwide aging population is also claiming for solutions to extend active lifestyles throughout all life, including additional years that medicine is allowing humanity to have \citep{Lutz2008}.
-At the same time,  the digital gadgets and the digital revolution have modified the way people interact with all their devices.
-A new emerging digital society is consolidating and smart wearable sensors are starting to be used routinely to obtain bio-signals, in an active or passive way (Wolpaw Communication and control)
-Although this communication is based on muscular movement \citep{Guger2017}, all these trends are precisely pushing this boundary beyond the confines of the body and beyond the limitation of human movement.
-A new form of Human Machine communication which directly connects the Central Nervous System to a machine is currently being under development: Brain Machine Interfaces.
-Moreover, as long as computers are being used inside every machine that humanity posses \citep{Domingo2012} this discipline is specializing as Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).
-On the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG) \citep{Schomer2010}.
+%People with disabilities needs technological solutions for inclusion
+
+%Old people is more likely to be disabled.
+
+%The society is getting old.
+
+%We also have digital gadgets everywhere, so the society needs new ways to interact with those devices.
+
+%So far the interaction with devices has been based on muscular movement, but these trends are pushing the boundary beyond the confines of the body.
+
+Current society is demanding technology to provide the means to realize the utopia of social inclusion for people with disabilities\citep{Wolpaw2002}.
+Additionally, as societies are aging \citep{Lutz2008} the incidence of neuromuscular atrophies, strokes and other invalidating diseases is increasing.
+Concurrently, the digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices \citep{Domingo2012}.
+All this communication is based on muscular movement \citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human movement.
+A new form of Human Machine communication which directly connects the Central Nervous System (CNS) to a machine or computer device is currently being developed: Brain Machine Interfaces (BMI), Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).
+
+At the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG) \citep{Schomer2010}.
 
 EEG sensors are wearable \citep{Puce2017} non-invasive, portable and mobile \citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution \citep{Hartman2005}.
-This humble diagnosis device has been transformed into currently the best approach that we posses to precisely detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive our cars, steer our drones, write our emails, and control our wheelchairs \citep{Yuste2017}.
+This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, or control wheelchairs \citep{Yuste2017}.
 
-The clinical and historical approach to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph (Atlas of EEG patterns): multichannel signals were extracted and plotted over a piece of paper, continuously.
+The clinical and historical tactic to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph\citep{Hartman2005}: multichannel signals were extracted and continuously plotted over a piece of paper.
 Electroencephalographers or Electroencephalography technician have decoded and detected patterns along the signals by visually inspecting them \citep{Schomer2010}.
-Even nowadays clinical EEG remains a visually interpreted test \citep{Hartman2005}.
+Nowadays clinical EEG still remains a visually interpreted test \citep{Hartman2005}.
 
-Automatic processing, or quantitave EEG, was based first on analog electronic devices and later on computerized digital processing methods \citep{Jansen1991}.
-They implemented mathematically and algorithmically complex procedures to decode the information with outstanding success \citep{Yuste2017}.
-The best materialization of the automatic processing of EEG signals lays in the BCI discipline, where $71.2\%$ is based on EEG \citep{Guger2017}.
+In contrast, automatic processing, or quantitative EEG, was based first on analog electronic devices and later on computerized digital processing methods \citep{Jansen1991}.
+They implemented mathematically and algorithmically complex procedures to decode the information with good results \citep{Yuste2017}.
+The best materialization of the automatic processing of EEG signals rests in the BCI discipline, where around $71.2\%$ is based on noninvasive EEG \citep{Guger2017}.
 
@@
 Hence, the traditional and knowledgeable approach was mainly overshadow particularly in BCI Research, and the waveform of the EEG was replaced by pragmatically sound procedures that were difficult to link to existing clinical EEG knowledge.
-We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and that can help to decode them in an automated way.
+We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and that can help to decode them in a supervised and semi-automated way.
 
@@
 The second is to evaluate and study these methods by analyzing its classification performance against a pseudo-real dataset.
-We aim to outline a better understanding of the different approaches.
 We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective \citep{Nijboer2009,Wei2017}.
@@
 
-The Electroencephalogram is one of the most widespread used device to capture brain signals.
-It was initially developed by Hans Berger in 1924 and has been extensively used for decades to diagnose neural diseases and other medical conditions.
+The Electroencephalography consists on the measurement of small variations of electrical currents over the scalp.
+It is one of the most widespread used method to capture brain signals and was initially developed by Hans Berger in 1924 and has been extensively used for decades to diagnose neural diseases and other medical conditions.
 
 The first characterization that Dr.
-Berger detected was the visual cortical alpha wave, the \textit{Berger Rythm} \citep{Jansen1991}.
+Berger detected was the Visual Cortical Alpha Wave, the \textit{Berger Rythm} \citep{Jansen1991}.
 He understood that the amplitude and shape of this rhythm was coherently associated to a cognitive action (eyes closing).
-We should ask ourselves if the EEG research explosion that came after that discovery would have happened, if it weren't so evident that the shape alteration was due to a very simple and verifiable cognitive process.
+We should ask ourselves if the research advancement that came after that discovery would have happened if it weren't so evident that the shape alteration was due to a very simple and verifiable cognitive process.
 
-The EEG signal is highly complex.
+The EEG signal is a highly complex multi-channel time-series.
 It can be modeled as a linear stochastic process with great similarities to noise \citep{Thakor2004}.
-They are measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.
+It is measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.
 
@@
 \centering
-\includegraphics[width=6cm]{eegsample.png}
+\includegraphics[height=6cm,width=12cm]{images/eegsample.png}
 \caption{Sample EEG signal captured and plotted on a 2D image using the popular and commercial EPOC Emotiv using EEGLAB.}
@@
 
-The device that captures these small variations in the current potentials over the scalp is called the Electroencephalograph.
-An  explanation on the particularities and modelling of EEG can be obtained from \citep{Jackson2014}, as well as in the description of the electrophysiological aspects of the method \citep{Haberman2012}.
+The device that captures these small variations in current potentials over the scalp is called the Electroencephalograph.
+Electrodes are located in predetermined positions over the head, usually embedded in saline solutions to facilitate the electrophysiological interface and are connected to a differential amplifier with a high gain which allowed the measurement of tiny signals.
+Although initially analog devices were developed and used, nowadays digital versions connected directly to a computer are pervasive.
+A detailed explanation on the particularities and modeling of EEG can be obtained from \citep{Jackson2014}, and a description of its electrophysiological aspects from \citep{Haberman2012}.
 
-EEG signals have the following general components:
+Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The following components regularly  characterize EEG signals:
 
 \begin{itemize}
-\item Artefacts: endogeneous exogeneous as well as physiological or non-physiological (EEG MEG REview).
+\item Artifacts:  These are signal sources which are not generated from the CNS, but can be detected from the EEG signal.
+These are called endogeneous or physiological when they are generated from a biological source like muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic noise\citep{Weeda2012}.
 \item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e.
 its mean and variance, and any other higher-order moments are not time-invariant \citep{Jansen1991}.
-\item DC drift and trending
-\item Basal EEG activity
-\item Inter-subject and intra-subject variability:It is known that EEG can be affected by the person behavior like sleep hygiene, caffeine intake, smoking habit, alcohol intake before EEG \citep{Farzan2017}.
+\item DC drift and trending: in EEG jargon, which is derived from concepts of electrical amplifiers theory, Direct Current (DC) refers to very low frequency components of the EEG signal which varies around a common center, usually the zero value.
+DC drift means that this center value drifts in time.
+Although sometimes considered as a nuisance that needs to get rid of, it is known that very important cognitive phenomena like Slow Cortical Potentials or Slow Activity Transients in innates do affect the drift and can be used to understand some particular brain functioning \cite{Schomer2010}.
+\item Basal EEG activity: the EEG is the compound summation of myriads of electrical sources from the CNS.
+These sources generate a baseline EEG which shows continuous activity with a small or null relation with any concurrent cognitive activity or task.
+\item Inter-subject and intra-subject variability: EEG can be affected by the person's behavior like sleep hygiene, caffeine intake, smoking habit or alcohol intake previously to the signal measuring procedure \citep{Farzan2017}.
 \end{itemize}
 
-Overall, EEG can be characterized by their phase, amplitude,  frequency and \textit{waveform}.
+Regarding how the EEG activity can be related to an external stimulus that is affecting the subject, it can be considered as
 
@@
 \item Spontaneous: generally treated as noise or basal EEG.
-\item Evoked: activity that is time-locked to an incoming stimulus or an executed motor action.  In contrast it is often called Induced activty
+\item Evoked: activity that is time-locked to an incoming stimulus or an executed motor action.  In contrast it is often called Induced activity
 \end{itemize}
 
-Understood according to the belongings or not to a repeated rhythm.
+\noindent Additionally, according to the existence of a repeated rhythm, the EEG activity can be understood as
 
 \begin{itemize}
-\item Rhythmic. EEG activity consisting in waves of approximately constant frequency.  It is often abbreviated RA (regular rythmic activity).
-\item Arrhythmic. EEG activity in which no stable rhythms are present.
-\item Dysrhythmic. Rhythms and/or patterns of EEG activity that characteristically appear in patient groups or rarely or seen in healthy subjects.
+\item Rhythmic: EEG activity consisting in waves of approximately constant frequency.  It is often abbreviated RA (regular rythmic activity).
+\item Arrhythmic: EEG activity in which no stable rhythms are present.
+\item Dysrhythmic: Rhythms and/or patterns of EEG activity that characteristically appear in patient groups or rarely or seen in healthy subjects.
 \end{itemize}
 
-Additionally, the number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: waveforms can be generalized, focal or lateralized (EEG).
+The number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: signal elements can be generalized, focal or lateralized, depending on in which channel (i.e.
+electrode) they are found.
 
-Finally, indexes can be derived as  CFC, Cross Frequency Coupling,  Phase-Amplitude Coupling, Phase-Phase Coupling.
-
-The traditional clinical approach consisted in analyzing the paper strip that was generated by the plot of the signal obtained from the device.
-Expert technician and physicians analyzed visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.
-Atlases and guidelines were created in order to help in the recognition of these complex patterns.
-Even Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .
-The EEG research has focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements} \citep{Schomer2010}.
-
-Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)  \citep{Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines.
-Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g.
-positive or negative peak polarity) or the location within the strip.
-It is regular to establish an amplitude difference between different waveforms from which a relation between them is established and structured index are created (e.g.
-sleep K-Complex is well characterized based on rates between positive vs negative amplitude) (REFERENCE).
-Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, K-complexes, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).
-
-Moreover, in Epilepsy EEG data acquisition is crucial during the assessment of patients with focal epilepsies for potential seizure surgery, where the origin of the seizure activity must be reliably detected.
-The determination of the Epileptic Seizure onset is defined as the first electrical change seen in the EEG rhythm as compared to baseline or any clinical sign indicating seizure onset (Clinical EEG).
-Most epilepsy patients also show characteristic interictal (or between-seizure) epileptiform discharges (IEDs) termed spike (<70 microsec duration), spike and wave, or sharp-wave (70–200 microsec duration) discharges.
-(Electroencephalography Introduction Canadian Epileptic Assosiation).
-All these characterization are strictly based on the waveform shape.
-
-%Seizures captured in their entirety typically show progression from low-voltage, high-frequency spikes to high-voltage, low-fre- quency spike-and-slow wave activity, before stopping abruptly and being replaced by background slowing or suppression (Fig.
-29.9).
-Usual morphologic features include typical rhythmic, gen- eralized, symmetric spike-and-waves or polyspikes and waves at 2 to 3.5 Hz; atypical spike and wave with lower frequency and less symmetry; multiple spike-and-wave (repetitive complexes of two or more spikes followed by a slow wave); and high-voltage, repetitive, rhythmic, focal or generalized delta activity with inter- mixed spikes, sharp waves, or sharp components (Fig.
-29.24) (45).
-Diagnosis is more difficult when the seizure (or SE) pre- cedes the beginning of the tracing and continues beyond its end.
-In such cases, rhythmic sharp features, typically faster than 1 Hz, may be seen, often with variability.
-
-Finally, there are specific EEG patterns that can be used to determine Depth of anestisia and aEEG, amplitude integrated electroencephalography or cerebral function monitoring (REFERENCE) is used to determine coma levels based on shapes obtained from shapes observed in the EEG.
-
-%Evoked Potentials (EPs) and Event Related Potentials (ERPs) are in general characterized by the waveforms.
-
-%The waveform of the EEG depends on the settings of the capturing device.
-The most important part to consider is that of montage.
-It could be bipolar or referencial.
-The traditional convention, somehow maintained in neuro research, downward polarity was considered negative and upward deflection for negative (Knott, 1985)  It is of utmost importance to remark that  EEG waveforms represent the differential voltage between a given electrode and the recording reference.
-It is therefore clear that the choice of reference completely determines EEG waveforms (Lehmann, 1987; Dien, 1998), an important methodological consideration that all too often is still not recognized in the EEG literature.
-For example, recording with a vertex (Cz) reference would lead to small EEG deflections in the proximity of Cz due to potential synchronization of firing activities within closely spaced brain regions and volume propagation of the EEG signal.
-Similarly, recording with mastoid or linked ears montage would lead to rather small waveforms at electrodes positioned over temporal brain regions (Pivik et al., 1993).
-Digital EEG can be rereferenced offline.
-
-%
-%BCI --> distinguishing the pertinent signal characteristics from extraneous content and representing them in a compact or menaingful form, amenable to interpretation by a human or computer (Wolpaw and Wolpaw)
-
-%Transient phenomena allows also to record occurrence and temporal sequence (mimicking spike analysis in neuro reserach)
-
-%The traditional approach do not consider waveform even though the brain oscilations are in general nonsinusoidal (reference)
-% Waveform characterization, on the other hand, has been extensively used in artifact detection and averaging.
+%Finally, indexes can be derived as  CFC, Cross Frequency Coupling,  Phase-Amplitude Coupling, Phase-Phase Coupling.
 
@@
 The shape of the signal, the waveform, can be defined as the graphed line that represents the signal's amplitude plotted against time.
-It can  also called EEG biomarker,  EEG pattern, signal shape, signal form and a morphological signal (CITAS DONDE SE LAS LLAMA ASI).
+It can also be called EEG biomarker,  EEG pattern, signal shape, signal form and a morphological signal \citep{Jansen1991}.
 
-Waveform characterization pays importance to context, both in a spatial and in a temporal sense (Jensen).
-On the basis of contextual consideration, some specific waveform can be considered as noise while in other context is precisely the element which has a cognitive functional implication.
+The signal context is crucial for waveform characterization, both in a spatial and in a temporal domain \citep{Jansen1991}.
+Depending on the context, some specific waveform can be considered as noise while in other cases is precisely the element which has a cognitive functional implication.
 
@@
 %phenomenological, the signal is treated as a black box.
-
-Shape or waveform analysis methods are considered as nonparametric methods (in opposition to statistical or dynamical models).
-They can explore the amplitude, energy or more complex like the teager energy operand and L-Z Lempel-Ziv complexity measurement (thakor qEEG).
 
@@
 
-Slight variations in weather slopes or other features are scored may be important to certain applications.
+A waveform can have a characteristic shape, a rising or falling phase, a pronounced plateau or it may be composed of ripples and wiggles.
+In order to describe them, they are characterized by its amplitude, the arch, whether they have (non)sinusoidal shape, by the presence of an oscillation or imitating a sawtooth (e.g.
+Motor Cortical Beta Oscillations).
+The characterization by their sharpness is also common, particularly in Epilepsy, and they can also be identified by their resemblance to spikes (e.g.
+Spike-wave discharge).
 
-It has
-\begin{itemize}
-\item characteristic shape, or waveform.
-\item rising phase
-\item falling phase
-\item pronounced plateau
-\item ripples and wiggles. 
-\end{itemize}
+Other depictions may include, subjective definitions of sharper, arch comb or wicket shape, rectangular, containing a decay phase or voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace.
+Derived ratios and indexes can be used as well like peak and trough sharpness ratio, symmetry between rise and decay phase and slope ratio (steepness of the rise period to that of the adjacent decay period).
+For instance,  wording like "Central trough is sharper and more negative that the adjacent troughs" are common in the literature.
+
+Other regular characterizations which are based on shape features may include:
 
 \begin{itemize}
-\item Amplitude
-\item Arch
-\item Frequency
-\item Phase
-\item Nonsinusoidal sinusoidal
-\item Oscilation
-\item Sawtooth: motor cortical beta oscillations
-\item Sharpness
-\item Spike-wave discharge
-\item Transient event
-\end{itemize}
-
-Other characterization include, subjective definitions of sharper, arch comb or wicket shape, sawtooth, rectangular, spike-wave like, decay phase, voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace, peak and trough sharpness ratio, symmetry between rise and decay phase, slope ratio (steepness of the rise period to that of the adjacent decay period.
-For instance,  descriptions like "Central trough is sharper and more negative that the adjacent troughs" are common in the literature.
-
-\begin{itemize}
-\item Attenuation (synonyms: suppression, depression).
+\item Attenuation: Also called suppression or depression.
 Reduction of amplitude of EEG activity resulting from decreased voltage.
@@
 The term implies an increase in the number of neural elements contributing to the rhythm.
-(Note: term is used in interpretative sense but as a descriptor of change in the EEG).
 \item Paroxysmal.
@@
 \item Polymorphic: distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
-\item Sinusoidal. Waves resembling sine waves. Monomorphic activity usually is sinusoidal.
 \item Transient. An isolated wave or pattern that is distinctly different from background activity.
@@
 
+The traditional clinical approach consisted in analyzing the paper strip that was generated by the plot of the signal obtained from the device.
+Expert technician and physicians analyzed visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.
+Atlases and guidelines were created in order to help in the recognition of these complex patterns.
+Even Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .
+The clinical EEG research has also focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements} \citep{Schomer2010}.
+
+Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)  \citep{Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines.
+Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g.
+positive or negative peak polarity) or the location within the strip.
+It is regular to establish an amplitude difference between different waveforms from which a relation between them is established and a structured index are created (e.g.
+sleep K-Complex is well characterized based on rates between positive vs negative amplitude) (REFERENCE).
+Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, K-complexes, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).
+
+Moreover, EEG data acquisition is a key procedure during the assessment of patients with focal Epilepsy for potential seizure surgery, where the source of the seizure activity must be reliably identified.
+The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.
+The interictal epileptiform discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges\citep{EEGIntro}.
+
+%Seizures captured in their entirety typically show progression from low-voltage, high-frequency spikes to high-voltage, low-fre- quency spike-and-slow wave activity, before stopping abruptly and being replaced by background slowing or suppression (Fig.
+29.9).
+Usual morphologic features include typical rhythmic, gen- eralized, symmetric spike-and-waves or polyspikes and waves at 2 to 3.5 Hz; atypical spike and wave with lower frequency and less symmetry; multiple spike-and-wave (repetitive complexes of two or more spikes followed by a slow wave); and high-voltage, repetitive, rhythmic, focal or generalized delta activity with inter- mixed spikes, sharp waves, or sharp components (Fig.
+29.24) (45).
+Diagnosis is more difficult when the seizure (or SE) pre- cedes the beginning of the tracing and continues beyond its end.
+In such cases, rhythmic sharp features, typically faster than 1 Hz, may be seen, often with variability.
+
+%Finally, there are specific EEG patterns that can be used to determine Depth of anestisia.
+
+%Evoked Potentials (EPs) and Event Related Potentials (ERPs) are in general characterized by the waveforms.
+
+%The waveform of the EEG depends on the settings of the capturing device.
+The most important part to consider is that of montage.
+It could be bipolar or referencial.
+The traditional convention, somehow maintained in neuro research, downward polarity was considered negative and upward deflection for negative (Knott, 1985)  It is of utmost importance to remark that  EEG waveforms represent the differential voltage between a given electrode and the recording reference.
+It is therefore clear that the choice of reference completely determines EEG waveforms (Lehmann, 1987; Dien, 1998), an important methodological consideration that all too often is still not recognized in the EEG literature.
+For example, recording with a vertex (Cz) reference would lead to small EEG deflections in the proximity of Cz due to potential synchronization of firing activities within closely spaced brain regions and volume propagation of the EEG signal.
+Similarly, recording with mastoid or linked ears montage would lead to rather small waveforms at electrodes positioned over temporal brain regions (Pivik et al., 1993).
+Digital EEG can be rereferenced offline.
+
+%
+%BCI --> distinguishing the pertinent signal characteristics from extraneous content and representing them in a compact or menaingful form, amenable to interpretation by a human or computer (Wolpaw and Wolpaw)
+
+%Transient phenomena allows also to record occurrence and temporal sequence (mimicking spike analysis in neuro reserach)
+
+%The traditional approach do not consider waveform even though the brain oscilations are in general nonsinusoidal (reference)
+% Waveform characterization, on the other hand, has been extensively used in artifact detection and averaging.
+
 \section{Materials and Methods}
 
-%To verify the validity of the proposed framework and method, the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by  IRCCS Fondazione Santa Lucia, was used to perform a binary classification task on the provided signals.
-The algorithm was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
-
-In order to determine which processing method to be considered as a  "waveform template matching" we restrict ourselves to the following criteria:
+The following criteria is proposed to identify methods which are based on the signal's waveform:
 
 \begin{enumerate}
+\item The analysis take account the shape of the plot of the signal.
 \item The pattern can be identified and verified by visual inspection.
 \item The pattern matching is performed in time-domain.
-\item The analysis take account the shape of the plot of the signal.
-\item The feature extraction procedure allows to create a Template database.
-\item A Single Channel Processing approach.
+\item The method encompass a feature extraction procedure.
+\item The feature extraction procedure allows to create a template dictionary.
+\item A single-channel processing scheme.
 \end{enumerate}
@@
 As described in (Mixed Domain Signal Analysis) the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the pattern.
+All these algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
 
-\subsection{Selected Algorithms}
+\subsection{EEG Waveform Analysis Algorithms}
+
+Shape or waveform analysis methods are considered as nonparametric (in opposition to statistical or dynamical models).
+They explore signal's time-domain metrics or even derive more complex indexes from it \citep{Thakor2009}.
+
+One of the earliest approach to automatically process EEG data is the Peak Picking method.
+Although of limited usability, peak picking has been used to determine latency of transient events in EEG \citep{Jaskowski2000,Zhang2011}.
+Straightforward in its implementation, it consists in selecting a component, a simple component based on the expected location of its more prominent deflection \citep{Ouyang2017}.
+Of regular use in ERP Research, the name of many of the EEG features reference directly a peak within the component, e.g.
+P300 or P3a P3b or N100.
+This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.
+The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).
+
+A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.
+This was even used in the seminal work of Farwell and Donchin on P300 \citep{Farwell1988,WolpawJonathanR2012}.
+Additionally, a logarithmic graph of the peak-to-peak amplitude which is called amplitude integrated EEG (aEEG) \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.
+
+Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature from the amplitude and frequency of its signal and its derivative in time-domain is used.
+Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied, and finally the seminal proposals of Fujimori, Uchida and the PAA algorithm as few of the earliest proposals where the idea of capturing the shape of the signal were established.
+
+%Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.
+
+ %(fujimori, PAA, etc).  Uchida 1999
+
+%\subsubsection{Averaging Methods}
+%
+%The methods that allow to identified waveforms are used to determine different alignments while averaging epochs.
+%
+%This long-standing problem has been tackled from different perspectives.  Woody's Template Matching is perhaps the best effort as well as Pham, and Tuan and others EML.
+%
+%Dictionary, template based method.  Many do not consider the waveform, albeit they do use a set of templates obtained from a dictionary.
+%
+%Cross Covariance with Template or Cross Correlation or ACF: Autocorrelation Function: PHAM METHOD
+%
+%Applying a FIR filter with templates 
+%
+%Krusienski et al 2007
+%Serby et al 2005
+%
+%extended to wavelent analysis.
+%
+%Dynamic Time Warping and other Warp Averaging methods were also used.
+%
+%Wrapping Analysis
+%
+%Warp averaging
 
@@
  \item Merging of Increasing and Decreasing Sequences
+\item Local Binary Patterns (1-D LNBP, 1D-LBP and LBP)
 \end{itemize}
 
-Although the term morphology has been used to identify this approach, We specifically excluded Mathematical Morphological based methods \cite{Yamaguchi2009}.
-On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm.
-(cita a esa parte del paper).
+All these methods provide a feature that can be used as a template, whereas all of them are based on metrics that can be extracted from the shape of the signal.
+These features can be used to create dictionaries or templates databases.
+These templates provide the basis for the pattern matching algorithm and offline classification.
 
-\subsection{Excluded Methods}
+\subsection{Matching Pursuit - MP 1 and MP 2}
 
-\subsubsection{Waveform features}
+Pursuit algorithms refer, in their many variants, as blind source separation(REF) techniques that assume that the EEG signal is a linear combination of different sources that comes from template's dictionaries.
+The underlying idea is to find the best template out of a dictionary that matches against the signal.
+The algorithm works iteratively, starting from the raw signal and then, the template is substracted from the signal in Graham-Schmidt approach and the process iterates.
+(ref Cohen book, Sanei book, Mallat and Zhang 1993).
 
-One of the earliest approach to automatically process EEG data is the Peak Picking method.
-Although of limited usability, peak picking has been used to determine latency of transient events in EEG \citep{Jaskowski2000,Zhang2011}.
-Straightforward in its implementation, it consists in selecting a component, particularly a simple component based on the expected location of its more prominent deflection \citep{Ouyang2017}.
-Particularly common in ERP Research, the name of many of the EEG features directly reference a peak within the component, e.g.
-P300 or P3a P3b or N100.
-This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.
-The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).
+On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm. (cita a esa parte del paper).
 
-A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.
-This was even used in the seminal work of Farwell and Donchin on P300 \citep{Farwell1988,WolpawJonathanR2012}.
-Additionally, a logarithmic graph of the peak-to-peak amplitude or aEEG \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.
+\subsection{Permutation Entropy - PE}
 
-Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms  (automatic detection of k complexes), (A Waveform Analyzer Applied to the Human EEG) where a feature from amplitude and frequency of its signal and its derivative in time-domain is used.
+Bond and Pompe Permutation Entropy has been extensively used in EEG processing, particularly for Epilepsy pre-ictal detection.
+This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the amount of entropy of each code within the signal.
+For instance, a pure random signal, will achieve the maximum entropy value due to the probability of each code being equal for all of them.
+This method is the best representative of Waveform Complexity: Lempel-Ziv method L-Z complexity are other variants which use a different definition (Permutation Entropy a new feature for brain computer interfaces).
 
-Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.
+\subsection{Slope Horizontal Chain Code - SHCC}
 
- (fujimori, PAA, etc).  Uchida 1999
+This method is an extension of Slope Chain Code (SCC) (ref a SCC paper y a SHCC paper), and a code of a sequence of sample points is generated based on the angle between the horizontal segment and any segment, one by one.
+This method can be encompass as a syntactic analysis technique, where the EEG segment is represented as a series of elementary patterns, similar to tokens, due to the quantization of angles.
 
-\subsubsection{Averaging Methods}
+\subsection{Histogram of Gradients Orientations - HGO}
 
-The methods that allow to identified waveforms are used to determine different alignments while averaging epochs.
+This Histogram of Gradient Orientations is based on using Computer Vision techniques to extract directly information from a signal based on the plot of a signal on a 2D image.
+This mimicks exactly what a person is doing by visually inspecting the plot, and the waveform.
+To do that, the region of the image where the signal is located is divided in a $4 x 4$ block and the signal bidimensional gradients is calculated.
+For each block within the patch, a histogram of gradient orientations, 8 circular orientations, is calculated.
+This histogram is concatenated for all the 16 blocks and a feature is thus formed.
+The details of the method can be found here \citep{Ramele2016}.
+This method was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries.
 
-This long-standing problem has been tackled from different perspectives.  Woody's Template Matching is perhaps the best effort as well as Pham, and Tuan and others EML.
+%\subsection{Merger of Increasing and Decreasing Sequences}
 
-Dictionary, template based method.  Many do not consider the waveform, albeit they do use a set of templates obtained from a dictionary.
-
-Cross Covariance with Template or Cross Correlation or ACF: Autocorrelation Function: PHAM METHOD
-
-Applying a FIR filter with templates
-
-Krusienski et al 2007
-Serby et al 2005
-
-extended to wavelent analysis.
-
-Dynamic Time Warping and other Warp Averaging methods were also used.
-
-Wrapping Analysis
-
-Warp averaging
-
-\subsection{Matching Pursuit}
-
-Pursuit algorithms refer, in their many variants, as blind source separation techniques that assume that the EEG signal is a linear combination of different sources that comes from template dictionaries.
-
-We include them here because they are traditional in terms for the traditional view of Morphological Analysis or Shape domain analysis, though they do not restrict to our own definition of "shape domain processing".
-
-However in terms of
-
-\subsection{Permutation Entropy}
-
-Bond and Pompe PE method.
-
-This method is the best representative of Waveform Complexity: Lempel-Ziv method L-Z complexity are other variants which use a different definition.
-
-Permutation Entropy a new feature for brain computer interfaces.
-
-\subsection{Slope Horizontal Chain Code}
-
-contour representation based on an adapted version of the Slope Chain Code (SCC)
-
-This method can be encompass as a syntactic analysis technique, where the EEG segment is represented as a series of elementary patterns, similar to tokens.
-
-\subsection{Histogram of Gradients}
-
-\subsection{Merger of Increasing and Decreasing Sequences}
-
-Stepwise downsampling
+%Stepwise downsampling
 
@@
 
-In order to verify each one of the approaches we will inject into a known dataset spurious p300 signal that we may be able to control.
-By implementing this pseudo-real approach, we can effectively control null-signals of this evoked potentials following the same approach of other similar works ( ref   Ouyang2017, el trabajo de estudio de templates y el trabajo de p3 y la tesis).
+Farwell and Donchin P300 Speller \citep{Farwell1988} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing in a virtual keyboard.
+This procedure exploits a cognitive phenomena called oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
+This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.
 
-Jaskowski and Verleger (2000)
+Higher accuracy algorithms can detect it in a good-enough single-trial approach [P300 Summary], but in general these algorithms are not related to the shape of the P300 ERP.
+Hence it is difficult to establish a backward mapping back to a certain visual pattern within the waveform.
 
-At the same time adding a mixed signal with varying degrees of SNR will allow us to determine robustness of the different methods.
+In order to verify each algorithm in a controlled procedure, a real ERP template is superimposed into a real EEG stream.
+This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on anyone in particular.
+Everything is there, except the P300 ERP component.
+By implementing this pseudo-real approach, it is possible to effectively control null-signals and the shape of this evoked potential following the same approach of other similar works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
 
-We will extend this results to offline processing a public dataset of real P300 patterns.
+The experiments are as follows:
 
-The original signal-to-noise ratio was calculated as Hue 2010.
+\begin{itemize}
+\item Pseudo-real dataset Classification Performance: the letter classification performance of each one of these methods.
+\item Latency Noise: Gaussian noise is added to the latency where the ERP template is injected into the EEG stream.
+\item Component Amplitude Noise: the different components of the ERP template are altered to make them less distinguishable.
+\end{itemize}
+
+Finally results for the same methods while offline processing a public dataset of real P300 patterns are also shown.
+
+\subsubsection{Pseudo-Real Dataset Generation}
+
+The template ERP is extracted from the Subject $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+Segments from the EEG signal are labeled as hit and are extracted for the trial number $2$, and they are point-to-point coherently averaged.
+This P300 ERP can be seen in Fig.
+\ref{fig:erptemplate1}.
+
+An EEG stream with null-P300 signal everywhere is obtained by the following procedure: 
+The participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
+No monetary compensation is handed out and he/she agree and sign a written informed consent.
+This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.
+The participant is healthy and have normal or corrected-to-normal vision and no history of neurological disorders.
+This voluntary is of unspecified gender, aged between 20-30 years old.
+EEG data is collected in a single recording session.
+She/He is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of him.
+The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.
+Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.
+Reference is set to the right ear lobe and ground is preset as the AFz position.
+Sampling frequency is set to 250 Hz, which is the closest possible to the one used with the other dataset.
+
+The participant is instructed to passively watch the flashing screen while not focusing on any particular letter.
+A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.
+
+%This P300 Speller protocol consist in the flashing of 35 trails of 35 letters (7 words of 5 letter) where the intensification sequence of 6 rows and 6 columns is repeated 10 times for each letter.
+More details can be found on the published work of \citep{Riccio2013}.
+
+Fig.
+\ref{fig:gaincheck} shows the result of superimposing the template signal into EEG stream, time-locked to the stimulus onset.
+These 12 point-to-point averaged segments correspond to the first trial of the EEG stream.
+
+%The original signal-to-noise ratio was calculated as Hue 2010.
 
@@
 \centering
-\includegraphics[width=6cm]{images/erptemplate1.eps}
-\caption{ERP Template obtained from the point-to-point averaging of a P300 component of a subject obtained from the public dataset.}
+\includegraphics[width=12cm]{images/erptemplate1.eps}
+\caption{ERP Template obtained from the coherent point-to-point ensemble average of subject eight obtained from the BNCI Horizon public dataset. The P1, N1 and P300 peaks can be seen.}
 \label{fig:erptemplate1}
@@
 
-\subsubsection{Classification}
+\begin{figure}[H]
+\centering
+\includegraphics[width=10cm]{images/doublegain.eps}
+\includegraphics[width=10cm]{images/triplegain.eps}
+\caption{8-Channel EEG signal superimposed with the ERP Template.  Left show the signal with single gain and right the same signal with double gain.}
+\label{fig:doubleandtriplegain}
+\end{figure}
 
-In order to classify we will use the same classifier for every method.  By doing this we will avoid
+\begin{figure}[H]
+\centering
+\includegraphics[width=15cm]{images/GainCheck.eps}
+\caption{These are the point-to-point averaged signals for the first trial.  The ERP was injected on class 3 and 9.}
+\label{fig:gaincheck}
+\end{figure}
 
-\subsubsection{Parameters}
+\subsubsection{Classification} \label{section:classification}
 
-Each method has its own set of parameters which are optimized.
-We will try to use an approach used in (matching pursuit paper for bci) where the best parameters are those which provide the best performance.
+To analyze how each method identifies the hidden P300 signal, the same classification algorithm is used.
+The original dataset is composed of 35 trials to decode 7 words of 5 letters  of the P300 matrix.
+Each trial is composed of 10 intensification sequences of the 6 columns and 6 rows \citep{Riccio2013}.
+The averaging procedure aims to extract the superimposed ERP Template and cancel out everything else on each segment.
 
-Hence, in order to obtain the right set of parameters for each method, an optimization problem was used fixing the injected signal to a gain of 3, and optimizing the parameters.
+This classification procedure identifies the row and column, matching to the template $T$ by computing
+
+\begin{equation}
+\hat{row} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q \in NN_T(f^{row}_u)}^{} \left\lVert q -  f^{row}_u \right\rVert ^2
+\label{eq:multiclassificationrow}
+\end{equation}
+
+\noindent and
+
+\begin{equation}
+\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q \in NN_T(f^{col}_u)}^{} \left\lVert q -  f^{col}_u \right\rVert ^2
+\label{eq:multiclassificationcol}
+\end{equation}
+
+\noindent where $NN_T(f^l_u),\;l\in\{row,{col}\}$  is the set of the $k$ nearest neighbors to $f^l_u$ and $q$ is a template feature that belongs to it.
+This set is obtained by sorting all the elements in the dictionary $T$ based on the distances between them and $f^l_u$, choosing the $k$ smaller elements.
+This procedure is a modification of the k-NBNN  algorithm~\citep{Boiman2008}.
+
+By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{row} $ and column $ \hat{col} $.
+%Figure~\ref{fig:classification} shows a scheme of this process.
+
+%\subsubsection{Parameters}
+
+%The parameters for each method were derived from bibliography were they are introduced
+
+%Each method has its own set of parameters. To define them an optimization problem was defined and the set of values for each method that achieve the maximum performance value for
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=15cm]{images/dictionary.eps}
+\caption{Coherently averaged signals containing the superimposed ERP.
+Each one is extracted from the 15 first trials (2 signals from each trial, one belonging to the column and the other to the row).
+These are the templates that are used by the classification algorithm described in \ref{section:classification}.}
+\label{fig:amplitudenoise}
+\end{figure}
 
@@
 
-Results are shown in Table 1 and in Figure 1.
-Table 1 shows the performance to identify each letter of the standard P300 Speller Matrix.
-Figure 2 shows how the performance increase when averaging is implemented.
+Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.
+Table \ref{section:results} shows the single-trial performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was obtained.
+Figure \ref{fig:performancetest} shows the performance curves for eight algorithms.
+Each one represents the percentage of letters that were actually predicted by the combination of feature plus a classification method.
+The data were divided in two, and the first 15 letters were used to derive the dictionary of templates $T$ while the remaining 20 letters were used to measure the classification accuracy.
+Figure \ref{fig:performancetestlatency} shows the same results obtained when a random uniform latency shift was added to each trial.
+Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained when the amplitude of the N1 and P3 component of the template are randomly reduced (uniform distribution).
 
 \begin{table}[H]
-\caption{Accuracy levels obtained by a 3-fold cross validation.
-The values reported by the dataset publishers for Cz are reproduced here for comparison.
-Additionally, BCI accuracies for channel Cz can be seen as well as performance levels and their standard deviation obtained for the BPC, the best performing channel for each subject.}
+\caption{Classification performance obtained for all the waveform-based algorithms.
+All the methods process the signal on a channel-by-channel basis, so the best performing channel is also shown.
+It can be understood as the channel that adds less noise to the ERP template.
+All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal that was used to obtain the feature that was finally classified.}
 \centering
 %% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{cc}
+\begin{tabular}{ccc}
 \toprule
-\textbf{Method}	& \textbf{Single Trial Performance}	\\
+\textbf{Method}	& \textbf{Channel} & \textbf{Single Trial Performance}	\\
 \midrule
-MP & 90 \\
-PE & 89 \\
-SHCC & 20 \\
+MP 1 & PO8 & $25\%$ \\
+MP 2 & PO7 & $10\%$ \\
+HGO  & PO8 & $25\%$ \\
+PE     & Cz & $20\%$ \\
+SHCC & P4 & $10\%$ \\
+RS     & Fz & $15\%$ \\
+RS m-c &   & $15\%$ \\
 \bottomrule
@@
 \centering
-\includegraphics[width=6cm]{images/performance.eps}
+\includegraphics[width=15cm]{images/PerformanceTest.eps}
 \caption{Performance obtained from the different methods.}
-\label{fig:performance}
+\label{fig:performancetest}
 \end{figure}
@@
 \centering
-\includegraphics[width=6cm]{images/P300Performance.eps}
-\caption{Performance obtained from the different methods while averaging different trials.}
-\label{fig:p300performance}
+\includegraphics[width=15cm]{images/PerformanceTestLatency.eps}
+\caption{Performance obtained for all the methods.  Latencies were artificially added to each single-trial segment.  Performance is significantly reduced for all methods.}
+\label{fig:performancetestlatency}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=15cm]{images/PerformanceTestAmplitude.eps}
+\caption{Performance obtained for all the methods.  The amplitudes of the N1 and P3 component of the superimposed ERP are randomly reduced.}
+\label{fig:performancetestamplitude}
 \end{figure}
@@
 
-In spite of all this, the conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.
+Findings
+
+The shape of the signal is absolutely related to the single trial latency.
+However the methods are more robust to changes in peak amplitudes.
+
+The multichannel approach provides better information, except for the raw signal.
+
+The conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.
 The need for more objective measurements pushed the adoption of more automated means of decoding the signals and demanded a need for replication \citep{Thakor2004}.
@@
 EKG, or Electrocardiogram, on the other hand, has been extensively processed and analyzed by waveform methods.
-P,Q,R,S,T peaks in EKG analysis (REFERENCE)
+P,Q,R,S,T peaks in EKG analysis (REFERENCE).  Additionally, the same problem can be found in Intelligent Character Recognition.
 
-Dimensionality reduction which is truly applied when the waveform is actually formed.
+Dimensionality reduction which is truly applied when the waveform is actually formed (curse of dimensionality).
 
-We left out the segmetation issue which is itself a huge problem.  Where should I look from within the EEG trace ?
+Segmetation or identification is not considered.  Where should I look from within the EEG trace ?  This is clear the problem of detection and not feature extraction.
 
@@
 
-The purpose of this work was twofold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to verify and compare those methods.
+The purpose of this work was twofold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain classification accuracies based exclusively on the signal's waveform.
 
@@
 
-Berger H: tiber das Elektrenkephalogramm des Menschen, Archiv Psychiatr, 87 (1929) pp. 527-570.
+%Berger H: tiber das Elektrenkephalogramm des Menschen, Archiv Psychiatr, 87 (1929) pp. 527-570.
 

### 2018-05-02 sensors.tex
First manuscript done version

@@
 When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".
-\item Hypersynchrony.
-Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range.
-The term implies an increase in the number of neural elements contributing to the rhythm.
-\item Paroxysmal.
-Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity.
+\item Hypersynchrony: Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range.
+The term suggest an increase in the number of neural elements contributing to the rhythm.
+\item Paroxysmal: Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity.
 Though the term does not directly imply abnormality, much abnormal activity is paroxysmal.
@@
 Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature from the amplitude and frequency of its signal and its derivative in time-domain is used.
-Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied, and finally the seminal proposals of Fujimori, Uchida and the PAA algorithm as few of the earliest proposals where the idea of capturing the shape of the signal were established.
+Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied, and finally the proposals of Fujimori, Uchida and the PAA algorithm as few of the earliest proposals where the idea of capturing the shape of the signal were established.
 
@@
 
-On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm. (cita a esa parte del paper).
+Two variants of this algorithm are evaluated.
+In \textit{MP 1} the dictionary is directly constructed with the normalized templates from Fig \ref{fig:dictionaryfig}.
+In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.
+To construct each feature a signal is decomposed in its coefficients and then they can be compared by any distance function.
+
+%On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm. (cita a esa parte del paper).
 
@@
 \centering
-\includegraphics[width=10cm]{images/doublegain.eps}
-\includegraphics[width=10cm]{images/triplegain.eps}
-\caption{8-Channel EEG signal superimposed with the ERP Template.  Left show the signal with single gain and right the same signal with double gain.}
+\includegraphics[width=5cm]{images/doublegain.eps}
+\includegraphics[width=5cm]{images/triplegain.eps}
+\caption{8-Channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the superimposed template.}
 \label{fig:doubleandtriplegain}
@@
 Each one is extracted from the 15 first trials (2 signals from each trial, one belonging to the column and the other to the row).
-These are the templates that are used by the classification algorithm described in \ref{section:classification}.}
-\label{fig:amplitudenoise}
+These are the templates used to build the dictionary $T$ and that are used by the classification algorithm described in \ref{section:classification}.}
+\label{fig:dictionaryfig}
 \end{figure}
@@
 Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.
-Table \ref{section:results} shows the single-trial performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was obtained.
+Table \ref{tab:results} shows the single-trial performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was obtained.
 Figure \ref{fig:performancetest} shows the performance curves for eight algorithms.
@@
 The data were divided in two, and the first 15 letters were used to derive the dictionary of templates $T$ while the remaining 20 letters were used to measure the classification accuracy.
-Figure \ref{fig:performancetestlatency} shows the same results obtained when a random uniform latency shift was added to each trial.
+Figure \ref{fig:performancetestlatency} shows the same results obtained when a Gaussian latency shift was added to each trial.
 Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained when the amplitude of the N1 and P3 component of the template are randomly reduced (uniform distribution).
@@
 \begin{table}[H]
-\caption{Classification performance obtained for all the waveform-based algorithms.
-All the methods process the signal on a channel-by-channel basis, so the best performing channel is also shown.
+\caption{Classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, HGO Histogram of Gradient Orientation, PE Permutation Entropy, SHCC Slope Horizontal Code Chain, RS Raw Signal, NN Neural Network, SVM Support Vector Machines.
+All the methods that process the signal on a channel-by-channel basis, so the best performing channel is also shown.
+The mark $m-c$ is used on the control algorithms to identify that the feature was constructed using all the channels.
 It can be understood as the channel that adds less noise to the ERP template.
@@
 RS m-c &   & $15\%$ \\
+NN s-c & Fz & $10\%$ \\
+NN m-c & Cz & $45\%$ \\
+SVM m-c & Cz & $50\%$ \\
 \bottomrule
@@
 
-Findings
+The classification methods Support Vector Machine SVM and Artificial Neural Networks NN are added for comparison using the feature of Kasinksky (REF).
+Additionally, the Kasinksky feature is used with the same classification algorithm as detailed in \ref{section:classification}.
 
-The shape of the signal is absolutely related to the single trial latency.
-However the methods are more robust to changes in peak amplitudes.
+All the methods show a significant reduction of performance when the latency noise is added, and at the same time all the methods show resistance to noise in peak amplitudes.
 
-The multichannel approach provides better information, except for the raw signal.
+Except for the raw signal, using a multichannel feature provides a better synthesis of the differentiation information that provides a better accuracy.
+This can be seen in Fig \ref{fig:performancetest} where the performance of the control algorithm NN is improved by using a multichannel feature.
+The poor raw signal performance while using a multichannel feasture is likely to the curse of dimensionality due to the fact that the feature is simple too large.
+
+Using a straightforward dictionary of templates for MP-1 proved more beneficial in terms of performance than the standard approach of using a Hilbert base of Wavelets atoms.
+
+The methods \textit{HGO} and \textit{SHCC} performed much better than the other algorithms including when the latency and the amplitudes are noisy.
+
+It is important to remark that Segmentation or identification is not considered in these experiments.
+The  problem of segmentation is related to the detection of a feature within an EEG trace and not to the feature extraction problem.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
 
@@
 
-This approach is relative common in chemical analysis (i.e.
+The purpose of this work was threefold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain classification accuracies based exclusively on the signal's waveform.
+
+One of the main goals of the BCI discipline is to provide assistance to patients and to provide alternative tools to be used in diagnostics and rehabilitation procedure.
+This requires a clinical focus which is often neglected in BCI research (ref).
+At the same time, BCI reliability and the avoidance of the black box phenomena are goals that are still unfulfilled in this discipline(WOkpaw and wolpaw).
+The benefits of analysing or including metrics about the shape of the EEG, are that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can lead to more robust implementation of BCI devices.
+
+Successful approaches in Computer Vision or pattern recognition in other areas use a set of different features to compound ensemble classifiers.
+All these methods had the advantage that they can truly map a visual component with a clinical meaning, assessed by a physician, and provide an automatic identification which can be constructed by the set of methods.
+
+More work has to be conducted in order to extend this methods to other patterns of waves.
+
+The procedure of analysing signals by their waveforms is relative common in chemical analysis (i.e.
 chemometrics Skoog, D.; West, D.; Holler, F.
 Analtyical Chemistry, An Introduction; Saunders: Philadelphia, 1994.), geology (sismic analysis), and quantitive financial analysis.
-EKG, or Electrocardiogram, on the other hand, has been extensively processed and analyzed by waveform methods.
-P,Q,R,S,T peaks in EKG analysis (REFERENCE).  Additionally, the same problem can be found in Intelligent Character Recognition.
-
-Dimensionality reduction which is truly applied when the waveform is actually formed (curse of dimensionality).
-
-Segmetation or identification is not considered.  Where should I look from within the EEG trace ?  This is clear the problem of detection and not feature extraction.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-
-The purpose of this work was twofold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain classification accuracies based exclusively on the signal's waveform.
-
-Invasive methods remains an issue (referencia del paper de nature que muestra como causan deficits los implantes)
-
-BCI as Assistive technology and BCI supplement natural motor output
-The value of clinical focus, 
-BCI reliability (WOkpaw and wolpaw).
-Clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can lead to more robust implementation of BCI devices.
-
-Let us now turn to the question posed in the title of this paper.
-
-Successful approaches in Computer Vision or pattern recognition in other areas use a set of different features to compount ensemble classifiers.  All these methods had the advantage 
-that they can truly map a visual component with a clinical meaning, assessed by a physician, and provide an automatic identification which can be constructed by the set of methods.
-We believe this approach can be used to provide tools which are more meaningful to the clinical set.
-
-Particularly considering BCI, this approach, in line with the general idea, can be used to foster BCI solutions that can help clinical settings.  Precisely for the same reasons.
-
-More work has to be conducted in order to extend this methods to other patterns of waves.
+EKG, or Electrocardiogram, on the other hand, has been extensively processed and analysed by waveform methods.
+Additionally, the same problem can be found in Intelligent Character Recognition.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{This project was supported by the ITBACyT-15 funding program issued by ITBA University.}
+\acknowledgments{This project was supported by the ITBACyT-15 funding program issued by ITBA University.
+We would like to thank to Dr.
+Jian Zhang for providing details of the MIDS algorithm, to Dr.
+Valentina Unakafova for providing the Permutation Entropy algorithm and to Dr.
+Montserrat-Alvarado González for providing the source code and a detailed description of the SHCC algorithm.}
 

### 2018-06-11 sensors.tex
Added better images

@@
 %=================================================================
-\documentclass[sensors,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
+\documentclass[brainsci,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
 
@@
 Concurrently, the digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices \citep{Domingo2012}.
-All this communication is based on muscular movement \citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human movement.
+All this communication is based on muscular movement \citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human motion.
 A new form of Human Machine communication which directly connects the Central Nervous System (CNS) to a machine or computer device is currently being developed: Brain Machine Interfaces (BMI), Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).
@@
 
-In contrast, automatic processing, or quantitative EEG, was based first on analog electronic devices and later on computerized digital processing methods \citep{Jansen1991}.
-They implemented mathematically and algorithmically complex procedures to decode the information with good results \citep{Yuste2017}.
+In contrast, automatic processing, or quantitative EEG (qEEG), was based first on analog electronic devices and later on computerized digital processing methods \citep{Jansen1991}.
+They implemented mathematically and algorithmically complex procedures to decode the information associated with a clinical condition, or even some higher cognitive state \citep{Yuste2017}.
 The best materialization of the automatic processing of EEG signals rests in the BCI discipline, where around $71.2\%$ is based on noninvasive EEG \citep{Guger2017}.
@@
 
-This article unfolds as follows: Section 1 will provide a brief introduction to EEG characterization and the particularities of the EEG waveform characterization.
-Section 3 will explain the algorithms based on the waveform that will be used to perform the benchmark.
-Section 4 the experimentation procedure will be explained.
-Results will be presented in Section 5 and finally Discussion and Conclusions will be established in the final sections.
+This article unfolds as follows: Section \ref{EEG} will provide a brief introduction to EEG and the particularities of the EEG waveform characterization.
+Section \ref{Algorithms} will explain the algorithms based on the waveform that will be analyzed.
+In Section \ref{Experimental} the experimentation procedure will be explained.
+Results will be presented in Section \ref{section:results} and finally Discussion and Conclusions will be established in the final sections.
 
 \section{Electroencephalography}
+\label{EEG}
 
 The Electroencephalography consists on the measurement of small variations of electrical currents over the scalp.
-It is one of the most widespread used method to capture brain signals and was initially developed by Hans Berger in 1924 and has been extensively used for decades to diagnose neural diseases and other medical conditions.
+It is one of the most widespread used methods to capture brain signals and was initially developed by Hans Berger in 1924 and has been extensively used for decades to diagnose neural diseases and other medical conditions.
 
@@
 \centering
-\includegraphics[height=6cm,width=12cm]{images/eegsample.png}
-\caption{Sample EEG signal captured and plotted on a 2D image using the popular and commercial EPOC Emotiv using EEGLAB.}
+\includegraphics[height=6cm,width=12cm]{images/sampleeeg.eps}
+\caption{Sample EEG signal obtained from g.Tec g.Nautilus.  Time axis is in seconds and five seconds are displayed.  The eight channels provided by this device are shown.}
 \label{fig:sampleeeg}
 \end{figure}
+
+%locations.eps
 
@@
 \item Artifacts:  These are signal sources which are not generated from the CNS, but can be detected from the EEG signal.
-These are called endogeneous or physiological when they are generated from a biological source like muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic noise\citep{Weeda2012}.
+They are called endogeneous or physiological when they are generated from a biological source like muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic noise\citep{Weeda2012}.
 \item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e.
@@
 DC drift means that this center value drifts in time.
-Although sometimes considered as a nuisance that needs to get rid of, it is known that very important cognitive phenomena like Slow Cortical Potentials or Slow Activity Transients in innates do affect the drift and can be used to understand some particular brain functioning \cite{Schomer2010}.
+Although sometimes considered as a nuisance that needs to get rid of, it is known that very important cognitive phenomena like Slow Cortical Potentials or Slow Activity Transients in infants do affect the drift and can be used to understand some particular brain functioning \cite{Schomer2010}.
 \item Basal EEG activity: the EEG is the compound summation of myriads of electrical sources from the CNS.
@@
 \item Spontaneous: generally treated as noise or basal EEG.
-\item Evoked: activity that is time-locked to an incoming stimulus or an executed motor action.  In contrast it is often called Induced activity
+\item Evoked: activity that can be detected synchronously after some specific amount of time after the onset of the stimulus.
+This is usually referred as time-locked.
+In contrast to the previous one, it is often called Induced activity.
 \end{itemize}
@@
 \begin{itemize}
-\item Rhythmic: EEG activity consisting in waves of approximately constant frequency.  It is often abbreviated RA (regular rythmic activity).
+\item Rhythmic: EEG activity consisting in waves of approximately constant frequency.
+It is often abbreviated RA (regular rythmic activity).
+They are loosely classified by their frequencies, and their naming convention was derived from the original naming used by Hans Berger himself, and after Alpha Waves (10 Hz), it came Delta (4 Hz), Theta (4-7 Hz), Sigma (12-16 Hz), Beta (12-30 Hz) and Gamma (30-100 Hz).
 \item Arrhythmic: EEG activity in which no stable rhythms are present.  
-\item Dysrhythmic: Rhythms and/or patterns of EEG activity that characteristically appear in patient groups or rarely or seen in healthy subjects.
+\item Dysrhythmic: Rhythms and/or patterns of EEG activity that characteristically appear in patient groups and rarely seen in healthy subjects.
 \end{itemize}
@@
 
-The traditional clinical approach consisted in analyzing the paper strip that was generated by the plot of the signal obtained from the device.
-Expert technician and physicians analyzed visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.
+The traditional clinical approach consists in analyzing the paper strip that is generated by the plot of the signal obtained from the device.
+Expert technician and physicians analyze visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.
 Atlases and guidelines were created in order to help in the recognition of these complex patterns.
@@
 It is regular to establish an amplitude difference between different waveforms from which a relation between them is established and a structured index are created (e.g.
-sleep K-Complex is well characterized based on rates between positive vs negative amplitude) (REFERENCE).
-Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, K-complexes, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).
+sleep K-Complex is well characterized based on rates between positive vs negative amplitude) \citep{Uchida1999}.
+Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).
 
@@
 %Finally, there are specific EEG patterns that can be used to determine Depth of anestisia.
-
-%Evoked Potentials (EPs) and Event Related Potentials (ERPs) are in general characterized by the waveforms.
 
@@
 
+The search of methods based on waveforms is conducted by following the PRISMA~\citep{Moher2009} guidelines.
+Search is performed on Google Scholar, Semantic Web and IEEE Xplore search engines by the terms "Waveforms" OR "Shape" OR "Morphology" OR "Visual inspection" + "EEG".
+
 The following criteria is proposed to identify methods which are based on the signal's waveform:
@@
 
-As described in (Mixed Domain Signal Analysis) the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the pattern.
+As described in \citep{allen2004signal} the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the prototype signal.
 All these algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
@@
 \subsection{EEG Waveform Analysis Algorithms}
+\label{Algorithms}
 
@@
 Straightforward in its implementation, it consists in selecting a component, a simple component based on the expected location of its more prominent deflection \citep{Ouyang2017}.
-Of regular use in ERP Research, the name of many of the EEG features reference directly a peak within the component, e.g.
+Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient component that may arise as a brain response to an external visual, tactile or auditory stimulus.
+Particularly, EPs are regularly used to assess auditory response in infants.
+They are precisely characterized in this manner, where the name of many of the EEG features evoke directly a peak within the component, e.g.
 P300 or P3a P3b or N100.
-This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.
+This leads to a natural procedure to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.
 The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).
@@
 Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature from the amplitude and frequency of its signal and its derivative in time-domain is used.
-Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied, and finally the proposals of Fujimori, Uchida and the PAA algorithm as few of the earliest proposals where the idea of capturing the shape of the signal were established.
+Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied.
+Finally the proposals of Fujimori, Uchida and the PAA \citep{Uchida1996} algorithm are few of the earliest proposals where the idea of capturing the shape of the signal were established.
 
@@
 \item Histogram of Gradient Orientations
- \item Merging of Increasing and Decreasing Sequences
-\item Local Binary Patterns (1-D LNBP, 1D-LBP and LBP)
+%\item Merging of Increasing and Decreasing Sequences
+%\item Local Binary Patterns (1-D LNBP, 1D-LBP and LBP)
 \end{itemize}
@@
 All these methods provide a feature that can be used as a template, whereas all of them are based on metrics that can be extracted from the shape of the signal.
-These features can be used to create dictionaries or templates databases.
+These features can be used to create dictionaries or template databases.
 These templates provide the basis for the pattern matching algorithm and offline classification.
@@
 \subsection{Experimental Protocol}
+\label{Experimental}
 
@@
 
-Higher accuracy algorithms can detect it in a good-enough single-trial approach [P300 Summary], but in general these algorithms are not related to the shape of the P300 ERP.
+%Algorithms based on qEEG can detect this component in a good-enough single-trial approach [P300 Summary], but at the same time, these algorithms do not establish a clear relationship with the P300 ERP.
 Hence it is difficult to establish a backward mapping back to a certain visual pattern within the waveform.
@@
 Everything is there, except the P300 ERP component.
-By implementing this pseudo-real approach, it is possible to effectively control null-signals and the shape of this evoked potential following the same approach of other similar works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
+By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to a similar procedure used in similar works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
 
@@
 \includegraphics[width=12cm]{images/erptemplate1.eps}
-\caption{ERP Template obtained from the coherent point-to-point ensemble average of subject eight obtained from the BNCI Horizon public dataset. The P1, N1 and P300 peaks can be seen.}
+\caption{ERP Template obtained from the coherent point-to-point ensemble average of subject Eight obtained from the BNCI Horizon public dataset.
+The template is $1s$ long which is 256 sample points, and the eight channels are superimposed with different colors.
+The P1, N1 and P300 peaks can be seen.}
 \label{fig:erptemplate1}
@@
 \centering
-\includegraphics[width=5cm]{images/doublegain.eps}
-\includegraphics[width=5cm]{images/triplegain.eps}
-\caption{8-Channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the superimposed template.}
+\includegraphics[width=10cm]{images/doublegain.eps}
+\includegraphics[width=10cm]{images/triplegain.eps}
+\caption{Eight-channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the added template.}
 \label{fig:doubleandtriplegain}
@@
 \begin{table}[H]
-\caption{Classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, HGO Histogram of Gradient Orientation, PE Permutation Entropy, SHCC Slope Horizontal Code Chain, RS Raw Signal, NN Neural Network, SVM Support Vector Machines.
-All the methods that process the signal on a channel-by-channel basis, so the best performing channel is also shown.
-The mark $m-c$ is used on the control algorithms to identify that the feature was constructed using all the channels.
-It can be understood as the channel that adds less noise to the ERP template.
-All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal that was used to obtain the feature that was finally classified.}
+\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, HGO Histogram of Gradient Orientation, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
+Additionally, three control algorithms were included: RS Raw Signal, NN Neural Network and SVM Support Vector Machines.
+All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
+The mark \textit{m-c} is used on the control algorithms to identify that the feature was constructed using all the channels.
+In this case with absence of null-signals, it can be interpreted as the channel that adds less noise to the ERP template.
+All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal.
+}
 \centering
@@
 \midrule
+SVM     & m-c  & $50\%$ \\
+NN       & m-c  & $45\%$ \\
+HGO  & PO8 & $25\%$ \\
 MP 1 & PO8 & $25\%$ \\
+PE     & Cz & $20\%$ \\
+RS     & Fz & $15\%$ \\
+RS     & m-c  & $15\%$ \\
 MP 2 & PO7 & $10\%$ \\
-HGO  & PO8 & $25\%$ \\
-PE     & Cz & $20\%$ \\
 SHCC & P4 & $10\%$ \\
-RS     & Fz & $15\%$ \\
-RS m-c &   & $15\%$ \\
 NN s-c & Fz & $10\%$ \\
-NN m-c & Cz & $45\%$ \\
-SVM m-c & Cz & $50\%$ \\
 \bottomrule
@@
 \includegraphics[width=15cm]{images/PerformanceTest.eps}
-\caption{Performance obtained from the different methods.}
+\caption{Speller performance obtained for each method.
+Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
 \label{fig:performancetest}
@@
 \includegraphics[width=15cm]{images/PerformanceTestLatency.eps}
-\caption{Performance obtained for all the methods.  Latencies were artificially added to each single-trial segment.  Performance is significantly reduced for all methods.}
+\caption{Speller performance obtained for each method while latencies were artificially added to each single-trial segment.  The achieved performance is significantly reduced for all methods.}
 \label{fig:performancetestlatency}
@@
 \includegraphics[width=15cm]{images/PerformanceTestAmplitude.eps}
-\caption{Performance obtained for all the methods.  The amplitudes of the N1 and P3 component of the superimposed ERP are randomly reduced.}
+\caption{Speller performance obtained while the amplitudes of the N1 and P3 component of the superimposed ERP are randomly reduced.}
 \label{fig:performancetestamplitude}
@@
 \section{Discussion}
+\label{section:discussion}
 
@@
 \section{Conclusion}
+\label{section:conclusion}
 
@@
 The need for more objective measurements pushed the adoption of more automated means of decoding the signals and demanded a need for replication \citep{Thakor2004}.
-This lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread.
+This lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}
 
@@
 We would like to thank to Dr.
-Jian Zhang for providing details of the MIDS algorithm, to Dr.
 Valentina Unakafova for providing the Permutation Entropy algorithm and to Dr.
 Montserrat-Alvarado González for providing the source code and a detailed description of the SHCC algorithm.}
+
+%We would like to thank to Dr. Jian Zhang for providing details of the MIDS algorithm, to Dr. Valentina Unakafova for providing the Permutation Entropy algorithm
 

### 2018-08-13 sensors.tex
Added some equations.

@@
 
+\begin{table}[H]
+\caption{Subjective EEG characterization methods found in the surveyed literature.}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+\begin{tabular}{ccc}
+\toprule
+\textbf{Method}	& \textbf{Phenomena} & \textbf{Reference}	\\
+\midrule
+SVM     & m-c  & $50\%$ \\
+NN       & m-c  & $45\%$ \\
+HGO  & PO8 & $25\%$ \\
+MP 1 & PO8 & $25\%$ \\
+PE     & Cz & $20\%$ \\
+RS     & Fz & $15\%$ \\
+RS     & m-c  & $15\%$ \\
+MP 2 & PO7 & $10\%$ \\
+SHCC & P4 & $10\%$ \\
+NN s-c & Fz & $10\%$ \\
+\bottomrule
+\end{tabular}
+\label{tab:results}
+\end{table}
+
 \section{Materials and Methods}
@@
 
-Pursuit algorithms refer, in their many variants, as blind source separation(REF) techniques that assume that the EEG signal is a linear combination of different sources that comes from template's dictionaries.
-The underlying idea is to find the best template out of a dictionary that matches against the signal.
-The algorithm works iteratively, starting from the raw signal and then, the template is substracted from the signal in Graham-Schmidt approach and the process iterates.
-(ref Cohen book, Sanei book, Mallat and Zhang 1993).
+\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Comon2010} techniques that assume that the EEG signal is a linear combination of different sparse sources extracted from a template's dictionaries.
+\textit{MP}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.
+Given a signal, this optimization technique, tries to find the indexes of atoms and their weights (contributions) that minimize,
+
+\begin{equation}
+\varepsilon =  \left\lVert   x(n,c) - \sum_{i=1}^{m} w_i g_{yi}   \right\rVert
+\label{eq:mperror}
+\end{equation}
+
+which is the error from the signal approximation constructed by the weighted atoms.  The algorithm goes by first setting the approximation signal by the original signal itself,
+
+\begin{equation}
+\tilde{x_{0}}(n,c) = x(n,c), k \leftarrow 1
+\label{eq:multiclassificationrow}
+\end{equation}
+
+Hence, it finds the best template, out of the dictionary, that matches current approximation (original signal).
+
+\begin{equation}
+g_{yk}(n) = \arg \max_{g_{yi}} \left\lVert < \tilde{x_{k}}(n,c), g_{yi} > \right\rVert 
+\label{eq:multiclassificationrow}
+\end{equation}
+
+where $g_{yi}$ are all the available scaled, translated and modulated atoms from the dictionary.  This determines the atom selection process, and their contribution is calculating based on
+
+\begin{equation}
+w_{k} =  \frac{< \tilde{x_{k}}(n,c), g_{yk} >}{g_{yk}}
+\label{eq:multiclassificationrow}
+\end{equation}
+
+with $k$ representing the index of the selected atom $g_{yk}$.
+Finally the contribution of each atom is subtracted from the next approximation, refining the approximation process \citep{Cohen2008,Sanei1997, Mallat1993}
+
+\begin{equation}
+\tilde{x_{k+1}}(n,c)=  \tilde{x_k}(n,c) - w_{k} g_{yk} (n), k \leftarrow k + 1
+\label{eq:multiclassificationrow}
+\end{equation}
 
@@
 In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.
-To construct each feature a signal is decomposed in its coefficients and then they can be compared by any distance function.
+The first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter to craft each feature is crafted by decomposing the signal in its coefficients and building a vector with them.
+
+\begin{equation}
+f =  {\bigg ( w_{i} \bigg )}_{1}^{m} 
+\label{eq:multiclassificationrow}
+\end{equation}
 
@@
 
-Bond and Pompe Permutation Entropy has been extensively used in EEG processing, particularly for Epilepsy pre-ictal detection.
+Bond and Pompe Permutation Entropy has been extensively used in EEG processing, particularly for Epilepsy pre-ictal detection \citep{Bandt2002}.
 This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the amount of entropy of each code within the signal.
@@
 This method is the best representative of Waveform Complexity: Lempel-Ziv method L-Z complexity are other variants which use a different definition (Permutation Entropy a new feature for brain computer interfaces).
+
+A signal represented by
+
+\begin{equation}
+(x_1,x_2,...,x_{N})
+\label{eq:pesignal}
+\end{equation}
+
+is resampled by $\tau$ values doing
+
+\begin{equation}
+(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(N-1)\tau})
+\label{eq:pesignal}
+\end{equation}
 
@@
 This method can be encompass as a syntactic analysis technique, where the EEG segment is represented as a series of elementary patterns, similar to tokens, due to the quantization of angles.
+
+A signal, can be represented by a list of ordered-pairs $e$,
+
+\begin{equation}
+e = \left[ (x,y)_{1}, (x,y)_{2}, ..., (x,y)_{N} \right]
+\label{eq:shccdelta}
+\end{equation}
+
+\noindent and it can be divided into $G$ different blocks.  These blocks are obtained by resampling the original signal from the index
+
+\begin{equation}
+g = \lfloor n + ( m \Delta ) + 0.5 \rfloor
+\label{eq:shccdelta}
+\end{equation}
+
+\noindent with $n$ being the original sampling index, and $ 1 \leq t \leq N $.  On the other hand, $\Delta$ can be obtained by
+
+\begin{equation}
+\Delta = \bigg \lceil \frac{N}{G+1} \bigg \rceil
+\label{eq:shccdelta}
+\end{equation}
+
+\noindent with $ G < N $. The value $m$ can be derived from
+
+\begin{equation}
+m = sign \bigg (  \frac{N-1}{\Delta} \bigg )  \bigg \lfloor \left\lvert \frac{T-1}{\Delta} \right\lvert \bigg \rfloor
+\label{eq:shccdelta}
+\end{equation}
+
+This resampling produces a new sequence of values,
+
+\begin{equation}
+e = \left[ (x',y')_{1}, (x',y')_{s}, ..., (x',y')_{G} \right]
+\label{eq:shccdelta}
+\end{equation}
+
+Next step is the normalization of each paired-values according to
+
+\begin{equation}
+\textbf{x} = \frac{1}{\max(x') - min(x')} ( \textbf{x'} - \min(x') \textbf{1} )
+\label{eq:shccdelta}
+\end{equation}
+
+\begin{equation}
+\textbf{y} = \frac{1}{\max(y') - min(y')} ( \textbf{y'} - \min(y') \textbf{1} )
+\label{eq:shccdelta}
+\end{equation}
+
+\noindent effectively yielding $x,y \in [0,1]$.
+
+Finally, the feature is constructed by calculating the point-to-point slope against the horizontal plane,
+
+\begin{equation}
+f = \bigg (  \frac{y_{s}-y_{s-1}}{x_{s}-x_{s-1}}  \bigg )_{2}^{G} 
+\label{eq:shccdelta}
+\end{equation}
 
@@
 This method was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries.
+
+\begin{equation}
+f = {\bigg ( h(i,j,\theta) \bigg )}_{i,j,\theta}
+\label{eq:multiclassificationrow}
+\end{equation}
 

### 2018-08-14 sensors.tex
Added math formulae

@@
 
-Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The following components regularly  characterize EEG signals:
+Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The following elements regularly  characterize EEG signals:
 
@@
 \item Polymorphic: distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
-\item Transient. An isolated wave or pattern that is distinctly different from background activity.
+\item Transient. An isolated wave or pattern that is distinctly different from background activity.  Define component, source and sensor space.
 \end{itemize}
@@
 \midrule
-SVM     & m-c  & $50\%$ \\
-NN       & m-c  & $45\%$ \\
-HGO  & PO8 & $25\%$ \\
-MP 1 & PO8 & $25\%$ \\
-PE     & Cz & $20\%$ \\
-RS     & Fz & $15\%$ \\
-RS     & m-c  & $15\%$ \\
-MP 2 & PO7 & $10\%$ \\
-SHCC & P4 & $10\%$ \\
-NN s-c & Fz & $10\%$ \\
+Characteristic Shape           & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Rising and Falling Phase      & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Pronounced plateau      & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Ripples and Wiggles     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Sinusoidal Shape        & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Sawtooth                     & Motor Imagery, Sleep &  \citep{Cacioppo2007, Mak2010} \\
+Sharpness or Spike-like     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Rectangular     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Wicket Shape     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Peak and Trough Sharpness Ratio     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Symmetry between rise and decay phase     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Slope Ratio    & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
+Positive vs Negative Ratio    & Sleep K-Complex &  \citep{Cacioppo2007, Mak2010} \\
+Base-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007, Mak2010} \\
+Peak-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007, Mak2010} \\
+Latency                                 & ERP  & \citep{Cacioppo2007, Mak2010}  \\
+Integrated Activity               & ERP, Epilepsy & \citep{Cacioppo2007, Shah2015} \\
+Cross-Correlation                & ERP, Epilepsy, Sleep & \citep{Cacioppo2007, Shah2015} \\
+Coupling \\ Cross Frequency,  Phase-Amplitude, Phase-Phase     & Sleep & $20\%$ \\
+Period Amplitude Analysis  & ERP, Epilepsy & \citep{Cacioppo2007, Shah2015} \\
 \bottomrule
@@
 Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied.
-Finally the proposals of Fujimori, Uchida and the PAA \citep{Uchida1996} algorithm are few of the earliest proposals where the idea of capturing the shape of the signal were established.
+Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA) \citep{Uchida1996} algorithm are few of the earliest proposals where the idea of capturing the shape of the signal were established.
 
@@
 
-\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Comon2010} techniques that assume that the EEG signal is a linear combination of different sparse sources extracted from a template's dictionaries.
-\textit{MP}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.
+\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Vincent2010} techniques that assume that the EEG signal is a linear combination of different sparse sources extracted from a template's dictionaries.
+Matching Pursuit \textit{MP}~\citep{Mallat1993}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.
 Given a signal, this optimization technique, tries to find the indexes of atoms and their weights (contributions) that minimize,
@@
 
-which is the error from the signal approximation constructed by the weighted atoms.  The algorithm goes by first setting the approximation signal by the original signal itself,
+which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{yi}$.
+The algorithm goes by first setting the approximation signal by the original signal itself,
 
@@
 
-Hence, it finds the best template, out of the dictionary, that matches current approximation (original signal).
+Hence, it finds the best template, out of the dictionary, that matches current approximation.
 
@@
 
-where $g_{yi}$ are all the available scaled, translated and modulated atoms from the dictionary.  This determines the atom selection process, and their contribution is calculating based on
+where $g_{yi}$ are all the available scaled, translated and modulated atoms from the dictionary.  This determines the atom selection process, and their contribution is calculated based on
 
@@
 
-with $k$ representing the index of the selected atom $g_{yk}$.
-Finally the contribution of each atom is subtracted from the next approximation, refining the approximation process \citep{Cohen2008,Sanei1997, Mallat1993}
+with $k$ represents the index of the selected atom $g_{yk}$.
+Finally the contribution of each atom is subtracted from the next approximation,  extracting the approximating weights and atoms for this iteration \citep{Cohen2014,Sanei2007, Mallat1993}
 
@@
 In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.
-The first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter to craft each feature is crafted by decomposing the signal in its coefficients and building a vector with them.
-
-\begin{equation}
-f =  {\bigg ( w_{i} \bigg )}_{1}^{m} 
+The first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building a vector with them.
+
+\begin{equation}
+f =  {\bigg ( w_{i} \bigg )}_{1}^{D} 
 \label{eq:multiclassificationrow}
@@
 
+\noindent where $D$ is the size of the dictionary.
+
 %On the other hand, Morphological Component Analysis is a variant form of Blind Source Separation which can be considered the extension of Matching Pursuit algorithm. (cita a esa parte del paper).
@@
 
-Bond and Pompe Permutation Entropy has been extensively used in EEG processing, particularly for Epilepsy pre-ictal detection \citep{Bandt2002}.
+Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and particularly for Epilepsy pre-ictal detection \citep{Bandt2002}.
 This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the amount of entropy of each code within the signal.
-For instance, a pure random signal, will achieve the maximum entropy value due to the probability of each code being equal for all of them.
-This method is the best representative of Waveform Complexity: Lempel-Ziv method L-Z complexity are other variants which use a different definition (Permutation Entropy a new feature for brain computer interfaces).
-
-A signal represented by
-
-\begin{equation}
-(x_1,x_2,...,x_{N})
+For instance, a pure random signal, will achieve the maximum entropy value due to the probability of each code being equal for all of them~\citep{Nicolaou2010}.
+
+A signal on a window of length $W$ is represented by
+
+\begin{equation}
+(x_1,x_2,...,x_{W})
 \label{eq:pesignal}
@@
 
-is resampled by $\tau$ values doing
-
-\begin{equation}
-(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(N-1)\tau})
+and it can be resampled by $\tau$ intervals doing
+
+\begin{equation}
+(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(m-1)\tau})
 \label{eq:pesignal}
@@
 
+This sequence of order $m$ corresponds to one ordinal pattern $\pi_{n}$, for $n$ varying between $1$ and $W$.
+Within the signal window there can be at most $k$ different ordinal (and overlapping) patterns,
+
+\begin{equation}
+(\pi_{1},\pi_{2},...,\pi_{k})
+\label{eq:pesignal}
+\end{equation}
+
+\noindent with $k = W-(m-1) \tau$.  The probability density function for all the available permutations of order $m$ should be $ p = (p_1,p_2,...,p_{m!}) $ with $ \left\lVert p \right\lVert = 1 $.
+
+Hence, the set of time samples map to a new set of $k$ ordinal patterns, and the \textit{pdf} can be estimated by the empirical permutation entropy,
+
+\begin{equation}
+P_i = \frac{1}{k} \sum_{s=1}^{k} \left[ \pi_{s} = \pi_{i} \right]
+\label{eq:pesignal}
+\end{equation}
+
+\noindent with $1 \leq i \leq m!$.
+By using the Iverson Bracket, for each $i$ only those ordinal patterns that were effectively found are counted to estimate $p_i$.
+The empirical permutation entropy can be estimated from this histogram as,
+
+\begin{equation}
+H(p) = \sum_{i=1}^{m!} p_{i} \; log \frac{1}{p_{i}}
+\label{eq:pesignal}
+\end{equation}
+
+The implemented code was derived from~\citep{Unakafova2013}, and the model description from \citep{Berger2017}.
+This procedure will produce a scalar number for the given signal window of size $W$.
+To derive a feature out of it, a sliding window procedure must be implemented.
+Thus, the length of the feature is $N - (W + \tau (m - 1)) + 1$.
+
+\begin{equation}
+f =  {\bigg ( H(p)_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
+\label{eq:multiclassificationrow}
+\end{equation}
+
+\noindent with $u$ varying on a sample by sample basis along the signal but excluding the first window.
+
 \subsection{Slope Horizontal Chain Code - SHCC}
 
-This method is an extension of Slope Chain Code (SCC) (ref a SCC paper y a SHCC paper), and a code of a sequence of sample points is generated based on the angle between the horizontal segment and any segment, one by one.
-This method can be encompass as a syntactic analysis technique, where the EEG segment is represented as a series of elementary patterns, similar to tokens, due to the quantization of angles.
+This method is an extension of Slope Chain Code (SCC)~\citep{Alvarado-Gonzalez2016}.
+A code is generated from a sequence of sample points based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.
 
@@
 
-\noindent with $n$ being the original sampling index, and $ 1 \leq t \leq N $.  On the other hand, $\Delta$ can be obtained by
+\noindent with $n$ being the original sampling index, and $ 1 \leq n \leq N $.  On the other hand, $\Delta$ can be obtained by
 
@@
 
-\noindent with $ G < N $. The value $m$ can be derived from
+\noindent with $ G < N $. Lastly, the value $m$ can be derived from
 
@@
 
-Next step is the normalization of each paired-values according to
+The next step is the normalization of each ordered-pair according to
 
@@
 
-\noindent effectively yielding $x,y \in [0,1]$.
+\noindent effectively yielding $x,y \in [0,1]$. \\
 
@@
 This mimicks exactly what a person is doing by visually inspecting the plot, and the waveform.
-To do that, the region of the image where the signal is located is divided in a $4 x 4$ block and the signal bidimensional gradients is calculated.
+To do that, the region of the image where the signal is located is divided in a $4 x 4$ block and the signal's bidimensional gradients is calculated.
 For each block within the patch, a histogram of gradient orientations, 8 circular orientations, is calculated.
@@
 
+intelligible property
+
 The conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.

### 2018-08-16 sensors.tex
Update with a little modification about someting I remember

@@
 
-In contrast, automatic processing, or quantitative EEG (qEEG), was based first on analog electronic devices and later on computerized digital processing methods \citep{Jansen1991}.
+In contrast, automatic processing, or quantitative EEG (qEEG), was based first on analog electronic devices~\citep{Jestico1977} and later on computerized digital processing methods \citep{Jansen1991}.
 They implemented mathematically and algorithmically complex procedures to decode the information associated with a clinical condition, or even some higher cognitive state \citep{Yuste2017}.
@@
 \midrule
+Rounded                                & $\alpha$-Waves & \citep{Schomer2010} \\
 Characteristic Shape           & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\

### 2018-08-22 sensors.tex
Pushing everything into a deliverable version basic, but publishable version

@@
 
-Other depictions may include, subjective definitions of sharper, arch comb or wicket shape, rectangular, containing a decay phase or voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace.
+Depictions may include subjective definitions of sharper, arch comb or wicket shape, rectangular, containing a decay phase or voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace.
 Derived ratios and indexes can be used as well like peak and trough sharpness ratio, symmetry between rise and decay phase and slope ratio (steepness of the rise period to that of the adjacent decay period).
-For instance,  wording like "Central trough is sharper and more negative that the adjacent troughs" are common in the literature.
-
-Other regular characterizations which are based on shape features may include:
+For instance,  wording like "Central trough is sharper and more negative that the adjacent troughs"~\citep{Cole2017} are common in the literature.
+
+Other regular characterizations which are based on shape features may encompass:
 
@@
 \item Paroxysmal: Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity.
-Though the term does not directly imply abnormality, much abnormal activity is paroxysmal.
+This term does not directly imply abnormality per se, much abnormal activity is paroxysmal.
+\item Monomorphic: Activity appearing to be composed of one dominant waveform pattern.
+\item Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.
+\item Transient: An isolated wave or pattern that is distinctly different from background activity. 
 \end{itemize}
 
-\begin{itemize}
-\item Monomorphic: Distinct EEG activity appearing to be composed of one dominant activity
-\item Polymorphic: distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
-\item Transient. An isolated wave or pattern that is distinctly different from background activity.  Define component, source and sensor space.
-\end{itemize}
+%  Define component, source and sensor space.
 
@@
 
+% Non-exhaustive list of EEG Waveforms, their phenomena and the literature reference where they are described.
+
 \begin{table}[H]
-\caption{Subjective EEG characterization methods found in the surveyed literature.}
+\caption{EEG characterization descriptions found in the surveyed literature.}
 \centering
@@
 \midrule
-Rounded                                & $\alpha$-Waves & \citep{Schomer2010} \\
-Characteristic Shape           & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Rising and Falling Phase      & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Pronounced plateau      & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Ripples and Wiggles     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Sinusoidal Shape        & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Sawtooth                     & Motor Imagery, Sleep &  \citep{Cacioppo2007, Mak2010} \\
-Sharpness or Spike-like     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Rectangular     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Wicket Shape     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Peak and Trough Sharpness Ratio     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Symmetry between rise and decay phase     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Slope Ratio    & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
-Positive vs Negative Ratio    & Sleep K-Complex &  \citep{Cacioppo2007, Mak2010} \\
-Base-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007, Mak2010} \\
-Peak-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007, Mak2010} \\
-Latency                                 & ERP  & \citep{Cacioppo2007, Mak2010}  \\
+Positive Rounded Component                    & $\alpha$-Waves, Epilepsy & \citep{Schomer2010,Tatum2008} \\
+Rising and Falling Phase      & Epilepsy &  \citep{Thakor2004,Tatum2008} \\
+Terminal plateau      & Epilepsy &  \citep{Thakor2004} \\
+Ripples and Wiggles     & Epilepsy, ERP &  \citep{EEGIntro, Thakor2004,Cacioppo2007,Kappenman2012} \\
+Sinusoidal Shape        & Epilepsy &  \citep{Cacioppo2007,Tatum2008,Ouyang2017,Kappenman2012} \\
+Sawtooth                     & Motor Imagery, Sleep &  \citep{EEGIntro,Rodenbeck2006,Tatum2008} \\
+Sharpness or Spike-like     & Epilepsy &  \citep{Cacioppo2007} \\
+Rectangular     & Epilepsy &  \citep{Cacioppo2007} \\
+Wicket Shape     & Epilepsy &  \citep{Cacioppo2007} \\
+Peak and Trough Sharpness Ratio     & Epilepsy &  \citep{Cacioppo2007} \\
+Symmetry between rise and decay phase     & Epilepsy &  \citep{Cacioppo2007} \\
+Slope Ratio    & Epilepsy &  \citep{Cacioppo2007} \\
+Positive/Negative Peak Amplitude & ERP & \citep{Mak2012} \\
+Positive vs Negative Ratio    & Sleep K-Complex &  \citep{Cacioppo2007} \\
+Base-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007} \\
+Peak-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007, Mak2012} \\
+Positive/Negative Peak Latency                                 & ERP  & \citep{Cacioppo2007, Mak2012}  \\
 Integrated Activity               & ERP, Epilepsy & \citep{Cacioppo2007, Shah2015} \\

### 2018-08-24 sensors.tex
Ready for roll-out

@@
 
-Other regular characterizations which are based on shape features may encompass:
+Other regular characterizations which are based on the waveform shape may encompass:
 
@@
 \begin{table}[H]
-\caption{EEG characterization descriptions found in the surveyed literature.}
+\caption{EEG waveforms descriptions found in the surveyed literature.}
 \centering
@@
 Ripples and Wiggles     & Epilepsy, ERP &  \citep{EEGIntro, Thakor2004,Cacioppo2007,Kappenman2012} \\
-Sinusoidal Shape        & Epilepsy &  \citep{Cacioppo2007,Tatum2008,Ouyang2017,Kappenman2012} \\
+Sinusoidal Shape        & Epilepsy &  \citep{Cacioppo2007,Tatum2008,Ouyang2017,Kappenman2012,Cole2017} \\
 Sawtooth                     & Motor Imagery, Sleep &  \citep{EEGIntro,Rodenbeck2006,Tatum2008} \\
-Sharpness or Spike-like     & Epilepsy &  \citep{Cacioppo2007} \\
-Rectangular     & Epilepsy &  \citep{Cacioppo2007} \\
-Wicket Shape     & Epilepsy &  \citep{Cacioppo2007} \\
-Peak and Trough Sharpness Ratio     & Epilepsy &  \citep{Cacioppo2007} \\
-Symmetry between rise and decay phase     & Epilepsy &  \citep{Cacioppo2007} \\
-Slope Ratio    & Epilepsy &  \citep{Cacioppo2007} \\
-Positive/Negative Peak Amplitude & ERP & \citep{Mak2012} \\
-Positive vs Negative Ratio    & Sleep K-Complex &  \citep{Cacioppo2007} \\
-Base-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007} \\
-Peak-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007, Mak2012} \\
-Positive/Negative Peak Latency                                 & ERP  & \citep{Cacioppo2007, Mak2012}  \\
-Integrated Activity               & ERP, Epilepsy & \citep{Cacioppo2007, Shah2015} \\
+Sharpness or Spike-like     & Epilepsy &  \citep{Thakor2004,Hartman2005,Sanei2007,EEGIntro} \\
+Rectangular     & Epilepsy &  \citep{Thakor2004,Cole2017} \\
+Wicket Shape     & Epilepsy &  \citep{EEGIntro,Hartman2005,Sanei2007,Tatum2008,Schomer2010,Cole2017} \\
+Peak and Trough Sharpness Ratio     & Epilepsy &  \citep{Hartman2005,Sanei2007,Lawrence2010,Cole2017} \\
+Symmetry between rise and decay phase     & Epilepsy &  \citep{Hartman2005,Cole2017} \\
+Slope Ratio    & Sleep &  \citep{Subha2010} \\
+Positive/Negative Peak Amplitude & ERP & \citep{Thakor2004,Hartman2005,Tatum2008,Mak2012,MullerPutz2015,Cole2017} \\
+Positive vs Negative Ratio    & Sleep K-Complex &  \citep{EEGIntro} \\
+Base-to-Peak Amplitude     & ERP  &  \citep{Cole2017} \\
+Peak-to-Peak Amplitude     & ERP  &  \citep{Mak2012} \\
+Positive/Negative Peak Latency                                 & ERP  & \citep{Mak2012}  \\
+Integrated Activity               & ERP, Epilepsy,ICU & \citep{Uchida1999, Shah2015} \\
 Cross-Correlation                & ERP, Epilepsy, Sleep & \citep{Cacioppo2007, Shah2015} \\
-Coupling \\ Cross Frequency,  Phase-Amplitude, Phase-Phase     & Sleep & $20\%$ \\
-Period Amplitude Analysis  & ERP, Epilepsy & \citep{Cacioppo2007, Shah2015} \\
+Coupling \\ Cross Frequency,  Phase-Amplitude, Phase-Phase     & Sleep & \citep{Cole2017} \\
+Period Amplitude Analysis  & ERP, Epilepsy & \citep{Uchida1999,Cacioppo2007, Shah2015} \\
 \bottomrule
@@
 
-As described in \citep{allen2004signal} the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the prototype signal.
+As described in \citep{allen2004signal} the Pattern Matching problem in Signal processing is finding a signal given the region that best describes the structure of the prototype signal template.
+On the other hand, a \textit{feature} is a meaningful quantification, usually a multidimensional vector, that synthesize the information of a given signal or signal segment~\citep{WolpawJonathanR2012}.
 All these algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
@@
 Shape or waveform analysis methods are considered as nonparametric (in opposition to statistical or dynamical models).
-They explore signal's time-domain metrics or even derive more complex indexes from it \citep{Thakor2009}.
+They explore signal's time-domain metrics or even derive more complex indexes or features from it \citep{Thakor2009}.
 
@@
 Although of limited usability, peak picking has been used to determine latency of transient events in EEG \citep{Jaskowski2000,Zhang2011}.
-Straightforward in its implementation, it consists in selecting a component, a simple component based on the expected location of its more prominent deflection \citep{Ouyang2017}.
+Straightforward in its implementation, it consists in assigning a component to a simple waveform element based on the expected location of its more prominent deflection \citep{Ouyang2017}.
 Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient component that may arise as a brain response to an external visual, tactile or auditory stimulus.
-Particularly, EPs are regularly used to assess auditory response in infants.
-They are precisely characterized in this manner, where the name of many of the EEG features evoke directly a peak within the component, e.g.
+Particularly, ERPs are regularly used to assess auditory response in infants.
+They are characterized in this manner, where the name of many of the EEG features evoke directly a peak within the waveform, e.g.
 P300 or P3a P3b or N100.
@@
 Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature from the amplitude and frequency of its signal and its derivative in time-domain is used.
-Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied.
-Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA) \citep{Uchida1996} algorithm are few of the earliest proposals where the idea of capturing the shape of the signal were established.
+Moreover, other approaches explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied.
+Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA) \citep{Uchida1996} algorithm are few of the earliest depictions where the idea of capturing the shape of the signal were established.
 
@@
 
-All these methods provide a feature that can be used as a template, whereas all of them are based on metrics that can be extracted from the shape of the signal.
-These features can be used to create dictionaries or template databases.
-These templates provide the basis for the pattern matching algorithm and offline classification.
+All these methods provide a feature that can be used as a template, whereas all of them are based on metrics that are extracted from the shape of the signal.
+These features are used to create dictionaries or template databases.
+Finally, these templates provide the basis for the pattern matching algorithm and offline classification.
 
@@
 
-\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Vincent2010} techniques that assume that the EEG signal is a linear combination of different sparse sources extracted from a template's dictionaries.
+\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Vincent2010} techniques that assume the EEG signal as a linear combination of different sparse sources extracted from a template's dictionaries.
 Matching Pursuit \textit{MP}~\citep{Mallat1993}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.
@@
 which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{yi}$.
-The algorithm goes by first setting the approximation signal by the original signal itself,
+The algorithm goes by first setting the approximating signal $\tilde{x_{0}}$  as the original signal itself,
 
@@
 
-Hence, it finds the best template, out of the dictionary, that matches current approximation.
+\noindent and setting the iterative counter $k$ as $1$. Hence, it searches recurrently the best template out of the dictionary  that matches current approximation.
 
@@
 
-with $k$ represents the index of the selected atom $g_{yk}$.
-Finally the contribution of each atom is subtracted from the next approximation,  extracting the approximating weights and atoms for this iteration \citep{Cohen2014,Sanei2007, Mallat1993}
+with $k$ representing the index of the selected atom $g_{yk}$.  Finally the projection of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
 
@@
 Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and particularly for Epilepsy pre-ictal detection \citep{Bandt2002}.
-This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the amount of entropy of each code within the signal.
-For instance, a pure random signal, will achieve the maximum entropy value due to the probability of each code being equal for all of them~\citep{Nicolaou2010}.
-
+This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found on the signal synthesized as numeric value called entropy~\citep{Nicolaou2010}.
 A signal on a window of length $W$ is represented by
@@
 This sequence of order $m$ corresponds to one ordinal pattern $\pi_{n}$, for $n$ varying between $1$ and $W$.
-Within the signal window there can be at most $k$ different ordinal (and overlapping) patterns,
+Along the signal window there can be at most $k$ different ordinal (and overlapping) patterns,
 
@@
 
-Hence, the set of time samples map to a new set of $k$ ordinal patterns, and the \textit{pdf} can be estimated by the empirical permutation entropy,
+Hence, the set of time samples map to a new set of $k$ ordinal patterns, and the \textit{pdf} can be calculated by the empirical permutation entropy,
 
@@
 This procedure will produce a scalar number for the given signal window of size $W$.
-To derive a feature out of it, a sliding window procedure must be implemented.
+To derive a feature, a sliding window procedure must be implemented.
 Thus, the length of the feature is $N - (W + \tau (m - 1)) + 1$.
@@
 This method is an extension of Slope Chain Code (SCC)~\citep{Alvarado-Gonzalez2016}.
-A code is generated from a sequence of sample points based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.
+A coding scheme is generated from a sequence of sample points based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.
 

### 2018-08-28 sensors.tex
Added a new set of images. Completed math for every method.

@@
 \documentclass[brainsci,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
+
+\usepackage{subfigure}
 
@@
 % Full title of the paper (Capitalized)
-\Title{EEG Waveform Analysis with applications to Brain Computer Interfaces}
+\Title{EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces}
 
@@
 
-All these methods provide a feature that can be used as a template, whereas all of them are based on metrics that are extracted from the shape of the signal.
+All these methods provide a feature $ f $  that can be used as a template, whereas all of them are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length of the EEG segment in sample points.
 These features are used to create dictionaries or template databases.
@@
 \begin{equation}
-\varepsilon =  \left\lVert   x(n,c) - \sum_{i=1}^{m} w_i g_{yi}   \right\rVert
+\varepsilon =  \left\lVert   x(n) - \sum_{i=1}^{m} w_i g_{yi}(n)   \right\rVert
 \label{eq:mperror}
@@
 \begin{equation}
-\tilde{x_{0}}(n,c) = x(n,c), k \leftarrow 1
-\label{eq:multiclassificationrow}
+\tilde{x_{0}}(n) = x(n), k \leftarrow 1
+\label{eq:mp2}
 \end{equation}
@@
 \begin{equation}
-g_{yk}(n) = \arg \max_{g_{yi}} \left\lVert < \tilde{x_{k}}(n,c), g_{yi} > \right\rVert 
-\label{eq:multiclassificationrow}
+g_{yk}(n) = \arg \max_{g_{yi}} \left\lVert < \tilde{x_{k}}(n), g_{yi}(n) > \right\rVert 
+\label{eq:mp3}
 \end{equation}
@@
 \begin{equation}
-w_{k} =  \frac{< \tilde{x_{k}}(n,c), g_{yk} >}{g_{yk}}
-\label{eq:multiclassificationrow}
+w_{k} =  \frac{< \tilde{x_{k}}(n), g_{yk}(n) >}{  | g_{yk} | }
+\label{eq:mp4}
 \end{equation}
@@
 \begin{equation}
-\tilde{x_{k+1}}(n,c)=  \tilde{x_k}(n,c) - w_{k} g_{yk} (n), k \leftarrow k + 1
-\label{eq:multiclassificationrow}
+\tilde{x_{k+1}}(n)=  \tilde{x_k}(n) - w_{k} g_{yk} (n), k \leftarrow k + 1
+\label{eq:mp5}
 \end{equation}
@@
 In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.
-The first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building a vector with them.
+For the first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building, an eventually sparse, vector with them.
 
@@
 f =  {\bigg ( w_{i} \bigg )}_{1}^{D} 
-\label{eq:multiclassificationrow}
+\label{eq:mp6}
 \end{equation}
@@
 
-Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and particularly for Epilepsy pre-ictal detection \citep{Bandt2002}.
-This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found on the signal synthesized as numeric value called entropy~\citep{Nicolaou2010}.
+Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and significantly for Epilepsy pre-ictal detection \citep{Bandt2002}.
+This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found on the signal synthesized as a numeric value called entropy~\citep{Nicolaou2010}.
 A signal on a window of length $W$ is represented by
@@
 (x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(m-1)\tau})
-\label{eq:pesignal}
+\label{eq:pe2}
 \end{equation}
@@
 (\pi_{1},\pi_{2},...,\pi_{k})
-\label{eq:pesignal}
+\label{eq:pe3}
 \end{equation}
@@
 P_i = \frac{1}{k} \sum_{s=1}^{k} \left[ \pi_{s} = \pi_{i} \right]
-\label{eq:pesignal}
+\label{eq:pe4}
 \end{equation}
@@
 H(p) = \sum_{i=1}^{m!} p_{i} \; log \frac{1}{p_{i}}
-\label{eq:pesignal}
+\label{eq:pe5}
 \end{equation}
@@
 f =  {\bigg ( H(p)_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
-\label{eq:multiclassificationrow}
+\label{eq:pe6}
 \end{equation}
@@
 This method is an extension of Slope Chain Code (SCC)~\citep{Alvarado-Gonzalez2016}.
-A coding scheme is generated from a sequence of sample points based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.
+This algorithm proceeds by generating a coding scheme from a sequence of sample points.
+This encoding is based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.
 
@@
 g = \lfloor n + ( m \Delta ) + 0.5 \rfloor
-\label{eq:shccdelta}
+\label{eq:shcc2}
 \end{equation}
@@
 \Delta = \bigg \lceil \frac{N}{G+1} \bigg \rceil
-\label{eq:shccdelta}
+\label{eq:shcc3}
 \end{equation}
@@
 m = sign \bigg (  \frac{N-1}{\Delta} \bigg )  \bigg \lfloor \left\lvert \frac{T-1}{\Delta} \right\lvert \bigg \rfloor
-\label{eq:shccdelta}
+\label{eq:shcc4}
 \end{equation}
@@
 e = \left[ (x',y')_{1}, (x',y')_{s}, ..., (x',y')_{G} \right]
-\label{eq:shccdelta}
+\label{eq:shcc5}
 \end{equation}
@@
 \textbf{x} = \frac{1}{\max(x') - min(x')} ( \textbf{x'} - \min(x') \textbf{1} )
-\label{eq:shccdelta}
+\label{eq:shcc6}
 \end{equation}
@@
 \textbf{y} = \frac{1}{\max(y') - min(y')} ( \textbf{y'} - \min(y') \textbf{1} )
-\label{eq:shccdelta}
+\label{eq:shcc7}
 \end{equation}
@@
 f = \bigg (  \frac{y_{s}-y_{s-1}}{x_{s}-x_{s-1}}  \bigg )_{2}^{G} 
-\label{eq:shccdelta}
-\end{equation}
-
-\subsection{Histogram of Gradients Orientations - HGO}
-
-This Histogram of Gradient Orientations is based on using Computer Vision techniques to extract directly information from a signal based on the plot of a signal on a 2D image.
-This mimicks exactly what a person is doing by visually inspecting the plot, and the waveform.
-To do that, the region of the image where the signal is located is divided in a $4 x 4$ block and the signal's bidimensional gradients is calculated.
-For each block within the patch, a histogram of gradient orientations, 8 circular orientations, is calculated.
-This histogram is concatenated for all the 16 blocks and a feature is thus formed.
-The details of the method can be found here \citep{Ramele2016}.
-This method was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries.
+\label{eq:shcc8}
+\end{equation}
+
+\subsection{Scale Invariant Feature Transform - SIFT}
+
+SIFT~\citep{Lowe2004} is a very successful feature extraction technique from Computer Vision .
+It has a biomimetic inspiration on how the visual cortex analyze images based on orientations.
+This method has been used in \citep{Ramele2016} to analyze EEG signals based on their plots on 2D images.
+
+The first step of the algorithm is plot generation based on single-channel EEG segments $ x(n) $.
+
+\begin{equation}
+\tilde{x}(n) = \left\lfloor \frac{\delta ( x(n) - \bar{x})}{\sigma_{x}} \right\rceil 
+\label{eq:sift1}
+\end{equation}
+
+\noindent with $\delta$ being the signal magnification factor and $\bar{x}$ and $\sigma_{x}$ the mean and standard deviation of $x$ on the signal segment.
+The width of the image is determined straightforward based on the $1$-second length in sample units.
+This corresponds to the sampling frequency $F_s$ of the EEG signal segment.
+The width is adjusted by multiplying by a magnification factor $\delta$,
+
+\begin{equation}
+w = \delta \; F_s
+\label{eq:sift2}
+\end{equation}
+
+\noindent whereas the height is calculated based on the peak-to-peak amplitude of the signal within the segment,
+
+\begin{equation}
+h = ( \max_{n} \tilde{x}(n) - \min_{n} \tilde{x}(n))
+\label{eq:sift4}
+\end{equation}
+
+Equation \ref{eq:sift5} determines the height of the image where the signal's zero value will be located.
+
+\begin{equation}
+z = \left\lfloor \frac{h}{2} \right\rfloor - \left\lfloor \frac{\max_{n} \tilde{x}(n) + \min_{n} \tilde{x}(n)}{2} \right\rfloor
+\label{eq:sift5}
+\end{equation}
+
+Finally, a binary, black-and-white image plot is generated based on
+
+\begin{equation}
+I(z_1,z_2) = \left\{ \begin{array}{rl}
+255 & \text{if} \,  z_1 = \delta \  n; \! z_2 = \tilde{x}(n) + z \\
+0   & \mbox{otherwise}
+\end{array}\right.
+\label{eq:sift6}
+\end{equation}
+
+Once the plot is generated, the plot's center is used to localize the SIFT patch.
+This region of the image, where the signal's most important salient waveform should be located, is divided in a grid of $4 x 4$ block and the bidimensional gradient vectors are calculated on each one of them.
+At the same time, for each block within the patch, a histogram of the gradient orientations, 8 circular orientations, are calculated.
+This histogram is concatenated for all the 16 blocks and a feature is thus formed:
 
@@
 f = {\bigg ( h(i,j,\theta) \bigg )}_{i,j,\theta}
-\label{eq:multiclassificationrow}
-\end{equation}
+\label{eq:sift7}
+\end{equation}
+
+\noindent with $i$ and $j$ varying between $0$ and $3$, localizing the 16 blocks within the grid.
+The angles $\theta$ are the eight possible values between $0$ and $315$.
+This vector is normalized, clapped to $0.2$, and re-normalized again.
+Details of the method can be found on \citep{Ramele2016,Lowe2004}.
+It was implemented using the VLFeat  \citep{Vedaldi2010} public Computer Vision libraries.
 
@@
 Farwell and Donchin P300 Speller \citep{Farwell1988} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing in a virtual keyboard.
-This procedure exploits a cognitive phenomena called oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
+This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
 This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.
 
-%Algorithms based on qEEG can detect this component in a good-enough single-trial approach [P300 Summary], but at the same time, these algorithms do not establish a clear relationship with the P300 ERP.
-Hence it is difficult to establish a backward mapping back to a certain visual pattern within the waveform.
-
-In order to verify each algorithm in a controlled procedure, a real ERP template is superimposed into a real EEG stream.
+How much is it possible to classify and implement a valid P300 ERP Speller using only the waveform structure of the signal and which algorithm perform better at this even when single trial latency variations and component amplitudes issues are present ?
+In order to verify this in a controlled procedure, a real P300 ERP template, obtained from a public dataset, is superimposed into a different EEG stream.
 This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on anyone in particular.
@@
 \item Pseudo-real dataset Classification Performance: the letter classification performance of each one of these methods.
-\item Latency Noise: Gaussian noise is added to the latency where the ERP template is injected into the EEG stream.
+\item Latency Noise: Gaussian noise is added to the ERP template latency, i.e. the temporal location where it is injected into the EEG stream.
 \item Component Amplitude Noise: the different components of the ERP template are altered to make them less distinguishable.
@@
 
-Finally results for the same methods while offline processing a public dataset of real P300 patterns are also shown.
+\textbf{Borrar o hacer: Finally results for the same methods while offline processing a public dataset of real P300 patterns are also shown.}
 
@@
 
-The template ERP is extracted from the Subject $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+The template ERP is extracted from the Subject Number $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
 Segments from the EEG signal are labeled as hit and are extracted for the trial number $2$, and they are point-to-point coherently averaged.
@@
 
-An EEG stream with null-P300 signal everywhere is obtained by the following procedure: 
+An EEG stream with null-P300 signal is obtained by the following procedure: 
 The participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
@@
 
-\begin{figure}[H]
+\begin{figure}[htb]
 \centering
-\includegraphics[width=10cm]{images/doublegain.eps}
-\includegraphics[width=10cm]{images/triplegain.eps}
-\caption{Eight-channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the added template.}
+\subfigure[EEG trace of the original signal.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
+\subfigure[The same signal segment with the added template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
+\subfigure[EEG sample of Cz channel of the original EEG trace.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
+\subfigure[The same segment with the added template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
+\caption{Eight-channel EEG signal superimposed with the ERP Template.}
 \label{fig:doubleandtriplegain}
 \end{figure}
+
+%\begin{figure}[H]
+%\centering
+%\includegraphics[width=12cm]{images/nogain.eps}
+%\includegraphics[width=12cm]{images/singlegain.eps}
+%\caption{Eight-channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the added template.}
+%\label{fig:doubleandtriplegain}
+%\end{figure}
 

### 2018-08-29 sensors.tex
Small change

@@
 SIFT~\citep{Lowe2004} is a very successful feature extraction technique from Computer Vision .
-It has a biomimetic inspiration on how the visual cortex analyze images based on orientations.
+It has a biomimetic inspiration on how the visual cortex analyze images based on orientations~\citep{cogprints561}.
 This method has been used in \citep{Ramele2016} to analyze EEG signals based on their plots on 2D images.
 
-The first step of the algorithm is plot generation based on single-channel EEG segments $ x(n) $.
+The first step of the algorithm is the plot generation based on single-channel EEG segments $ x(n) $.
 
@@
 
-\noindent with $\delta$ being the signal magnification factor and $\bar{x}$ and $\sigma_{x}$ the mean and standard deviation of $x$ on the signal segment.
-The width of the image is determined straightforward based on the $1$-second length in sample units.
+\noindent with $\delta$ being the signal magnification factor and $\bar{x}$ and $\sigma_{x}$, the mean and standard deviation of $x$ on the signal segment.
+The width of the image is determined based on the $1$-second length size of the segment window in sample units.
 This corresponds to the sampling frequency $F_s$ of the EEG signal segment.
-The width is adjusted by multiplying by a magnification factor $\delta$,
+The width is adjusted by multiplying by the magnification factor $\delta$,
 
@@
 
-Equation \ref{eq:sift5} determines the height of the image where the signal's zero value will be located.
+Equation \ref{eq:sift5} determines the vertical position of the image where the signal's zero value will be located.
 

### 2018-08-31 sensors.tex
Results and Classification. Conclussion and Discussion only remains

@@
 
-Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature from the amplitude and frequency of its signal and its derivative in time-domain is used.
+Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.
 Moreover, other approaches explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied.
@@
 
+\noindent where $z_1$ and $z_2$ are the image coordinates values.  These points are interpolated using the Bresenham algorithm.
+
 Once the plot is generated, the plot's center is used to localize the SIFT patch.
@@
 
-How much is it possible to classify and implement a valid P300 ERP Speller using only the waveform structure of the signal and which algorithm perform better at this even when single trial latency variations and component amplitudes issues are present ?
+How much is it possible to classify and implement a P300 ERP Speller using only the waveform structure of the signal ?
+Which algorithm perform better at this even when single trial latency variations and component amplitudes issues are present ?
+
 In order to verify this in a controlled procedure, a real P300 ERP template, obtained from a public dataset, is superimposed into a different EEG stream.
-This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on anyone in particular.
+This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on any letter in particular.
 Everything is there, except the P300 ERP component.
-By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to a similar procedure used in similar works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
+By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to similar procedures used in other works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
 
@@
 \item Latency Noise: Gaussian noise is added to the ERP template latency, i.e. the temporal location where it is injected into the EEG stream.
-\item Component Amplitude Noise: the different components of the ERP template are altered to make them less distinguishable.
+\item Component Amplitude Noise: the amplitude of the main P3b  components of the ERP template is altered based on uniform distribution.
 \end{itemize}
 
-\textbf{Borrar o hacer: Finally results for the same methods while offline processing a public dataset of real P300 patterns are also shown.}
+%\textbf{Borrar o hacer: Finally results for the same methods while offline processing a public dataset of real P300 patterns are also shown.}
 
@@
 An EEG stream with null-P300 signal is obtained by the following procedure: 
-The participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
+A subject participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
 No monetary compensation is handed out and he/she agree and sign a written informed consent.
@@
 The participant is healthy and have normal or corrected-to-normal vision and no history of neurological disorders.
-This voluntary is of unspecified gender, aged between 20-30 years old.
+This voluntary subject is of unspecified gender, aged between 20-30 years old.
 EEG data is collected in a single recording session.
-She/He is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of him.
+She/He is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.
 The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.
@@
 Reference is set to the right ear lobe and ground is preset as the AFz position.
-Sampling frequency is set to 250 Hz, which is the closest possible to the one used with the other dataset.
+Sampling frequency is set to 250 Hz.
 
@@
 Fig.
-\ref{fig:gaincheck} shows the result of superimposing the template signal into EEG stream, time-locked to the stimulus onset.
+\ref{fig:gains} shows a $5s$ sample of the EEG trace.
+Channel $S$ represents the twelve different stimulus markers (columns or rows) while channel $L$ represent the label (hit vs nohit).
+Subfigure (a) shows the signals before the ERP template is superimposed while subfigure (b) shows the same signals with the superimposed ERP template.
+At first sight differences are really hard to spot visually.
+Subfigures (c) and (d) show only one second of the same segment only channel Cz and Label.
+The superimposed ERP can be divised around $31.5$s where in (d) the peak is bigger.
+Fig.
+\ref{fig:gaincheck} shows the obtained ensemble average ERPs as result of superimposing the template signal into EEG stream, time-locked to the stimulus onset.
 These 12 point-to-point averaged segments correspond to the first trial of the EEG stream.
@@
 \includegraphics[width=12cm]{images/erptemplate1.eps}
-\caption{ERP Template obtained from the coherent point-to-point ensemble average of subject Eight obtained from the BNCI Horizon public dataset.
+\caption{ERP Template obtained from the coherent point-to-point ensemble average of subject Eight obtained from the BNCI Horizon public dataset 008-2014.
 The template is $1s$ long which is 256 sample points, and the eight channels are superimposed with different colors.
@@
 \subfigure[EEG trace of the original signal.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
-\subfigure[The same signal segment with the added template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
+\subfigure[The same signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
 \subfigure[EEG sample of Cz channel of the original EEG trace.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
-\subfigure[The same segment with the added template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
+\subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
 \caption{Eight-channel EEG signal superimposed with the ERP Template.}
-\label{fig:doubleandtriplegain}
+\label{fig:gains}
 \end{figure}
@@
 \centering
-\includegraphics[width=15cm]{images/GainCheck.eps}
-\caption{These are the point-to-point averaged signals for the first trial.  The ERP was injected on class 3 and 9.}
+\includegraphics[width=1.0\linewidth]{images/GainCheck.eps}
+\caption{Point-to-point averaged signals for the first trial.  The ERP was superimposed on classes 3 and 9.}
 \label{fig:gaincheck}
@@
 
+%The averaging procedure aims to extract the superimposed ERP Template and cancel out everything else on each segment.
+
 \subsubsection{Classification} \label{section:classification}
 
-To analyze how each method identifies the hidden P300 signal, the same classification algorithm is used.
-The original dataset is composed of 35 trials to decode 7 words of 5 letters  of the P300 matrix.
-Each trial is composed of 10 intensification sequences of the 6 columns and 6 rows \citep{Riccio2013}.
-The averaging procedure aims to extract the superimposed ERP Template and cancel out everything else on each segment.
-
-This classification procedure identifies the row and column, matching to the template $T$ by computing
-
-\begin{equation}
-\hat{row} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q \in NN_T(f^{row}_u)}^{} \left\lVert q -  f^{row}_u \right\rVert ^2
+The same classification algorithm based on k-nearest neighbors is used for all the methods~\citep{Boiman2008}.
+The experiment is composed of 35 trials to spell 7 words of 5 letters each.
+Each trial is composed of 10 intensification sequences of the 6 columns and 6 rows of the Speller Matrix.
+The first $15$ trials are used to build the dictionary of templates, extracting the EEG segment for the row and column that contain the P300 ERP, hence shielding $30$ different templates (Fig.
+\ref{fig:dictionaryfig}).
+
+Described algorithms produce a feature $f$ for each EEG segment.
+The aim of the classification procedure is to identify for the remaining $20$ trials which of the 6 features $f$ that were obtained for the row/column intensifications is the one that elicited the P300 response on the averaged EEG segment.
+The row number of the matrix can be obtained by doing
+
+\begin{equation}
+\hat{row} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q_i=1}^{k} \left\lVert f_u - q_i \right\rVert ^2
 \label{eq:multiclassificationrow}
@@
 
-\noindent and
-
-\begin{equation}
-\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q \in NN_T(f^{col}_u)}^{} \left\lVert q -  f^{col}_u \right\rVert ^2
+\noindent with $q_i$ being the set of k-nearest neighbors of the feature $f_u$ with $u$ varying from $1$ to $6$.
+The parameter $k$ represents the number of neighbors chosen from the dictionary of templates.
+The column can be obtained by the same way,
+
+\begin{equation}
+\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q_i=1}^{k} \left\lVert f_u - q_i \right\rVert ^2
 \label{eq:multiclassificationcol}
@@
 
-\noindent where $NN_T(f^l_u),\;l\in\{row,{col}\}$  is the set of the $k$ nearest neighbors to $f^l_u$ and $q$ is a template feature that belongs to it.
-This set is obtained by sorting all the elements in the dictionary $T$ based on the distances between them and $f^l_u$, choosing the $k$ smaller elements.
-This procedure is a modification of the k-NBNN  algorithm~\citep{Boiman2008}.
-
-By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{row} $ and column $ \hat{col} $.
-%Figure~\ref{fig:classification} shows a scheme of this process.
+Thus, the letter identification performance can be obtained by measuring the accuracy at identifying the correct letter on the matrix, coordinated by $ \hat{row} $ and $ \hat{col} $.
 
@@
 \caption{Coherently averaged signals containing the superimposed ERP.
-Each one is extracted from the 15 first trials (2 signals from each trial, one belonging to the column and the other to the row).
-These are the templates used to build the dictionary $T$ and that are used by the classification algorithm described in \ref{section:classification}.}
+Each one is extracted from the 15 first trials (2 signals from each trial, one belonging to the column flashing and the other to row flashing).
+These are the templates used to build the dictionary and that are used by the classification algorithm described in \ref{section:classification}.}
 \label{fig:dictionaryfig}
@@
 Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.
-Table \ref{tab:results} shows the single-trial performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was obtained.
+Table \ref{tab:results} shows the single-intensification performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was attained.
 Figure \ref{fig:performancetest} shows the performance curves for eight algorithms.
 Each one represents the percentage of letters that were actually predicted by the combination of feature plus a classification method.
-The data were divided in two, and the first 15 letters were used to derive the dictionary of templates $T$ while the remaining 20 letters were used to measure the classification accuracy.
-Figure \ref{fig:performancetestlatency} shows the same results obtained when a Gaussian latency shift was added to each trial.
+As previously described the data were divided in two, and the first 15 letters were used to derive the dictionary of templates while the remaining 20 letters were used to measure the classification performance.
+Figure \ref{fig:performancetestlatency} shows the same results obtained when a Gaussian latency shift was included when the ERP was superimposed.
 Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained when the amplitude of the N1 and P3 component of the template are randomly reduced (uniform distribution).
@@
 \begin{table}[H]
-\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, HGO Histogram of Gradient Orientation, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
+\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
 Additionally, three control algorithms were included: RS Raw Signal, NN Neural Network and SVM Support Vector Machines.
 All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
-The mark \textit{m-c} is used on the control algorithms to identify that the feature was constructed using all the channels.
 In this case with absence of null-signals, it can be interpreted as the channel that adds less noise to the ERP template.
+The mark \textit{m-c} is used on the control algorithms to identify that the feature was crafted concatenating all the channels.
 All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal.
@@
 NN       & m-c  & $45\%$ \\
-HGO  & PO8 & $25\%$ \\
+SIFT  & PO8 & $25\%$ \\
 MP 1 & PO8 & $25\%$ \\
@@
 \includegraphics[width=15cm]{images/PerformanceTestLatency.eps}
-\caption{Speller performance obtained for each method while latencies were artificially added to each single-trial segment.  The achieved performance is significantly reduced for all methods.}
+\caption{Speller performance obtained for each method while latencies were artificially added to each single-intensification segment.
+The achieved performance is significantly reduced for all methods.}
 \label{fig:performancetestlatency}
@@
 
-The methods \textit{HGO} and \textit{SHCC} performed much better than the other algorithms including when the latency and the amplitudes are noisy.
+The methods \textit{SIFT} and \textit{SHCC} performed much better than the other algorithms including when the latency and the amplitudes are noisy.
 

### 2018-09-03 sensors.tex
A candidate version

@@
 \end{tabular}
-\label{tab:results}
+\label{tab:methods}
 \end{table}
@@
 \item Slope Horizontal Chain Code
-\item Histogram of Gradient Orientations
+\item Scale Invariant Feature Transform
 %\item Merging of Increasing and Decreasing Sequences
@@
 How much is it possible to classify and implement a P300 ERP Speller using only the waveform structure of the signal ?
-Which algorithm perform better at this even when single trial latency variations and component amplitudes issues are present ?
+Which algorithm perform better at this even when segment latency variations and component amplitudes issues are present ?
 
@@
 
+%\begin{table}[H]
+%\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
+Additionally, the control algorithms SVM Support Vector Machines was included.
+All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
+In this case with absence of null-signals, it can be interpreted as the channel that adds less noise to the ERP template.
+All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal.
+}
+%\centering
+%%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+%\begin{tabular}{cccc}
+%\toprule
+%\textbf{Method}	& \textbf{Channel} & \textbf{Single Intensification Performance} &  \textbf{Performance} 	\\
+%\midrule
+%SVM     & PO8  & $50\%$ & $70\%$ \\
+%SIFT  & PO8 & $25\%$ & $95\%$ \\
+%MP 1 & PO8 & $25\%$ & $70\%$ \\
+%PE     & Cz & $20\%$ & $55\%$ \\
+%MP 2 & PO7 & $10\%$ & $30\%$\\
+%SHCC & P4 & $10\%$ & $100\%$\\
+%\bottomrule
+%\end{tabular}
+%\label{tab:results}
+%\end{table}
+
 \begin{table}[H]
 \caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
-Additionally, three control algorithms were included: RS Raw Signal, NN Neural Network and SVM Support Vector Machines.
+Additionally, the control algorithms SVM Support Vector Machines was included.
 All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
 In this case with absence of null-signals, it can be interpreted as the channel that adds less noise to the ERP template.
-The mark \textit{m-c} is used on the control algorithms to identify that the feature was crafted concatenating all the channels.
 All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal.
@@
 %% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccc}
+\begin{tabular}{ccccc}
 \toprule
-\textbf{Method}	& \textbf{Channel} & \textbf{Single Trial Performance}	\\
+\textbf{Method}	& \textbf{Channel} &  \textbf{Performance} & \textbf{Perf w/Latency noise}	& \textbf{Perf w/Amplitude Noise}\\
 \midrule
-SVM     & m-c  & $50\%$ \\
-NN       & m-c  & $45\%$ \\
-SIFT  & PO8 & $25\%$ \\
-MP 1 & PO8 & $25\%$ \\
-PE     & Cz & $20\%$ \\
-RS     & Fz & $15\%$ \\
-RS     & m-c  & $15\%$ \\
-MP 2 & PO7 & $10\%$ \\
-SHCC & P4 & $10\%$ \\
-NN s-c & Fz & $10\%$ \\
+MP 1 & PO8  & $70\%$ & $25\%$ & $50\%$\\
+MP 2 & PO7 & $30\%$ & $15\%$ & $10\%$\\
+SIFT  & PO8 & $95\%$ & $30\%$ & $65\%$\\
+PE     & Cz & $55\%$ & $10\%$ & $40\%$\\
+SHCC & P4 & $100\%$ & $20\%$ & $75\%$\\
+SVM     & PO8  & $85\%$ & $10\%$ & $60\%$\\
 \bottomrule
@@
 
-The classification methods Support Vector Machine SVM and Artificial Neural Networks NN are added for comparison using the feature of Kasinksky (REF).
-Additionally, the Kasinksky feature is used with the same classification algorithm as detailed in \ref{section:classification}.
-
-All the methods show a significant reduction of performance when the latency noise is added, and at the same time all the methods show resistance to noise in peak amplitudes.
-
-Except for the raw signal, using a multichannel feature provides a better synthesis of the differentiation information that provides a better accuracy.
-This can be seen in Fig \ref{fig:performancetest} where the performance of the control algorithm NN is improved by using a multichannel feature.
-The poor raw signal performance while using a multichannel feasture is likely to the curse of dimensionality due to the fact that the feature is simple too large.
-
-Using a straightforward dictionary of templates for MP-1 proved more beneficial in terms of performance than the standard approach of using a Hilbert base of Wavelets atoms.
-
-The methods \textit{SIFT} and \textit{SHCC} performed much better than the other algorithms including when the latency and the amplitudes are noisy.
-
-It is important to remark that Segmentation or identification is not considered in these experiments.
-The  problem of segmentation is related to the detection of a feature within an EEG trace and not to the feature extraction problem.
+The classification method Support Vector Machine SVM is added for comparison using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
+
+All the methods show a significant reduction of performance when latency noise is added, and at the same time all the methods show some resistance to noise in peak amplitudes of the main components.
+
+Using a straightforward dictionary of templates for \textit{MP-1} proved more beneficial in terms of performance than the standard approach of using a Hilbert base of Wavelets atoms on \textit{MP-2}.
+Either applying latency noise or amplitude noise, the method based on the signal's templates instead of using their coefficient proved to achieve much better performance values.
+
+Results obtained for algorithms MP-1, SIFT, SHCC show them to be comparable with the results obtained for SVM.
+Particularly, the methods \textit{SIFT} and \textit{SHCC} performed much better than the other algorithms even when the latency and the amplitudes are noisy.
+Latency noise on the other hand, reduces the information contained in the signal, mainly due to the invalidation of the SNR enhancement performed by the averaging procedure.
+This reduction impacts on the obtained performance for all the methods, almost equally.
 
@@
 
-intelligible property
+The purpose of this work was threefold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain acceptable classification accuracies based exclusively on the signal's waveform.
+
+We verified that similar performance results are obtained for the methods \textit{MP-1}, \textit{SIFT} and \textit{SHCC} and that it is possible to obtain discriminating information from the underlying signal based exclusively on an automated method of processing the waveforms.
+This open the possibility to use these techniques to implement automatic detection procedures with a strong intelligible property, due to the fact that they are based on metrics which can be visually verified.
+
+%How much is it possible to classify and implement a P300 ERP Speller using only the waveform structure of the signal ?
+Which algorithm perform better at this even when segment latency variations and component amplitudes issues are present ?
+
+One of the main goals of the BCI discipline is to provide assistance to patients and to provide alternative tools to be used in diagnostics and rehabilitation procedure.
+This requires a clinical focus which is often neglected in BCI research~\citep{Chavarriaga2017}.
+At the same time, BCI reliability is yet an unfulfilled goal in this discipline~\citep{WolpawJonathanR2012}.
+The benefits of analyzing or including metrics about the shape of the EEG, are that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can lead to more robust implementation of BCI devices.
 
@@
 The need for more objective measurements pushed the adoption of more automated means of decoding the signals and demanded a need for replication \citep{Thakor2004}.
-This lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}
-
-The purpose of this work was threefold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain classification accuracies based exclusively on the signal's waveform.
-
-One of the main goals of the BCI discipline is to provide assistance to patients and to provide alternative tools to be used in diagnostics and rehabilitation procedure.
-This requires a clinical focus which is often neglected in BCI research (ref).
-At the same time, BCI reliability and the avoidance of the black box phenomena are goals that are still unfulfilled in this discipline(WOkpaw and wolpaw).
-The benefits of analysing or including metrics about the shape of the EEG, are that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can lead to more robust implementation of BCI devices.
+This lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}.
+We believe that the adoption of \textit{hybrid} methodology which can process the signal automatically, but at the same time, mantains an intelligible property that can be mapped to existing procedures, and above all, can mantain the clinician control of the procedure will be benefical to clinical practice, Neuroscience and BCI research.
 
@@
 All these methods had the advantage that they can truly map a visual component with a clinical meaning, assessed by a physician, and provide an automatic identification which can be constructed by the set of methods.
-
-More work has to be conducted in order to extend this methods to other patterns of waves.
-
-The procedure of analysing signals by their waveforms is relative common in chemical analysis (i.e.
-chemometrics Skoog, D.; West, D.; Holler, F.
-Analtyical Chemistry, An Introduction; Saunders: Philadelphia, 1994.), geology (sismic analysis), and quantitive financial analysis.
-EKG, or Electrocardiogram, on the other hand, has been extensively processed and analysed by waveform methods.
-Additionally, the same problem can be found in Intelligent Character Recognition.
+More work has to be conducted in order to extend this methods to other wave patterns.
+
+Another benefit of these approaches is that they have a potential universal applicability.
+As they are only analyzing waveform, they can be explored in other disciplines where the structure or shape of the waveform is of relevance.
+Analyzing signals by their waveforms is relative common in chemical analysis~\citep{Skoog2000}, seismic analysis in Geology~\citep{Owens1984}, and quantitative financial analysis.
+Electrocardiogram EKG, on the other hand, has been extensively processed and studying the waveform structure~\citep{Stockman1976}.
+%Additionally, the same problem can be found in Intelligent Character Recognition.
 
@@
 BCI: Brain Computer Interfaces\\
+BMI: Brain Machine Interfaces\\
+BNCI: Brain-Neural Computer Interfaces\\
 SNR: Signal to Noise Ratio\\
 CNS: Central Nervous System\\
-ALS: Amyotrophic Lateral Sclerosis\\
+DC: Direct Current\\
 ERP: Event-Related Potential\\
@@
 SIFT: Scale Invariant Feature Transform\\
-HOG: Histogram Of Gradients}
+SHCC: Slope Horizontal Chain Code\\
+PE: Permutation Entropy\\
+MP: Matching Pursuit\\
+ICU: Intensive Care Unit\\
+EKG: Electrocardiogram\\
+PAA:Period Amplitude Analysis
+}
 

### 2018-09-10 sensors.tex
Subo una nueva version super curada

@@
 The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community which was based intensively on EEG waveforms and the structure of signal plots.
-Hence, the purpose of this work is to help to start to fill this gap by doing a review and description of the procedures that have been used to detect patterns in the waveforms, and to perform a benchmarking analysis of them.
+Hence, the purpose of this work is to establish a bridge to fill this gap by doing a review and description of the procedures that have been used to detect patterns in the electroencephalographic waveforms, and to perform a benchmarking analysis of them.
 }
@@
 
-Current society is demanding technology to provide the means to realize the utopia of social inclusion for people with disabilities\citep{Wolpaw2002}.
-Additionally, as societies are aging \citep{Lutz2008} the incidence of neuromuscular atrophies, strokes and other invalidating diseases is increasing.
-Concurrently, the digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices \citep{Domingo2012}.
-All this communication is based on muscular movement \citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human motion.
-A new form of Human Machine communication which directly connects the Central Nervous System (CNS) to a machine or computer device is currently being developed: Brain Machine Interfaces (BMI), Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).
-
-At the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG) \citep{Schomer2010}.
-
-EEG sensors are wearable \citep{Puce2017} non-invasive, portable and mobile \citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution \citep{Hartman2005}.
-This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, or control wheelchairs \citep{Yuste2017}.
+Current society is demanding technology to provide the means to realize the utopia of social inclusion for people with disabilities~\citep{Wolpaw2002}.
+Additionally, as societies are aging~\citep{Lutz2008} the incidence of neuromuscular atrophies, strokes and other invalidating diseases is increasing.
+Concurrently, the digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices~\citep{Domingo2012}.
+All this human computer interaction is based on muscular movement~\citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human motion.
+A new form of human machine communication which directly connects the Central Nervous System (CNS) to a machine or computer device is currently being developed: Brain Machine Interfaces (BMI), Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).
+
+At the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG)~\citep{Schomer2010}.
+
+EEG sensors are wearable~\citep{Puce2017} non-invasive, portable and mobile~\citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution~\citep{Hartman2005}.
+This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, or control wheelchairs~\citep{Yuste2017}.
 
 The clinical and historical tactic to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph\citep{Hartman2005}: multichannel signals were extracted and continuously plotted over a piece of paper.
-Electroencephalographers or Electroencephalography technician have decoded and detected patterns along the signals by visually inspecting them \citep{Schomer2010}.
-Nowadays clinical EEG still remains a visually interpreted test \citep{Hartman2005}.
-
-In contrast, automatic processing, or quantitative EEG (qEEG), was based first on analog electronic devices~\citep{Jestico1977} and later on computerized digital processing methods \citep{Jansen1991}.
-They implemented mathematically and algorithmically complex procedures to decode the information associated with a clinical condition, or even some higher cognitive state \citep{Yuste2017}.
-The best materialization of the automatic processing of EEG signals rests in the BCI discipline, where around $71.2\%$ is based on noninvasive EEG \citep{Guger2017}.
+Electroencephalographers or Electroencephalography technician have decoded and detected patterns along the signals by visually inspecting them~\citep{Schomer2010}.
+Nowadays clinical EEG still remains a visually interpreted test~\citep{Hartman2005}.
+
+In contrast, automatic processing, or quantitative EEG (qEEG), was based first on analog electronic devices~\citep{Jestico1977} and later on computerized digital processing methods~\citep{Jansen1991}.
+They implemented algorithmically complex procedures to decode the information associated with a clinical condition, or even some higher cognitive state \citep{Yuste2017}.
+The best materialization of the automatic processing of EEG signals rests in the BCI discipline, where around $71.2\%$ is based on noninvasive EEG~\citep{Guger2017}.
 
@@
 
-Hence, the traditional and knowledgeable approach was mainly overshadow particularly in BCI Research, and the waveform of the EEG was replaced by pragmatically sound procedures that were difficult to link to existing clinical EEG knowledge.
+Hence, the traditional and knowledgeable approach was mainly overshadow particularly in BCI Research, and the waveform of the EEG was replaced by pragmatic procedures that were difficult to link to existing clinical EEG knowledge.
 We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and that can help to decode them in a supervised and semi-automated way.
@@
 The second is to evaluate and study these methods by analyzing its classification performance against a pseudo-real dataset.
-We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective \citep{Nijboer2009,Wei2017}.
+We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective~\citep{Nijboer2009,Wei2017}.
 
@@
 The first characterization that Dr.
-Berger detected was the Visual Cortical Alpha Wave, the \textit{Berger Rythm} \citep{Jansen1991}.
+Berger detected was the Visual Cortical Alpha Wave, the \textit{Berger Rythm}~\citep{Jansen1991}.
 He understood that the amplitude and shape of this rhythm was coherently associated to a cognitive action (eyes closing).
@@
 The EEG signal is a highly complex multi-channel time-series.
-It can be modeled as a linear stochastic process with great similarities to noise \citep{Thakor2004}.
+It can be modeled as a linear stochastic process with great similarities to noise ~\citep{Thakor2004}.
 It is measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.
+Figure \ref{fig:sampleeeg} show $5$ seconds of a sample 8-channel EEG signal.
 
@@
 
-The device that captures these small variations in current potentials over the scalp is called the Electroencephalograph.
-Electrodes are located in predetermined positions over the head, usually embedded in saline solutions to facilitate the electrophysiological interface and are connected to a differential amplifier with a high gain which allowed the measurement of tiny signals.
+The device that captures these small variations in potential differences over the scalp is called the Electroencephalograph.
+Electrodes are located in predetermined positions over the head, usually embedded in saline solutions to facilitate the electrophysiological interface and are connected to a differential amplifier with a high gain which allows the measurement of tiny signals.
 Although initially analog devices were developed and used, nowadays digital versions connected directly to a computer are pervasive.
@@
 \item Artifacts:  These are signal sources which are not generated from the CNS, but can be detected from the EEG signal.
-They are called endogeneous or physiological when they are generated from a biological source like muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic noise\citep{Weeda2012}.
+They are called endogeneous or physiological when they are generated from a biological source like face muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic noise\citep{Weeda2012}.
 \item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e.
@@
 \item Hypersynchrony: Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range.
-The term suggest an increase in the number of neural elements contributing to the rhythm.
+The term suggest an increase in the number of neural elements contributing to the rhythm, or in the synchronization of different neurons with the same discharge pattern~\citep{Buzsaki2012}.
 \item Paroxysmal: Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity. 
-This term does not directly imply abnormality per se, much abnormal activity is paroxysmal.
 \item Monomorphic: Activity appearing to be composed of one dominant waveform pattern.
@@
 The interictal epileptiform discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges\citep{EEGIntro}.
+
+\textbf{Waveform characterization is of quite importance in terms of Event Related Potentials.
+Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.
+Particularly, ERPs are regularly used to assess auditory response in infants.
+They are extensively used and studied in Cognitive Neuroscience ~\citep{Luck2005}.
+ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.
+The P300 is one of the most studied ERP component in BCI because it can be used to implement a Speller application.
+We will start with this process now.}
+
+%The clinical EEG patterns are identified by the shape of waveform amplitude plotted against time.
+
+%
+%Non-exhaustive list of transient events
+%
 
@@
 \begin{enumerate}
-\item The analysis take account the shape of the plot of the signal.
+\item The analysis consider the shape of the plot of the signal.
 \item The pattern can be identified and verified by visual inspection.
@@
 \item The feature extraction procedure allows to create a template dictionary.
-\item A single-channel processing scheme.
 \end{enumerate}
@@
 On the other hand, a \textit{feature} is a meaningful quantification, usually a multidimensional vector, that synthesize the information of a given signal or signal segment~\citep{WolpawJonathanR2012}.
-All these algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
 
@@
 
-Shape or waveform analysis methods are considered as nonparametric (in opposition to statistical or dynamical models).
-They explore signal's time-domain metrics or even derive more complex indexes or features from it \citep{Thakor2009}.
+Shape or waveform analysis methods are considered as nonparametric methods.  They explore signal's time-domain metrics or even derive more complex indexes or features from it \citep{Thakor2009}.
 
@@
 
-All these methods provide a feature $ f $  that can be used as a template, whereas all of them are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length of the EEG segment in sample points.
+All these methods provide a feature $ f $  that can be used as a template.
+The notation $()_{1}^{n} $ will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f$.
+Additionally, thy are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.
 These features are used to create dictionaries or template databases.
 Finally, these templates provide the basis for the pattern matching algorithm and offline classification.
+These algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
 
@@
 \begin{equation}
-\varepsilon =  \left\lVert   x(n) - \sum_{i=1}^{m} w_i g_{yi}(n)   \right\rVert
+\varepsilon =  \left\lVert   x(n) - \sum_{i=1}^{m} w_i g_{i}(n)   \right\rVert
 \label{eq:mperror}
@@
 
-which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{yi}$.
-The algorithm goes by first setting the approximating signal $\tilde{x_{0}}$  as the original signal itself,
-
-\begin{equation}
-\tilde{x_{0}}(n) = x(n), k \leftarrow 1
+which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.
+The algorithm goes by first setting the approximating signal $\tilde{x}_{0}$  as the original signal itself,
+
+\begin{equation}
+\tilde{x}_{0}(n) = x(n)
 \label{eq:mp2}
@@
 \begin{equation}
-g_{yk}(n) = \arg \max_{g_{yi}} \left\lVert < \tilde{x_{k}}(n), g_{yi}(n) > \right\rVert 
+g_{k} = \arg \max_{g_{i}} \left\lvert  \sum_{n=1}^{N} \tilde{x}_{k-1}(n) \; g_{i}(n) \right\rvert 
 \label{eq:mp3}
@@
 
-where $g_{yi}$ are all the available scaled, translated and modulated atoms from the dictionary.  This determines the atom selection process, and their contribution is calculated based on
-
-\begin{equation}
-w_{k} =  \frac{< \tilde{x_{k}}(n), g_{yk}(n) >}{  | g_{yk} | }
+where $g_{i}$ are all the available scaled, translated and modulated atoms from the dictionary.
+The operation $\left\lvert \cdot \right\rvert$ corresponds to the absolute value of the inner product.
+This step determines the atom selection process, and their contribution is calculated based on
+
+%The operation $\left\lvert < \cdot,\cdot > \right\rvert$ corresponds to the absolute value of the inner product.
+
+\begin{equation}
+w_{k} =  \frac{\sum_{n=1}^{N} \tilde{x}_{k-1}(n) \; g_{k}(n)}{  {\left\lVert  g_{k} \right\rVert}^{2} }
 \label{eq:mp4}
@@
 
-with $k$ representing the index of the selected atom $g_{yk}$.  Finally the projection of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
-
-\begin{equation}
-\tilde{x_{k+1}}(n)=  \tilde{x_k}(n) - w_{k} g_{yk} (n), k \leftarrow k + 1
+with $k$ representing the index of the selected atom $g_{k}$.  Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
+
+\begin{equation}
+\tilde{x}_{k+1}(n)=  \tilde{x}_{k}(n) - w_{k} g_{k} (n)
 \label{eq:mp5}
@@
 
+The stopping criteria can be established based on a limiting threshold on Equation \ref{eq:mperror} or based on a predetermined number of steps and selected atoms.
 Two variants of this algorithm are evaluated.
-In \textit{MP 1} the dictionary is directly constructed with the normalized templates from Fig \ref{fig:dictionaryfig}.
-In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.
-For the first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building, an eventually sparse, vector with them.
+In \textit{MP 1} the dictionary is directly constructed with the normalized templates directly extracted from the real signal segments which is a straightforward implementation of the pattern matching technique.
+In \textit{MP 2} the coefficients of Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary~\citep{Vareka2012}.
+For the first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building, an eventually sparse, vector with them:
 
@@
 
-Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and significantly for Epilepsy pre-ictal detection \citep{Bandt2002}.
-This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found on the signal synthesized as a numeric value called entropy~\citep{Nicolaou2010}.
-A signal on a window of length $W$ is represented by
+Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and increasingly for Epilepsy pre-ictal detection \citep{Bandt2002}.
+This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found on the signal.
+This numeric value can be calculated as information entropy~\citep{Nicolaou2010}.
+Let's consider a signal on a window of length $W$ represented by
 
@@
 
-and it can be resampled by $\tau$ intervals doing
+and resampled by $\tau$ intervals and starting from the sampling point $n$, doing
 
@@
 
-This sequence of order $m$ corresponds to one ordinal pattern $\pi_{n}$, for $n$ varying between $1$ and $W$.
-Along the signal window there can be at most $k$ different ordinal (and overlapping) patterns,
+This sequence is of order $m$, which is the number of sample points used to derive the ordinal element called $\pi$.
+There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a strict order relationship.
+For example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(3,1,2)$.
+Thus, along the signal window there can be at most $k$ different ordinal (and overlapping) elements $\pi_{s}$
 
@@
 
-\noindent with $k = W-(m-1) \tau$.  The probability density function for all the available permutations of order $m$ should be $ p = (p_1,p_2,...,p_{m!}) $ with $ \left\lVert p \right\lVert = 1 $.
-
-Hence, the set of time samples map to a new set of $k$ ordinal patterns, and the \textit{pdf} can be calculated by the empirical permutation entropy,
-
-\begin{equation}
-P_i = \frac{1}{k} \sum_{s=1}^{k} \left[ \pi_{s} = \pi_{i} \right]
+\noindent with $k = W-(m-1) \tau$.
+The probability density function \textit{pdf} for all the available permutations of order $m$ should be $ \textbf{p} = (p_1,p_2,...,p_{m!}) $ with $ \sum_{1}^{m!} p_{i} = 1 $.
+
+Hence, the time series window is mapped to a new set of $k$ ordinal elements, and the \textit{pdf} can be calculated by the empirical permutation entropy,
+
+\begin{equation}
+p_i = \frac{1}{k} \sum_{s=1}^{k} \left[ \pi_{s} = \pi_{i} \right]
 \label{eq:pe4}
@@
 \noindent with $1 \leq i \leq m!$.
-By using the Iverson Bracket, for each $i$ only those ordinal patterns that were effectively found are counted to estimate $p_i$.
+The Iverson Bracket $ \left[ \cdot \right] $ resolves to $1$ when their logical proposition argument is true, $0$ otherwise.
+Therefore, for each $i$ only those ordinal elements $\pi_{s}$ that were effectively found along the signal are counted to estimate $p_i$, and zero elsewhere.
 The empirical permutation entropy can be estimated from this histogram as,
@@
 \begin{equation}
-H(p) = \sum_{i=1}^{m!} p_{i} \; log \frac{1}{p_{i}}
+H(\textbf{p}) = \sum_{i=1}^{m!} p_{i} \; log \frac{1}{p_{i}}.
 \label{eq:pe5}
@@
 \begin{equation}
-f =  {\bigg ( H(p)_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
+f =  {\bigg ( H(\textbf{p})_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
 \label{eq:pe6}
@@
 
-A signal, can be represented by a list of ordered-pairs $e$,
+A signal of length $N$, can be represented by a list of ordered-pairs $e$,
 
@@
 \begin{equation}
-g = \lfloor n + ( m \Delta ) + 0.5 \rfloor
+G = \lfloor n + ( m \Delta ) + 0.5 \rfloor
 \label{eq:shcc2}
@@
 
-\noindent with $n$ being the original sampling index, and $ 1 \leq n \leq N $.  On the other hand, $\Delta$ can be obtained by
+\noindent with $n$ being the original sampling index on $ 1 \leq n \leq N $ and $\lfloor \cdot \rfloor$ being the floor operation, i.e.
+rounding of the number argument to the closest smaller integer number.
+On the other hand, $\Delta$ can be obtained by
 
@@
 
-\noindent with $ G < N $. Lastly, the value $m$ can be derived from
-
-\begin{equation}
-m = sign \bigg (  \frac{N-1}{\Delta} \bigg )  \bigg \lfloor \left\lvert \frac{T-1}{\Delta} \right\lvert \bigg \rfloor
+\noindent with $ G < N $ and using instead $\lceil \cdot \rceil$ as the ceil operation, the rounding to the closest bigger integer number. Lastly, the value $m$ can be derived from
+
+\begin{equation}
+m = sign \bigg (  \frac{N-1}{\Delta} \bigg )  \bigg \lfloor \left\lvert \frac{N-1}{\Delta} \right\lvert \bigg \rfloor
 \label{eq:shcc4}
@@
 \begin{equation}
-e = \left[ (x',y')_{1}, (x',y')_{s}, ..., (x',y')_{G} \right]
+e = \left[ (x',y')_{1}, ...,(x',y')_{s}, ..., (x',y')_{G} \right]
 \label{eq:shcc5}
@@
 
-The next step is the normalization of each ordered-pair according to
-
-\begin{equation}
-\textbf{x} = \frac{1}{\max(x') - min(x')} ( \textbf{x'} - \min(x') \textbf{1} )
+The next step is the normalization of each ordered-pair as vectors $\textbf{x'} = (x'_{1},...,x'_{G})$ and $\textbf{y'} = (y'_{1},...,y'_{G})$ according to
+
+\begin{equation}
+\hat{\textbf{x}} = \frac{( \textbf{x'} - \min(\textbf{x'}) \textbf{1} )}{\max(\textbf{x'}) - \min(\textbf{x'})} 
 \label{eq:shcc6}
@@
 \begin{equation}
-\textbf{y} = \frac{1}{\max(y') - min(y')} ( \textbf{y'} - \min(y') \textbf{1} )
+\hat{\textbf{y}} = \frac{( \textbf{y'} - \min(\textbf{y'}) \textbf{1} )}{\max(\textbf{y'}) - \min(\textbf{y'})} 
 \label{eq:shcc7}
@@
 
-\noindent effectively yielding $x,y \in [0,1]$. \\
+\noindent with $\textbf{1}$ being the vector of length $G$ with all their components equal to $1$.  This operation effectively normalize the components $\hat{x}_{s},\hat{y}_{s} \in [0,1]$.\\
 
@@
 \begin{equation}
-f = \bigg (  \frac{y_{s}-y_{s-1}}{x_{s}-x_{s-1}}  \bigg )_{2}^{G} 
+f = \bigg (  \frac{\hat{y}_{s}-\hat{y}_{s-1}}{\hat{x}_{s}-\hat{x}_{s-1}}  \bigg )_{2}^{G} 
 \label{eq:shcc8}
@@
 
-The first step of the algorithm is the plot generation based on single-channel EEG segments $ x(n) $.
+The first step of the algorithm is the plot generation based on single-channel EEG segments $ x(n) $.  Hence, this signal is normalized by the z-score~\citep{Zhang2013}:
 
@@
 \begin{equation}
-h = ( \max_{n} \tilde{x}(n) - \min_{n} \tilde{x}(n))
+h = ( \max_{n} \tilde{x}(n) - \min_{n} \tilde{x}(n)).
 \label{eq:sift4}
@@
 
-Once the plot is generated, the plot's center is used to localize the SIFT patch.
+Once the plot is generated, the plot's center is used to localize the center of the SIFT patch.
 This region of the image, where the signal's most important salient waveform should be located, is divided in a grid of $4 x 4$ block and the bidimensional gradient vectors are calculated on each one of them.
-At the same time, for each block within the patch, a histogram of the gradient orientations, 8 circular orientations, are calculated.
+Therefore, for each block within the patch, a histogram of the gradient orientations, 8 circular orientations, are calculated.
 This histogram is concatenated for all the 16 blocks and a feature is thus formed:
@@
 \begin{equation}
-f = {\bigg ( h(i,j,\theta) \bigg )}_{i,j,\theta}
+f = {\bigg ( h(i,j,\theta) \bigg )}_{(0,0,0)}^{(3,3,315)}
 \label{eq:sift7}
@@
 \noindent with $i$ and $j$ varying between $0$ and $3$, localizing the 16 blocks within the grid.
-The angles $\theta$ are the eight possible values between $0$ and $315$.
-This vector is normalized, clapped to $0.2$, and re-normalized again.
+The angles $\theta$ are the eight possible equidistant values between $0$ and $315$.
+This vector is normalized, clamped to $0.2$, and re-normalized again.
 Details of the method can be found on \citep{Ramele2016,Lowe2004}.
@@
 \begin{itemize}
-\item Pseudo-real dataset Classification Performance: the letter classification performance of each one of these methods.
-\item Latency Noise: Gaussian noise is added to the ERP template latency, i.e. the temporal location where it is injected into the EEG stream.
+\item Pseudo-Real Dataset Classification Performance: the letter classification performance of each one of these methods applied on the artificially generated pseudo-real dataset.
+\item Latency Noise: Uniform noise is added to the ERP template latency, i.e.
+the temporal location where it is injected into the EEG stream.
+The latency range from $0$ to $40\%$ of the $1s$ of the window size.
 \item Component Amplitude Noise: the amplitude of the main P3b components of the ERP template is altered based on uniform distribution.
+This component is completely substracted from the original template or not substracted at all.
+The substraction percentage is obtained from a uniform distribution.
+Additionally the substraction curve is warped in a gaussian window of 800 ms.
 \end{itemize}
@@
 A subject participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
-No monetary compensation is handed out and he/she agree and sign a written informed consent.
+No monetary compensation is handed out and she/he agrees and signs a written informed consent.
 This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.
 The participant is healthy and have normal or corrected-to-normal vision and no history of neurological disorders.
-This voluntary subject is of unspecified gender, aged between 20-30 years old.
+This voluntary subject is aged between 20-30 years old.
 EEG data is collected in a single recording session.
@@
 Channel $S$ represents the twelve different stimulus markers (columns or rows) while channel $L$ represent the label (hit vs nohit).
+Nohit are marked as a square signal and hit are markers are identified by the double-amplitude square signals.
 Subfigure (a) shows the signals before the ERP template is superimposed while subfigure (b) shows the same signals with the superimposed ERP template.
@@
 \includegraphics[width=1.0\linewidth]{images/GainCheck.eps}
-\caption{Point-to-point averaged signals for the first trial.  The ERP was superimposed on classes 3 and 9.}
+\caption{Point-to-point averaged signals for the first letter identification trial.
+The ERP was superimposed on classes 3 and 9.
+Class 3 was obtained while averaging the segments where the row of the speller matrix was intensified whereas class 9 was calculated from the intensification of the corresponding column.}
 \label{fig:gaincheck}
@@
 Described algorithms produce a feature $f$ for each EEG segment.
-The aim of the classification procedure is to identify for the remaining $20$ trials which of the 6 features $f$ that were obtained for the row/column intensifications is the one that elicited the P300 response on the averaged EEG segment.
+The aim of the classification procedure is to identify for the remaining $20$ trials which of the 6 features $f$ that were obtained for row intensification and which of the 6 features for column intensifications are the ones that elicited the P300 response on the averaged EEG segment.
 The row number of the matrix can be obtained by doing
@@
 \begin{equation}
-\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q_i=1}^{k} \left\lVert f_u - q_i \right\rVert ^2
+\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q_i=1}^{k} \left\lVert f_u - q_i \right\rVert ^2.
 \label{eq:multiclassificationcol}
@@
 
+Finallly, the classification method Support Vector Machine SVM is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
+This method is bla bla bla due to the reason bla bla bla.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 As previously described the data were divided in two, and the first 15 letters were used to derive the dictionary of templates while the remaining 20 letters were used to measure the classification performance.
-Figure \ref{fig:performancetestlatency} shows the same results obtained when a Gaussian latency shift was included when the ERP was superimposed.
-Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained when the amplitude of the N1 and P3 component of the template are randomly reduced (uniform distribution).
+Figure \ref{fig:performancetestlatency} shows the same results obtained when the ERP was superimposed with a noisy latency shift.
+Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained when the amplitude of the P3b component of the template are randomly attenuated.
 
@@
 \caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
-Additionally, the control algorithms SVM Support Vector Machines was included.
+Additionally, the control algorithms SVM Support Vector Machines was included, with a linear kernel.
 All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
@@
 \midrule
-MP 1 & PO8  & $70\%$ & $25\%$ & $50\%$\\
-MP 2 & PO7 & $30\%$ & $15\%$ & $10\%$\\
-SIFT  & PO8 & $95\%$ & $30\%$ & $65\%$\\
-PE     & Cz & $55\%$ & $10\%$ & $40\%$\\
-SHCC & P4 & $100\%$ & $20\%$ & $75\%$\\
-SVM     & PO8  & $85\%$ & $10\%$ & $60\%$\\
+MP 1 & PO8  & $67\%$ & $15\%$ & $50\%$\\
+MP 2 & PO7 & $24\%$ & $6\%$ & $10\%$\\
+SIFT  & PO8 & $91\%$ & $18\%$ & $66\%$\\
+PE     & Cz & $61\%$ & $9\%$ & $32\%$\\
+SHCC & P4 & $98\%$ & $31\%$ & $80\%$\\
+SVM     & PO8  & $78\%$ & $7\%$ & $53\%$\\
 \bottomrule
@@
 \centering
-\includegraphics[width=15cm]{images/PerformanceTest.eps}
+\includegraphics[width=15cm]{images/CrossPerformanceTest.eps}
 \caption{Speller performance obtained for each method.
@@
 \centering
-\includegraphics[width=15cm]{images/PerformanceTestLatency.eps}
+\includegraphics[width=15cm]{images/CrossPerformanceTestLatency.eps}
 \caption{Speller performance obtained for each method while latencies were artificially added to each single-intensification segment.
-The achieved performance is significantly reduced for all methods.}
+The achieved performance is significantly reduced for all methods.
+Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
 \label{fig:performancetestlatency}
@@
 \centering
-\includegraphics[width=15cm]{images/PerformanceTestAmplitude.eps}
-\caption{Speller performance obtained while the amplitudes of the N1 and P3 component of the superimposed ERP are randomly reduced.}
+\includegraphics[width=15cm]{images/CrossPerformanceTestAmplitude.eps}
+\caption{Speller performance obtained while the amplitudes of the N1 and P3 component of the superimposed ERP are randomly reduced.
+Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
 \label{fig:performancetestamplitude}
@@
 \label{section:discussion}
-
-The classification method Support Vector Machine SVM is added for comparison using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
 
@@
 This lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}.
-We believe that the adoption of \textit{hybrid} methodology which can process the signal automatically, but at the same time, mantains an intelligible property that can be mapped to existing procedures, and above all, can mantain the clinician control of the procedure will be benefical to clinical practice, Neuroscience and BCI research.
-
-Successful approaches in Computer Vision or pattern recognition in other areas use a set of different features to compound ensemble classifiers.
-All these methods had the advantage that they can truly map a visual component with a clinical meaning, assessed by a physician, and provide an automatic identification which can be constructed by the set of methods.
-More work has to be conducted in order to extend this methods to other wave patterns.
+We believe that the adoption of \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an intelligible property that can be mapped to existing procedures, and above all, can mantain the clinician control of the procedure will be beneficial to clinical practice, Neuroscience and BCI research.
+
+These methods have the advantage that they can map a visual component with a clinical meaning to feature with a clear mathematical representation.
+Successful approaches in Computer Vision and Pattern Recognition in other areas use an ensemble of features to compound classifiers~\citep{Criminisi2013}, with a significant enhancement of the classification performances~\citep{Gu2012}.
 
@@
 Analyzing signals by their waveforms is relative common in chemical analysis~\citep{Skoog2000}, seismic analysis in Geology~\citep{Owens1984}, and quantitative financial analysis.
-Electrocardiogram EKG, on the other hand, has been extensively processed and studying the waveform structure~\citep{Stockman1976}.
+Electrocardiogram EKG, on the other hand, has been extensively processed and studied analyzing the waveform structure~\citep{Stockman1976}.
 %Additionally, the same problem can be found in Intelligent Character Recognition.

### 2018-09-14 sensors.tex
Cambios
Small changes

@@
 
+%\usepackage{draftwatermark}
+%\SetWatermarkText{DRAFT}
+%\SetWatermarkScale{1}
 \usepackage{subfigure}
@@
 
-\textbf{Waveform characterization is of quite importance in terms of Event Related Potentials.
+Waveform characterization is the method in which analysis has been performed for Event Related Potentials.
 Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.
 Particularly, ERPs are regularly used to assess auditory response in infants.
-They are extensively used and studied in Cognitive Neuroscience ~\citep{Luck2005}.
+They are extensively used and studied in Cognitive Neuroscience~\citep{Luck2005}.
 ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.
 The P300 is one of the most studied ERP component in BCI because it can be used to implement a Speller application.
-We will start with this process now.}
+Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.
 
@@
 \begin{itemize}
-\item Pseudo-Real Dataset Classification Performance: the letter classification performance of each one of these methods applied on the artificially generated pseudo-real dataset.
-\item Latency Noise: Uniform noise is added to the ERP template latency, i.e.
+\item Experiment 1 - Pseudo-Real Dataset Classification Performance: the letter classification performance of each one of these methods applied on the artificially generated pseudo-real dataset.
+\item Experiment 2 - Latency Noise: Uniform noise is added to the ERP template latency, i.e.
 the temporal location where it is injected into the EEG stream.
 The latency range from $0$ to $40\%$ of the $1s$ of the window size.
-\item Component Amplitude Noise: the amplitude of the main P3b components of the ERP template is altered based on uniform distribution.
+\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b components of the ERP template is altered based on uniform distribution.
 This component is completely substracted from the original template or not substracted at all.

### 2018-09-17 sensors.tex
Confirmed section

@@
 \pubvolume{xx}
-\pubyear{2017}
-\copyrightyear{2017}
+\pubyear{2018}
+\copyrightyear{2018}
 \externaleditor{Academic Editor: name}
@@
 Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.
-Particularly, ERPs are regularly used to assess auditory response in infants.
+ERPs are regularly used to assess auditory response in infants.
 They are extensively used and studied in Cognitive Neuroscience~\citep{Luck2005}.
@@
 Straightforward in its implementation, it consists in assigning a component to a simple waveform element based on the expected location of its more prominent deflection \citep{Ouyang2017}.
-Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient component that may arise as a brain response to an external visual, tactile or auditory stimulus.
+Of regular use in ERP Research, the name of many of the EEG features reference directly a peak within the component, e.g.
+P300 or P3a P3b or N100.
+This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.
+The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).
+
+%Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient component that may arise as a brain response to an external visual, tactile or auditory stimulus.
 Particularly, ERPs are regularly used to assess auditory response in infants.
@@
 All these methods provide a feature $ f $  that can be used as a template.
-The notation $()_{1}^{n} $ will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f$.
-Additionally, thy are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.
+The notation $\{\}_{1}^{n} $ will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f$.
+These algorithms are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.
 These features are used to create dictionaries or template databases.
@@
 Matching Pursuit \textit{MP}~\citep{Mallat1993}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.
-Given a signal, this optimization technique, tries to find the indexes of atoms and their weights (contributions) that minimize,
+Given a signal, this optimization technique, tries to find the indexes of $m$ atoms and their weights (contributions) that minimize,
 
@@
 
-which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.
+\noindent which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.
 The algorithm goes by first setting the approximating signal $\tilde{x}_{0}$  as the original signal itself,
@@
 
-where $g_{i}$ are all the available scaled, translated and modulated atoms from the dictionary.
+\noindent where $g_{i}$ are all the available scaled, translated and modulated atoms from the dictionary.
 The operation $\left\lvert \cdot \right\rvert$ corresponds to the absolute value of the inner product.
@@
 
-with $k$ representing the index of the selected atom $g_{k}$.  Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
-
-\begin{equation}
-\tilde{x}_{k+1}(n)=  \tilde{x}_{k}(n) - w_{k} g_{k} (n)
+\noindent with $k$ representing the index of the selected atom $g_{k}$ and $\left\lVert \cdot \right\rVert$ its norm.
+Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
+
+\begin{equation}
+\tilde{x}_{k}(n)=  \tilde{x}_{k-1}(n) - w_{k} g_{k} (n)
 \label{eq:mp5}
@@
 \begin{equation}
-f =  {\bigg ( w_{i} \bigg )}_{1}^{D} 
+f =  {\bigg \{ w_{i} \bigg \}}_{1}^{D} 
 \label{eq:mp6}
@@
 \begin{equation}
-f =  {\bigg ( H(\textbf{p})_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
+f =  {\bigg \{ H(\textbf{p})_{u} \bigg \}}_{1}^{N - (W + \tau (m - 1)) + 1}
 \label{eq:pe6}
@@
 \begin{equation}
-f = \bigg (  \frac{\hat{y}_{s}-\hat{y}_{s-1}}{\hat{x}_{s}-\hat{x}_{s-1}}  \bigg )_{2}^{G} 
+f = \bigg \{  \frac{\hat{y}_{s}-\hat{y}_{s-1}}{\hat{x}_{s}-\hat{x}_{s-1}}  \bigg \}_{2}^{G} 
 \label{eq:shcc8}
@@
 \begin{equation}
-f = {\bigg ( h(i,j,\theta) \bigg )}_{(0,0,0)}^{(3,3,315)}
+f = {\bigg \{ h(i,j,\theta) \bigg \}}_{(0,0,0)}^{(3,3,315)}
 \label{eq:sift7}

### 2018-09-18 sensors.tex
Few changes

@@
 \item Experiment 2 - Latency Noise: Uniform noise is added to the ERP template latency, i.e.
-the temporal location where it is injected into the EEG stream.
-The latency range from $0$ to $40\%$ of the $1s$ of the window size.
-\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b components of the ERP template is altered based on uniform distribution.
-This component is completely substracted from the original template or not substracted at all.
-The substraction percentage is obtained from a uniform distribution.
-Additionally the substraction curve is warped in a gaussian window of 800 ms.
+to the temporal location where it is injected into the EEG stream.
+The latency lag ranges from $0$ to $40\%$ of the $1s$ of the window size.
+\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b components of the ERP template is randomly altered.
+This component is subtracted from the original template.
+The subtraction percentage is obtained from a uniform distribution.
+Additionally the subtracted curve is warped in a Gaussian window of 800 ms.
 \end{itemize}
+
+These are executed using 3-fold cross validation procedure.
 

### 2018-09-18 sensors.tex
Added the BCI Competition results

@@
 Hence, the traditional and knowledgeable approach was mainly overshadow particularly in BCI Research, and the waveform of the EEG was replaced by pragmatic procedures that were difficult to link to existing clinical EEG knowledge.
-We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and that can help to decode them in a supervised and semi-automated way.
+We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and which can decode them in a supervised and semi-automated procedure.
 
@@
 The second is to evaluate and study these methods by analyzing its classification performance against a pseudo-real dataset.
-We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective~\citep{Nijboer2009,Wei2017}.
+We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal waveform, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective~\citep{Nijboer2009,Wei2017}.
 
@@
 It is measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.
-Figure \ref{fig:sampleeeg} show $5$ seconds of a sample 8-channel EEG signal.
+Figure \ref{fig:sampleeeg} shows $5$ seconds of a sample 8-channel EEG signal.
 
@@
 positive or negative peak polarity) or the location within the strip.
-It is regular to establish an amplitude difference between different waveforms from which a relation between them is established and a structured index are created (e.g.
+It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g.
 sleep K-Complex is well characterized based on rates between positive vs negative amplitude) \citep{Uchida1999}.
@@
 \begin{enumerate}
-\item The analysis consider the shape of the plot of the signal.
+\item The analysis considers the shape of the plot of the signal.
 \item The pattern can be identified and verified by visual inspection.
@@
 
-These are executed using 3-fold cross validation procedure.
-
-%\textbf{Borrar o hacer: Finally results for the same methods while offline processing a public dataset of real P300 patterns are also shown.}
+These are executed using 3-fold cross validation procedure. Finally results for the same methods while performing an offline BCI Simulation of the dataset  IIb of the BCI Competition II (2003).
 
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=15cm]{images/PerformanceBCICompetition.eps}
+\caption{Speller performance obtained for the dataset IIb of the BCI Competition II (2003) for each one of the algorithms.
+An offline BCI Simulation is performed using the first $42$ trials as training and the remaining $31$ as testing.}
+\label{fig:performancebcicompetition}
+\end{figure}
+
+\begin{table}[H]
+\caption{Speller classification performance obtained for the BCI Competition II (2003) for each one of the algorithms using 15 repetitions of intensification sequences.
+The first $42$ trials are used for training to build the template dictionary and the remaining $31$ for testing.
+}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+\begin{tabular}{ccc}
+\toprule
+\textbf{Method}	& \textbf{Channel} &  \textbf{Performance} \\
+\midrule
+MP 1 & FC2  & $50\%$ \\
+MP 2 & CPz & $22\%$ \\
+SIFT  & Cz & $67\%$ \\
+PE     & PO8 & $22\%$ \\
+SHCC & Cz & $61\%$ \\
+SVM     & C1  & $32\%$ \\
+\bottomrule
+\end{tabular}
+\label{tab:results}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-09-19 sensors.tex
Added info on SVM

@@
 Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and increasingly for Epilepsy pre-ictal detection \citep{Bandt2002}.
-This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found on the signal.
+This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found along the signal.
 This numeric value can be calculated as information entropy~\citep{Nicolaou2010}.
@@
 Therefore, for each $i$ only those ordinal elements $\pi_{s}$ that were effectively found along the signal are counted to estimate $p_i$, and zero elsewhere.
-The empirical permutation entropy can be estimated from this histogram as,
+The empirical permutation entropy can be calculated from this histogram as,
 
@@
 The implemented code was derived from~\citep{Unakafova2013}, and the model description from \citep{Berger2017}.
-This procedure will produce a scalar number for the given signal window of size $W$.
+This procedure produces a scalar number for the given signal window of size $W$.
 To derive a feature, a sliding window procedure must be implemented.
@@
 \noindent with $\delta$ being the signal magnification factor and $\bar{x}$ and $\sigma_{x}$, the mean and standard deviation of $x$ on the signal segment.
-The width of the image is determined based on the $1$-second length size of the segment window in sample units.
+The width of the image is determined based on the $1$-second length size of the segment in sample units.
 This corresponds to the sampling frequency $F_s$ of the EEG signal segment.
@@
 
-\noindent where $z_1$ and $z_2$ are the image coordinates values.  These points are interpolated using the Bresenham algorithm.
-
-Once the plot is generated, the plot's center is used to localize the center of the SIFT patch.
+\noindent where $z_1$ and $z_2$ are the image coordinates values.  These points are interpolated using the Bresenham algorithm~\citep{Ramele2016}.
+
+Once the plot is generated, its center is used to localize the center of the SIFT patch.
 This region of the image, where the signal's most important salient waveform should be located, is divided in a grid of $4 x 4$ block and the bidimensional gradient vectors are calculated on each one of them.
@@
 \begin{itemize}
-\item Experiment 1 - Pseudo-Real Dataset Classification Performance: the letter classification performance of each one of these methods applied on the artificially generated pseudo-real dataset.
-\item Experiment 2 - Latency Noise: Uniform noise is added to the ERP template latency, i.e.
-to the temporal location where it is injected into the EEG stream.
-The latency lag ranges from $0$ to $40\%$ of the $1s$ of the window size.
-\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b components of the ERP template is randomly altered.
+\item Experiment 1 - Pseudo-Real Dataset Classification Performance: the letter classification performance of each one of these methods on the artificially generated pseudo-real dataset.
+\item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERP over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
+This value is obtained from a uniform distribution ranging from $0$ to $40\%$ of the $1s$ of the window size~\citep{DaPelo2018}.
+\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.
 This component is subtracted from the original template.
@@
 Additionally the subtracted curve is warped in a Gaussian window of 800 ms.
+
 \end{itemize}
 
-These are executed using 3-fold cross validation procedure. Finally results for the same methods while performing an offline BCI Simulation of the dataset  IIb of the BCI Competition II (2003).
+The classification method Support Vector Machine SVM~\citep{Kaper2004} is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
+This method has been proved efficient in decoding P300 in several BCI Competitions.
+
+All these experiments are executed using 3-fold cross validation procedure that preserves the structure of the letter identification trials.
+Spelling letters are scrambled while the order and group of each intensification sequence is preserved.
+
+Finally the performance at letter identifications for these same methods is evaluated by performing an offline BCI Simulation on the Dataset IIb of the BCI Competition II (2003)~\citep{Blankertz2002}.
 
@@
 \end{figure}
-
-Finallly, the classification method Support Vector Machine SVM is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
-This method is bla bla bla due to the reason bla bla bla.
 

### 2018-09-19 sensors.tex
Some figures were modified. Fixed many sections

@@
 Rectangular     & Epilepsy &  \citep{Thakor2004,Cole2017} \\
+Line length       & Anomaly Detection & \citep{Wulsin2011} \\
+Root Mean Square & Anomaly Detection & \citep{Wulsin2011} \\
 Wicket Shape     & Epilepsy &  \citep{EEGIntro,Hartman2005,Sanei2007,Tatum2008,Schomer2010,Cole2017} \\
@@
 Base-to-Peak Amplitude     & ERP  &  \citep{Cole2017} \\
-Peak-to-Peak Amplitude     & ERP  &  \citep{Mak2012} \\
+Peak-to-Peak Amplitude     & ERP  &  \citep{Wulsin2011,Mak2012} \\
 Positive/Negative Peak Latency                                 & ERP  & \citep{Mak2012}  \\
-Integrated Activity               & ERP, Epilepsy,ICU & \citep{Uchida1999, Shah2015} \\
+Integrated Activity               & ERP, Epilepsy,ICU & \citep{Wulsin2011,Uchida1999, Shah2015} \\
 Cross-Correlation                & ERP, Epilepsy, Sleep & \citep{Cacioppo2007, Shah2015} \\
@@
 
-How much is it possible to classify and implement a P300 ERP Speller using only the waveform structure of the signal ?
+Is it possible to classify and implement a P300 ERP Speller using only the waveform structure of the signal ?
 Which algorithm perform better at this even when segment latency variations and component amplitudes issues are present ?
 
-In order to verify this in a controlled procedure, a real P300 ERP template, obtained from a public dataset, is superimposed into a different EEG stream.
+In order to verify this in a controlled procedure, a real P300 ERP template, obtained from a public dataset, is superimposed into a distinct EEG stream.
 This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on any letter in particular.
@@
 \item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERP over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
-This value is obtained from a uniform distribution ranging from $0$ to $40\%$ of the $1s$ of the window size~\citep{DaPelo2018}.
+The lagging value is picked from a uniform distribution ranging from $0$ to $40\%$ of the $1s$ segment size~\citep{DaPelo2018}.
 \item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.
 This component is subtracted from the original template.
-The subtraction percentage is obtained from a uniform distribution.
-Additionally the subtracted curve is warped in a Gaussian window of 800 ms.
-
+The subtraction percentage is drawn from a uniform distribution.
+Additionally the subtracted curve is multiplied by a Gaussian window of $800$ ms.
+This avoids adding any discontinuity into the artificial generated signal.
 \end{itemize}
@@
 
-All these experiments are executed using 3-fold cross validation procedure that preserves the structure of the letter identification trials.
+All these experiments are executed using two fold cross validation procedure that preserves the structure of the letter identification trials.
 Spelling letters are scrambled while the order and group of each intensification sequence is preserved.
@@
 The template ERP is extracted from the Subject Number $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
-Segments from the EEG signal are labeled as hit and are extracted for the trial number $2$, and they are point-to-point coherently averaged.
-This P300 ERP can be seen in Fig.
-\ref{fig:erptemplate1}.
+Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.
+This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.
 
@@
 
-Fig.
-\ref{fig:gains} shows a $5s$ sample of the EEG trace.
-Channel $S$ represents the twelve different stimulus markers (columns or rows) while channel $L$ represent the label (hit vs nohit).
-Nohit are marked as a square signal and hit are markers are identified by the double-amplitude square signals.
+Along the EEG stream, the markers information was used to localize the \textit{True} segments where the P300 should have been found, and those timing locations are used to superimpose the extracted ERP waveform.
+
+Figure~\ref{fig:gains} shows a $5s$ sample of the EEG trace obtained with the MNE library~\citep{Gramfort2013}.
+Channel $S$ represents the twelve different stimulus markers (columns or rows) while channel $L$ represent the label (\textit{True} vs \textit{False}).
+\textit{False} segments are marked as a square signal and \textit{True} segments are identified by the double-amplitude square signals.
 Subfigure (a) shows the signals before the ERP template is superimposed while subfigure (b) shows the same signals with the superimposed ERP template.
-At first sight differences are really hard to spot visually.
+At first-sight, differences are really hard to spot visually.
 Subfigures (c) and (d) show only one second of the same segment only channel Cz and Label.
-The superimposed ERP can be divised around $31.5$s where in (d) the peak is bigger.
-Fig.
-\ref{fig:gaincheck} shows the obtained ensemble average ERPs as result of superimposing the template signal into EEG stream, time-locked to the stimulus onset.
+The superimposed ERP can be devised enclosed by the vertical bars, around $31.5$s, where in (d) the peak is slightly bigger.
+Figure~\ref{fig:gaincheck} shows the obtained ensemble average ERPs as result of superimposing the template signal into EEG stream, time-locked to the stimulus onset.
 These 12 point-to-point averaged segments correspond to the first trial of the EEG stream.
@@
 The template is $1s$ long which is 256 sample points, and the eight channels are superimposed with different colors.
-The P1, N1 and P300 peaks can be seen.}
+The P3b component can be seen around the sample index $150$ and $200$.}
 \label{fig:erptemplate1}
@@
 \subfigure[EEG trace of the original signal.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
-\subfigure[The same signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
-\subfigure[EEG sample of Cz channel of the original EEG trace.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
+\subfigure[The same eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
+\subfigure[EEG sample of Cz and L channel of the original EEG trace.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
 \subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
-\caption{Eight-channel EEG signal superimposed with the ERP Template.}
+\caption{Eight-channel EEG signal without and with the superimposed ERP Template.
+The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented.
+On (c) and (d) the small variation that was introduced by the superimposition of the ERP can be seen enclosed by the vertical bars, where the slope of the bump on subfigure (d) is slightly bigger}
 \label{fig:gains}
@@
 The same classification algorithm based on k-nearest neighbors is used for all the methods~\citep{Boiman2008}.
-The experiment is composed of 35 trials to spell 7 words of 5 letters each.
-Each trial is composed of 10 intensification sequences of the 6 columns and 6 rows of the Speller Matrix.
-The first $15$ trials are used to build the dictionary of templates, extracting the EEG segment for the row and column that contain the P300 ERP, hence shielding $30$ different templates (Fig.
-\ref{fig:dictionaryfig}).
+The experiment is composed of $35$ trials to spell $7$ words of $5$ letters each.
+Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.
+The first $15$ trials are used to build the dictionary of templates, extracting the averaged EEG segments for the row and column that already contain the P300 ERP, hence shielding $30$ different templates per channel.
+Figure~\ref{fig:dictionaryfig} shows the set of templates while using the first $15$ trials of the dataset.
 
@@
 
-Thus, the letter identification performance can be obtained by measuring the accuracy at identifying the correct letter on the matrix, coordinated by $ \hat{row} $ and $ \hat{col} $.
+Thus, the letter identification performance can be obtained by measuring the accuracy channel-by-channel at identifying the correct letter on the matrix, coordinated by $ \hat{row} $ and $ \hat{col} $.
 
@@
 Each one is extracted from the 15 first trials (2 signals from each trial, one belonging to the column flashing and the other to row flashing).
-These are the templates used to build the dictionary and that are used by the classification algorithm described in \ref{section:classification}.}
+These are the templates used to build a dictionary per channel and that are used by the classification algorithm described in \ref{section:classification}.}
 \label{fig:dictionaryfig}
@@
 Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.
-Table \ref{tab:results} shows the single-intensification performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was attained.
-Figure \ref{fig:performancetest} shows the performance curves for eight algorithms.
-Each one represents the percentage of letters that were actually predicted by the combination of feature plus a classification method.
-As previously described the data were divided in two, and the first 15 letters were used to derive the dictionary of templates while the remaining 20 letters were used to measure the classification performance.
-Figure \ref{fig:performancetestlatency} shows the same results obtained when the ERP was superimposed with a noisy latency shift.
-Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained when the amplitude of the P3b component of the template are randomly attenuated.
-
-%\begin{table}[H]
-%\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
-Additionally, the control algorithms SVM Support Vector Machines was included.
-All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
-In this case with absence of null-signals, it can be interpreted as the channel that adds less noise to the ERP template.
-All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal.
-}
-%\centering
-%%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-%\begin{tabular}{cccc}
-%\toprule
-%\textbf{Method}	& \textbf{Channel} & \textbf{Single Intensification Performance} &  \textbf{Performance} 	\\
-%\midrule
-%SVM     & PO8  & $50\%$ & $70\%$ \\
-%SIFT  & PO8 & $25\%$ & $95\%$ \\
-%MP 1 & PO8 & $25\%$ & $70\%$ \\
-%PE     & Cz & $20\%$ & $55\%$ \\
-%MP 2 & PO7 & $10\%$ & $30\%$\\
-%SHCC & P4 & $10\%$ & $100\%$\\
-%\bottomrule
-%\end{tabular}
-%\label{tab:results}
-%\end{table}
+Table \ref{tab:results} shows the  performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was attained.
+Figure \ref{fig:performancetest} shows the performance curves for six algorithms.
+Each one represents the percentage of letters that were actually predicted by the algorithms using a cross-validation procedure.
+As previously described the data were continuously divided in two folds, where the first 15 letters were used to derive the dictionary of templates while the remaining 20 letters were used to measure the classification performance.
+This is repeated one hundred times, and performances averaged.
+Figure \ref{fig:performancetestlatency} shows the same results for the Experiment 2, where a noisy latency lag was included.
+Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template are randomly attenuated.
+Furthermore, results obtained for the dataset BCI Competition 2003 IIb are shown in Figures \ref{fig:performancebcicompetition} and in Table~\ref{tab:bcicompetitionresults}.
 
@@
 \begin{table}[H]
-\caption{Speller classification performance obtained for the BCI Competition II (2003) for each one of the algorithms using 15 repetitions of intensification sequences.
+\caption{Speller classification performance obtained for the dataset IIb of the BCI Competition II (2003) for each one of the algorithms using $15$ repetitions of intensification sequences.
 The first $42$ trials are used for training to build the template dictionary and the remaining $31$ for testing.
@@
 \end{tabular}
-\label{tab:results}
+\label{tab:bcicompetitionresults}
 \end{table}
@@
 Using a straightforward dictionary of templates for \textit{MP-1} proved more beneficial in terms of performance than the standard approach of using a Hilbert base of Wavelets atoms on \textit{MP-2}.
-Either applying latency noise or amplitude noise, the method based on the signal's templates instead of using their coefficient proved to achieve much better performance values.
+Either applying latency noise or amplitude noise, the method based on the signal's templates instead of using their coefficients achieve much better performance values.
 
@@
 
+Regarding results against a public and real dataset of P300 ERP from the Berlin BCI Competition 2003, the obtained performance is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$.
+It is also similar to the performance obtained for the Experiment 3, which shows a more realistic scenario.
+
+However, it is bellow of those values obtained by the winners of that competition (perfect performance).
+Despite of this, it is important to remark that the algorithms presented here analyze the single-channel waveform structure of the signal which cannot be easily done in a multichannel approach.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The purpose of this work was threefold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain acceptable classification accuracies based exclusively on the signal's waveform.
+The purpose of this work was threefold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain acceptable classification performances based exclusively on the signal's waveform.
 
 We verified that similar performance results are obtained for the methods \textit{MP-1}, \textit{SIFT} and \textit{SHCC} and that it is possible to obtain discriminating information from the underlying signal based exclusively on an automated method of processing the waveforms.
-This open the possibility to use these techniques to implement automatic detection procedures with a strong intelligible property, due to the fact that they are based on metrics which can be visually verified.
+This brings the possibility to use these techniques to implement intelligible automatic detection procedures due to the fact that they are based on metrics which can be visually verified.
 
@@
 At the same time, BCI reliability is yet an unfulfilled goal in this discipline~\citep{WolpawJonathanR2012}.
-The benefits of analyzing or including metrics about the shape of the EEG, are that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can lead to more robust implementation of BCI devices.
+The convenience of analyzing or including metrics about the shape of the EEG, is that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can lead to more robust implementation of BCI devices.
 
 The conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.
-The need for more objective measurements pushed the adoption of more automated means of decoding the signals and demanded a need for replication \citep{Thakor2004}.
-This lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}.
-We believe that the adoption of \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an intelligible property that can be mapped to existing procedures, and above all, can mantain the clinician control of the procedure will be beneficial to clinical practice, Neuroscience and BCI research.
-
-These methods have the advantage that they can map a visual component with a clinical meaning to feature with a clear mathematical representation.
-Successful approaches in Computer Vision and Pattern Recognition in other areas use an ensemble of features to compound classifiers~\citep{Criminisi2013}, with a significant enhancement of the classification performances~\citep{Gu2012}.
-
-Another benefit of these approaches is that they have a potential universal applicability.
+At the same time, it is a subjective time-consuming task, with long-learning curves, requires specialized personnel, and it has significant error rates~\citep{Tjepkema-Cloostermans2018}.
+These problems has pushed for the adoption of more automated means of decoding the signals~\citep{Thakor2004}.
+This trend lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}~\citep{Wulsin2011,Tjepkema-Cloostermans2018}.
+
+We believe that the adoption of a \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an intelligible property~\citep{j2018challenge} that can be mapped to existing procedures, and above all, can maintain the clinician trust on the system behavior is beneficial to Clinical Practice, Neuroscience and BCI research.
+
+These methods have the advantage that they can map a visual component with a clinical meaning to a feature with an objective representation.
+Compound classifiers or ensemble of features can be explored.
+Successful approaches in Computer Vision or Pattern Recognition in other areas use them~\citep{Criminisi2013} with a significant enhancement of the classification performances~\citep{Gu2012}.
+
+% This is an important area for future study.
+
+Another benefit of these methodologies is that they have a potential universal applicability.
 As they are only analyzing waveform, they can be explored in other disciplines where the structure or shape of the waveform is of relevance.
@@
 EKG: Electrocardiogram\\
-PAA:Period Amplitude Analysis
+PAA:Period Amplitude Analysis\\
+SVM:Support Vector Machine
 }

### 2018-09-19 sensors.tex
Almost ready.

@@
 The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community which was based intensively on EEG waveforms and the structure of signal plots.
-Hence, the purpose of this work is to establish a bridge to fill this gap by doing a review and description of the procedures that have been used to detect patterns in the electroencephalographic waveforms, and to perform a benchmarking analysis of them.
-}
+Hence, the purpose of this work is to establish a bridge to fill this gap by reviewing and describing the procedures that have been used to detect patterns in the electroencephalographic waveforms, benchmarking them on a controlled pseudo-real dataset of an ERP waveform and verifying their performance on a public dataset of a BCI Competition.}
 
@@
 
-The aim of this study is twofold: first to review current literature of EEG processing techniques which are based on analysis of the waveform.
+The aim of this study is threefold: first to review current literature of EEG processing techniques which are based on analysis of the waveform.
 The second is to evaluate and study these methods by analyzing its classification performance against a pseudo-real dataset.
+And third, to verify their applicability to a real and public dataset.
 We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal waveform, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective~\citep{Nijboer2009,Wei2017}.
@@
 The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.
-The interictal epileptiform discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges\citep{EEGIntro}.
+The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges\citep{EEGIntro}.
 
@@
 
-All these methods provide a feature $ f $  that can be used as a template.
-The notation $\{\}_{1}^{n} $ will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f$.
+All these methods provide a feature $ f $~\footnote{The notation $\{\}_{1}^{n} $ will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f$}  that can be used as a template.
 These algorithms are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.
@@
 Two variants of this algorithm are evaluated.
-In \textit{MP 1} the dictionary is directly constructed with the normalized templates directly extracted from the real signal segments which is a straightforward implementation of the pattern matching technique.
+In \textit{MP 1} the dictionary is constructed with the normalized templates directly extracted from the real signal segments which is a straightforward implementation of the pattern matching technique.
 In \textit{MP 2} the coefficients of Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary~\citep{Vareka2012}.
@@
 This numeric value can be calculated as information entropy~\citep{Nicolaou2010}.
-Let's consider a signal on a window of length $W$ represented by
+Let's consider a signal on a window of length $W$ represented by the sample points
 
@@
 
-and resampled by $\tau$ intervals and starting from the sampling point $n$, doing
-
-\begin{equation}
-(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(m-1)\tau})
+and resampled by $\tau$ intervals, starting from the sampling point $n$, doing
+
+\begin{equation}
+(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(m-1)\tau}).
 \label{eq:pe2}
@@
 Therefore, for each $i$ only those ordinal elements $\pi_{s}$ that were effectively found along the signal are counted to estimate $p_i$, and zero elsewhere.
-The empirical permutation entropy can be calculated from this histogram as,
+The empirical permutation entropy can be calculated from the histogram as,
 
@@
 This procedure produces a scalar number for the given signal window of size $W$.
-To derive a feature, a sliding window procedure must be implemented.
+To derive a feature, a sliding window procedure must be implemented to cover an entire segment of length $N$.
 Thus, the length of the feature is $N - (W + \tau (m - 1)) + 1$.
@@
 
-This method is an extension of Slope Chain Code (SCC)~\citep{Alvarado-Gonzalez2016}.
-This algorithm proceeds by generating a coding scheme from a sequence of sample points.
+This algorithm~\citep{Alvarado-Gonzalez2016} proceeds by generating a coding scheme from a sequence of sample points.
 This encoding is based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.
@@
 \begin{equation}
-m = sign \bigg (  \frac{N-1}{\Delta} \bigg )  \bigg \lfloor \left\lvert \frac{N-1}{\Delta} \right\lvert \bigg \rfloor
+m = sign \bigg (  \frac{N-1}{\Delta} \bigg )  \bigg \lfloor \left\lvert \frac{N-1}{\Delta} \right\lvert \bigg \rfloor.
 \label{eq:shcc4}
@@
 \begin{equation}
-e = \left[ (x',y')_{1}, ...,(x',y')_{s}, ..., (x',y')_{G} \right]
+e = \left[ (x',y')_{1}, ...,(x',y')_{s}, ..., (x',y')_{G} \right].
 \label{eq:shcc5}
@@
 \begin{equation}
-z = \left\lfloor \frac{h}{2} \right\rfloor - \left\lfloor \frac{\max_{n} \tilde{x}(n) + \min_{n} \tilde{x}(n)}{2} \right\rfloor
+z = \left\lfloor \frac{h}{2} \right\rfloor - \left\lfloor \frac{\max_{n} \tilde{x}(n) + \min_{n} \tilde{x}(n)}{2} \right\rfloor.
 \label{eq:sift5}
@@
 
-Farwell and Donchin P300 Speller \citep{Farwell1988} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing in a virtual keyboard.
+Farwell and Donchin P300 Speller \citep{Farwell1988} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
 This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
@@
 Everything is there, except the P300 ERP component.
+Along the EEG stream, the markers information was used to localize the \textit{True} segments where the P300 should have been found, and those timing locations are used to superimpose the extracted ERP waveform.
 By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to similar procedures used in other works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
@@
 \begin{itemize}
-\item Experiment 1 - Pseudo-Real Dataset Classification Performance: the letter classification performance of each one of these methods on the artificially generated pseudo-real dataset.
+\item Experiment 1 - Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.
 \item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERP over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
@@
 
-The classification method Support Vector Machine SVM~\citep{Kaper2004} is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
+The classification method Support Vector Machine SVM~\citep{Kaper2004} with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
 This method has been proved efficient in decoding P300 in several BCI Competitions.
@@
 
-The template ERP is extracted from the Subject Number $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+The template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
 Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.
@@
 
-Along the EEG stream, the markers information was used to localize the \textit{True} segments where the P300 should have been found, and those timing locations are used to superimpose the extracted ERP waveform.
-
 Figure~\ref{fig:gains} shows a $5s$ sample of the EEG trace obtained with the MNE library~\citep{Gramfort2013}.
@@
 At first-sight, differences are really hard to spot visually.
-Subfigures (c) and (d) show only one second of the same segment only channel Cz and Label.
+Subfigures (c) and (d) show only one second of channels Cz and L from the same segment.
 The superimposed ERP can be devised enclosed by the vertical bars, around $31.5$s, where in (d) the peak is slightly bigger.
-Figure~\ref{fig:gaincheck} shows the obtained ensemble average ERPs as result of superimposing the template signal into EEG stream, time-locked to the stimulus onset.
+Figure~\ref{fig:gaincheck} shows the obtained ensemble average ERPs as result of superimposing the template signal into the EEG stream, time-locked to the stimulus onset.
 These 12 point-to-point averaged segments correspond to the first trial of the EEG stream.
@@
 \includegraphics[width=12cm]{images/erptemplate1.eps}
-\caption{ERP Template obtained from the coherent point-to-point ensemble average of subject Eight obtained from the BNCI Horizon public dataset 008-2014.
+\caption{ERP Template obtained from the coherent point-to-point ensemble average from the signals of Subject Number Eight of the BNCI Horizon public dataset 008-2014.
 The template is $1s$ long which is 256 sample points, and the eight channels are superimposed with different colors.
@@
 Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.
-The first $15$ trials are used to build the dictionary of templates, extracting the averaged EEG segments for the row and column that already contain the P300 ERP, hence shielding $30$ different templates per channel.
+Fifteen trials are used to build the dictionary of templates, extracting the averaged EEG segments for the row and column that already contain the P300 ERP, hence shielding $30$ different templates per channel.
 Figure~\ref{fig:dictionaryfig} shows the set of templates while using the first $15$ trials of the dataset.
@@
 The parameter $k$ represents the number of neighbors chosen from the dictionary of templates.
-The column can be obtained by the same way,
+The column can be obtained in the same way,
 
@@
 Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.
-Table \ref{tab:results} shows the  performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance was attained.
+Table \ref{tab:results} shows the  performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance is attained.
 Figure \ref{fig:performancetest} shows the performance curves for six algorithms.
-Each one represents the percentage of letters that were actually predicted by the algorithms using a cross-validation procedure.
-As previously described the data were continuously divided in two folds, where the first 15 letters were used to derive the dictionary of templates while the remaining 20 letters were used to measure the classification performance.
+Each one represents the percentage of letters that is actually predicted by the algorithms using a cross-validation procedure.
+As previously described the data is continuously divided in two folds, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance.
 This is repeated one hundred times, and performances averaged.
 Figure \ref{fig:performancetestlatency} shows the same results for the Experiment 2, where a noisy latency lag was included.
-Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template are randomly attenuated.
+Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template is randomly attenuated.
 Furthermore, results obtained for the dataset BCI Competition 2003 IIb are shown in Figures \ref{fig:performancebcicompetition} and in Table~\ref{tab:bcicompetitionresults}.
@@
 \caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
-Additionally, the control algorithms SVM Support Vector Machines was included, with a linear kernel.
+Additionally, the control algorithms SVM Support Vector Machines is included for comparison.
 All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
@@
 The achieved performance is significantly reduced for all methods.
-Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
+Y-axis shows letter identification performance while X-axis shows the number of intensification sequences used to calculate the ensemble average.}
 \label{fig:performancetestlatency}
@@
 \includegraphics[width=15cm]{images/CrossPerformanceTestAmplitude.eps}
-\caption{Speller performance obtained while the amplitudes of the N1 and P3 component of the superimposed ERP are randomly reduced.
-Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
+\caption{Speller performance obtained while the amplitudes of the P3b component of the superimposed ERP is randomly reduced.
+Y-axis shows performance accuracy while X-axis shows the number of intensification sequences.}
 \label{fig:performancetestamplitude}
@@
 \caption{Speller performance obtained for the dataset IIb of the BCI Competition II (2003) for each one of the algorithms.
-An offline BCI Simulation is performed using the first $42$ trials as training and the remaining $31$ as testing.}
+An offline BCI Simulation is performed using the first $42$ trials as training and the remaining $31$ as testing.
+The horizontal axis show the number of intensification sequences, from $0$ to $15$ for this dataset, while the vertical axis show the performance rate.}
 \label{fig:performancebcicompetition}
@@
 The first $42$ trials are used for training to build the template dictionary and the remaining $31$ for testing.
+The channel where the best performance is attained, is also shown.
 }
@@
 
-All the methods show a significant reduction of performance when latency noise is added, and at the same time all the methods show some resistance to noise in peak amplitudes of the main components.
+All the methods show a significant reduction of performance when latency noise is added, and at the same time all the methods show some resistance to noise in peak amplitudes of the main component.
 
@@
 
-Results obtained for algorithms MP-1, SIFT, SHCC show them to be comparable with the results obtained for SVM.
-Particularly, the methods \textit{SIFT} and \textit{SHCC} performed much better than the other algorithms even when the latency and the amplitudes are noisy.
-Latency noise on the other hand, reduces the information contained in the signal, mainly due to the invalidation of the SNR enhancement performed by the averaging procedure.
+The methods \textit{MP-1}, \textit{SIFT}, \textit{SHCC} performed much better than the other algorithms, including \textit{SVM}, even when the latency and the amplitudes are altered.
+Latency noise on the other hand, reduces the information contained in the signal, mainly due to the invalidation of the SNR enhancement performed by the signal averaging procedure.
 This reduction impacts on the obtained performance for all the methods, almost equally.
 
-Regarding results against a public and real dataset of P300 ERP from the Berlin BCI Competition 2003, the obtained performance is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$.
-It is also similar to the performance obtained for the Experiment 3, which shows a more realistic scenario.
-
-However, it is bellow of those values obtained by the winners of that competition (perfect performance).
-Despite of this, it is important to remark that the algorithms presented here analyze the single-channel waveform structure of the signal which cannot be easily done in a multichannel approach.
+Regarding results produced for the public and real dataset of P300 ERP from the Berlin BCI Competition II (2003), the obtained performance is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$.
+It is also similar to the performance obtained for the Experiment 3, which represents coincidentally a more realistic scenario.
+
+However, it is bellow of values obtained by the winners of that competition (perfect performance).
+Despite of this, it is important to remark that the algorithms presented here analyze the single-channel waveform structure of the signal which cannot be easily done in a multichannel approach~\citep{Gribonval2008}.
 
@@
 
-The purpose of this work was threefold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain acceptable classification performances based exclusively on the signal's waveform.
+The purpose of this work is threefold, (1) raise awareness about the utility of using automatic waveform-based methods to study EEG signals, (2) to provide an overview of the state-of-the-art of those methods, and (3) to compare those methods and verify if it is possible to obtain acceptable classification performances based exclusively on the signal's waveform.
 
@@
 At the same time, BCI reliability is yet an unfulfilled goal in this discipline~\citep{WolpawJonathanR2012}.
-The convenience of analyzing or including metrics about the shape of the EEG, is that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can lead to more robust implementation of BCI devices.
+The convenience of analyzing or including metrics about the shape of the EEG, is that clinical EEG diagnosis may support a vast set of already understood knowledge which is based on identifying EEG patterns by their shape and that can steer towards a more robust implementation of BCI devices.
 
@@
 These problems has pushed for the adoption of more automated means of decoding the signals~\citep{Thakor2004}.
-This trend lead to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}~\citep{Wulsin2011,Tjepkema-Cloostermans2018}.
-
-We believe that the adoption of a \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an intelligible property~\citep{j2018challenge} that can be mapped to existing procedures, and above all, can maintain the clinician trust on the system behavior is beneficial to Clinical Practice, Neuroscience and BCI research.
+This trend pointed to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}~\citep{Wulsin2011,Tjepkema-Cloostermans2018}.
+
+We believe that the adoption of a \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an inherent intelligible property~\citep{j2018challenge} that can be mapped to existing procedures, and above all, can maintain the clinician trust on the system behavior is beneficial to Clinical Practice, Neuroscience and BCI research.
 

### 2018-09-24 sensors.tex
Added some modifications and small fixing on notation.  The details of the BCI experiment were enhanced as well.

@@
 
-All these methods provide a feature $ f $~\footnote{The notation $\{\}_{1}^{n} $ will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f$}  that can be used as a template.
+All these methods provide a feature $ f $~\footnote{The notation $f=\{f_i\}_{1}^{n} $ or $f=\{f_i\}_{i \in J}^{} $  will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f=\{f_1,f_2,...,f_n\}$.}  that can be used as a template.
 These algorithms are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.
@@
 \textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Vincent2010} techniques that assume the EEG signal as a linear combination of different sparse sources extracted from a template's dictionaries.
-Matching Pursuit \textit{MP}~\citep{Mallat1993}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.
+Matching Pursuit \textit{MP}~\citep{Mallat1993}, the most representative of these algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.
 Given a signal, this optimization technique, tries to find the indexes of $m$ atoms and their weights (contributions) that minimize,
@@
 
-\noindent which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.
+\noindent which is the error between the signal and its approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.
 The algorithm goes by first setting the approximating signal $\tilde{x}_{0}$  as the original signal itself,
@@
 
-\noindent with $k$ representing the index of the selected atom $g_{k}$ and $\left\lVert \cdot \right\rVert$ its norm.
+\noindent with $k$ representing the index of the selected atom $g_{k}$ and $\left\lVert \cdot \right\rVert$ its euclidean norm.
 Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
@@
 
-and resampled by $\tau$ intervals, starting from the sampling point $n$, doing
+\noindent and resampled by $\tau$ intervals, starting from the sampling point $n$, doing
 
@@
 \noindent with $k = W-(m-1) \tau$.
-The probability density function \textit{pdf} for all the available permutations of order $m$ should be $ \textbf{p} = (p_1,p_2,...,p_{m!}) $ with $ \sum_{1}^{m!} p_{i} = 1 $.
+The probability density function \textit{pdf} for all the available permutations of order $m$ should be $ \textbf{p} = (p_1,p_2,...,p_{m!}) $ with $ \sum_{i=1}^{m!} p_{i} = 1 $.
 
@@
 To derive a feature, a sliding window procedure must be implemented to cover an entire segment of length $N$.
-Thus, the length of the feature is $N - (W + \tau (m - 1)) + 1$.
-
-\begin{equation}
-f =  {\bigg \{ H(\textbf{p})_{u} \bigg \}}_{1}^{N - (W + \tau (m - 1)) + 1}
+Thus, the length of the feature is $N - (W + \tau (m - 1))$.
+
+\begin{equation}
+f =  {\bigg \{ H(\textbf{p})_{u} \bigg \}}_{W + \tau  m}^{N}.
 \label{eq:pe6}
@@
 
-\noindent with $u$ varying on a sample by sample basis along the signal but excluding the first window.
+\noindent with $u$ varying on a sample by sample basis along the signal, starting from the specified index.
 
@@
 \begin{equation}
-e = \left[ (x',y')_{1}, ...,(x',y')_{s}, ..., (x',y')_{G} \right].
+e' = \left[ (x',y')_{1}, ...,(x',y')_{s}, ..., (x',y')_{G} \right].
 \label{eq:shcc5}
@@
 \begin{equation}
-\hat{\textbf{x}} = \frac{( \textbf{x'} - \min(\textbf{x'}) \textbf{1} )}{\max(\textbf{x'}) - \min(\textbf{x'})} 
+\hat{\textbf{x}} = \frac{\textbf{x'} - \min(\textbf{x'}) \textbf{1} }{\max(\textbf{x'}) - \min(\textbf{x'})} 
 \label{eq:shcc6}
@@
 \begin{equation}
-\hat{\textbf{y}} = \frac{( \textbf{y'} - \min(\textbf{y'}) \textbf{1} )}{\max(\textbf{y'}) - \min(\textbf{y'})} 
+\hat{\textbf{y}} = \frac{ \textbf{y'} - \min(\textbf{y'}) \textbf{1}}{\max(\textbf{y'}) - \min(\textbf{y'})} 
 \label{eq:shcc7}
@@
 
-\noindent with $\textbf{1}$ being the vector of length $G$ with all their components equal to $1$.  This operation effectively normalize the components $\hat{x}_{s},\hat{y}_{s} \in [0,1]$.\\
+\noindent with $\textbf{1}$ being the vector of length $G$ with all their components equal to $1$.
+Hence, the scalar components $\hat{x}_{s}$ of $\hat{\textbf{x}}$, and $\hat{y}_{s}$ of $\hat{\textbf{y}}$, with $s$ varying between $1$ and $G$ are effectively normalized to $\hat{x}_{s},\hat{y}_{s} \in [0,1]$.
+\\
 
@@
 \begin{equation}
-h = ( \max_{n} \tilde{x}(n) - \min_{n} \tilde{x}(n)).
+h = \max_{n} \tilde{x}(n) - \min_{n} \tilde{x}(n).
 \label{eq:sift4}
@@
 
-\noindent where $z_1$ and $z_2$ are the image coordinates values.  These points are interpolated using the Bresenham algorithm~\citep{Ramele2016}.
+\noindent where $z_1$ and $z_2$ are the image coordinates values, $255$ represents white and $0$ is the background black color of the plot.
+These points are interpolated using the Bresenham algorithm~\citep{Ramele2016}.
 
@@
 \begin{equation}
-f = {\bigg \{ h(i,j,\theta) \bigg \}}_{(0,0,0)}^{(3,3,315)}
+f = {\bigg \{ \bigg \{  \bigg \{ h(i,j,\theta) \bigg \}_{i \in I }  \bigg \}_{j \in I} \bigg \}_{\theta \in \Theta} }
 \label{eq:sift7}
@@
 
-\noindent with $i$ and $j$ varying between $0$ and $3$, localizing the 16 blocks within the grid.
-The angles $\theta$ are the eight possible equidistant values between $0$ and $315$.
+\noindent with $i$ and $j$ varying belonging to $I = \left[0,3 \right] $ localizing the 16 blocks within the grid.
+The angles $\theta$ that belong to $\Theta$ are the eight possible equidistant values between $0$ and $315$.
 This vector is normalized, clamped to $0.2$, and re-normalized again.
@@
 \item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERP over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
-The lagging value is picked from a uniform distribution ranging from $0$ to $40\%$ of the $1s$ segment size~\citep{DaPelo2018}.
+The lagging value is picked from a uniform distribution $U(0,0.4)$ [s] ranging from $0$ to $0.4$ of the $1s$ segment size~\citep{DaPelo2018}.
 \item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.
-This component is subtracted from the original template.
-The subtraction percentage is drawn from a uniform distribution.
-Additionally the subtracted curve is multiplied by a Gaussian window of $800$ ms.
+This component is defined to be located from the sample point $37$ up to $249$ which is around $840$ ms.
+This component multiplied by a gain factor is subtracted from the original template.
+This gain factor between $0$ and $1$ is drawn from a uniform distribution $U(0,1)$.
+Additionally the subtracted curve is multiplied by a Gaussian window with a support of the same length~\citep{Harris1978}.
 This avoids adding any discontinuity into the artificial generated signal.
@@
 
-The classification method Support Vector Machine SVM~\citep{Kaper2004} with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
-This method has been proved efficient in decoding P300 in several BCI Competitions.
-
-All these experiments are executed using two fold cross validation procedure that preserves the structure of the letter identification trials.
+The classification method Support Vector Machine SVM with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
+This method has been proved efficient in decoding P300 in several BCI Competitions~\citep{Kaper2004}.
+
+All these experiments are executed using cross validation procedure dividing the letter to spell in two sets, preserving the structure of the letter identification trials.
 Spelling letters are scrambled while the order and group of each intensification sequence is preserved.
@@
 Finally the performance at letter identifications for these same methods is evaluated by performing an offline BCI Simulation on the Dataset IIb of the BCI Competition II (2003)~\citep{Blankertz2002}.
+The protocol of this dataset is very similar to what was used to obtain the pseudo-real dataset.
+The sampling frequency of this dataset is $240$, the number of letters are $73$ where the first $42$ are used to create the template dictionary for all the methods and the remaining $31$ are used to test the character recognition rate performance.
+Additionally, in this dataset the number of available intensification number sequences is $15$.
 
@@
 Channel $S$ represents the twelve different stimulus markers (columns or rows) while channel $L$ represent the label (\textit{True} vs \textit{False}).
-\textit{False} segments are marked as a square signal and \textit{True} segments are identified by the double-amplitude square signals.
+Labels are represented by square signals.
+\textit{False} segments are marked with single amplitude square signals while \textit{True} segments are identified by double-amplitude square signals.
 Subfigure (a) shows the signals before the ERP template is superimposed while subfigure (b) shows the same signals with the superimposed ERP template.
@@
 \caption{Point-to-point averaged signals for the first letter identification trial.
-The ERP was superimposed on classes 3 and 9.
-Class 3 was obtained while averaging the segments where the row of the speller matrix was intensified whereas class 9 was calculated from the intensification of the corresponding column.}
+The ERP is superimposed on classes 3 and 9.
+Class 3 is obtained while averaging the segments where the row of the speller matrix is intensified whereas class 9 is calculated from the intensification of the corresponding column.}
 \label{fig:gaincheck}
@@
 Described algorithms produce a feature $f$ for each EEG segment.
-The aim of the classification procedure is to identify for the remaining $20$ trials which of the 6 features $f$ that were obtained for row intensification and which of the 6 features for column intensifications are the ones that elicited the P300 response on the averaged EEG segment.
+The aim of the classification procedure is to identify for the remaining $20$ trials which of the 6 features $f$ that were obtained for row intensification, labeled by $\{ 1,...,6 \}$, and which of the 6 features for column intensification, named $\{ 7,...,12 \}$ are the ones that elicited the P300 response on the averaged EEG segment.
 The row number of the matrix can be obtained by doing
@@
 \begin{equation}
-\hat{row} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q_i=1}^{k} \left\lVert f_u - q_i \right\rVert ^2
+\hat{row} = \arg \min_{u \in \{1,\dots,6\}} \sum_{i=1}^{k} \left\lVert f_u - q_i \right\rVert ^2
 \label{eq:multiclassificationrow}
@@
 \begin{equation}
-\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q_i=1}^{k} \left\lVert f_u - q_i \right\rVert ^2.
+\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{i=1}^{k} \left\lVert f_u - q_i \right\rVert ^2.
 \label{eq:multiclassificationcol}
@@
 Each one represents the percentage of letters that is actually predicted by the algorithms using a cross-validation procedure.
-As previously described the data is continuously divided in two folds, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance.
+As previously described the data is continuously divided in two sets, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance.
 This is repeated one hundred times, and performances averaged.
@@
 \caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
-Additionally, the control algorithms SVM Support Vector Machines is included for comparison.
+Additionally, the control algorithm SVM Support Vector Machines is included for comparison.
 All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
@@
 \toprule
-\textbf{Method}	& \textbf{Channel} &  \textbf{Performance} & \textbf{Perf w/Latency noise}	& \textbf{Perf w/Amplitude Noise}\\
+\textbf{Method}	& \textbf{Channel} &   \multicolumn{3}{c}{Performance} \\
+%\cline{3-5} \\
+ 	&  &  \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}\\
 \midrule
@@
 It is also similar to the performance obtained for the Experiment 3, which represents coincidentally a more realistic scenario.
-
-However, it is bellow of values obtained by the winners of that competition (perfect performance).
-Despite of this, it is important to remark that the algorithms presented here analyze the single-channel waveform structure of the signal which cannot be easily done in a multichannel approach~\citep{Gribonval2008}.
+It is important to remark that the algorithms presented here analyze the single-channel waveform structure of the signal~\citep{Gribonval2008}.
 

### 2018-09-24 sensors.tex
The abstract was modified

@@
 \abstract{The Electroencephalography (EEG) is not just a mere clinical tool anymore.
-It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time and translating or decoding brain signals that can be used to diagnose disease or implement Human Computer Interaction devices.
-The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community which was based intensively on EEG waveforms and the structure of signal plots.
-Hence, the purpose of this work is to establish a bridge to fill this gap by reviewing and describing the procedures that have been used to detect patterns in the electroencephalographic waveforms, benchmarking them on a controlled pseudo-real dataset of an ERP waveform and verifying their performance on a public dataset of a BCI Competition.}
+It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time.
+It is now being used to translate or decode brain signals to diagnose disease or implement Human Computer Interaction devices.
+The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community.
+Clinical EEG is based intensively on waveforms and the structure of signal plots.
+Hence, the purpose of this work is to establish a bridge to fill this gap by reviewing and describing the procedures that have been used to detect patterns in the electroencephalographic waveforms, benchmarking them on a controlled pseudo-real dataset of an EEG Event Related Potentials waveform and verifying their performance on a public dataset of a BCI Competition.}
 

### 2018-09-25 sensors.tex
Added statistical tests.

@@
 
-\noindent with $i$ and $j$ varying belonging to $I = \left[0,3 \right] $ localizing the 16 blocks within the grid.
+\noindent with $i$ and $j$ varying belonging to $I = \{ 0,1,2,3 \} $ localizing the 16 blocks within the grid.
 The angles $\theta$ that belong to $\Theta$ are the eight possible equidistant values between $0$ and $315$.
@@
 \item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.
-This component is defined to be located from the sample point $37$ up to $249$ which is around $840$ ms.
-This component multiplied by a gain factor is subtracted from the original template.
+This component is defined to be located from the stimulus onset between $148$ ms up to $996$ ms which is around $840$ ms long.
+This waveform element, multiplied by a gain factor, is subtracted from the original template.
 This gain factor between $0$ and $1$ is drawn from a uniform distribution $U(0,1)$.
-Additionally the subtracted curve is multiplied by a Gaussian window with a support of the same length~\citep{Harris1978}.
+Additionally this subtracted waveform is multiplied by a Gaussian window with a support of the same length~\citep{Harris1978}.
 This avoids adding any discontinuity into the artificial generated signal.
@@
 Furthermore, results obtained for the dataset BCI Competition 2003 IIb are shown in Figures \ref{fig:performancebcicompetition} and in Table~\ref{tab:bcicompetitionresults}.
+For this experiment the number of available intensification sequences is 15.
 
@@
 \includegraphics[width=15cm]{images/CrossPerformanceTest.eps}
-\caption{Speller performance obtained for each method.
+\caption{Speller performance obtained for each method for the Experiment 1.
 Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
@@
 \includegraphics[width=15cm]{images/CrossPerformanceTestLatency.eps}
-\caption{Speller performance obtained for each method while latencies were artificially added to each single-intensification segment.
+\caption{Speller performance obtained for each method while latencies are artificially added to each single-intensification segment corresponding to the Experiment 2.
 The achieved performance is significantly reduced for all methods.
@@
 \includegraphics[width=15cm]{images/CrossPerformanceTestAmplitude.eps}
-\caption{Speller performance obtained while the amplitudes of the P3b component of the superimposed ERP is randomly reduced.
+\caption{Speller performance obtained for the Experiment 3 while the amplitudes of the P3b component of the superimposed ERP is randomly reduced.
 Y-axis shows performance accuracy while X-axis shows the number of intensification sequences.}
@@
 \includegraphics[width=15cm]{images/PerformanceBCICompetition.eps}
-\caption{Speller performance obtained for the dataset IIb of the BCI Competition II (2003) for each one of the algorithms.
+\caption{Speller performance obtained for the Dataset IIb of the BCI Competition II (2003) for each one of the algorithms.
 An offline BCI Simulation is performed using the first $42$ trials as training and the remaining $31$ as testing.
@@
 
-All the methods show a significant reduction of performance when latency noise is added, and at the same time all the methods show some resistance to noise in peak amplitudes of the main component.
-
-Using a straightforward dictionary of templates for \textit{MP-1} proved more beneficial in terms of performance than the standard approach of using a Hilbert base of Wavelets atoms on \textit{MP-2}.
-Either applying latency noise or amplitude noise, the method based on the signal's templates instead of using their coefficients achieve much better performance values.
-
-The methods \textit{MP-1}, \textit{SIFT}, \textit{SHCC} performed much better than the other algorithms, including \textit{SVM}, even when the latency and the amplitudes are altered.
-Latency noise on the other hand, reduces the information contained in the signal, mainly due to the invalidation of the SNR enhancement performed by the signal averaging procedure.
-This reduction impacts on the obtained performance for all the methods, almost equally.
-
-Regarding results produced for the public and real dataset of P300 ERP from the Berlin BCI Competition II (2003), the obtained performance is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$.
-It is also similar to the performance obtained for the Experiment 3, which represents coincidentally a more realistic scenario.
-It is important to remark that the algorithms presented here analyze the single-channel waveform structure of the signal~\citep{Gribonval2008}.
+A significant reduction of performance was found when latency noise is added (Experiment 1 vs 2, Wilcoxon rank sum test, one tail, rejected towards greater with $p=0.0022$).
+The latency noise reduces the information contained in the averaged signal, mainly due to the invalidation of the SNR enhancement performed by the signal averaging procedure.
+This reduction alters the obtained shape of the waveform of the ERP and impacts on the performances regardless of the method.
+On the other hand, all the algorithms show some resistance to noise in peak amplitudes of the main component (Experiment 1 vs 3,Wilcoxon rank sum test, one tail, accepted with $p=0.05$).
+
+A two-way balanced Quade test was performed to evaluate the performance's differences for the three experiments and for the six methods.
+Difference among the method's performances were found with $p=0.0019$.
+The methods \textit{SHCC}, \textit{SIFT}, \textit{MP-1} performed much better than the other algorithms, including \textit{SVM}.
+With multiple comparisons to a level of $0.05$ significant differences are found between \textit{SHCC} vs MP-1,MP-2, PE and SVM, between \textit{SIFT} vs.
+MP-2 and PE, and between \textit{SVM} and MP-2.
+
+Additionally, using a straightforward dictionary of templates for \textit{MP-1} proved more beneficial in terms of performance than the approach of using a Hilbert base of Wavelets atoms on \textit{MP-2} (significant differences between MP-1 vs MP-2).
+Either applying latency noise or amplitude noise, the method based on the signal's templates instead of using their coefficients achieve much better character identification rates.
+
+Regarding results produced for the public and real dataset of P300 ERP from the Berlin BCI Competition II (2003), the obtained performance is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.
+It is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.
+It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Gribonval2008}.
 
@@
 
-We verified that similar performance results are obtained for the methods \textit{MP-1}, \textit{SIFT} and \textit{SHCC} and that it is possible to obtain discriminating information from the underlying signal based exclusively on an automated method of processing the waveforms.
+We verified that similar performance results are obtained for the methods \textit{SHCC}, \textit{SIFT} and \textit{MP-1} and that it is possible to obtain discriminating information from the underlying signal based exclusively on an automated method of processing the waveforms.
 This brings the possibility to use these techniques to implement intelligible automatic detection procedures due to the fact that they are based on metrics which can be visually verified.
+
+Further work should be conducted in terms of a multichannel meaningful extension of these waveform-based methods~\citep{Gribonval2008}.
+Moreover, the possibilities of finding overcomplete dictionaries for matching pursuit sparse representation based on obtained signal templates, could also be considered an area of future improvement.
 

### 2018-09-25 statistics/performance2.m

@@ -1,0 +1,70 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+% C?digo Matlab
+
+% valores tabla 2
+T2= 0.01*[67  15  50
+24  6  10
+91  18  66
+61  9  32
+98  31  80
+78  7  53];
+
+% compara si las performances de los m?todos son distintas para los 3
+% experimentos
+% tratamiento: m?todo - bloque: experimento
+quadetest(T2')
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% QUADE TEST FOR IDENTICAL TREATMENT EFFECTS:
+% TWO-WAY BALANCED, COMPLETE BLOCK DESIGNS
+% --------------------------------------------------------------------------------
+% Number of observation: 18
+% Number of blocks: 3
+% Number of treatments: 6
+% --------------------------------------------------------------------------------
+% F-statistic approximation
+% Quade test statistic W: 8.8889
+% F=W df-num=5 df-denom=10 - p-value (2 tailed): 0.0019
+% --------------------------------------------------------------------------------
+%  
+% POST-HOC MULTIPLE COMPARISONS
+% --------------------------------------------------------------------------------
+% Critical value: 11.5777
+% Absolute difference among mean ranks
+%      0     0     0     0     0     0
+%     13     0     0     0     0     0
+%     11    24     0     0     0     0
+%      6     7    17     0     0     0
+%     17    30     6    23     0     0
+%      3    16     8     9    14     0
+% Absolute difference > Critical Value
+%    0   0   0   0   0   0
+%    1   0   0   0   0   0
+%    0   1   0   0   0   0
+%    0   0   1   0   0   0
+%    1   1   0   1   0   0
+%    0   1   0   0   1   0
+% 
+% Conclusiones:
+% 
+% pvalue: 0.0019 => (chiquito) => se rechaza H0 => hay diferencias entre las performances
+% 
+% ?Para cu?les?
[…]

### 2018-09-25 statistics/performances.m

@@ -1,0 +1,51 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+% C?digo Matlab
+
+% valores tabla 2
+T2= 0.01*[67  15  50
+24  6  10
+91  18  66
+61  9  32
+98  31  80
+78  7  53];
+
+% Wilcoxon rank sum test
+% para ver si las perf. de exp 1 son > que las del exp 2
+[p,h,stats] = ranksum(T2(:,1),T2(:,2),'tail','right','method','exact')
+% para ver si las perf. de exp 1 son > que las del exp 3
+[p,h,stats] = ranksum(T2(:,1),T2(:,3),'tail','right','method','exact')
+
+% 
+% Resultados: (Wilcoxon rank sum test a 1 cola H0: = vs Ha: >) 
+% 
+% perf. de exp 1 son > que las del exp 2
+% ---------------------------------------
+% p =
+%     0.0022
+% h =
+%   logical
+%    1
+% stats =
+%   struct with fields:
+%     ranksum: 56
+% 
+% => el resultado que devuelve de rechazo de H0 lo hace el matlab a nivel 0.05, pero el pvalues es chiquito
+% => se rechaza H0 => lo que postul?s est? OK.
+% 
+% 
+% perf. de exp 1 son > que las del exp 3
+% ---------------------------------------
+% >> [p,h,stats] = ranksum(T2(:,1),T2(:,3),'tail','right','method','exact')
+% p =
+%     0.0898
+% h =
+%   logical
+%    0
+% stats =
+%   struct with fields:
+%     ranksum: 48
+% 
+% El pvalue > 0.05 => no se rechazar?a H0 (ah?, ah?...) pero va acorde con lo que postul?s en el paper (que las performances no son mayores, ie, que son resistentes).
+% 
+% ***************************
+% Para los datos del BCI, dejemos nom?s lo que postul?s en el paper sin ning?n test porque es un sujeto solito (n=1) :(

### 2018-09-25 sensors.tex
Some minor adjustments

@@
 The same classification algorithm based on k-nearest neighbors is used for all the methods~\citep{Boiman2008}.
-The experiment is composed of $35$ trials to spell $7$ words of $5$ letters each.
+The experimental protocol used to generated the pseudo-real dataset used in the experiments 1 to 3 is composed of $35$ trials to spell $7$ words of $5$ letters each.
 Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.
@@
 
-A significant reduction of performance was found when latency noise is added (Experiment 1 vs 2, Wilcoxon rank sum test, one tail, rejected towards greater with $p=0.0022$).
+A significant reduction of performance was found when latency noise is added (Experiment 1 vs 2, Wilcoxon rank sum test, one tail, with $p=0.0022$).
 The latency noise reduces the information contained in the averaged signal, mainly due to the invalidation of the SNR enhancement performed by the signal averaging procedure.
 This reduction alters the obtained shape of the waveform of the ERP and impacts on the performances regardless of the method.
-On the other hand, all the algorithms show some resistance to noise in peak amplitudes of the main component (Experiment 1 vs 3,Wilcoxon rank sum test, one tail, accepted with $p=0.05$).
+On the other hand, all the algorithms show some resistance to noise in peak amplitudes of the main component.
+This is shown by the similarities of obtained results between the Experiment 1 and 2 (Wilcoxon rank sum test, two tails, not rejected with $p=0.1797$).
 
 A two-way balanced Quade test was performed to evaluate the performance's differences for the three experiments and for the six methods.
-Difference among the method's performances were found with $p=0.0019$.
+Difference among the method's letter identification rates were found with $p=0.0019$.
 The methods \textit{SHCC}, \textit{SIFT}, \textit{MP-1} performed much better than the other algorithms, including \textit{SVM}.
-With multiple comparisons to a level of $0.05$ significant differences are found between \textit{SHCC} vs MP-1,MP-2, PE and SVM, between \textit{SIFT} vs.
+By making multiple comparisons at a level  $0.05$, significant differences are found between \textit{SHCC} vs MP-1,MP-2, PE and SVM, between \textit{SIFT} vs.
 MP-2 and PE, and between \textit{SVM} and MP-2.
@@
 
-Regarding results produced for the public and real dataset of P300 ERP from the Berlin BCI Competition II (2003), the obtained performance is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.
+Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.
 It is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.

### 2018-09-25 sensors.tex
Fix typo

@@
 On the other hand, all the algorithms show some resistance to noise in peak amplitudes of the main component.
-This is shown by the similarities of obtained results between the Experiment 1 and 2 (Wilcoxon rank sum test, two tails, not rejected with $p=0.1797$).
+This is shown by the similarities of obtained results between the Experiment 1 and 3 (Wilcoxon rank sum test, two tails, not rejected with $p=0.1797$).
 

### 2018-10-22 BrainScience/response.letter.1.tex
Start with reviewers modifications

\documentclass[journal,onecolumn,12pt]{IEEEtran}

\usepackage{amsmath,amssymb,bm}
\usepackage{amsthm, amsfonts}	
\usepackage{bm,bbm}
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{fancybox}
\usepackage{url,booktabs}
\usepackage[round]{natbib}

\usepackage{xr}

\title{Reply to Reviewer's Comments on\\
''EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces''}
\author{}

\begin{document}

\maketitle
\pagenumbering{roman}
\setcounter{page}{1}

We are grateful to the reviewer for pointing out relevant issues in our manuscript.

In the following, we discuss how we dealt with each raised issue.

\vskip+1ex
\noindent \dotfill

\section*{\fbox{Reviewer \#1 Transcript:}}

The paper is an interesting research paper and presents an analysis of the P300 ERP. The authors provide a literature review of the current methods used to detect EEG patterns and conduct three experiments to identify the P300 ERP in EEG recordings. The paper is well-organized. The paper should incorporate the following comments before its acceptance for publication:

1. The major concern is that the experiments have been conducted with only one participant. The authors should collect more data in this study. In other studies cited in the paper by authors,  two and 16 participants were recruited.

2. It is strongly recommended that the authors should re-check some paragraphs of the manuscript, correct any grammatical and syntactic errors and improve the scientific language. In some cases, it was hard to follow the meaning. (e.g. Page 2, Paragraph 4, Line 51-54, Page 11, Paragraph 3, Line 343-345)

3. In the Introduction section the authors refer to BCI applications; however, no scientific studies are reported. Also, the literature section should be revised with recent publication. Below are some scientific studies that is suggested to be included for BCI applications:

a. Cruz, A., Pires, G., & Nunes, U. J. (2018). Double ErrP Detection for Automatic Error Correction in an ERP-Based BCI Speller. IEEE Transactions on Neural Systems and Rehabilitation Engineering, 26(1), 26-36.

b. Tzimourta, K.D., Tsoulos, I., Bilero, T., Tzallas, A., Tsipouras, M., Giannakeas, N. (2018). Direct Assessment of Alcohol Consumption in Mental State Using Brain Computer Interfaces and Grammatical Evolution. Inventions, 3(3), 51.

c. Kevric, J., & Subasi, A. (2016). The impact of Mspca signal de-noising in real-time wireless brain computer interface system. Southeast Europe Journal of Soft Computing, 4(1).

and for sleep studies:

d. Boostani R, Karimzadeh F, Nami M (2017) A comparative review on sleep stage classification methods in patients and healthy individuals. Computer methods and programs in biomedicine 140: 77-91.

e. Dimitriadis, S. I., Salis, C., & Linden, D. (2018). A novel, fast and efficient single-sensor automatic sleep-stage classification based on complementary cross-frequency coupling estimates. Clinical Neurophysiology, 129(4), 815-828.

4. In the literature section, Table I briefly presents the proposed methods followed by other authors; however, there is no text describing the table. The authors should include a paragraph describing each method and the recent findings concisely, in order to present better the literature review.

5. In Figure 2 please insert the y-axis title to clearly present ERP template.

6. The Results section is not well-presented. The authors are advised to include a paragraph before Table 2, Table 3 and Fig. 6-8 presenting the obtained results (best/worst channel performance and algorithm performance for each experiment) and then discuss the results in the Discussion section. Please revise excessive descriptions in the captions of Table 2, Table 3 and Fig. 6-8.

7. “the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of 70%” (Page 19, Paragraph 2, Line 465-466) What is the obtained character identification rate in other scientific studies? Is the obtained result (70%) of this study comparable to others?

8. In the Discussion section a table presenting the statistical analysis results would present better the results and would facilitate the reader.

\subsection*{\ovalbox{General Comments}}
I have read the manuscript "EEG characterization and classification based on Histogram of Gradients of Signal Plots" with great interest since its title promised a solution to an important problem. In spite of presenting an interesting methodology and some interesting issues related to EEG characterization, I consider the manuscript is not ready for publication. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your comments and observations. We have modified several points according to your recommendations, we hope the article is now better suited for publication. 
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2.1) the title is misleading since a first, and naive, read suggest being able to classify and characterize any EEG with their different components (i.e., P200, P300, P600, N100, N200, N400 to name a few). Clearly, the title does not match the content since the authors focus on the P300

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation. The title of the article was changed accordingly to your suggestion.  
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2.2) there is evidence of the benefits of working with these signals P300, but, is the proposed method extensible to other signals?, this is not clear from the manuscript.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you very much for your comment.  We published in~\citep{Ramele2016} the application of the same method to identify rhythmic EEG events like Visual Occipital Alpha Waves and to classify Motor Imagery.  We are also working on unpublished material where we are analyzing the same approach for the detection of K-Complexes and classification of  SSVEP patterns but the results are in progress.  We added this information on the article (Introduction and Conclusion, section 3).  We hope this new information enriches the article and clarifies this point.

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
\subsection*{\ovalbox{Third Paragraph}}

3.1) Another issue with the manuscript is its mathematical vagueness. For example, the variable T is used as a vector or two-tuple but its notation is different from the notation the authors use for vectors

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  We worked hard to improve all the mathematical description of the method. The issue with variable T was fixed and we improved on the font styling to differentiate vectors and scalars.  Section 1.1 was modified as well as section 1.2.3.  The mathematical terminology were corrected all along the article. We hope we have full addressed this point. }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3.2) Moreover, there is something wrong in equation (4)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We appreciate a lot your truthful comment.  Equation (4) is very important for us because it is the essence of the method.  %We based that equation on the description of how the histogram of oriented gradients is calculated published in Vedaldi and Szeliski, Computer Vision and Algorithms.  
We reviewed the equation, corrected their terms and we hope it is now more clear.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3.3) and something inconsistent between equations (7) and (8)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We have revised the manuscript and modified equations (7) and (8) (now (8) and (9)) which are located on Section 1.2.3 (Parameters).  We added more information on the text and specified subscript indexes to avoid the inconsistency.
We thank you for noticing this.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3.4) Also, the authors go back and forth the digital space and the continuous space and although there is an equivalence, the manuscript could benefit from a clearer explanation.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  We modified all equations to emphasize the point that they are calculated in digital space, particularly on equation (4). We also added that image gradients are also calculated using finite differences, and added information about the support of each equation to explain that they are actually calculated in digital space.
With this, we hope we have full addressed your point.
 }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

\subsection*{\ovalbox{Fourth Paragraph}}

4.1) Another issue that I find worrisome is the design of experiments. It seems that the number of subjects is rather low, it would be interesting to know why the authors decided to use such a subset.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your comment. The original dataset was published as the 008-2014 dataset at the BNCI-Horizon website, which is very important in BCI Research.  As far as we are aware of, experiments on this dataset of ALS patients has not been published by any independent group.
In order to address this issue, we included a new own dataset with the same experimental conditions that we generated ourselves, on healthy subjects.  We published this dataset on the Kaggle platform for replication.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

4.3) Also, the authors speculate that their results are not that good because they did not take into account the latency and magnitude of the signals, there is previous evidence in the literature that the P300 suffers from these, and other, problems and, thus, one wonders why the authors did not anticipate such problems.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your comments.  We added more detailed information that address this issue. First in Section 1.1.4  we included information about the signal averaging procedure.  In Section 1.2.3 we added the parameters that were used.  More importantly, in Section 2 we added information and references about the tests that we did to address latency and amplitude issues.  We included in Section 3 information about future and current works where we are also dealing with these problems.
We hope we full addressed your remarks.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

\subsection*{\ovalbox{Fifth Paragraph}}

5.1) it seems the authors assume that the readers have some previous deep knowledge of some topics such as that some channels should carry more or better information without a previous reference.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your suggestion. The Results section was modified to clarify the expectation about the EEG channels where a visual P300 oddball paradigm generates stronger signals.  We also added references.
We hope we  have addressed your point with this.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

5.2)  issue has to do with the way the manuscript is written

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you a lot for this important comment.  We re-read all the manuscript and asked  colleagues to proofread the article.  They gave us interesting suggestions that helped us to improve different sections.  We would be very pleased if you find it now easier to read it.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

5.3) sometimes is very difficult to follow the ideas the authors want to convey, an example can be found in the paragraph containing equation (5)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you very much for this information.  We reviewed completely that equation (it is now equation (6) and equation (7)). We included a step-by-step procedure that describes the proposed algorithm and we added a schematic chart in Figure 3.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
\bibliographystyle{mdpi}
\bibliography{article}

\end{document}

### 2018-10-22 BrainScience/response.letter.2.tex
Start with reviewers modifications

This manuscript presents findings that contrast the performance of a selection of BCI classification methods in a single subject database. All in all, I find several aspects of this paper quite interesting, particularly its attempt to assess the impact of latency and component amplitude variation on classification performance. However, overall, I fear the range/scope of the paper is too limited to make a significant contribution to the literature.

1. Abstract, while automated analysis of EEG data certainly has significant potential with regards to the future of clinical EEG research, it is too soon to unqualifiedly state that these methods as they are “outshine the research done by the clinical EEG community.” Some aspects of the introduction also veer towards hyperbole on occasion and may be better received if they were more conservatively stated.

2. Several notable papers in the domain of P300-BCI work should be referenced (e.g., Rakotomamonjy & Guigue, 2008) as well as other prior efforts to evaluate P300-BCI classification methods (e.g., Manyakov, Chumerin, Combaz & Hulle, 2011; Oweis, Hamdi, Ghazali & Lwissy, 2013).

4. Page 12, Lines 383-394. It should be noted that the dataset from which the template ERP subject was drawn from was an ALS patient dataset. It is currently unclear why Subject Number 8 and trial number 2 was selected out of all the subjects and trials in the BNCI-Horizon database.

5. Integrating the template ERP signal into a Null-EEG stream and assuming everything is there except the P300 ERP component is probably an untenable assumption. As indicated in the manuscript itself, for instance, the null-EEG stream is a passive task. While the non-P300 trials in an active task contain additional activity associated with processes including target stimuli anticipation, active task focus, cognitive fatigue, etc. This is not true of the Null-EEG stream. Furthermore, combining and subsequently classifying EEG data from two different subjects is questionable at best given the physiological differences between the two. Differences in recording condition, equipment and task stimuli should also be considered. Collectively, these factors make it difficult to assess the value of the classification results observed on this study.

6. The separate manipulation of latency, as well as amplitude noise, is interesting and if extended may in and of itself prove to be an interesting manuscript in and of itself (e.g. contrasting the impact of a range of different latency variations and amplitude noise on BCI classification ability), true EEG data almost invariantly involves not only a mix of both noise sources but other sources such as cognitive fatigue, systematic task stimuli adaptation etc. Thus all three experiments represent assessments of isolated aspects of noise which would be interesting if expanded on in a more expansive simulation set (see point 4 and summary) but currently are too limited in scope to draw strong conclusions regarding the expected inter-component performance in real-world data.

It seems as though the entire study is built on the classification of a single active P300 BCI-Speller template from an ALS participant into a single null-EEG stream of a healthy subject performing a passive-viewing version of a BCI-Speller task. The generalizability of these results are thus difficult, if not impossible to assess. Given that assessments of classification methods for P300 BCI interfaces are fairly common (e.g. Manyakov et al., 2011; Oweis et al. 2013; Krusienski et al. 2010), it is unclear what the value of another contrast of classification methods is to the field. The assessment of the impact of sources of noise is interesting (see point 6), but would need to be expanded on considerably to be of practical value. Furthermore, while this manuscript frequently emphasizes the value of BCI to the clinical field. However, most clinicians are generally uninterested in classifying the presence/absence of a P300 response, arguably the most obvious cognitive ERP component extant. Instead, the focus in that area is instead on subtler dynamics in the EEG signal such as frequency slowing, triphasic waves, and generalized ictal activity. Indeed, the introduction of this manuscript mentions several EEG components of value which notably does not include the P300 in its list. If demonstrating clinical utility is of interest, it would seem more appropriate to target the classification of these subtler elements with a larger sample set and broader response ERP pool.

### 2018-10-22 BrainScience/brainscience.tex
Ready to send it
Added reference to Code Ocean
Start with reviewers modifications

--- 2018-09-25 sensors.tex
@@
 % Keywords
-\keyword{electroencephalography (EEG); ERP,BCI,waveform, signal structure}
+\keyword{electroencephalography; EEG; waveform; ERP ;BCI;P300}
 
@@
 And third, to verify their applicability to a real and public dataset.
-We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal waveform, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective~\citep{Nijboer2009,Wei2017}.
+
+%We believe that the importance of waveform analysis methods, as described here, is that by using this methodology, collaboration could be fostered because there is a clear description and characterization of the signal waveform, where the extensive literature which explores clinical EEG can be reviewed from the same shared perspective~\citep{Nijboer2009,Wei2017}.
 
@@
 These algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
+To maintain reproducibility, the data and the source code has been made available in the online repository of the Code Ocean platform under the name \textit{EEGWave}.
 
@@
 SIFT~\citep{Lowe2004} is a very successful feature extraction technique from Computer Vision .
-It has a biomimetic inspiration on how the visual cortex analyze images based on orientations~\citep{cogprints561}.
+It has a biomimetic inspiration on how the visual cortex analyze images based on orientations~\citep{Edelman1997}.
 This method has been used in \citep{Ramele2016} to analyze EEG signals based on their plots on 2D images.
@@
 This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.
-
-Is it possible to classify and implement a P300 ERP Speller using only the waveform structure of the signal ?
-Which algorithm perform better at this even when segment latency variations and component amplitudes issues are present ?
 
@@
 %=====================================
-\bibliography{sensors}
+\bibliography{brainscience}
 

### 2018-10-22 BrainScience/response.letter.1.tex
Added the response letter and the statistical letters

@@
 a.
-Cruz, A., Pires, G., & Nunes, U.
+Cruz, A., Pires, G., \& Nunes, U.
 J.
@@
 
-c. Kevric, J., & Subasi, A. (2016). The impact of Mspca signal de-noising in real-time wireless brain computer interface system. Southeast Europe Journal of Soft Computing, 4(1).
+c. Kevric, J., \& Subasi, A. (2016). The impact of Mspca signal de-noising in real-time wireless brain computer interface system. Southeast Europe Journal of Soft Computing, 4(1).
 
@@
 Dimitriadis, S.
-I., Salis, C., & Linden, D.
+I., Salis, C., \& Linden, D.
 (2018).
@@
 7.
-“the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of 70%” (Page 19, Paragraph 2, Line 465-466) What is the obtained character identification rate in other scientific studies?
-Is the obtained result (70%) of this study comparable to others?
+“the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of 70\%” (Page 19, Paragraph 2, Line 465-466) What is the obtained character identification rate in other scientific studies?
+Is the obtained result (70\%) of this study comparable to others?
 
@@
 \subsection*{\ovalbox{General Comments}}
-I have read the manuscript "EEG characterization and classification based on Histogram of Gradients of Signal Plots" with great interest since its title promised a solution to an important problem.
-In spite of presenting an interesting methodology and some interesting issues related to EEG characterization, I consider the manuscript is not ready for publication.
+The paper is an interesting research paper and presents an analysis of the P300 ERP.
+The authors provide a literature review of the current methods used to detect EEG patterns and conduct three experiments to identify the P300 ERP in EEG recordings.
+The paper is well-organized.
+The paper should incorporate the following comments before its acceptance for publication:
 \vskip+1ex
@@
 {\color{blue}
-Thank you for your comments and observations. We have modified several points according to your recommendations, we hope the article is now better suited for publication. 
+Thank you for your comments and for this high quality and detailed review.
+We have modified several points according to your recommendations, we hope the article is now better suited for publication.
 }
@@
 
-2.1) the title is misleading since a first, and naive, read suggest being able to classify and characterize any EEG with their different components (i.e., P200, P300, P600, N100, N200, N400 to name a few).
-Clearly, the title does not match the content since the authors focus on the P300
+1.
+The major concern is that the experiments have been conducted with only one participant.
+The authors should collect more data in this study.
+In other studies cited in the paper by authors,  two and 16 participants were recruited.
 
@@
 {\color{blue}
-Thank you for the observation. The title of the article was changed accordingly to your suggestion.  
+We have now included more participants to this experiment.
 }
@@
 
-2.2) there is evidence of the benefits of working with these signals P300, but, is the proposed method extensible to other signals?, this is not clear from the manuscript.
+2.
+It is strongly recommended that the authors should re-check some paragraphs of the manuscript, correct any grammatical and syntactic errors and improve the scientific language.
+In some cases, it was hard to follow the meaning.
+(e.g.
+Page 2, Paragraph 4, Line 51-54, Page 11, Paragraph 3, Line 343-345)
 
@@
 \vskip+1ex
-\subsection*{\ovalbox{Third Paragraph}}
 
-3.1) Another issue with the manuscript is its mathematical vagueness.
-For example, the variable T is used as a vector or two-tuple but its notation is different from the notation the authors use for vectors
+3.
+In the Introduction section the authors refer to BCI applications; however, no scientific studies are reported.
+Also, the literature section should be revised with recent publication.
+Below are some scientific studies that is suggested to be included for BCI applications:
 
@@
 {\color{blue}
-Thank you for the observation.
-We worked hard to improve all the mathematical description of the method.
-The issue with variable T was fixed and we improved on the font styling to differentiate vectors and scalars.
-Section 1.1 was modified as well as section 1.2.3.
-The mathematical terminology were corrected all along the article.
-We hope we have full addressed this point.
+
 }
@@
 
-3.2) Moreover, there is something wrong in equation (4)
+4.
+In the literature section, Table I briefly presents the proposed methods followed by other authors; however, there is no text describing the table.
+The authors should include a paragraph describing each method and the recent findings concisely, in order to present better the literature review.
 
@@
 {\color{blue}
-We appreciate a lot your truthful comment.
-Equation (4) is very important for us because it is the essence of the method.
-%We based that equation on the description of how the histogram of oriented gradients is calculated published in Vedaldi and Szeliski, Computer Vision and Algorithms.
-We reviewed the equation, corrected their terms and we hope it is now more clear.
+
 }
@@
 
-3.3) and something inconsistent between equations (7) and (8)
+5. In Figure 2 please insert the y-axis title to clearly present ERP template.
 
@@
 {\color{blue}
-We have revised the manuscript and modified equations (7) and (8) (now (8) and (9)) which are located on Section 1.2.3 (Parameters).
-We added more information on the text and specified subscript indexes to avoid the inconsistency.
-We thank you for noticing this.
+
 }
@@
 
-3.4) Also, the authors go back and forth the digital space and the continuous space and although there is an equivalence, the manuscript could benefit from a clearer explanation.
+6.
+The Results section is not well-presented.
+The authors are advised to include a paragraph before Table 2, Table 3 and Fig.
+6-8 presenting the obtained results (best/worst channel performance and algorithm performance for each experiment) and then discuss the results in the Discussion section.
+Please revise excessive descriptions in the captions of Table 2, Table 3 and Fig.
+6-8.
 
@@
 {\color{blue}
-Thank you for the observation.
-We modified all equations to emphasize the point that they are calculated in digital space, particularly on equation (4).
-We also added that image gradients are also calculated using finite differences, and added information about the support of each equation to explain that they are actually calculated in digital space.
+
 With this, we hope we have full addressed your point.
@@
 
-\subsection*{\ovalbox{Fourth Paragraph}}
-
-4.1) Another issue that I find worrisome is the design of experiments. It seems that the number of subjects is rather low, it would be interesting to know why the authors decided to use such a subset.
+7.
+“the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of 70\%” (Page 19, Paragraph 2, Line 465-466) What is the obtained character identification rate in other scientific studies?
+Is the obtained result (70\%) of this study comparable to others?
 
@@
 
-4.3) Also, the authors speculate that their results are not that good because they did not take into account the latency and magnitude of the signals, there is previous evidence in the literature that the P300 suffers from these, and other, problems and, thus, one wonders why the authors did not anticipate such problems.
+8. In the Discussion section a table presenting the statistical analysis results would present better the results and would facilitate the reader.
 
-\vskip+1ex
-\noindent \dotfill
-\vskip+1ex
-%
-\begin{quotation}
-{\color{blue}
-Thank you for your comments.
-We added more detailed information that address this issue.
-First in Section 1.1.4  we included information about the signal averaging procedure.
-In Section 1.2.3 we added the parameters that were used.
-More importantly, in Section 2 we added information and references about the tests that we did to address latency and amplitude issues.
-We included in Section 3 information about future and current works where we are also dealing with these problems.
-We hope we full addressed your remarks.
-}
-\end{quotation}
-%
-\vskip+1ex
-\noindent \dotfill
-\vskip+1ex
-
-\subsection*{\ovalbox{Fifth Paragraph}}
-
-5.1) it seems the authors assume that the readers have some previous deep knowledge of some topics such as that some channels should carry more or better information without a previous reference.
-
-\vskip+1ex
-\noindent \dotfill
-\vskip+1ex
-%
-\begin{quotation}
-{\color{blue}
-Thank you for your suggestion.
-The Results section was modified to clarify the expectation about the EEG channels where a visual P300 oddball paradigm generates stronger signals.
-We also added references.
-We hope we  have addressed your point with this.
-}
-\end{quotation}
-%
-\vskip+1ex
-\noindent \dotfill
-\vskip+1ex
-
-5.2)  issue has to do with the way the manuscript is written
-
-\vskip+1ex
-\noindent \dotfill
-\vskip+1ex
-%
-\begin{quotation}
-{\color{blue}
-Thank you a lot for this important comment.
-We re-read all the manuscript and asked  colleagues to proofread the article.
-They gave us interesting suggestions that helped us to improve different sections.
-We would be very pleased if you find it now easier to read it.
-}
-\end{quotation}
-%
-\vskip+1ex
-\noindent \dotfill
-\vskip+1ex
-
-5.3) sometimes is very difficult to follow the ideas the authors want to convey, an example can be found in the paragraph containing equation (5)
-
-\vskip+1ex
-\noindent \dotfill
-\vskip+1ex
-%
-\begin{quotation}
-{\color{blue}
-Thank you very much for this information.
-We reviewed completely that equation (it is now equation (6) and equation (7)).
-We included a step-by-step procedure that describes the proposed algorithm and we added a schematic chart in Figure 3.
-}
-\end{quotation}
-%
-\vskip+1ex
-\noindent \dotfill
-\vskip+1ex
 \bibliographystyle{mdpi}

### 2018-10-23 BrainScience/response.letter.1.tex
Some small changes

@@
 
+Absolutely thank you very much for providing
+
 }
@@
 
+We have modified the text highlighting the connections with the information provided in this table and rearranged the way the information was being shown.
+
 }
@@
 {\color{blue}
-
+Our applogize for this mistake.  We have corrected it and revised all the other figures.
 }
@@
 {\color{blue}
-
+We have included the best/worst channel which is much better for the understanding of the performance obtained for the algorithm.
+We have modified the captions of the tables and figures, thank you very much for raising this issue.
 With this, we hope we have full addressed your point.
@@
 {\color{blue}
-Thank you for your comment.
-The original dataset was published as the 008-2014 dataset at the BNCI-Horizon website, which is very important in BCI Research.
-As far as we are aware of, experiments on this dataset of ALS patients has not been published by any independent group.
-In order to address this issue, we included a new own dataset with the same experimental conditions that we generated ourselves, on healthy subjects.
-We published this dataset on the Kaggle platform for replication.
+Studies on this dataset reported performance values of perfect classification though they used handcrafted algorithms aimed to this dataset.
 }
@@
 
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+We have added a new table explaining all the statistical tests performed, including the information provided for the new experiments.
+}
+\end{quotation}
+%
+
 \bibliographystyle{mdpi}

### 2018-10-23 BrainScience/response.letter.2.tex
Some small changes

@@
+\documentclass[journal,onecolumn,12pt]{IEEEtran}
+
+\usepackage{amsmath,amssymb,bm}
+\usepackage{amsthm, amsfonts}	
+\usepackage{bm,bbm}
+\usepackage[normalem]{ulem}
+\usepackage{color}
+\usepackage{fancybox}
+\usepackage{url,booktabs}
+\usepackage[round]{natbib}
+
+\usepackage{xr}
+
+\title{Reply to Reviewer's Comments on\\
+''EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces''}
+\author{}
+
+\begin{document}
+
+\maketitle
+\pagenumbering{roman}
+\setcounter{page}{1}
+
+We are grateful to the reviewer for pointing out relevant issues in our manuscript.
+
+In the following, we discuss how we dealt with each raised issue.
+
+\vskip+1ex
+\noindent \dotfill
+
+\section*{\fbox{Reviewer \#2 Transcript:}}
+
 This manuscript presents findings that contrast the performance of a selection of BCI classification methods in a single subject database.
@@
 2.
-Several notable papers in the domain of P300-BCI work should be referenced (e.g., Rakotomamonjy & Guigue, 2008) as well as other prior efforts to evaluate P300-BCI classification methods (e.g., Manyakov, Chumerin, Combaz & Hulle, 2011; Oweis, Hamdi, Ghazali & Lwissy, 2013).
+Several notable papers in the domain of P300-BCI work should be referenced (e.g., Rakotomamonjy \& Guigue, 2008) as well as other prior efforts to evaluate P300-BCI classification methods (e.g., Manyakov, Chumerin, Combaz \& Hulle, 2011; Oweis, Hamdi, Ghazali \& Lwissy, 2013).
 
@@
 If demonstrating clinical utility is of interest, it would seem more appropriate to target the classification of these subtler elements with a larger sample set and broader response ERP pool.
+
+\subsection*{\ovalbox{General Comments}}
+This manuscript presents findings that contrast the performance of a selection of BCI classification methods in a single subject database.
+All in all, I find several aspects of this paper quite interesting, particularly its attempt to assess the impact of latency and component amplitude variation on classification performance.
+However, overall, I fear the range/scope of the paper is too limited to make a significant contribution to the literature.
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+
+}
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+
+1.
+Abstract, while automated analysis of EEG data certainly has significant potential with regards to the future of clinical EEG research, it is too soon to unqualifiedly state that these methods as they are “outshine the research done by the clinical EEG community.” Some aspects of the introduction also veer towards hyperbole on occasion and may be better received if they were more conservatively stated.
+
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+We have now included more participants to this experiment.
+}
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+
+2.
+Several notable papers in the domain of P300-BCI work should be referenced (e.g., Rakotomamonjy \& Guigue, 2008) as well as other prior efforts to evaluate P300-BCI classification methods (e.g., Manyakov, Chumerin, Combaz \& Hulle, 2011; Oweis, Hamdi, Ghazali \& Lwissy, 2013).
+
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+Thank you very much for your comment.
+We published in~\citep{Ramele2016} the application of the same method to identify rhythmic EEG events like Visual Occipital Alpha Waves and to classify Motor Imagery.
+We are also working on unpublished material where we are analyzing the same approach for the detection of K-Complexes and classification of  SSVEP patterns but the results are in progress.
+We added this information on the article (Introduction and Conclusion, section 3).
+We hope this new information enriches the article and clarifies this point.
+
+}
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+
+4.
+Page 12, Lines 383-394.
+It should be noted that the dataset from which the template ERP subject was drawn from was an ALS patient dataset.
+It is currently unclear why Subject Number 8 and trial number 2 was selected out of all the subjects and trials in the BNCI-Horizon database.
+
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+
+}
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+
+5.
+Integrating the template ERP signal into a Null-EEG stream and assuming everything is there except the P300 ERP component is probably an untenable assumption.
+As indicated in the manuscript itself, for instance, the null-EEG stream is a passive task.
+While the non-P300 trials in an active task contain additional activity associated with processes including target stimuli anticipation, active task focus, cognitive fatigue, etc.
+This is not true of the Null-EEG stream.
+Furthermore, combining and subsequently classifying EEG data from two different subjects is questionable at best given the physiological differences between the two.
+Differences in recording condition, equipment and task stimuli should also be considered.
+Collectively, these factors make it difficult to assess the value of the classification results observed on this study.
+
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+
+}
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+
+6.
+The separate manipulation of latency, as well as amplitude noise, is interesting and if extended may in and of itself prove to be an interesting manuscript in and of itself (e.g.
+contrasting the impact of a range of different latency variations and amplitude noise on BCI classification ability), true EEG data almost invariantly involves not only a mix of both noise sources but other sources such as cognitive fatigue, systematic task stimuli adaptation etc.
+Thus all three experiments represent assessments of isolated aspects of noise which would be interesting if expanded on in a more expansive simulation set (see point 4 and summary) but currently are too limited in scope to draw strong conclusions regarding the expected inter-component performance in real-world data.
+
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+
+}
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+
+It seems as though the entire study is built on the classification of a single active P300 BCI-Speller template from an ALS participant into a single null-EEG stream of a healthy subject performing a passive-viewing version of a BCI-Speller task.
+The generalizability of these results are thus difficult, if not impossible to assess.
+Given that assessments of classification methods for P300 BCI interfaces are fairly common (e.g.
+Manyakov et al., 2011; Oweis et al.
+2013; Krusienski et al.
+2010), it is unclear what the value of another contrast of classification methods is to the field.
+The assessment of the impact of sources of noise is interesting (see point 6), but would need to be expanded on considerably to be of practical value.
+Furthermore, while this manuscript frequently emphasizes the value of BCI to the clinical field.
+However, most clinicians are generally uninterested in classifying the presence/absence of a P300 response, arguably the most obvious cognitive ERP component extant.
+Instead, the focus in that area is instead on subtler dynamics in the EEG signal such as frequency slowing, triphasic waves, and generalized ictal activity.
+Indeed, the introduction of this manuscript mentions several EEG components of value which notably does not include the P300 in its list.
+If demonstrating clinical utility is of interest, it would seem more appropriate to target the classification of these subtler elements with a larger sample set and broader response ERP pool.
+
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+
+With this, we hope we have full addressed your point.
+ }
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
+
+\bibliographystyle{mdpi}
+\bibliography{article}
+
+\end{document}

### 2018-10-23 BrainScience/brainscience.tex
Added the response letter and the statistical letters
Some small changes

@@
 It is now being used to translate or decode brain signals to diagnose disease or implement Human Computer Interaction devices.
-The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, outshines the research done by the clinical EEG community.
+The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, overshadow the research done by the clinical EEG community.
 Clinical EEG is based intensively on waveforms and the structure of signal plots.
@@
 The best materialization of the automatic processing of EEG signals rests in the BCI discipline, where around $71.2\%$ is based on noninvasive EEG~\citep{Guger2017}.
+
+Important applications have been developed in Spelling areas like this and that, in passive BCI with alcohol consumption monitoring.
+
+KevricJasmin2016
+Tzimourta2018
+Cruz2018
+
+Measure the flickering effect on the original vs the modified signal.
 
@@
 
+\begin{table}[H]
+\caption{Statistical Analysis Tests }
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+\begin{tabular}{ccc}
+\toprule
+\textbf{Comparison}	& \textbf{Statistical Method} &   \textbf{p-value} \\
+\midrule
+Experiment 1 vs. 2 & Wilcoxon rank sum test, one tail  & $0.0022$ \\
+Experiment 1 vs. 3 & Wilcoxon rank sum test, two tails not rejected & $0.1797$ \\
+\bottomrule
+\end{tabular}
+\label{tab:statisticalanalysis1}
+\end{table}
+
+\begin{table}[H]
+\caption{Statistical Analysis Tests  - Multiple comparisons}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+\begin{tabular}{ccc}
+\toprule
+\textbf{Comparison}	& \textbf{Statistical Method} &   \textbf{p-value} \\
+\midrule
+SHCC & MP-1, MP-2  & $0.05$ \\
+SIFT   & MP-2, PE       & $0.1797$ \\
+SVM   & MP-2             & $0.1797$ \\
+\bottomrule
+\end{tabular}
+\label{tab:statisticalanalysis2}
+\end{table}
+
 A two-way balanced Quade test was performed to evaluate the performance's differences for the three experiments and for the six methods.

### 2018-10-25 BrainScience/response.letter.1.tex
Subo un monton de cambios previos a poner todo otra vez

@@
 Thank you for your comments and for this high quality and detailed review.
-We have modified several points according to your recommendations, we hope the article is now better suited for publication.
+We have modified several points according to your recommendations.
+We hope the article is now better suited for publication.
 }
@@
 Thank you very much for your comment.
-We published in~\citep{Ramele2016} the application of the same method to identify rhythmic EEG events like Visual Occipital Alpha Waves and to classify Motor Imagery.
-We are also working on unpublished material where we are analyzing the same approach for the detection of K-Complexes and classification of  SSVEP patterns but the results are in progress.
-We added this information on the article (Introduction and Conclusion, section 3).
-We hope this new information enriches the article and clarifies this point.
-
+We have modified the specific paragraphs that the Reviewer kindly offered as example and we have revised the entire manuscript modifying different parts, simplifying sections.
 }

### 2018-10-25 BrainScience/brainscience.tex
Subo un monton de cambios previos a poner todo otra vez

@@
 It is now being used to translate or decode brain signals to diagnose disease or implement Human Computer Interaction devices.
-The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, overshadow the research done by the clinical EEG community.
+The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, neglected the research done by the clinical EEG community.
 Clinical EEG is based intensively on waveforms and the structure of signal plots.
@@
 EEG sensors are wearable~\citep{Puce2017} non-invasive, portable and mobile~\citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution~\citep{Hartman2005}.
-This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, or control wheelchairs~\citep{Yuste2017}.
+This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, or control wheelchairs, alcohol consumption monitoring~\citep{Yuste2017,KevricJasmin2016,Tzimourta2018,Cruz2018}.
 
@@
 The best materialization of the automatic processing of EEG signals rests in the BCI discipline, where around $71.2\%$ is based on noninvasive EEG~\citep{Guger2017}.
-
-Important applications have been developed in Spelling areas like this and that, in passive BCI with alcohol consumption monitoring.
-
-KevricJasmin2016
-Tzimourta2018
-Cruz2018
-
-Measure the flickering effect on the original vs the modified signal.
 
@@
 
-Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)  \citep{Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines.
+Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}.
 Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g.
@@
 Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.
+Results for this competition show perfect classification with tailored algorithm.
+Even comparing the results with SVM multichannel, which uses all the available information from all channels show comparable results.
+
 It is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.
-It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Gribonval2008}.
+It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.
 

### 2018-10-25 BrainScience/brainscience.tex
Small changes

@@
 It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time.
-It is now being used to translate or decode brain signals to diagnose disease or implement Human Computer Interaction devices.
-The automatic processing approach which is based on using quantitative algorithms to detect the cloaked information buried in the signal, neglected the research done by the clinical EEG community.
-Clinical EEG is based intensively on waveforms and the structure of signal plots.
+It is now being used to translate or decode brain signals to diagnose disease or implement Brain Computer Interface (BCI) devices to transmit volitional information.
+The automatic decoding is mainly implemented by using quantitative algorithms to detect the cloaked information buried in the signal.
+However, clinical EEG is based intensively on waveforms and the structure of signal plots.
 Hence, the purpose of this work is to establish a bridge to fill this gap by reviewing and describing the procedures that have been used to detect patterns in the electroencephalographic waveforms, benchmarking them on a controlled pseudo-real dataset of an EEG Event Related Potentials waveform and verifying their performance on a public dataset of a BCI Competition.}
@@
 % Keywords
-\keyword{electroencephalography; EEG; waveform; ERP ;BCI;P300}
+\keyword{electroencephalography; BCI; EEG; waveform; ERP ;P300}
 
@@
 EEG sensors are wearable~\citep{Puce2017} non-invasive, portable and mobile~\citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution~\citep{Hartman2005}.
-This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, or control wheelchairs, alcohol consumption monitoring~\citep{Yuste2017,KevricJasmin2016,Tzimourta2018,Cruz2018}.
-
-The clinical and historical tactic to analyze EEG signals were based on detecting visual patterns out of the EEG trace or polygraph\citep{Hartman2005}: multichannel signals were extracted and continuously plotted over a piece of paper.
-Electroencephalographers or Electroencephalography technician have decoded and detected patterns along the signals by visually inspecting them~\citep{Schomer2010}.
+This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, control wheelchairs or to assess alcohol consumption~\citep{Yuste2017,KevricJasmin2016,Tzimourta2018,Cruz2018}.
+
+The clinical and historical tactic to analyze EEG signals is based on detecting visual patterns out of the EEG trace or polygraph\citep{Hartman2005}: multichannel signals are extracted and continuously plotted over a piece of paper.
+Electroencephalographers or Electroencephalography technician decode and detect patterns along the signals by visually inspecting them~\citep{Schomer2010}.
 Nowadays clinical EEG still remains a visually interpreted test~\citep{Hartman2005}.
 
-In contrast, automatic processing, or quantitative EEG (qEEG), was based first on analog electronic devices~\citep{Jestico1977} and later on computerized digital processing methods~\citep{Jansen1991}.
-They implemented algorithmically complex procedures to decode the information associated with a clinical condition, or even some higher cognitive state \citep{Yuste2017}.
+%In contrast, automatic processing, or quantitative EEG (qEEG), was based first on analog electronic devices~\citep{Jestico1977} and later on computerized digital processing methods~\citep{Jansen1991}.
+Complex procedures to decode the information associated with a clinical condition, or even some higher cognitive state \citep{Yuste2017}.
 The best materialization of the automatic processing of EEG signals rests in the BCI discipline, where around $71.2\%$ is based on noninvasive EEG~\citep{Guger2017}.
 
-%rich clinical literature
-
-Hence, the traditional and knowledgeable approach was mainly overshadow particularly in BCI Research, and the waveform of the EEG was replaced by pragmatic procedures that were difficult to link to existing clinical EEG knowledge.
+The need of quantitative procedures to automate the decoding of EEG signals has been materialized in BCI where around $71.2\%$ is based on noninvasive EEG~\citep{Guger2017}.
+However, methods of decoding signals based on the detection of waveforms has been scarce.
+Hence, this traditional and knowledgeable approach has been neglected particularly in BCI Research.
 We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and which can decode them in a supervised and semi-automated procedure.

### 2018-10-26 BrainScience/response.letter.2.tex
Draft

@@
 
+We would like to honestly thank the Reviewer for this absolutely excellent and detailed review.
+We have modified the experiments and several parts of the manuscript in order to make it more suitable for publication.
+The list of changes can be found at the end of this letter.
+
 }
@@
 1.
-Abstract, while automated analysis of EEG data certainly has significant potential with regards to the future of clinical EEG research, it is too soon to unqualifiedly state that these methods as they are “outshine the research done by the clinical EEG community.” Some aspects of the introduction also veer towards hyperbole on occasion and may be better received if they were more conservatively stated.
+Abstract, while automated analysis of EEG data certainly has significant potential with regards to the future of clinical EEG research, it is too soon to unqualifiedly state that these methods as they are "outshine the research done by the clinical EEG community." Some aspects of the introduction also veer towards hyperbole on occasion and may be better received if they were more conservatively stated.
 
@@
 {\color{blue}
-We have now included more participants to this experiment.
+We apologize for this mistake.
+We definitely selected an incorrect wording because the message that we aimed to convey was exactly the opposite.
+We have modified the abstract to emphasize the message and the idea that we failed to transmit in our original manuscript.
 }
@@
 {\color{blue}
-Thank you very much for your comment.
-We published in~\citep{Ramele2016} the application of the same method to identify rhythmic EEG events like Visual Occipital Alpha Waves and to classify Motor Imagery.
-We are also working on unpublished material where we are analyzing the same approach for the detection of K-Complexes and classification of  SSVEP patterns but the results are in progress.
-We added this information on the article (Introduction and Conclusion, section 3).
-We hope this new information enriches the article and clarifies this point.
-
+These important references were added to the manuscript.  Thank you for pointing them out.
 }
@@
 {\color{blue}
+
+This is a very important comment.
+The reason we selected that particular trial and subject was because our manuscript stressed the importance of the waveform, the shape of the ERP.
+The obtained P300 waveform of Subject Number 8, trial 2 was the most prototypical that we've found along the BNCI-Horizon dataset of ALS patients.
+The analysis was performed by visual observation.
+We have now added also 4 more templates for the Active part of the pseudo-real dataset.
+
+We have now included this reason on the manuscript which was mistakenly omitted on our first version.
 
@@
 
+The recording conditions of both datasets are exactly the same.
+The EEG device, though not the same exact model, is from the same manufacturer and the sampling frequency is as close as possible.
+All the preprocessing stage was implemented according to~\citep{Riccio2013}.
+
+Following Reviewer comments, the pseudo-real dataset is now divided in two.
+First, we've replicated the experiment conducted by~\cite{Riccio2013} et al on 4 healthy subjects performing an active-modality P300 copy-spelling task while using the P300-Based BCI Speller.
+We also extended the original single null-EEG stream to 4 subjects while performing the passive-modality where the letter to copy-spell is not informed to the user and the subject is unaware of how the matrix works.
+
+Regarding the active-modality, for each subject a the most prototypical P300 waveform template is selected and extracted.
+This template is superimposed to each segment marked as hit on the EEG stream.
+The baseline was obtained by picking randomly any no hit segment for the same subject.
+
 }
@@
 {\color{blue}
+
+The purpose of these experiments were to assess the incidence of the most common source of noise in the generation of the P300, to the shape of the obtained waveform of the P300, "how the averaged signal looked-like".
 
@@
 
-With this, we hope we have full addressed your point.
+We would like to emphasize that the purpose of this work was to raise awareness about the automatic procedures that can analyze EEG signals by their waveforms, mimicking what traditionally EEG physicians performed on the clinical EEG.
+As far as we are aware of, no work has addressed this topic in this form and we humbly believe it could be a contribution.
+The Reviewer is right, but the objective of this manuscript was to establish the connection, and to focus on the goal of the BCI discipline to foster interaction between the clinical and BCI community.
+
+We have modified the manuscript to emphasize these aspects.
+
  }
@@
 \bibliographystyle{mdpi}
-\bibliography{article}
+\bibliography{brainscience}
 

### 2018-10-26 BrainScience/brainscience.tex
Draft

@@
 However, methods of decoding signals based on the detection of waveforms has been scarce.
-Hence, this traditional and knowledgeable approach has been neglected particularly in BCI Research.
+Hence, the traditional and knowledgeable approach has been neglected particularly in BCI Research.
 We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and which can decode them in a supervised and semi-automated procedure.
@@
 Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.
+
+Table~\ref{tab:methods} shows a list of depictions used to describe waveforms in the surveyed literature.
+Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.
+For ERPs on the other hand, more elaborate indexes has been provided establishing relations between signal components amplitudes.
+Finally, in sleep studies and ICU research areas is where the most complex indexes has been derived, particularly coupling signal properties like phase, amplitude and frequency.
 
@@
 
-Farwell and Donchin P300 Speller \citep{Farwell1988} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
+Farwell and Donchin P300 Speller \citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
 This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
@@
 
-The template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+\paragraph{Passive Modality}
+
+In this case the template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
 Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.
 This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.
+This template is selected due to its shape which more closely resemble the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
 
 An EEG stream with null-P300 signal is obtained by the following procedure: 
-A subject participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
-No monetary compensation is handed out and she/he agrees and signs a written informed consent.
+Four (4) participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
+No monetary compensation is handed out and they agree and sign a written informed consent.
 This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.
-The participant is healthy and have normal or corrected-to-normal vision and no history of neurological disorders.
-This voluntary subject is aged between 20-30 years old.
+The participants are healthy and have normal or corrected-to-normal vision and no history of neurological disorders.
+These voluntary subjects are aged between 20-40 years old.
 EEG data is collected in a single recording session.
-She/He is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.
+Each subject is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.
 The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.
@@
 
-The participant is instructed to passively watch the flashing screen while not focusing on any particular letter.
+All participants are instructed to passively watch the flashing screen while not focusing on any particular letter.
+None of them have any experience with a BCI device.
 A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.
@@
 
+\paragraph{Active modality}
+
+The experiment conducted by~\citep{Riccio2013} is replicated on 4 healthy subjects, performing a copy-spelling task with feedback.
+In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.
+The most prototypical P300 waveform template is selected by visual observation as well.
+This template is superimposed to each segment marked as hit on the EEG stream.
+The baseline of each segment, on the other hand, is obtained by picking randomly any no hit segment for the same subject.
+
 \subsubsection{Classification} \label{section:classification}
@@
 This is repeated one hundred times, and performances averaged.
-Figure \ref{fig:performancetestlatency} shows the same results for the Experiment 2, where a noisy latency lag was included.
-Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template is randomly attenuated.
-Furthermore, results obtained for the dataset BCI Competition 2003 IIb are shown in Figures \ref{fig:performancebcicompetition} and in Table~\ref{tab:bcicompetitionresults}.
-For this experiment the number of available intensification sequences is 15.
 
@@
 
+Figure \ref{fig:performancetestlatency} shows the same results for the Experiment 2, where a noisy latency lag was included.
+Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template is randomly attenuated.
+
 \begin{figure}[H]
@@
 \end{figure}
+
+Furthermore, results obtained for the dataset BCI Competition 2003 IIb are shown in Figures \ref{fig:performancebcicompetition} and in Table~\ref{tab:bcicompetitionresults}.
+For this experiment the number of available intensification sequences is 15.
 
@@
 Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.
-Results for this competition show perfect classification with tailored algorithm.
-Even comparing the results with SVM multichannel, which uses all the available information from all channels show comparable results.
-
-It is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.
+When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.
+Results for this competition have shown perfect classification with tailored algorithm.
+This level is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.
 It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.
@@
 
-The conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.
+%The conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.
 At the same time, it is a subjective time-consuming task, with long-learning curves, requires specialized personnel, and it has significant error rates~\citep{Tjepkema-Cloostermans2018}.
@@
 
-These methods have the advantage that they can map a visual component with a clinical meaning to a feature with an objective representation.
+%These methods have the advantage that they can map a visual component with a clinical meaning to a feature with an objective representation.
 Compound classifiers or ensemble of features can be explored.

### 2018-10-26 BrainScience/response.letter.1.tex

@@
 {\color{blue}
-Thank you for your comments and for this high quality and detailed review.
-We have modified several points according to your recommendations.
-We hope the article is now better suited for publication.
+Thank you for your encouraging and thoughtful comments. We have modified several points according to your recommendations. We hope the article is now better suited for publication. 
 }
@@
 {\color{blue}
-We have now included more participants to this experiment.
+We have now included more participants to this experiment and it was divided in two modalities.
+The first, which is now called passive-modality, includes now 4 subjects.
+The second, an active-modality, now includes 4 more participants who were engaged in a copy-spelling task on a P300-Based BCI Speller.
+For this subjects, the pseudo-real dataset is constructed by superimposed a P300 template obtained for each subject into segments marked as hit.
+The baseline for each segment is constructed from a random no-hit segment for the same subject.
 }
@@
 
-Absolutely thank you very much for providing
+Absolutely thank you very much for providing these very important references.  We added all of them.
 
@@
 
-We have modified the text highlighting the connections with the information provided in this table and rearranged the way the information was being shown.
+We have included a new paragraph describing the depictions which are shown in the table and in which areas are they used for.
 
@@
 {\color{blue}
-We have included the best/worst channel which is much better for the understanding of the performance obtained for the algorithm.
-We have modified the captions of the tables and figures, thank you very much for raising this issue.
+We have included the best/worst channel obtained for each algorithm, subject and experiment.
+This enhancement contributes greatly to the understanding of the obtained values.
+We have also modified the captions of the tables and figures, thank you very much for raising this issue.
 With this, we hope we have full addressed your point.
@@
 {\color{blue}
-Studies on this dataset reported performance values of perfect classification though they used handcrafted algorithms aimed to this dataset.
+Studies on this dataset reported performance values of perfect classification.
+However in this case, as we are emphasizing the waveforms of the signals, the classification uses only single-channel activity, which is less information.
+Regardless of this, the obtained results for some of the algorithms can be used to implement usable P300-Based Speller applications due to the usage of word predictive schemes.
 }

### 2018-10-26 BrainScience/response.letter.1.tex
Upload a new revised version

@@
 We have included the best/worst channel obtained for each algorithm, subject and experiment.
-This enhancement contributes greatly to the understanding of the obtained values.
+We hope this enhancement contributes greatly to the understanding of the obtained values.
 We have also modified the captions of the tables and figures, thank you very much for raising this issue.
@@
 {\color{blue}
-Studies on this dataset reported performance values of perfect classification.
+The reported binary classification accuracy for this BCI Competition was around 95\%.
 However in this case, as we are emphasizing the waveforms of the signals, the classification uses only single-channel activity, which is less information.

### 2018-10-26 BrainScience/response.letter.2.tex
Upload a new revised version

@@
 
-This is a very important comment.
 The reason we selected that particular trial and subject was because our manuscript stressed the importance of the waveform, the shape of the ERP.
@@
 The analysis was performed by visual observation.
-We have now added also 4 more templates for the Active part of the pseudo-real dataset.
+We have now added also 4 more for the Active part of the pseudo-real dataset.
 
@@
 
-Regarding the active-modality, for each subject a the most prototypical P300 waveform template is selected and extracted.
-This template is superimposed to each segment marked as hit on the EEG stream.
-The baseline was obtained by picking randomly any no hit segment for the same subject.
+Regarding the active-modality, for each subject, the most prototypical P300 waveform template is selected and extracted.
+This template is superimposed on each preprocessed segment marked as hit.
+The baseline was obtained by picking randomly any no-hit segment for the same subject.
 
@@
 
-The purpose of these experiments were to assess the incidence of the most common source of noise in the generation of the P300, to the shape of the obtained waveform of the P300, "how the averaged signal looked-like".
+We appreciate a lot this insight which we also found interest to look after and analyze more in detail.
+However, we would like to clarify that the purpose of these experiments were to assess the incidence of the most common source of noise in the generation of the P300, to the shape of the obtained waveform of the P300, "how the averaged signal looked-like".
 
@@
 As far as we are aware of, no work has addressed this topic in this form and we humbly believe it could be a contribution.
-The Reviewer is right, but the objective of this manuscript was to establish the connection, and to focus on the goal of the BCI discipline to foster interaction between the clinical and BCI community.
+The Reviewer is right, but the objective of this manuscript was to establish the connection, and to focus on the goal of the BCI discipline to foster interaction between the clinical and BCI community~\citep{Chavarriaga2017}.
 

### 2018-10-26 BrainScience/brainscience.tex
Upload a new revised version

@@
 
-\begin{table}[H]
-\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
+%\begin{table}[H]
+%\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
 Additionally, the control algorithm SVM Support Vector Machines is included for comparison.
@@
 }
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccccc}
-\toprule
-\textbf{Method}	& \textbf{Channel} &   \multicolumn{3}{c}{Performance} \\
-%\cline{3-5} \\
- 	&  &  \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}\\
-\midrule
-MP 1 & PO8  & $67\%$ & $15\%$ & $50\%$\\
-MP 2 & PO7 & $24\%$ & $6\%$ & $10\%$\\
-SIFT  & PO8 & $91\%$ & $18\%$ & $66\%$\\
-PE     & Cz & $61\%$ & $9\%$ & $32\%$\\
-SHCC & P4 & $98\%$ & $31\%$ & $80\%$\\
-SVM     & PO8  & $78\%$ & $7\%$ & $53\%$\\
-\bottomrule
-\end{tabular}
-\label{tab:results}
-\end{table}
+%\centering
+%%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+%\begin{tabular}{c|cccc|cccc|cccc|cccc|cccc|cccc|cccc|cccc}
+%\toprule
+%\textbf{Method}	& \multicolumn{4}{c}{Subject 1} & \multicolumn{4}{c}{Subject 2} & \multicolumn{4}{c}{Subject 3} & \multicolumn{4}{c}{Subject 4} & \multicolumn{4}{c}{Subject 5} & \multicolumn{4}{c}{Subject 6} & \multicolumn{4}{c}{Subject 7} & \multicolumn{4}{c}{Subject 8} \\
+%%\cline{3-5} \\
+% 	&  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}\\
+%\midrule
+%MP 1 & PO8  & $67\%$ & $15\%$ & $50\%$  & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$   \\
+%MP 2 & PO7 & $24\%$ & $6\%$ & $10\%$  & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
+%SIFT  & PO8 & $91\%$ & $18\%$ & $66\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
+%PE     & Cz & $61\%$ & $9\%$ & $32\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
+%SHCC & P4 & $98\%$ & $31\%$ & $80\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
+%SVM     & PO8  & $78\%$ & $7\%$ & $53\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
+%\bottomrule
+%\end{tabular}
+%\label{tab:results}
+%\end{table}
 
@@
 \centering
-\includegraphics[width=15cm]{images/CrossPerformanceTest.eps}
+\includegraphics[width=7.5cm]{images/1-1.eps}
+\includegraphics[width=7.5cm]{images/2-1.eps}
 \caption{Speller performance obtained for each method for the Experiment 1.
@@
 \centering
-\includegraphics[width=15cm]{images/CrossPerformanceTestLatency.eps}
+\includegraphics[width=7.5cm]{images/1-2.eps}
+\includegraphics[width=7.5cm]{images/2-2.eps}
 \caption{Speller performance obtained for each method while latencies are artificially added to each single-intensification segment corresponding to the Experiment 2.
@@
 \centering
-\includegraphics[width=15cm]{images/CrossPerformanceTestAmplitude.eps}
+\includegraphics[width=7.5cm]{images/1-3.eps}
+\includegraphics[width=7.5cm]{images/2-3.eps}
 \caption{Speller performance obtained for the Experiment 3 while the amplitudes of the P3b component of the superimposed ERP is randomly reduced.

### 2018-10-27 BrainScience/response.letter.1.tex
Submitted version

@@
 The second, an active-modality, now includes 4 more participants who were engaged in a copy-spelling task on a P300-Based BCI Speller.
-For this subjects, the pseudo-real dataset is constructed by superimposed a P300 template obtained for each subject into segments marked as hit.
+For these subjects, the pseudo-real dataset is constructed by superimposed a P300 template obtained for each subject into segments marked as hit.
 The baseline for each segment is constructed from a random no-hit segment for the same subject.
@@
 We have also modified the captions of the tables and figures, thank you very much for raising this issue.
-With this, we hope we have full addressed your point.
+
  }
@@
 {\color{blue}
-The reported binary classification accuracy for this BCI Competition was around 95\%.
+The reported binary classification accuracy for this BCI Competition was around 100\%.
 However in this case, as we are emphasizing the waveforms of the signals, the classification uses only single-channel activity, which is less information.
-Regardless of this, the obtained results for some of the algorithms can be used to implement usable P300-Based Speller applications due to the usage of word predictive schemes.
+Regardless of this, the obtained results for some of the algorithms can be used to implement usable P300-Based Speller applications due to the utilization of word predictive schemes.
 }
@@
 {\color{blue}
-We have added a new table explaining all the statistical tests performed, including the information provided for the new experiments.
+We have now included more subjects and decided to remove the statistical analysis sections which was prepared for only one subject.
 }

### 2018-11-06 BrainScience/brainscience.tex
Changes for the round 2

--- 2018-10-26 BrainScience/brainscience.tex
@@
 
-The clinical and historical tactic to analyze EEG signals is based on detecting visual patterns out of the EEG trace or polygraph\citep{Hartman2005}: multichannel signals are extracted and continuously plotted over a piece of paper.
+The clinical and historical tactic to analyze EEG signals is based on detecting visual patterns out of the EEG trace or polygraph~\citep{Hartman2005}: multichannel signals are extracted and continuously plotted over a piece of paper.
 Electroencephalographers or Electroencephalography technician decode and detect patterns along the signals by visually inspecting them~\citep{Schomer2010}.
@@
 
-This article unfolds as follows: Section \ref{EEG} will provide a brief introduction to EEG and the particularities of the EEG waveform characterization.
-Section \ref{Algorithms} will explain the algorithms based on the waveform that will be analyzed.
-In Section \ref{Experimental} the experimentation procedure will be explained.
-Results will be presented in Section \ref{section:results} and finally Discussion and Conclusions will be established in the final sections.
+This article unfolds as follows: Section \ref{EEG} provides a brief introduction to EEG and the particularities of the EEG waveform characterization.
+Section \ref{Algorithms} explains the algorithms waveform-based algorithms that are analyzed.
+In Section \ref{Experimental} the experimentation procedure is explained.
+Results are presented in Section \ref{section:results} and finally Discussion and Conclusions are expounded in the final sections.
 
@@
 Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.
-For ERPs on the other hand, more elaborate indexes has been provided establishing relations between signal components amplitudes.
-Finally, in sleep studies and ICU research areas is where the most complex indexes has been derived, particularly coupling signal properties like phase, amplitude and frequency.
+For ERPs on the other hand, more elaborate indexes has been provided, establishing relations between amplitudes of signal components.
+Finally, Sleep studies and ICU research are areas where the most complex indexes has been derived, particularly the coupling of signal properties like phase, amplitude and frequency.
 
@@
 
-In order to verify this in a controlled procedure, a real P300 ERP template, obtained from a public dataset, is superimposed into a distinct EEG stream.
+The objective of the following experiments is to assess the performance of the algorithms that aim to recognize the shape of the P300 waveform, obtained after averaging signal segments.
+This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.
+
+First for a passive modality, a real P300 ERP template, obtained from a public dataset, is superimposed into a distinct EEG stream.
 This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on any letter in particular.
@@
 By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to similar procedures used in other works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
+
+Second, an active-modality is also used, where a P300-Based BCI experiment is performed on four subjects, but the signal segments are modified to produce the desired P300 response, but based on information extracted from the same subject.
 
@@
 
-The classification method Support Vector Machine SVM with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
-This method has been proved efficient in decoding P300 in several BCI Competitions~\citep{Kaper2004}.
-
 All these experiments are executed using cross validation procedure dividing the letter to spell in two sets, preserving the structure of the letter identification trials.
@@
 Additionally, in this dataset the number of available intensification number sequences is $15$.
+The classification method Support Vector Machine SVM with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
+This method has been proved efficient in decoding P300 in several BCI Competitions~\citep{Kaper2004}.
 
 \subsubsection{Pseudo-Real Dataset Generation}
+
+BCI researcers seem to ignore the fact that these systems consist of a closed loop between the human and the mahine.
+the human is typically considered as a static compliant entity that always performs what the experimenters asks for in a precise and consistent way.
 
@@
 \subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
-\caption{Eight-channel EEG signal without and with the superimposed ERP Template.
+\caption{Eight-channel EEG signal for Subject Number 1 of the psuedo-real dataset without and with the superimposed ERP Template.
 The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented.
@@
 \includegraphics[width=1.0\linewidth]{images/GainCheck.eps}
-\caption{Point-to-point averaged signals for the first letter identification trial.
+\caption{Point-to-point averaged signals.
+These are extracted from the first letter identification trial of the Subject 1 of the psuedo-real dataset.
 The ERP is superimposed on classes 3 and 9.
@@
 In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.
-The most prototypical P300 waveform template is selected by visual observation as well.
-This template is superimposed to each segment marked as hit on the EEG stream.
-The baseline of each segment, on the other hand, is obtained by picking randomly any no hit segment for the same subject.
+The most prototypical P300 waveform template is selected by visual observation.
+This template is superimposed to each segment marked as \textit{True} on the EEG stream.
+The baseline of each segment, on the other hand, is obtained by picking randomly any \textit{False} segment for the same subject.
 
@@
 \caption{Coherently averaged signals containing the superimposed ERP.
-Each one is extracted from the 15 first trials (2 signals from each trial, one belonging to the column flashing and the other to row flashing).
-These are the templates used to build a dictionary per channel and that are used by the classification algorithm described in \ref{section:classification}.}
+Each one is extracted from the EEG signal of the Subject 1 of the pseudo-real dataset.
+These averaged signals correspond to the 15 first trials (2 signals from each trial, one belonging to the column flashing and the other to row flashing).
+These are the templates used to build a dictionary per channel per subject and are used by the classification algorithm described in \ref{section:classification}.}
 \label{fig:dictionaryfig}
@@
 
-Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.
-Table \ref{tab:results} shows the  performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance is attained.
-Figure \ref{fig:performancetest} shows the performance curves for six algorithms.
+Results for the first experiment are shown in Figures~\ref{fig:1-1}  and \ref{fig:2-1}.
+The performance while identifying each letter of the standard P300 Speller Matrix, and the channels where the best and worst performance is attained, is shown.
 Each one represents the percentage of letters that is actually predicted by the algorithms using a cross-validation procedure.
@@
 This is repeated one hundred times, and performances averaged.
-
-\begin{table}[H]
-\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain.
-Additionally, the control algorithm SVM Support Vector Machines is included for comparison.
-All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown.
-In this case with absence of null-signals, it can be interpreted as the channel that adds less noise to the ERP template.
-All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal.
+Figure~\ref{fig:1-1} show the results for the passive modality while Figure~\ref{fig:2-1} shows the results for the active modality.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=15cm]{images/1-1.eps}
+\caption{\textbf{Passive Modality - Experiment 1}:Speller performance curves obtained for each method for the four subjects that performed the active modality protocol.
+Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.
+The two curves show the performance for the best and worst performing channel.
 }
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccccc}
-\toprule
-\textbf{Method}	& \textbf{Channel} &   \multicolumn{3}{c}{Performance} \\
-%\cline{3-5} \\
- 	&  &  \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}\\
-\midrule
-MP 1 & PO8  & $67\%$ & $15\%$ & $50\%$\\
-MP 2 & PO7 & $24\%$ & $6\%$ & $10\%$\\
-SIFT  & PO8 & $91\%$ & $18\%$ & $66\%$\\
-PE     & Cz & $61\%$ & $9\%$ & $32\%$\\
-SHCC & P4 & $98\%$ & $31\%$ & $80\%$\\
-SVM     & PO8  & $78\%$ & $7\%$ & $53\%$\\
-\bottomrule
-\end{tabular}
-\label{tab:results}
-\end{table}
-
-Figure \ref{fig:performancetestlatency} shows the same results for the Experiment 2, where a noisy latency lag was included.
-Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template is randomly attenuated.
+\label{fig:1-1}
+\end{figure}
 
@@
 \centering
-\includegraphics[width=15cm]{images/CrossPerformanceTest.eps}
-\caption{Speller performance obtained for each method for the Experiment 1.
-Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
-\label{fig:performancetest}
+\includegraphics[width=15cm]{images/2-1.eps}
+\caption{\textbf{Active Modality - Experiment 1}:Speller performance curves obtained for each method for the four subjects that performed the active modality protocol.
+Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.
+The two curves show the performance for the best and worst performing channel.
+}
+\label{fig:2-1}
 \end{figure}
 
+Figures~\ref{fig:1-2} and ~\ref{fig:2-2}  shows the performance curves for five algorithms for the second experiment, where a noisy latency lag is included. Best and worst channels are also shown.
+
 \begin{figure}[H]
 \centering
-\includegraphics[width=15cm]{images/CrossPerformanceTestLatency.eps}
-\caption{Speller performance obtained for each method while latencies are artificially added to each single-intensification segment corresponding to the Experiment 2.
-The achieved performance is significantly reduced for all methods.
-Y-axis shows letter identification performance while X-axis shows the number of intensification sequences used to calculate the ensemble average.}
-\label{fig:performancetestlatency}
+\includegraphics[width=15cm]{images/1-2.eps}
+\caption{\textbf{Passive Modality - Experiment 2}: Performance curves for four subjects for the five algorithms while a random latency is included while superimposed the P300 signal template.}
+\label{fig:1-2}
 \end{figure}
@@
 \centering
-\includegraphics[width=15cm]{images/CrossPerformanceTestAmplitude.eps}
-\caption{Speller performance obtained for the Experiment 3 while the amplitudes of the P3b component of the superimposed ERP is randomly reduced.
-Y-axis shows performance accuracy while X-axis shows the number of intensification sequences.}
-\label{fig:performancetestamplitude}
+\includegraphics[width=15cm]{images/2-2.eps}
+\caption{\textbf{Active Modality - Experiment 2}: Performance curves for four subjects for the five algorithms while a random latency is included while superimposed the P300 signal template.}
+\label{fig:2-2}
+\end{figure}
+
+Finally, Figure \ref{fig:1-3} and \ref{fig:2-3} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template is randomly attenuated.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=15cm]{images/1-3.eps}
+\caption{\textbf{Passive Modality - Experiment 3}:Performance curves for four subjects for the five algorithms when he amplitude of the P3b component of the template is randomly attenuated. }
+\label{fig:1-3}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=15cm]{images/2-3.eps}
+\caption{\textbf{Active Modality - Experiment 3}:Performance curves for four subjects for the five algorithms when the amplitude of the P3b component of the template is randomly attenuated. }
+\label{fig:2-3}
 \end{figure}
@@
 
-A significant reduction of performance was found when latency noise is added (Experiment 1 vs 2, Wilcoxon rank sum test, one tail, with $p=0.0022$).
+%A significant reduction of performance was found when latency noise is added (Experiment 1 vs 2, Wilcoxon rank sum test, one tail, with $p=0.0022$).
 The latency noise reduces the information contained in the averaged signal, mainly due to the invalidation of the SNR enhancement performed by the signal averaging procedure.
@@
 
-\begin{table}[H]
-\caption{Statistical Analysis Tests }
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccc}
-\toprule
-\textbf{Comparison}	& \textbf{Statistical Method} &   \textbf{p-value} \\
-\midrule
-Experiment 1 vs. 2 & Wilcoxon rank sum test, one tail  & $0.0022$ \\
-Experiment 1 vs. 3 & Wilcoxon rank sum test, two tails not rejected & $0.1797$ \\
-\bottomrule
-\end{tabular}
-\label{tab:statisticalanalysis1}
-\end{table}
-
-\begin{table}[H]
-\caption{Statistical Analysis Tests  - Multiple comparisons}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccc}
-\toprule
-\textbf{Comparison}	& \textbf{Statistical Method} &   \textbf{p-value} \\
-\midrule
-SHCC & MP-1, MP-2  & $0.05$ \\
-SIFT   & MP-2, PE       & $0.1797$ \\
-SVM   & MP-2             & $0.1797$ \\
-\bottomrule
-\end{tabular}
-\label{tab:statisticalanalysis2}
-\end{table}
-
-A two-way balanced Quade test was performed to evaluate the performance's differences for the three experiments and for the six methods.
+A significant reduction of performance was found when latency noise is added.
+The latency noise reduces the information contained in the averaged signal, mainly due to the invalidation of the SNR enhancement performed by the signal averaging procedure.
+This reduction alters the obtained shape of the waveform of the ERP and impacts on the performances regardless of the method.
+On the other hand, all the algorithms show some resistance to noise in peak amplitudes of the main component.
+This is shown by the similarities of obtained results between the Experiment 1 and 3.
+
+%\begin{table}[H]
+%\caption{Statistical Analysis Tests }
+%\centering
+%%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+%\begin{tabular}{ccc}
+%\toprule
+%\textbf{Comparison}	& \textbf{Statistical Method} &   \textbf{p-value} \\
+%\midrule
+%Experiment 1 vs. 2 & Wilcoxon rank sum test, one tail  & $0.0022$ \\
+%Experiment 1 vs. 3 & Wilcoxon rank sum test, two tails not rejected & $0.1797$ \\
+%\bottomrule
+%\end{tabular}
+%\label{tab:statisticalanalysis1}
+%\end{table}
+
+%\begin{table}[H]
+%\caption{Statistical Analysis Tests  - Multiple comparisons}
+%\centering
+%%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+%\begin{tabular}{ccc}
+%\toprule
+%\textbf{Comparison}	& \textbf{Statistical Method} &   \textbf{p-value} \\
+%\midrule
+%SHCC & MP-1, MP-2  & $0.05$ \\
+%SIFT   & MP-2, PE       & $0.1797$ \\
+%SVM   & MP-2             & $0.1797$ \\
+%\bottomrule
+%\end{tabular}
+%\label{tab:statisticalanalysis2}
+%\end{table}
+
+%A two-way balanced Quade test was performed to evaluate the performance's differences for the three experiments and for the six methods.
 Difference among the method's letter identification rates were found with $p=0.0019$.
@@
 
-Additionally, using a straightforward dictionary of templates for \textit{MP-1} proved more beneficial in terms of performance than the approach of using a Hilbert base of Wavelets atoms on \textit{MP-2} (significant differences between MP-1 vs MP-2).
+Using a straightforward dictionary of templates for \textit{MP-1} proved more beneficial in terms of performance than the approach of using a Hilbert base of Wavelets atoms on \textit{MP-2}.
 Either applying latency noise or amplitude noise, the method based on the signal's templates instead of using their coefficients achieve much better character identification rates.

### 2018-11-07 BrainScience/response.letter.3.tex
Added the response letter for the Reviewer 2, the second round

\documentclass[journal,onecolumn,12pt]{IEEEtran}

\usepackage{amsmath,amssymb,bm}
\usepackage{amsthm, amsfonts}	
\usepackage{bm,bbm}
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{fancybox}
\usepackage{url,booktabs}
\usepackage[round]{natbib}

\usepackage{xr}

\title{Reply to Reviewer's Comments on\\
''EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces''}
\author{}

\begin{document}

\maketitle
\pagenumbering{roman}
\setcounter{page}{1}

We are grateful to the reviewer for pointing out relevant issues in our manuscript.

In the following, we discuss how we dealt with each raised issue.

\vskip+1ex
\noindent \dotfill

\section*{\fbox{Reviewer \#2 Round 2 - Transcript:}}

The authors have addressed many of my concerns regarding this manuscript. There are however a few points that could benefit from clarification.

1) While the authors have addressed the issue of having just a single null-EEG stream from a single subject, it appears to be the case still that a single P300 ERP template is being used. This would seem to either assuming that the P300 is generally invariant across trials (which is unlikely) or that the result of this assessment is restricted to the case in which the ERP response of interest does not vary between trials. This should at least be discussed as a limitation in the conclusion section of the manuscript.

2) The write-up on the active modality section is currently slightly vague. For instance, the reader should not be required to consult Riccio et al. 2013 to obtain information on what "feedback" constitutes. Also, perhaps I am missing something, but I am slightly confused as to why pseudo-real datasets involving superimposition of a selected P300 waveform template are being constructed here. If this is a replication of Riccio, doesn't the original stream itself contains naturally occurring P300s which the authors could contrast the detection performance of their selected classification methods of choice on?

3) As indicated in my initial summary section, the value of this work to the clinical community remains unclear. The value of BCI in clinical work has a long and well-established history (e.g., Neuper et al. 2003). The reference cited in support of this in the response document (Chavarriaga et al., 2017) focuses on the challenges of long-term independent BCI use by intended end users which the current study does not seem to focus on.

\subsection*{\ovalbox{General Comments}}
The authors have addressed many of my concerns regarding this manuscript. There are however a few points that could benefit from clarification. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We thank the Reviewer for

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
1) While the authors have addressed the issue of having just a single null-EEG stream from a single subject, it appears to be the case still that a single P300 ERP template is being used. This would seem to either assuming that the P300 is generally invariant across trials (which is unlikely) or that the result of this assessment is restricted to the case in which the ERP response of interest does not vary between trials. This should at least be discussed as a limitation in the conclusion section of the manuscript. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We apologize for this mistake.  We definitely selected an incorrect wording because the message that we aimed to convey was exactly the opposite.  We have modified the abstract to emphasize the message and the idea that we failed to transmit in our original manuscript.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2) The write-up on the active modality section is currently slightly vague. For instance, the reader should not be required to consult Riccio et al. 2013 to obtain information on what "feedback" constitutes. Also, perhaps I am missing something, but I am slightly confused as to why pseudo-real datasets involving superimposition of a selected P300 waveform template are being constructed here. If this is a replication of Riccio, doesn't the original stream itself contains naturally occurring P300s which the authors could contrast the detection performance of their selected classification methods of choice on?

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
These important references were added to the manuscript.  Thank you for pointing them out.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3) As indicated in my initial summary section, the value of this work to the clinical community remains unclear. The value of BCI in clinical work has a long and well-established history (e.g., Neuper et al. 2003). The reference cited in support of this in the response document (Chavarriaga et al., 2017) focuses on the challenges of long-term independent BCI use by intended end users which the current study does not seem to focus on.

\vskip+1ex
\noindent \dotfill
\vskip+1ex

\bibliographystyle{mdpi}
\bibliography{brainscience}

\end{document}

### 2018-11-07 BrainScience/response.letter.3.tex
Adding modifications per Reviewers comments

@@
 
-We are grateful to the reviewer for pointing out relevant issues in our manuscript.
+We are grateful to the reviewer for giving us very valuable information about issues in our manuscript.
 
@@
 
-We thank the Reviewer for
+We thank again the Reviewer for the detailed information.
 
@@
 {\color{blue}
-We apologize for this mistake.
-We definitely selected an incorrect wording because the message that we aimed to convey was exactly the opposite.
-We have modified the abstract to emphasize the message and the idea that we failed to transmit in our original manuscript.
+
+We have modified the experiments to use all the available templates.
+On the passive-modality, we are using 70 averaged templates obtained from Subject Number 8 of the BNCI-Horizon dataset (Riccio et al).
+They are randomly selected and they are superimposed on the EEG stream.
+
+For the active-modality, first all 70 templates are extracted from each subject on an offline averaging procedure.
+These 70 templates are superimposed to a random non-hit preprocessed segment (see next issue).
+
+We agree with the Reviewer that our initial approach was naively optimistic in terms of adding just one template to the whole set was naively optimistic.
+
 }
@@
 {\color{blue}
-These important references were added to the manuscript.  Thank you for pointing them out.
+
+We have modified the Section 2.6 to clarify in detail the procedure.
+We are using a pseudo-real dataset to have certainty of the existence of the P300 response, which is generating by a subject and we are not sure if the response is there or not.
+
 }
@@
 \vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+Thank you very much for this comment.
+It pointed out something that we stated incorrectly and which required more thoughtful write-up.
+We refer to the approach suggested by Chavarriaga et al 2017 is the UCD, User-Centered Design, where they state that "close interaction between family caregivers and medical staff tremendously facilitates the porocess of transferring an AT to the end-user's home as not only can the needs and requirements of the end user can be taken into account but also the device can be adjusted to the needs and requiremens twho support the end user.
+A multidisciplinary approach would allow for improvement of end-user training, for example, by taking into account expert opinion on how to create learning paradigms that promote success but also facilitate BCI use for the end-user." (page 4).
+
+The message that we aimed to transmit, and that we hope this new version of the manuscript now incorporates, is that a waveform-based BCI system, which has an inherent intelligible property (REF), will foster a collaboration in a multidisciplinary environment and will ease the acceptance, and the usability and usefulness of a BCI device.
+The reason being, for caregivers and medical staff, particularly those with the traditional knowledge of the clinical EEG which is based on waveforms, they will feel more natural the understanding of how the system is detecting signals.
+
+We have modified the 2.6 section and the Conclusions.
+}
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
 

### 2018-11-07 BrainScience/brainscience.tex
Adding modifications per Reviewers comments

@@
 \usepackage{subfigure}
+
+\subfigcapmargin = 4pt
 
@@
 
-\noindent with $k$ representing the index of the selected atom $g_{k}$ and $\left\lVert \cdot \right\rVert$ its euclidean norm.
+\noindent with $k$ representing the index of the selected atom $g_{k}$ and ${\left\lVert \cdot \right\rVert}_{2}$ its euclidean norm.
 Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
@@
 This sequence is of order $m$, which is the number of sample points used to derive the ordinal element called $\pi$.
-There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a strict order relationship.
-For example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(3,1,2)$.
+There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a decreasing order relationship~\citep{Keller2017}.
+For example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(1,3,2)$.
 Thus, along the signal window there can be at most $k$ different ordinal (and overlapping) elements $\pi_{s}$
@@
 
+The problem with any P300 experiment is that users are part of a closed loop with the BCI system and the human is typically considered as a static compliant entity that always performs what the experimenter asks for in a precise and consistent way.
+Therefore, P300 experiments are often mined with null-signals.
+These are EEG streams which are marked as hit but, because the subject was not particularly focused, or concentrated or so on, the real signal is not generated.
+This lack of certainty may be in detriment of analysis of algorithms and can be misleading.
+Similar experiments had address this same issue, particularly when benchmarking different algorithms.
+
 First for a passive modality, a real P300 ERP template, obtained from a public dataset, is superimposed into a distinct EEG stream.
@@
 
+\subsubsection{Experiments}
 The experiments are as follows:
@@
 \subsubsection{Pseudo-Real Dataset Generation}
-
-BCI researcers seem to ignore the fact that these systems consist of a closed loop between the human and the mahine.
-the human is typically considered as a static compliant entity that always performs what the experimenters asks for in a precise and consistent way.
 
@@
 \centering
-\subfigure[EEG trace of the original signal.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
-\subfigure[The same eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
-\subfigure[EEG sample of Cz and L channel of the original EEG trace.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
+\subfigure[EEG trace of the original signal.
+The horizontal axis represents $10$ seconds of the EEG stream, from the $28$th second up to the $38$th.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
+\subfigure[The same $10$-seconds eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
+\subfigure[EEG sample of Cz and L channel of the original EEG trace. Only $1$ second is shown here, at the $31$th second.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
 \subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
-\caption{Eight-channel EEG signal for Subject Number 1 of the psuedo-real dataset without and with the superimposed ERP Template.
+\caption{Eight-channel EEG signal for Subject Number 1 of the pseudo-real dataset without and with the superimposed ERP Template.
 The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented.
@@
 \caption{Point-to-point averaged signals.
-These are extracted from the first letter identification trial of the Subject 1 of the psuedo-real dataset.
+These are extracted from the first letter identification trial of the Subject 1 of the pseudo-real dataset.
 The ERP is superimposed on classes 3 and 9.
@@
 \includegraphics[width=15cm]{images/dictionary.eps}
-\caption{Coherently averaged signals containing the superimposed ERP.
+\caption{Coherently averaged signals segments of $1$-second length containing the superimposed ERP.
+Vertical axis unit is $\mu V$.
 Each one is extracted from the EEG signal of the Subject 1 of the pseudo-real dataset.
@@
 
-Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.
+Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Neuper2003,Kathner2017}.
 When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.
@@
 Which algorithm perform better at this even when segment latency variations and component amplitudes issues are present ?
+
+waveform-based BCI system, which has an inherent intelligible property (REF), will foster a collaboration in a multidisciplinary environment and will ease the acceptance, and the usability and usefulness of a BCI device.
+The reason being, for caregivers and medical staff, particularly those with the traditional knowledge of the clinical EEG which is based on waveforms, they will feel more natural the understanding of how the system is detecting signals.
 

### 2018-11-07 BrainScience/brainscience.tex
Adding modifications per Reviewers comments

@@
 
-The problem with any P300 experiment is that users are part of a closed loop with the BCI system and the human is typically considered as a static compliant entity that always performs what the experimenter asks for in a precise and consistent way.
+The problem with any P300 experiment is that users are part of a closed loop with the BCI system and the human is typically considered as a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.
 Therefore, P300 experiments are often mined with null-signals.

### 2018-11-08 BrainScience/brainscience.tex
Write-up active and passive modalities

--- 2018-11-07 BrainScience/brainscience.tex
@@
 \item Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.
-\item Transient: An isolated wave or pattern that is distinctly different from background activity. 
+\item Transient/Component: An isolated wave or pattern that is distinctly different from background activity. 
 \end{itemize}
@@
 Atlases and guidelines were created in order to help in the recognition of these complex patterns.
-Even Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .
-The clinical EEG research has also focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements} \citep{Schomer2010}.
-
-Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}.
-Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g.
+Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .
+The clinical EEG research has also focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements}~\citep{Schomer2010}.
+
+Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}.
+The different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}.
+Visual characterization includes the identification or classification of certain waveform components based on a subjective characterization (e.g.
 positive or negative peak polarity) or the location within the strip.
 It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g.
-sleep K-Complex is well characterized based on rates between positive vs negative amplitude) \citep{Uchida1999}.
+sleep K-Complex is well characterized based on rates between positive vs negative amplitude)~\citep{Uchida1999}.
 Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves, sleep spindles, polysplindles, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).
@@
 The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.
-The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges\citep{EEGIntro}.
-
-Waveform characterization is the method in which analysis has been performed for Event Related Potentials.
-Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.
+The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges~\citep{EEGIntro}.
+
+Waveform characterization is the method in which analysis has been performed for Event Related Potentials (ERP).
+These are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.
 ERPs are regularly used to assess auditory response in infants.
@@
 ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.
-The P300 is one of the most studied ERP component in BCI because it can be used to implement a Speller application.
+One of the most studied ERP is the P300, discovered in 1965 by Sutton, Braren, Zubin and John.
+This component is a positive deflection of a subject's EEG signal that arises when an unexpected and infrequent stimulus appears~\citep{Wolpaw2002}.
+The P300 is widely utilized in BCI because it can be harnessed to implement a Speller application.
 Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.
 
-Table~\ref{tab:methods} shows a list of depictions used to describe waveforms in the surveyed literature.
+Table~\ref{tab:methods} summarize a list of depictions used to describe waveforms in the surveyed literature.
 Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.
@@
 
-The search of methods based on waveforms is conducted by following the PRISMA~\citep{Moher2009} guidelines.
+The exploration of methods based on waveforms is conducted by following the PRISMA~\citep{Moher2009} guidelines.
 Search is performed on Google Scholar, Semantic Web and IEEE Xplore search engines by the terms "Waveforms" OR "Shape" OR "Morphology" OR "Visual inspection" + "EEG".
@@
 A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.
-This was even used in the seminal work of Farwell and Donchin on P300 \citep{Farwell1988,WolpawJonathanR2012}.
+This was even used in the seminal work of Farwell and Donchin on the P300 Speller~\citep{Farwell1988,WolpawJonathanR2012}.
 Additionally, a logarithmic graph of the peak-to-peak amplitude which is called amplitude integrated EEG (aEEG) \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.
 
-Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.
-Moreover, other approaches explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied.
-Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA) \citep{Uchida1996} algorithm are few of the earliest depictions where the idea of capturing the shape of the signal were established.
+Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms.
+In \cite{Klein1976} a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.
+Moreover, \citep{Yamaguchi2009} explored the use of Mathematical Morphology, where the time-domain structure of contractions and dilations were studied.
+Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA)~\citep{Uchida1996} algorithm are few of the earliest depictions where the idea of capturing the shape of the signal were established.
 
@@
 This region of the image, where the signal's most important salient waveform should be located, is divided in a grid of $4 x 4$ block and the bidimensional gradient vectors are calculated on each one of them.
-Therefore, for each block within the patch, a histogram of the gradient orientations, 8 circular orientations, are calculated.
+Therefore, for each block $(i,j)$ within the patch, a histogram $h(i,j,\theta)$ of the gradient orientations, for 8 circular orientations $\theta$, are calculated.
 This histogram is concatenated for all the 16 blocks and a feature is thus formed:
@@
 
-\noindent with $i$ and $j$ varying belonging to $I = \{ 0,1,2,3 \} $ localizing the 16 blocks within the grid.
+\noindent with $i$ and $j$ belonging to $I = \{ 0,1,2,3 \} $ and localizing the 16 blocks within the grid.
 The angles $\theta$ that belong to $\Theta$ are the eight possible equidistant values between $0$ and $315$.
@@
 
-Farwell and Donchin P300 Speller \citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
-This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
-This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.
+\begin{itemize}
+\item EEG Stream: the raw EEG streams for both modalities
+\item Templates: P300 templates obtained either from the Subject Number 8 or from the same subject.
+\item Pseudo-real dataset
+\item null-signals
+\end{itemize}
 
@@
 This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.
-
-The problem with any P300 experiment is that users are part of a closed loop with the BCI system and the human is typically considered as a static compliant entity that always performs what the experimenter asks for in a precise and consistent way.
-Therefore, P300 experiments are often mined with null-signals.
-These are EEG streams which are marked as hit but, because the subject was not particularly focused, or concentrated or so on, the real signal is not generated.
-This lack of certainty may be in detriment of analysis of algorithms and can be misleading.
+The experiments are performed by offline evaluation of the performance on a P300-Based BCI Speller application.
+
+Farwell and Donchin P300 Speller~\citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
+This procedure exploits a cognitive phenomena raised by the oddball paradigm~\citep{Luck2005}: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
+This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.
+
+The problem with any P300 experiment is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.
+Therefore, P300 experiments are often mined with \textit{null-signals}.
+These are EEG streams which are marked as hit but, because the subject was not particularly focused, or concentrated or so on, and the real signal is not generated.
+This lack of certainty may be in detriment of analysis of algorithms and can be misleading or difficult to dealt with.
 Similar experiments had address this same issue, particularly when benchmarking different algorithms.
+
+\subsubsection{Passive Modality}
 
@@
 
+In this case the template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.
+This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.
+This template is selected due to its shape which more closely resemble the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
+
+After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha
+
+\subsubsection{Active Modality}
+
 Second, an active-modality is also used, where a P300-Based BCI experiment is performed on four subjects, but the signal segments are modified to produce the desired P300 response, but based on information extracted from the same subject.
+In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.
+The most prototypical P300 waveform template is selected by visual observation.
+This template is superimposed to each segment marked as \textit{True} on the EEG stream.
+The baseline of each segment, on the other hand, is obtained by picking randomly any \textit{False} segment for the same subject.
 
@@
 \item Experiment 1 - Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.
-\item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERP over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
+The pool of $70$ P300 ERP waveforms, either obtained from the same subject in the passive-modality or from each subject in the active-modality are used to seed the EEG streams.
+The templates are randomly selected.
+\item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERPs over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
 The lagging value is picked from a uniform distribution $U(0,0.4)$ [s] ranging from $0$ to $0.4$ of the $1s$ segment size~\citep{DaPelo2018}.
@@
 
-\subsubsection{Pseudo-Real Dataset Generation}
-
-\paragraph{Passive Modality}
-
-In this case the template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
-Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.
-This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.
-This template is selected due to its shape which more closely resemble the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
+\subsubsection{EEG Stream Generation}
 
 An EEG stream with null-P300 signal is obtained by the following procedure:
-Four (4) participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
+
+Eight (8) healthy participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
 No monetary compensation is handed out and they agree and sign a written informed consent.
@@
 
-All participants are instructed to passively watch the flashing screen while not focusing on any particular letter.
+The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.
+Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.
+This yields exactly $120$ intensifications of rows and columns per trial.
+The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.
+This provides a $4$ Hz frequency of flashes on the screen.
+The initial pause and the inter-trial pauses are set to $20$ seconds.
+The whole experiment lasts for around $1400$ seconds.
+
+Four out of the eight participants are instructed to passively watch the flashing screen while not focusing on any particular letter.
+They do not receive any extra information on the screen.
 None of them have any experience with a BCI device.
 A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.
+
+The remaining four participants perform a copy-spelling task where the computer monitor suggests by highlighting the target letter, which letter the subject needs to focus.
+During each selection letter trial, the target letter is suggested at the bottom of the screen.
 
@@
 %The averaging procedure aims to extract the superimposed ERP Template and cancel out everything else on each segment.
-
-\paragraph{Active modality}
-
-The experiment conducted by~\citep{Riccio2013} is replicated on 4 healthy subjects, performing a copy-spelling task with feedback.
-In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.
-The most prototypical P300 waveform template is selected by visual observation.
-This template is superimposed to each segment marked as \textit{True} on the EEG stream.
-The baseline of each segment, on the other hand, is obtained by picking randomly any \textit{False} segment for the same subject.
 

### 2018-11-08 BrainScience/response.letter.3.tex
More changes

@@
 
-We thank again the Reviewer for the detailed information.
+We thank again the Reviewer for all the detailed information provided on this fruitful Review.
 
@@
 We have modified the experiments to use all the available templates.
-On the passive-modality, we are using 70 averaged templates obtained from Subject Number 8 of the BNCI-Horizon dataset (Riccio et al).
+On the passive-modality, we are using now 70 averaged templates obtained from Subject Number 8 of the BNCI-Horizon dataset (Riccio et al).
 They are randomly selected and they are superimposed on the EEG stream.
+This subject was selected due to the fact that the waveforms of their averaged P300 responses have a very distinctive and prototypical shape.
 
-For the active-modality, first all 70 templates are extracted from each subject on an offline averaging procedure.
+For the active-modality, first all 70 templates are extracted from each subject by an offline averaging procedure.
 These 70 templates are superimposed to a random non-hit preprocessed segment (see next issue).
-
-We agree with the Reviewer that our initial approach was naively optimistic in terms of adding just one template to the whole set was naively optimistic.
 
@@
 We are using a pseudo-real dataset to have certainty of the existence of the P300 response, which is generating by a subject and we are not sure if the response is there or not.
+This is the reason we followed a similar procedure performed by Varlaager in REF while trying to analyze different alternatives of alignment of P3 trials.
+But instead of using a costumized completely artificial P300 waveform we used one obtained from a public dataset of a real person, and from a particular group which is targetted by BCI applications.
 

### 2018-11-08 BrainScience/brainscience.tex
More changes

@@
 
-\begin{itemize}
-\item EEG Stream: the raw EEG streams for both modalities
-\item Templates: P300 templates obtained either from the Subject Number 8 or from the same subject.
-\item Pseudo-real dataset
-\item null-signals
-\end{itemize}
+%
+%\begin{itemize}
+%\item EEG Stream: the raw EEG streams for both modalities
+%\item Templates: P300 templates obtained either from the Subject Number 8 or from the same subject.
+%\item Pseudo-real dataset
+%\item null-signals
+%\end{itemize}
+%
 
@@
 This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.
-The experiments are performed by offline evaluation of the performance on a P300-Based BCI Speller application.
+The experiments are performed by the offline evaluation of the performance on a P300-Based BCI Speller application.
 
@@
 
-The problem with any P300 experiment is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.
+A problem with any P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.
 Therefore, P300 experiments are often mined with \textit{null-signals}.
-These are EEG streams which are marked as hit but, because the subject was not particularly focused, or concentrated or so on, and the real signal is not generated.
-This lack of certainty may be in detriment of analysis of algorithms and can be misleading or difficult to dealt with.
-Similar experiments had address this same issue, particularly when benchmarking different algorithms.
+These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the real signal is not generated.
+This lack of certainty may be in detriment of analysis of algorithms and can be misleading or difficult to deal with.
+Similar experiments had addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
+
+To solve this issue a pseudo-real dataset is generated under two different modalities.  A passive modality and an active modality.
 
@@
 
-First for a passive modality, a real P300 ERP template, obtained from a public dataset, is superimposed into a distinct EEG stream.
-This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on any letter in particular.
+First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into a distinct EEG stream.
+A set of template ERPs are acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+On the other hand, the EEG traces where these templates are superimposed, are experimentally obtained by subjects which are observing the flashing of the stimulus matrix during a P300 Speller procedure but they do not engage in focusing on any letter in particular.
 Everything is there, except the P300 ERP component.
-Along the EEG stream, the markers information was used to localize the \textit{True} segments where the P300 should have been found, and those timing locations are used to superimpose the extracted ERP waveform.
-By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to similar procedures used in other works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
-
-In this case the template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
-Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.
-This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.
-This template is selected due to its shape which more closely resemble the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
+Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.
+By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential.
+
+A sample P300 ERP obtained from the trial number $2$ of Subject 8 can be seen in Figure  \ref{fig:erptemplate1}.
+The templates obtained by this processing this Subject's EEG signals are selected due to their shapes more closely resembling the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
+Segments are extracted from the original dataset, and they are point-to-point coherently averaged.
 

### 2018-11-08 BrainScience/brainscience.tex
Toques

@@
 Therefore, P300 experiments are often mined with \textit{null-signals}.
-These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the real signal is not generated.
-This lack of certainty may be in detriment of analysis of algorithms and can be misleading or difficult to deal with.
-Similar experiments had addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
-
-To solve this issue a pseudo-real dataset is generated under two different modalities.  A passive modality and an active modality.
+These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the expected signal component is not generated.
+This lack of certainty may be in detriment of any conducted analysis and can be misleading or difficult to deal with.
+Previous works have addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
+
+In order to tackle this problem, a pseudo-real dataset based on an EEG stream is generated under two different modalities.  A passive modality and an active modality.
+
+\subsubsection{EEG Stream Generation}
+\label{EEGStream}
+
+Eight (8) healthy participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
+No monetary compensation is handed out and they agree and sign a written informed consent.
+This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.
+The participants are healthy and have normal or corrected-to-normal vision and no history of neurological disorders.
+These voluntary subjects are aged between 20-40 years old.
+EEG data is collected in a single recording session.
+Each subject is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.
+The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.
+Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.
+Reference is set to the right ear lobe and ground is preset as the AFz position.
+Sampling frequency is set to 250 Hz.
+
+The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.
+Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.
+This yields exactly $120$ intensifications of rows and columns per trial.
+The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.
+This provides a $4$ Hz frequency of flashes on the screen.
+The initial pause and the inter-trial pauses are set to $20$ seconds.
+The whole experiment lasts for around $1400$ seconds.
+This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as True and the remaining $700$ as False.
+
+Four out of the eight participants are instructed to passively watch the flashing screen while not focusing on any particular letter.
+They do not receive any extra information on the screen.
+None of them have any experience with a BCI device.
+A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.
+
+The remaining four participants perform a copy-spelling task where the computer monitor highlights the target letter, which is the one that the subject needs to focus.
+Across the duration of the trial, the current target letter is informed at the bottom of the screen.
 
@@
 A set of template ERPs are acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+The experimental protocol implemented to produce this dataset is the same as the one described in Section~\ref{EEGStream}.
 On the other hand, the EEG traces where these templates are superimposed, are experimentally obtained by subjects which are observing the flashing of the stimulus matrix during a P300 Speller procedure but they do not engage in focusing on any letter in particular.
@@
 A sample P300 ERP obtained from the trial number $2$ of Subject 8 can be seen in Figure  \ref{fig:erptemplate1}.
-The templates obtained by this processing this Subject's EEG signals are selected due to their shapes more closely resembling the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
-Segments are extracted from the original dataset, and they are point-to-point coherently averaged.
-
-After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha
+These templates are selected due to their shapes more closely resembling the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
+They are produced by extracting segments for this subject and by point-to-point coherently average them.
+
+%After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha
 
@@
 
-Second, an active-modality is also used, where a P300-Based BCI experiment is performed on four subjects, but the signal segments are modified to produce the desired P300 response, but based on information extracted from the same subject.
-In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.
-The most prototypical P300 waveform template is selected by visual observation.
-This template is superimposed to each segment marked as \textit{True} on the EEG stream.
-The baseline of each segment, on the other hand, is obtained by picking randomly any \textit{False} segment for the same subject.
+Second, an active modality is also implemented, where a P300-Based BCI Speller experiment is performed on four subjects.
+For this scenario, the signal segments are modified to guarantee the inclusion of a P300 component.
+However, in this case the templates are extracted from the same subject.
+Hence, the EEG signal is preprocessed and labeled segments are extracted.
+Segments labeled \textit{True} are coherently point-to-point averaged, and $70$ templates are produced from the whole set of $35$ trials.
+
+Once templates are procured, a random \textit{False} segment for the same subject is obtained.
+This is used as a baseline signal and is added to the template, conforming a new segment which contains one of the templates.
+This procedure continues until the $700$ segments marked as \textit{True} are completed.
+
+%The EEG stream contains $4200$ segments, which correspond to the preprocessed EEG data that was obtained coincidentally with the intensification of a row or column on the Speller Matrix.
 
@@
 This method has been proved efficient in decoding P300 in several BCI Competitions~\citep{Kaper2004}.
-
-\subsubsection{EEG Stream Generation}
-
-An EEG stream with null-P300 signal is obtained by the following procedure:
-
-Eight (8) healthy participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
-No monetary compensation is handed out and they agree and sign a written informed consent.
-This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.
-The participants are healthy and have normal or corrected-to-normal vision and no history of neurological disorders.
-These voluntary subjects are aged between 20-40 years old.
-EEG data is collected in a single recording session.
-Each subject is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.
-The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.
-Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.
-Reference is set to the right ear lobe and ground is preset as the AFz position.
-Sampling frequency is set to 250 Hz.
-
-The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.
-Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.
-This yields exactly $120$ intensifications of rows and columns per trial.
-The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.
-This provides a $4$ Hz frequency of flashes on the screen.
-The initial pause and the inter-trial pauses are set to $20$ seconds.
-The whole experiment lasts for around $1400$ seconds.
-
-Four out of the eight participants are instructed to passively watch the flashing screen while not focusing on any particular letter.
-They do not receive any extra information on the screen.
-None of them have any experience with a BCI device.
-A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.
-
-The remaining four participants perform a copy-spelling task where the computer monitor suggests by highlighting the target letter, which letter the subject needs to focus.
-During each selection letter trial, the target letter is suggested at the bottom of the screen.
 

### 2018-11-09 BrainScience/brainscience.tex
Update a new version.  This is the first draft with the general idea of what we were going to try to explain

@@
 
-A problem with any P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.
-Therefore, P300 experiments are often mined with \textit{null-signals}.
-These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the expected signal component is not generated.
+A problem with the information produced by a P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.
+Therefore, P300 experiments data is often mined with \textit{null-signals}.
+These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the expected signal element is not generated.
 This lack of certainty may be in detriment of any conducted analysis and can be misleading or difficult to deal with.
@@
 The whole experiment lasts for around $1400$ seconds.
-This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as True and the remaining $700$ as False.
+This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as \textit{True} and the remaining $700$ as \textit{False}.
 
@@
 
-First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into a distinct EEG stream.
-A set of template ERPs are acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into the generated EEG stream of 4 subjects.
+A set of template ERPs is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website~\citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
 The experimental protocol implemented to produce this dataset is the same as the one described in Section~\ref{EEGStream}.
@@
 Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.
-By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential.
+By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of the evoked potential.
 
@@
 %After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=12cm]{images/erptemplate1.eps}
+\caption{ERP Template obtained from the coherent point-to-point ensemble average from the signals of Subject Number Eight of the BNCI Horizon public dataset 008-2014.
+The template is $1s$ long which is 256 sample points, and the eight channels are superimposed with different colors.
+The P3b component can be seen around the sample index $150$ and $200$.}
+\label{fig:erptemplate1}
+\end{figure}
 
@@
 Once templates are procured, a random \textit{False} segment for the same subject is obtained.
-This is used as a baseline signal and is added to the template, conforming a new segment which contains one of the templates.
+This is used as a baseline signal and is added to the template, conforming a new segment which has a superimposed P300 template.
 This procedure continues until the $700$ segments marked as \textit{True} are completed.
@@
 %The EEG stream contains $4200$ segments, which correspond to the preprocessed EEG data that was obtained coincidentally with the intensification of a row or column on the Speller Matrix.
+
+Figure~\ref{fig:gains} shows a $5s$ sample of the EEG trace obtained with the MNE library~\citep{Gramfort2013}.
+Channel $S$ represents the twelve different stimulus markers (columns or rows) while channel $L$ represent the label (\textit{True} vs \textit{False}).
+Labels are represented by square signals.
+\textit{False} segments are marked with single amplitude square signals while \textit{True} segments are identified by double-amplitude square signals.
+Subfigure (a) shows the signals before the ERP template is superimposed while subfigure (b) shows the same signals with the superimposed ERP template.
+At first-sight, differences are really hard to spot visually.
+Subfigures (c) and (d) show only one second of channels Cz and L from the same segment.
+The superimposed ERP can be devised enclosed by the vertical bars, around $31.5$s, where in (d) the peak is slightly bigger.
+Figure~\ref{fig:gaincheck} shows the obtained ensemble average ERPs as result of superimposing the template signal into the EEG stream, time-locked to the stimulus onset.
+These 12 point-to-point averaged segments correspond to the first trial of the EEG stream.
+
+\begin{figure}[htb]
+\centering
+\subfigure[EEG trace of the original signal.
+The horizontal axis represents $10$ seconds of the EEG stream, from the $28$th second up to the $38$th.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
+\subfigure[The same $10$-seconds eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
+\subfigure[EEG sample of Cz and L channel of the original EEG trace. Only $1$ second is shown here, at the $31$th second.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
+\subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
+\caption{Eight-channel EEG signal for Subject Number 1 of the pseudo-real dataset without and with the superimposed ERP Template.
+The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented.
+On (c) and (d) the small variation that was introduced by the superimposition of the ERP can be seen enclosed by the vertical bars, where the slope of the bump on subfigure (d) is slightly bigger}
+\label{fig:gains}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=1.0\linewidth]{images/GainCheck.eps}
+\caption{Point-to-point averaged signals.
+These are extracted from the first letter identification trial of the Subject 1 of the pseudo-real dataset.
+The ERP is superimposed on classes 3 and 9.
+Class 3 is obtained while averaging the segments where the row of the speller matrix is intensified whereas class 9 is calculated from the intensification of the corresponding column.}
+\label{fig:gaincheck}
+\end{figure}
 
@@
 
-Finally the performance at letter identifications for these same methods is evaluated by performing an offline BCI Simulation on the Dataset IIb of the BCI Competition II (2003)~\citep{Blankertz2002}.
+Finally the performance at letter identifications for these same methods is evaluated by running an offline BCI Simulation on the Dataset IIb of the BCI Competition II (2003)~\citep{Blankertz2002}.
 The protocol of this dataset is very similar to what was used to obtain the pseudo-real dataset.
@@
 More details can be found on the published work of \citep{Riccio2013}.
-
-Figure~\ref{fig:gains} shows a $5s$ sample of the EEG trace obtained with the MNE library~\citep{Gramfort2013}.
-Channel $S$ represents the twelve different stimulus markers (columns or rows) while channel $L$ represent the label (\textit{True} vs \textit{False}).
-Labels are represented by square signals.
-\textit{False} segments are marked with single amplitude square signals while \textit{True} segments are identified by double-amplitude square signals.
-Subfigure (a) shows the signals before the ERP template is superimposed while subfigure (b) shows the same signals with the superimposed ERP template.
-At first-sight, differences are really hard to spot visually.
-Subfigures (c) and (d) show only one second of channels Cz and L from the same segment.
-The superimposed ERP can be devised enclosed by the vertical bars, around $31.5$s, where in (d) the peak is slightly bigger.
-Figure~\ref{fig:gaincheck} shows the obtained ensemble average ERPs as result of superimposing the template signal into the EEG stream, time-locked to the stimulus onset.
-These 12 point-to-point averaged segments correspond to the first trial of the EEG stream.
-
-%The original signal-to-noise ratio was calculated as Hue 2010.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=12cm]{images/erptemplate1.eps}
-\caption{ERP Template obtained from the coherent point-to-point ensemble average from the signals of Subject Number Eight of the BNCI Horizon public dataset 008-2014.
-The template is $1s$ long which is 256 sample points, and the eight channels are superimposed with different colors.
-The P3b component can be seen around the sample index $150$ and $200$.}
-\label{fig:erptemplate1}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\subfigure[EEG trace of the original signal.
-The horizontal axis represents $10$ seconds of the EEG stream, from the $28$th second up to the $38$th.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
-\subfigure[The same $10$-seconds eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
-\subfigure[EEG sample of Cz and L channel of the original EEG trace. Only $1$ second is shown here, at the $31$th second.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
-\subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
-\caption{Eight-channel EEG signal for Subject Number 1 of the pseudo-real dataset without and with the superimposed ERP Template.
-The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented.
-On (c) and (d) the small variation that was introduced by the superimposition of the ERP can be seen enclosed by the vertical bars, where the slope of the bump on subfigure (d) is slightly bigger}
-\label{fig:gains}
-\end{figure}
-
-%\begin{figure}[H]
-%\centering
-%\includegraphics[width=12cm]{images/nogain.eps}
-%\includegraphics[width=12cm]{images/singlegain.eps}
-%\caption{Eight-channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the added template.}
-%\label{fig:doubleandtriplegain}
-%\end{figure}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=1.0\linewidth]{images/GainCheck.eps}
-\caption{Point-to-point averaged signals.
-These are extracted from the first letter identification trial of the Subject 1 of the pseudo-real dataset.
-The ERP is superimposed on classes 3 and 9.
-Class 3 is obtained while averaging the segments where the row of the speller matrix is intensified whereas class 9 is calculated from the intensification of the corresponding column.}
-\label{fig:gaincheck}
-\end{figure}
 
@@
 As previously described the data is continuously divided in two sets, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance.
-This is repeated one hundred times, and performances averaged.
+This is repeated one ten times, and performances averaged.
 Figure~\ref{fig:1-1} show the results for the passive modality while Figure~\ref{fig:2-1} shows the results for the active modality.
@@
 \includegraphics[width=15cm]{images/1-2.eps}
-\caption{\textbf{Passive Modality - Experiment 2}: Performance curves for four subjects for the five algorithms while a random latency is included while superimposed the P300 signal template.}
+\caption{\textbf{Passive Modality - Experiment 2}: Performance curves for four subjects for the five algorithms when a random latency is included when superimposing the P300 signal template.}
 \label{fig:1-2}
@@
 \includegraphics[width=15cm]{images/2-2.eps}
-\caption{\textbf{Active Modality - Experiment 2}: Performance curves for four subjects for the five algorithms while a random latency is included while superimposed the P300 signal template.}
+\caption{\textbf{Active Modality - Experiment 2}: Performance curves for the four subjects for the five algorithms.  A random latency is included while superimposing the P300 signal template.}
 \label{fig:2-2}
@@
 \midrule
-MP 1 & FC2  & $50\%$ \\
-MP 2 & CPz & $22\%$ \\
-SIFT  & Cz & $67\%$ \\
-PE     & PO8 & $22\%$ \\
+MP 1 & Cz  & $50\%$ \\
+MP 2 & FC1 & $22\%$ \\
+SIFT  & FC1 & $67\%$ \\
+PE     & CP1 & $22\%$ \\
 SHCC & Cz & $61\%$ \\
@@
 When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.
-Results for this competition have shown perfect classification with tailored algorithm.
+Results for this competition have shown perfect classification with tailored algorithms~\citep{Kundu2018}.
 This level is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.
@@
 
-waveform-based BCI system, which has an inherent intelligible property (REF), will foster a collaboration in a multidisciplinary environment and will ease the acceptance, and the usability and usefulness of a BCI device.
-The reason being, for caregivers and medical staff, particularly those with the traditional knowledge of the clinical EEG which is based on waveforms, they will feel more natural the understanding of how the system is detecting signals.
-
-One of the main goals of the BCI discipline is to provide assistance to patients and to provide alternative tools to be used in diagnostics and rehabilitation procedure.
+%One of the main goals of the BCI discipline is to provide assistance to patients and to provide alternative tools to be used in diagnostics and rehabilitation procedure.
 This requires a clinical focus which is often neglected in BCI research~\citep{Chavarriaga2017}.
@@
 We believe that the adoption of a \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an inherent intelligible property~\citep{j2018challenge} that can be mapped to existing procedures, and above all, can maintain the clinician trust on the system behavior is beneficial to Clinical Practice, Neuroscience and BCI research.
+Additionally, this may foster collaboration in a multidisciplinary environment and may ease the acceptance and translation of BCI technology~\citep{Chavarriaga2017}.
+The reason being, for caregivers and medical staff, particularly those with the expertise of the clinical EEG which is based on waveforms, they may feel a natural understanding of how the system is performing.
 

### 2018-11-09 BrainScience/response.letter.3.tex
Submitted Version. Round 2

--- 2018-11-07 BrainScience/response.letter.3.tex
@@
 
-We thank the Reviewer for
+We thank again the Reviewer for all the detailed information provided on this fruitful Review.
 
@@
 {\color{blue}
-We apologize for this mistake.
-We definitely selected an incorrect wording because the message that we aimed to convey was exactly the opposite.
-We have modified the abstract to emphasize the message and the idea that we failed to transmit in our original manuscript.
+
+We have modified the experiments to use all the available templates.
+
+On the passive-modality, we are using now 70 averaged templates obtained from the 35 trials produced by Subject Number 8 of the BNCI-Horizon dataset\citep{Riccio2013}.
+They are randomly selected and they are superimposed on the EEG stream that we produced.
+This subject was selected due to the fact that the waveforms of their averaged P300 responses have a very distinctive and prototypical shape.
+
+Regarding the active-modality, first all 70 templates are extracted from each subject by an offline averaging procedure and from the segments marked as hits.
+These 70 templates are superimposed to a random non-hit preprocessed segment (see next issue) to create new segments that are guaranteed to have the artificial P300 response.
+
+We have modified completely Section 2.6 (Experimental Protocol) to provide details of these procedures.
+
 }
@@
 {\color{blue}
-These important references were added to the manuscript.  Thank you for pointing them out.
+
+We are using a pseudo-real dataset to have certainty of the existence of the P300 response, which is generated by a subject and we are not sure if the response is there or not.
+This is the reason we followed a similar procedure implemented by~\citep{Jaskowski2000} to analyze different alignment strategies of P3 trials.
+Instead of using a completely artificial P300 waveform, we opted to use templates obtained from a public dataset of a real person and which, at the same time, have a very prototypical waveform.
+Based on Reviewer suggestion we understood that the idea could be extended to use also the templates from the same subjects that performed a P300-Based BCI Speller experiment.
+
+We have modified the Section 2.6 (Experimental Protocol) to clarify in detail the active modality.
+We also modified the same section to emphasize the reason we opted for creating a pseudo-real dataset.
+We hope it is now more clearly explained.
+
 }
@@
 \vskip+1ex
+%
+\begin{quotation}
+{\color{blue}
+Thank you very much for this comment.  It pointed out something that we stated incorrectly and which required a more thoughtful write-up.
+
+We wanted to refer more specifically, to the recommendation suggested by \cite{Chavarriaga2017} to use the \textit{UCD, User-Centered Design}.
+In this work, authors state that "close interaction between family caregivers and medical staff tremendously facilitates the process of transferring an AT to the end-user's home as not only can the needs and requirements of the end user be taken into account but also the device can be adjusted to the needs and requirements of those who support the end user.
+A multidisciplinary approach would allow for improvement of end-user training, for example, by taking into account expert opinion on how to create learning paradigms that promote success but also facilitate BCI use for the end-user." (page 4).
+
+The message that we aimed to transmit, and that we hope this new version of the manuscript now incorporates, is that a waveform-based BCI system has an inherent intelligible property~\citep{j2018challenge}.
+We mean the ability of the system to emphasize clearly and noticeable what are the factors that caused the system action, decision or classification, in this case the waveform that can be visually inspected.
+We think such a system will help to foster collaboration in a multidisciplinary environment and will ease the acceptance, the usability and usefulness of a BCI device.
+The reason being, for caregivers and medical staff, particularly those with the traditional knowledge of the clinical EEG which is based on waveforms,  will feel more natural the understanding of how the system is detecting and analyzing signals.
+
+We have modified the Conclusions section to emphasize our point.
+}
+\end{quotation}
+%
+\vskip+1ex
+\noindent \dotfill
+\vskip+1ex
 

### 2018-11-09 BrainScience/brainscience.tex
Submitted Version. Round 2

@@
 However, clinical EEG is based intensively on waveforms and the structure of signal plots.
-Hence, the purpose of this work is to establish a bridge to fill this gap by reviewing and describing the procedures that have been used to detect patterns in the electroencephalographic waveforms, benchmarking them on a controlled pseudo-real dataset of an EEG Event Related Potentials waveform and verifying their performance on a public dataset of a BCI Competition.}
+Hence, the purpose of this work is to establish a bridge to fill this gap by reviewing and describing the procedures that have been used to detect patterns in the electroencephalographic waveforms, benchmarking them on a controlled pseudo-real dataset of a P300-Based BCI Speller and verifying their performance on a public dataset of a BCI Competition.}
 
@@
 % Keywords
-\keyword{electroencephalography; BCI; EEG; waveform; ERP ;P300}
+\keyword{electroencephalography; brain-computer interfaces; waveform;p300;SIFT;PE;MP;SHCC}
 
@@
 This article unfolds as follows: Section \ref{EEG} provides a brief introduction to EEG and the particularities of the EEG waveform characterization.
-Section \ref{Algorithms} explains the algorithms waveform-based algorithms that are analyzed.
+Section \ref{Algorithms} explains the waveform-based algorithms that are analyzed.
 In Section \ref{Experimental} the experimentation procedure is explained.
@@
 
-The Electroencephalography consists on the measurement of small variations of electrical currents over the scalp.
+The Electroencephalography consists on the measurement of small variations of electrical voltage over the scalp.
 It is one of the most widespread used methods to capture brain signals and was initially developed by Hans Berger in 1924 and has been extensively used for decades to diagnose neural diseases and other medical conditions.
@@
 Depictions may include subjective definitions of sharper, arch comb or wicket shape, rectangular, containing a decay phase or voltage rise, peaks and troughs, short term voltage change around each extrema in the raw trace.
-Derived ratios and indexes can be used as well like peak and trough sharpness ratio, symmetry between rise and decay phase and slope ratio (steepness of the rise period to that of the adjacent decay period).
+Derived ratios and indexes can be used as well, like peak and trough sharpness ratio, symmetry between rise and decay phase and slope ratio (steepness of the rise period to that of the adjacent decay period).
 For instance,  wording like "Central trough is sharper and more negative that the adjacent troughs"~\citep{Cole2017} are common in the literature.
@@
 The term suggest an increase in the number of neural elements contributing to the rhythm, or in the synchronization of different neurons with the same discharge pattern~\citep{Buzsaki2012}.
-\item Paroxysmal: Activity that emerges from background with a rapid onset, reaching (usually) quite high voltage and ending with an abrupt return to lower voltage activity. 
+\item Paroxysmal: Activity that emerges from background with a rapid onset, reaching frequently high voltage and ending with an abrupt return to lower voltage activity. 
 \item Monomorphic: Activity appearing to be composed of one dominant waveform pattern.
@@
 
-The traditional clinical approach consists in analyzing the paper strip that is generated by the plot of the signal obtained from the device.
+The traditional clinical approach to study electroencephalographic signals consists in analyzing the paper strip that is generated by the plot of the signal obtained from the device.
 Expert technician and physicians analyze visually the plots looking for specific patterns that may give a hint of the underlying cognitive process or pathology.
@@
 
-Table~\ref{tab:methods} summarize a list of depictions used to describe waveforms in the surveyed literature.
+Table~\ref{tab:methods} summarizes a list of depictions used to describe waveforms in the surveyed literature.
 Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.
@@
 These algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
-To maintain reproducibility, the data and the source code has been made available in the online repository of the Code Ocean platform under the name \textit{EEGWave}.
+To maintain reproducibility, the dataset described in Section~\ref{EEGStream} and the source code has been made available in the online repository of the Code Ocean platform under the name \textit{EEGWave}.
 
@@
 This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.
-The experiments are performed by the offline evaluation of the performance on a P300-Based BCI Speller application.
+The experiments are performed by the offline evaluation of the character identification rate of a Visual P300-Based BCI Speller application.
 
@@
 This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as \textit{True} and the remaining $700$ as \textit{False}.
+The extracted EEG signals are band-pass filtered using a 4th order Butterworth digital filter between $0.1$ and $10$ Hz and a $50$ Hz notch filter is applied to remove line AC noise.
+The EEG trace is finally downsampled to $16$ Hz.
+Segments of $1$-second length are extracted according to the markers information and those with variations larger than $70 \mu V$ are identified as artifacts and eliminated.
 
@@
 \item Experiment 1 - Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.
-The pool of $70$ P300 ERP waveforms, either obtained from the same subject in the passive-modality or from each subject in the active-modality are used to seed the EEG streams.
-The templates are randomly selected.
+The pool of $70$ P300 ERP waveforms, either obtained from the same subject in the passive-modality or from each subject in the active-modality are used to compose the artificial P300 wave in the pseudo-real dataset.
+Templates are randomly selected.
 \item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERPs over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
@@
 Each one is extracted from the EEG signal of the Subject 1 of the pseudo-real dataset.
-These averaged signals correspond to the 15 first trials (2 signals from each trial, one belonging to the column flashing and the other to row flashing).
+These averaged signals correspond to the 15 first trials (2 averaged signals from each trial, one belonging to the column flashing and the other to row flashing).
 These are the templates used to build a dictionary per channel per subject and are used by the classification algorithm described in \ref{section:classification}.}
@@
 Results for the first experiment are shown in Figures~\ref{fig:1-1}  and \ref{fig:2-1}.
-The performance while identifying each letter of the standard P300 Speller Matrix, and the channels where the best and worst performance is attained, is shown.
+The performance while identifying each letter of the standard P300 Speller Matrix, and the channels where the best and worst performance are attained, are shown.
 Each one represents the percentage of letters that is actually predicted by the algorithms using a cross-validation procedure.
 As previously described the data is continuously divided in two sets, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance.
-This is repeated one ten times, and performances averaged.
-Figure~\ref{fig:1-1} show the results for the passive modality while Figure~\ref{fig:2-1} shows the results for the active modality.
+This is repeated one hundred times, and performances averaged.
+Figure~\ref{fig:1-1} shows the results for the passive modality while Figure~\ref{fig:2-1} shows the results for the active modality.
 
@@
 
-We verified that similar performance results are obtained for the methods \textit{SHCC}, \textit{SIFT} and \textit{MP-1} and that it is possible to obtain discriminating information from the underlying signal based exclusively on an automated method of processing the waveforms.
-This brings the possibility to use these techniques to implement intelligible automatic detection procedures due to the fact that they are based on metrics which can be visually verified.
+The higher performance results are obtained for the methods \textit{SHCC} and \textit{SIFT} either on the pseudo-real dataset and on the BCI Competition.
+We verified that it is possible to obtain discriminating information from the underlying signal based exclusively on an automated method of processing the waveforms.
+This brings the possibility to use these techniques to implement intelligible~\citep{j2018challenge} automatic detection procedures, i.e.
+systems that are able to emphasize clearly and noticeable what are the factors that caused the system action, decision or classification.
+This is due to the fact that they are based on metrics which can be visually verified.
 
@@
 
-We believe that the adoption of a \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an inherent intelligible property~\citep{j2018challenge} that can be mapped to existing procedures, and above all, can maintain the clinician trust on the system behavior is beneficial to Clinical Practice, Neuroscience and BCI research.
+We believe that the adoption of a \textit{hybrid} methodology which can process the signal automatically, but at the same time, maintains an inherent intelligible property that can be mapped to existing procedures, and above all, can maintain the clinician trust on the system behavior is beneficial to Clinical Practice, Neuroscience and BCI research.
 Additionally, this may foster collaboration in a multidisciplinary environment and may ease the acceptance and translation of BCI technology~\citep{Chavarriaga2017}.
@@
 Another benefit of these methodologies is that they have a potential universal applicability.
-As they are only analyzing waveform, they can be explored in other disciplines where the structure or shape of the waveform is of relevance.
+As they are only analyzing waveforms, they can be explored in other disciplines where the structure or shape of the waveform is of relevance.
 Analyzing signals by their waveforms is relative common in chemical analysis~\citep{Skoog2000}, seismic analysis in Geology~\citep{Owens1984}, and quantitative financial analysis.
@@
 CNS: Central Nervous System\\
+AC: Alternating Current\\
 DC: Direct Current\\
 ERP: Event-Related Potential\\
-P300: Positive deflection of an Event-Related Potential which occurs 300 ms after onset of stimulus\\
+P300: Positive deflection at 300 ms\\
 ITR: Information Transfer Rate\\

### 2018-11-27 BrainScience/brainscience.old.tex
Final and published version

--- 2018-11-08 BrainScience/brainscience.tex
@@
 \usepackage{subfigure}
-
-\subfigcapmargin = 4pt
 
@@
 
-The clinical and historical tactic to analyze EEG signals is based on detecting visual patterns out of the EEG trace or polygraph~\citep{Hartman2005}: multichannel signals are extracted and continuously plotted over a piece of paper.
+The clinical and historical tactic to analyze EEG signals is based on detecting visual patterns out of the EEG trace or polygraph\citep{Hartman2005}: multichannel signals are extracted and continuously plotted over a piece of paper.
 Electroencephalographers or Electroencephalography technician decode and detect patterns along the signals by visually inspecting them~\citep{Schomer2010}.
@@
 
-This article unfolds as follows: Section \ref{EEG} provides a brief introduction to EEG and the particularities of the EEG waveform characterization.
-Section \ref{Algorithms} explains the algorithms waveform-based algorithms that are analyzed.
-In Section \ref{Experimental} the experimentation procedure is explained.
-Results are presented in Section \ref{section:results} and finally Discussion and Conclusions are expounded in the final sections.
+This article unfolds as follows: Section \ref{EEG} will provide a brief introduction to EEG and the particularities of the EEG waveform characterization.
+Section \ref{Algorithms} will explain the algorithms based on the waveform that will be analyzed.
+In Section \ref{Experimental} the experimentation procedure will be explained.
+Results will be presented in Section \ref{section:results} and finally Discussion and Conclusions will be established in the final sections.
 
@@
 \item Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.
-\item Transient/Component: An isolated wave or pattern that is distinctly different from background activity. 
+\item Transient: An isolated wave or pattern that is distinctly different from background activity. 
 \end{itemize}
@@
 Atlases and guidelines were created in order to help in the recognition of these complex patterns.
-Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .
-The clinical EEG research has also focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements}~\citep{Schomer2010}.
-
-Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}.
-The different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}.
-Visual characterization includes the identification or classification of certain waveform components based on a subjective characterization (e.g.
+Even Video-electroencephalography scalp recordings are routinely used as a diagnostic tools \citep{Giagante2003} .
+The clinical EEG research has also focused on temporal waveforms, and a whole branch of electrophenomenology has arisen around EEG \textit{graphoelements} \citep{Schomer2010}.
+
+Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}, where the different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}.
+Visual characterization includes the identification or classification of certain waveform components, or transient events, based on a subjective characterization (e.g.
 positive or negative peak polarity) or the location within the strip.
 It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g.
-sleep K-Complex is well characterized based on rates between positive vs negative amplitude)~\citep{Uchida1999}.
+sleep K-Complex is well characterized based on rates between positive vs negative amplitude) \citep{Uchida1999}.
 Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves,  sleep spindles, polysplindles, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).
@@
 The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.
-The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges~\citep{EEGIntro}.
-
-Waveform characterization is the method in which analysis has been performed for Event Related Potentials (ERP).
-These are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.
+The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges\citep{EEGIntro}.
+
+Waveform characterization is the method in which analysis has been performed for Event Related Potentials.
+Evoked Potentials (EPs) and Event Related Potentials (ERPs) are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.
 ERPs are regularly used to assess auditory response in infants.
@@
 ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.
-One of the most studied ERP is the P300, discovered in 1965 by Sutton, Braren, Zubin and John.
-This component is a positive deflection of a subject's EEG signal that arises when an unexpected and infrequent stimulus appears~\citep{Wolpaw2002}.
-The P300 is widely utilized in BCI because it can be harnessed to implement a Speller application.
+The P300 is one of the most studied ERP component in BCI because it can be used to implement a Speller application.
 Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.
 
-Table~\ref{tab:methods} summarize a list of depictions used to describe waveforms in the surveyed literature.
+Table~\ref{tab:methods} shows a list of depictions used to describe waveforms in the surveyed literature.
 Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.
@@
 
-The exploration of methods based on waveforms is conducted by following the PRISMA~\citep{Moher2009} guidelines.
+The search of methods based on waveforms is conducted by following the PRISMA~\citep{Moher2009} guidelines.
 Search is performed on Google Scholar, Semantic Web and IEEE Xplore search engines by the terms "Waveforms" OR "Shape" OR "Morphology" OR "Visual inspection" + "EEG".
@@
 A related method is used in \citep{Alvarado-Gonzalez2016} where the area under the curve of the EEG is sumarized to derive a feature.
-This was even used in the seminal work of Farwell and Donchin on the P300 Speller~\citep{Farwell1988,WolpawJonathanR2012}.
+This was even used in the seminal work of Farwell and Donchin on P300 \citep{Farwell1988,WolpawJonathanR2012}.
 Additionally, a logarithmic graph of the peak-to-peak amplitude which is called amplitude integrated EEG (aEEG) \citep{Shah2015} is used nowadays in Neonatal Intensive Care Units.
 
-Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms.
-In \cite{Klein1976} a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.
-Moreover, \citep{Yamaguchi2009} explored the use of Mathematical Morphology, where the time-domain structure of contractions and dilations were studied.
-Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA)~\citep{Uchida1996} algorithm are few of the earliest depictions where the idea of capturing the shape of the signal were established.
+Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.
+Moreover, other approaches explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied.
+Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA) \citep{Uchida1996} algorithm are few of the earliest depictions where the idea of capturing the shape of the signal were established.
 
@@
 
-\noindent with $k$ representing the index of the selected atom $g_{k}$ and ${\left\lVert \cdot \right\rVert}_{2}$ its euclidean norm.
+\noindent with $k$ representing the index of the selected atom $g_{k}$ and $\left\lVert \cdot \right\rVert$ its euclidean norm.
 Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
@@
 This sequence is of order $m$, which is the number of sample points used to derive the ordinal element called $\pi$.
-There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a decreasing order relationship~\citep{Keller2017}.
-For example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(1,3,2)$.
+There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a strict order relationship.
+For example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(3,1,2)$.
 Thus, along the signal window there can be at most $k$ different ordinal (and overlapping) elements $\pi_{s}$
@@
 This region of the image, where the signal's most important salient waveform should be located, is divided in a grid of $4 x 4$ block and the bidimensional gradient vectors are calculated on each one of them.
-Therefore, for each block $(i,j)$ within the patch, a histogram $h(i,j,\theta)$ of the gradient orientations, for 8 circular orientations $\theta$, are calculated.
+Therefore, for each block within the patch, a histogram of the gradient orientations, 8 circular orientations, are calculated.
 This histogram is concatenated for all the 16 blocks and a feature is thus formed:
@@
 
-\noindent with $i$ and $j$ belonging to $I = \{ 0,1,2,3 \} $ and localizing the 16 blocks within the grid.
+\noindent with $i$ and $j$ varying belonging to $I = \{ 0,1,2,3 \} $ localizing the 16 blocks within the grid.
 The angles $\theta$ that belong to $\Theta$ are the eight possible equidistant values between $0$ and $315$.
@@
 
-%
-%\begin{itemize}
-%\item EEG Stream: the raw EEG streams for both modalities
-%\item Templates: P300 templates obtained either from the Subject Number 8 or from the same subject.
-%\item Pseudo-real dataset
-%\item null-signals
-%\end{itemize}
-%
+Farwell and Donchin P300 Speller \citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
+This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
+This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.
 
@@
 This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.
-The experiments are performed by the offline evaluation of the performance on a P300-Based BCI Speller application.
-
-Farwell and Donchin P300 Speller~\citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
-This procedure exploits a cognitive phenomena raised by the oddball paradigm~\citep{Luck2005}: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
-This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.
-
-A problem with any P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.
-Therefore, P300 experiments are often mined with \textit{null-signals}.
-These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the expected signal component is not generated.
-This lack of certainty may be in detriment of any conducted analysis and can be misleading or difficult to deal with.
-Previous works have addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
-
-In order to tackle this problem, a pseudo-real dataset based on an EEG stream is generated under two different modalities.  A passive modality and an active modality.
-
-\subsubsection{EEG Stream Generation}
-\label{EEGStream}
-
-Eight (8) healthy participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
+
+First for a passive modality, a real P300 ERP template, obtained from a public dataset, is superimposed into a distinct EEG stream.
+This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on any letter in particular.
+Everything is there, except the P300 ERP component.
+Along the EEG stream, the markers information was used to localize the \textit{True} segments where the P300 should have been found, and those timing locations are used to superimpose the extracted ERP waveform.
+By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to similar procedures used in other works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.
+
+Second, an active-modality is also used, where a P300-Based BCI experiment is performed on four subjects, but the signal segments are modified to produce the desired P300 response, but based on information extracted from the same subject.
+
+The experiments are as follows:
+
+\begin{itemize}
+\item Experiment 1 - Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.
+\item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERP over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
+The lagging value is picked from a uniform distribution $U(0,0.4)$ [s] ranging from $0$ to $0.4$ of the $1s$ segment size~\citep{DaPelo2018}.
+\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.
+This component is defined to be located from the stimulus onset between $148$ ms up to $996$ ms which is around $840$ ms long.
+This waveform element, multiplied by a gain factor, is subtracted from the original template.
+This gain factor between $0$ and $1$ is drawn from a uniform distribution $U(0,1)$.
+Additionally this subtracted waveform is multiplied by a Gaussian window with a support of the same length~\citep{Harris1978}.
+This avoids adding any discontinuity into the artificial generated signal.
+\end{itemize}
+
+All these experiments are executed using cross validation procedure dividing the letter to spell in two sets, preserving the structure of the letter identification trials.
+Spelling letters are scrambled while the order and group of each intensification sequence is preserved.
+
+Finally the performance at letter identifications for these same methods is evaluated by performing an offline BCI Simulation on the Dataset IIb of the BCI Competition II (2003)~\citep{Blankertz2002}.
+The protocol of this dataset is very similar to what was used to obtain the pseudo-real dataset.
+The sampling frequency of this dataset is $240$, the number of letters are $73$ where the first $42$ are used to create the template dictionary for all the methods and the remaining $31$ are used to test the character recognition rate performance.
+Additionally, in this dataset the number of available intensification number sequences is $15$.
+The classification method Support Vector Machine SVM with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
+This method has been proved efficient in decoding P300 in several BCI Competitions~\citep{Kaper2004}.
+
+\subsubsection{Pseudo-Real Dataset Generation}
+
+\paragraph{Passive Modality}
+
+In this case the template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
+Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.
+This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.
+This template is selected due to its shape which more closely resemble the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
+
+An EEG stream with null-P300 signal is obtained by the following procedure: 
+Four (4) participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.
 No monetary compensation is handed out and they agree and sign a written informed consent.
@@
 
-The experimental protocol is composed of $35$ trials to spell $7$ words of $5$ letters each.
-Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.
-This yields exactly $120$ intensifications of rows and columns per trial.
-The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.
-This provides a $4$ Hz frequency of flashes on the screen.
-The initial pause and the inter-trial pauses are set to $20$ seconds.
-The whole experiment lasts for around $1400$ seconds.
-This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as True and the remaining $700$ as False.
-
-Four out of the eight participants are instructed to passively watch the flashing screen while not focusing on any particular letter.
-They do not receive any extra information on the screen.
+All participants are instructed to passively watch the flashing screen while not focusing on any particular letter.
 None of them have any experience with a BCI device.
 A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.
-
-The remaining four participants perform a copy-spelling task where the computer monitor highlights the target letter, which is the one that the subject needs to focus.
-Across the duration of the trial, the current target letter is informed at the bottom of the screen.
-
-\subsubsection{Passive Modality}
-
-First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into a distinct EEG stream.
-A set of template ERPs are acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia.
-The experimental protocol implemented to produce this dataset is the same as the one described in Section~\ref{EEGStream}.
-On the other hand, the EEG traces where these templates are superimposed, are experimentally obtained by subjects which are observing the flashing of the stimulus matrix during a P300 Speller procedure but they do not engage in focusing on any letter in particular.
-Everything is there, except the P300 ERP component.
-Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.
-By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential.
-
-A sample P300 ERP obtained from the trial number $2$ of Subject 8 can be seen in Figure  \ref{fig:erptemplate1}.
-These templates are selected due to their shapes more closely resembling the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
-They are produced by extracting segments for this subject and by point-to-point coherently average them.
-
-%After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha
-
-\subsubsection{Active Modality}
-
-Second, an active modality is also implemented, where a P300-Based BCI Speller experiment is performed on four subjects.
-For this scenario, the signal segments are modified to guarantee the inclusion of a P300 component.
-However, in this case the templates are extracted from the same subject.
-Hence, the EEG signal is preprocessed and labeled segments are extracted.
-Segments labeled \textit{True} are coherently point-to-point averaged, and $70$ templates are produced from the whole set of $35$ trials.
-
-Once templates are procured, a random \textit{False} segment for the same subject is obtained.
-This is used as a baseline signal and is added to the template, conforming a new segment which contains one of the templates.
-This procedure continues until the $700$ segments marked as \textit{True} are completed.
-
-%The EEG stream contains $4200$ segments, which correspond to the preprocessed EEG data that was obtained coincidentally with the intensification of a row or column on the Speller Matrix.
-
-\subsubsection{Experiments}
-The experiments are as follows:
-
-\begin{itemize}
-\item Experiment 1 - Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.
-The pool of $70$ P300 ERP waveforms, either obtained from the same subject in the passive-modality or from each subject in the active-modality are used to seed the EEG streams.
-The templates are randomly selected.
-\item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERPs over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
-The lagging value is picked from a uniform distribution $U(0,0.4)$ [s] ranging from $0$ to $0.4$ of the $1s$ segment size~\citep{DaPelo2018}.
-\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.
-This component is defined to be located from the stimulus onset between $148$ ms up to $996$ ms which is around $840$ ms long.
-This waveform element, multiplied by a gain factor, is subtracted from the original template.
-This gain factor between $0$ and $1$ is drawn from a uniform distribution $U(0,1)$.
-Additionally this subtracted waveform is multiplied by a Gaussian window with a support of the same length~\citep{Harris1978}.
-This avoids adding any discontinuity into the artificial generated signal.
-\end{itemize}
-
-All these experiments are executed using cross validation procedure dividing the letter to spell in two sets, preserving the structure of the letter identification trials.
-Spelling letters are scrambled while the order and group of each intensification sequence is preserved.
-
-Finally the performance at letter identifications for these same methods is evaluated by performing an offline BCI Simulation on the Dataset IIb of the BCI Competition II (2003)~\citep{Blankertz2002}.
-The protocol of this dataset is very similar to what was used to obtain the pseudo-real dataset.
-The sampling frequency of this dataset is $240$, the number of letters are $73$ where the first $42$ are used to create the template dictionary for all the methods and the remaining $31$ are used to test the character recognition rate performance.
-Additionally, in this dataset the number of available intensification number sequences is $15$.
-The classification method Support Vector Machine SVM with a linear kernel, is added for comparison as control using a feature $f$ constructed by normalizing the signal on each channel~\citep{Krusienski2006}.
-This method has been proved efficient in decoding P300 in several BCI Competitions~\citep{Kaper2004}.
 
@@
 \centering
-\subfigure[EEG trace of the original signal.
-The horizontal axis represents $10$ seconds of the EEG stream, from the $28$th second up to the $38$th.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
-\subfigure[The same $10$-seconds eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
-\subfigure[EEG sample of Cz and L channel of the original EEG trace. Only $1$ second is shown here, at the $31$th second.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
+\subfigure[EEG trace of the original signal.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
+\subfigure[The same eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
+\subfigure[EEG sample of Cz and L channel of the original EEG trace.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
 \subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
-\caption{Eight-channel EEG signal for Subject Number 1 of the pseudo-real dataset without and with the superimposed ERP Template.
+\caption{Eight-channel EEG signal for Subject Number 1 of the psuedo-real dataset without and with the superimposed ERP Template.
 The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented.
@@
 \caption{Point-to-point averaged signals.
-These are extracted from the first letter identification trial of the Subject 1 of the pseudo-real dataset.
+These are extracted from the first letter identification trial of the Subject 1 of the psuedo-real dataset.
 The ERP is superimposed on classes 3 and 9.
@@
 %The averaging procedure aims to extract the superimposed ERP Template and cancel out everything else on each segment.
+
+\paragraph{Active modality}
+
+The experiment conducted by~\citep{Riccio2013} is replicated on 4 healthy subjects, performing a copy-spelling task with feedback.
+In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.
+The most prototypical P300 waveform template is selected by visual observation.
+This template is superimposed to each segment marked as \textit{True} on the EEG stream.
+The baseline of each segment, on the other hand, is obtained by picking randomly any \textit{False} segment for the same subject.
 
@@
 \includegraphics[width=15cm]{images/dictionary.eps}
-\caption{Coherently averaged signals segments of $1$-second length containing the superimposed ERP.
-Vertical axis unit is $\mu V$.
+\caption{Coherently averaged signals containing the superimposed ERP.
 Each one is extracted from the EEG signal of the Subject 1 of the pseudo-real dataset.
@@
 
-Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Neuper2003,Kathner2017}.
+Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.
 When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.
@@
 Which algorithm perform better at this even when segment latency variations and component amplitudes issues are present ?
-
-waveform-based BCI system, which has an inherent intelligible property (REF), will foster a collaboration in a multidisciplinary environment and will ease the acceptance, and the usability and usefulness of a BCI device.
-The reason being, for caregivers and medical staff, particularly those with the traditional knowledge of the clinical EEG which is based on waveforms, they will feel more natural the understanding of how the system is detecting signals.
 

### 2018-11-27 brainsci-371413-done/brainsci-371413.tex
Final and published version

--- 2018-11-09 BrainScience/brainscience.tex
@@
 %=================================================================
-\documentclass[brainsci,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
+\documentclass[brainsci,article,accept,moreauthors,pdftex,10pt,a4paper]{mdpi}
 
@@
 % The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted.
-This will only make changes to the frontpage (e.g.
-the logo of the journal will get visible), the headings, and the copyright information.
+This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information.
 Also, line numbering will be removed.
@@
 \copyrightyear{2018}
-\externaleditor{Academic Editor: name}
+%\externaleditor{Academic Editor: name}
 \history{Received: date; Accepted: date; Published: date}
@@
 %\pdfoutput=1
-
+\usepackage{subfigure}
+\makeatletter
+\renewcommand{\@thesubfigure}{\normalsize(\textbf{\alph{subfigure}})}
+\makeatother
+\usepackage{upgreek}
 %=================================================================
@@
 % Full title of the paper (Capitalized)
-\Title{EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces}
-
+\Title{EEG Waveform Analysis of P300 ERP with Applications to Brain Computer Interfaces}
+%%Please carefully check the accuracy of names and affiliations. Changes will not be possible after proofreading. 
 % Authors, for the paper (add full first names)
-\Author{Rodrigo Ramele $^{1,\dagger}$, Ana Julia Villar $^{1}$ and Juan Miguel Santos  $^{1}$*}
+\Author{\hl{Rodrigo Ramele} *, Ana Julia Villar  and Juan Miguel Santos $^{\dagger}$}
 % Authors, for metadata in PDF
@@
 \address[1]{%
-$^{1}$ \quad Computer Engineering Department, Instituto Tecnológico de Buenos Aires (ITBA); info@itba.edu.ar}
-
+Computer Engineering Department, Instituto Tecnológico de Buenos Aires (ITBA), \hl{Buenos Aires}, Argentina;  jsantos@itba.edu.ar (J.M.S.); jvillar@itba.edu.ar (A.J.V.)}
+%%please add post code
 % Contact information of the corresponding author
-\corres{Correspondence: rramele@itba.edu.ar; Tel.: +54-9-11-4193-9382}
+\corres{Correspondence: rramele@itba.edu.ar}
 
 % Current address and/or shared authorship
-\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina}
+\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina.}
 
@@
 
-% Abstract (Do not use inserted blank lines, i.e. \\) 
+% Abstract (Do not use inserted blank lines, i.e., \\) 
 \abstract{The Electroencephalography (EEG) is not just a mere clinical tool anymore.
 It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time.
-It is now being used to translate or decode brain signals to diagnose disease or implement Brain Computer Interface (BCI) devices to transmit volitional information.
+It is now being used to translate or decode brain signals, to diagnose diseases or to implement Brain Computer Interface (BCI) devices.
 The automatic decoding is mainly implemented by using quantitative algorithms to detect the cloaked information buried in the signal.
@@
 % Keywords
-\keyword{electroencephalography; brain-computer interfaces; waveform;p300;SIFT;PE;MP;SHCC}
+\keyword{electroencephalography; brain-computer interfaces; waveform; p300; SIFT; PE; MP;~SHCC}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\setcounter{section}{-1} %% Remove this when starting to work on the template.
+%\setcounter{section}{-1} %% Remove this when starting to work on the template.
 
@@
 Additionally, as societies are aging~\citep{Lutz2008} the incidence of neuromuscular atrophies, strokes and other invalidating diseases is increasing.
-Concurrently, the digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices~\citep{Domingo2012}.
+Concurrently, the~digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices~\citep{Domingo2012}.
 All this human computer interaction is based on muscular movement~\citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human motion.
@@
 
-At the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG)~\citep{Schomer2010}.
+At the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the~Electroencephalogram (EEG)~\citep{Schomer2010}.
 
 EEG sensors are wearable~\citep{Puce2017} non-invasive, portable and mobile~\citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution~\citep{Hartman2005}.
-This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, control wheelchairs or to assess alcohol consumption~\citep{Yuste2017,KevricJasmin2016,Tzimourta2018,Cruz2018}.
+This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write~emails, control wheelchairs or to assess alcohol \hl{consumption}~\citep{Yuste2017,KevricJasmin2016,Tzimourta2018,Cruz2018}.%%there are missing some references, we can not find the references,  please modify, please check all the text.
 
@@
 It is measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.
-Figure \ref{fig:sampleeeg} shows $5$ seconds of a sample 8-channel EEG signal.
+Figure~\ref{fig:sampleeeg} shows $5$ s of a sample 8-channel EEG signal.
 
@@
 Although initially analog devices were developed and used, nowadays digital versions connected directly to a computer are pervasive.
-A detailed explanation on the particularities and modeling of EEG can be obtained from \citep{Jackson2014}, and a description of its electrophysiological aspects from \citep{Haberman2012}.
-
-Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The following elements regularly  characterize EEG signals:
-
-\begin{itemize}
+A detailed explanation on the particularities and modeling of EEG can be obtained from~\citep{Jackson2014}, and a description of its electrophysiological aspects from \citep{Haberman2012}.
+
+Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The~following elements regularly  characterize EEG signals:
+
+\begin{itemize}[leftmargin=*,labelsep=5.8mm]
 \item Artifacts:  These are signal sources which are not generated from the CNS, but can be detected from the EEG signal.
-They are called endogeneous or physiological when they are generated from a biological source like face muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic noise\citep{Weeda2012}.
-\item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e.
-its mean and variance, and any other higher-order moments are not time-invariant \citep{Jansen1991}.
+They are called endogeneous or physiological when they are generated from a biological source like face muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic~noise~\citep{Weeda2012}.
+\item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e., its mean and variance, and any other higher-order moments are not time-invariant \citep{Jansen1991}.
 \item DC drift and trending: in EEG jargon, which is derived from concepts of electrical amplifiers theory, Direct Current (DC) refers to very low frequency components of the EEG signal which varies around a common center, usually the zero value.
@@
 These sources generate a baseline EEG which shows continuous activity with a small or null relation with any concurrent cognitive activity or task.
-\item Inter-subject and intra-subject variability: EEG can be affected by the person's behavior like sleep hygiene, caffeine intake, smoking habit or alcohol intake previously to the signal measuring procedure \citep{Farzan2017}.
+\item Inter-subject and intra-subject variability: EEG can be affected by the person's behavior like sleep hygiene, caffeine intake, smoking habit or alcohol intake previously to the signal measuring~procedure \citep{Farzan2017}.
 \end{itemize}
@@
 
-\begin{itemize}
+\begin{itemize}[leftmargin=*,labelsep=5.8mm]
 \item Spontaneous: generally treated as noise or basal EEG.
@@
 
-\begin{itemize}
+\begin{itemize}[leftmargin=*,labelsep=5.8mm]
 \item Rhythmic: EEG activity consisting in waves of approximately constant frequency.
 It is often abbreviated RA (regular rythmic activity).
-They are loosely classified by their frequencies, and their naming convention was derived from the original naming used by Hans Berger himself, and after Alpha Waves (10 Hz), it came Delta (4 Hz), Theta (4-7 Hz), Sigma (12-16 Hz), Beta (12-30 Hz) and Gamma (30-100 Hz).
+They are loosely classified by their frequencies, and~their naming convention was derived from the original naming used by Hans Berger himself, and after Alpha Waves (10 Hz), it came Delta (4 Hz), Theta (4--7 Hz), Sigma (12--16~Hz), Beta~(12--30~Hz) and Gamma (30--100 Hz).
 \item Arrhythmic: EEG activity in which no stable rhythms are present.  
@@
 
-The number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: signal elements can be generalized, focal or lateralized, depending on in which channel (i.e.
-electrode) they are found.
+The number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: signal~elements can be generalized, focal or lateralized, depending on in which channel (i.e., electrode) they are found.
 
@@
 
-\subsection{EEG Waveform Characterization}
+\subsection*{EEG Waveform Characterization}
 
@@
 A waveform can have a characteristic shape, a rising or falling phase, a pronounced plateau or it may be composed of ripples and wiggles.
-In order to describe them, they are characterized by its amplitude, the arch, whether they have (non)sinusoidal shape, by the presence of an oscillation or imitating a sawtooth (e.g.
-Motor Cortical Beta Oscillations).
-The characterization by their sharpness is also common, particularly in Epilepsy, and they can also be identified by their resemblance to spikes (e.g.
-Spike-wave discharge).
+In order to describe them, they are characterized by its amplitude, the arch, whether they have (non)sinusoidal shape, by the presence of an oscillation or imitating a sawtooth (e.g., Motor Cortical Beta Oscillations).
+The characterization by their sharpness is also common, particularly in Epilepsy, and they can also be identified by their resemblance to spikes (e.g., Spike-wave discharge).
 
@@
 
-\begin{itemize}
+\begin{itemize}[leftmargin=*,labelsep=5.8mm]
 \item Attenuation: Also called suppression or depression.
@@
 \item Hypersynchrony: Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range.
-The term suggest an increase in the number of neural elements contributing to the rhythm, or in the synchronization of different neurons with the same discharge pattern~\citep{Buzsaki2012}.
+The term suggest an increase in the number of neural elements contributing to the rhythm, or in the synchronization of different neurons with the same discharge~pattern~\citep{Buzsaki2012}.
 \item Paroxysmal: Activity that emerges from background with a rapid onset, reaching frequently high voltage and ending with an abrupt return to lower voltage activity. 
 \item Monomorphic: Activity appearing to be composed of one dominant waveform pattern.
-\item Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.
-\item Transient/Component: An isolated wave or pattern that is distinctly different from background activity. 
+\item \textls[-15]{Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.}
+\item Transient/Component:~An isolated wave or pattern that is distinctly different from background~activity. 
 \end{itemize}
@@
 %
-%%(e.g. FIRDA - Frontal Intermittent Rhythmic Delta) and posteriorly in children e.g. OIRDA - Occipital Intermittent Rhythmic Delta).
+%%(e.g., FIRDA - Frontal Intermittent Rhythmic Delta) and posteriorly in children e.g., OIRDA - Occipital Intermittent Rhythmic Delta).
 %
@@
 
-Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}.
-The different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}.
-Visual characterization includes the identification or classification of certain waveform components based on a subjective characterization (e.g.
-positive or negative peak polarity) or the location within the strip.
-It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g.
-sleep K-Complex is well characterized based on rates between positive vs negative amplitude)~\citep{Uchida1999}.
+Sleep Research has been studied in this way by performing Polysomnographic recordings (\hl{PSG})~\citep{Boostani2017,Rodenbeck2006}.
+The different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized \hl{guidelines}~\citep{Dimitriadis2018}.
+Visual characterization includes the identification or classification of certain waveform components based on a subjective characterization (e.g., positive or negative peak polarity) or the location within the strip.
+It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g., sleep K-Complex is well characterized based on rates between positive vs negative amplitude)~\citep{Uchida1999}.
 Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves, sleep spindles, polysplindles, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).
@@
 The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.
-The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges~\citep{EEGIntro}.
+The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave~discharges~\citep{EEGIntro}.
 
@@
 This component is a positive deflection of a subject's EEG signal that arises when an unexpected and infrequent stimulus appears~\citep{Wolpaw2002}.
-The P300 is widely utilized in BCI because it can be harnessed to implement a Speller application.
-Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.
+The~P300 is widely utilized in BCI because it can be harnessed to implement a Speller application.
+Hence, P300~ERPs are a target phenomena to study by automatic waveform recognition methods.
 
@@
 Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.
-For ERPs on the other hand, more elaborate indexes has been provided, establishing relations between amplitudes of signal components.
+For ERPs on the other hand, more~elaborate indexes has been provided, establishing relations between amplitudes of signal components.
 Finally, Sleep studies and ICU research are areas where the most complex indexes has been derived, particularly the coupling of signal properties like phase, amplitude and frequency.
@@
 \centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+%% \tablesize{} %% You can specify the fontsize here, e.g.,  \tablesize{\footnotesize}. If commented out \small will be used.
 \begin{tabular}{ccc}
@@
 The exploration of methods based on waveforms is conducted by following the PRISMA~\citep{Moher2009} guidelines.
-Search is performed on Google Scholar, Semantic Web and IEEE Xplore search engines by the terms "Waveforms" OR "Shape" OR "Morphology" OR "Visual inspection" + "EEG".
+Search is performed on Google Scholar, Semantic Web and IEEE Xplore search engines by the terms ``Waveforms'' OR ``Shape'' OR ``Morphology'' OR ``Visual inspection'' + ``EEG''.
 
@@
 
-\begin{enumerate}
+\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
 \item The analysis considers the shape of the plot of the signal.
@@
 Straightforward in its implementation, it consists in assigning a component to a simple waveform element based on the expected location of its more prominent deflection \citep{Ouyang2017}.
-Of regular use in ERP Research, the name of many of the EEG features reference directly a peak within the component, e.g.
-P300 or P3a P3b or N100.
+Of~regular use in ERP Research, the name of many of the EEG features reference directly a peak within the component, e.g., P300 or P3a P3b or N100.
 This leads to a natural way to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.
-The letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).
+The~letter provides the polarity (Positive or Negative) and the numbering shows the time referencing the stimulus onset, or the ordinal position of each peak (first, second, etc).
 
@@
 Particularly, ERPs are regularly used to assess auditory response in infants.
-They are characterized in this manner, where the name of many of the EEG features evoke directly a peak within the waveform, e.g.
-P300 or P3a P3b or N100.
+They are characterized in this manner, where the name of many of the EEG features evoke directly a peak within the waveform, e.g., P300 or P3a P3b or N100.
 This leads to a natural procedure to classify them visually by selecting appropriate peaks and matching their positions and amplitudes in an orderly manner.
@@
 Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms.
-In \cite{Klein1976} a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.
+In~\cite{Klein1976} a feature derived from the amplitude and frequency of its signal and its derivative in time-domain is used.
 Moreover, \citep{Yamaguchi2009} explored the use of Mathematical Morphology, where the time-domain structure of contractions and dilations were studied.
@@
 
-\begin{itemize}
+\begin{itemize}[leftmargin=*,labelsep=5.8mm]
 \item Matching Pursuit
@@
 
-All these methods provide a feature $ f $~\footnote{The notation $f=\{f_i\}_{1}^{n} $ or $f=\{f_i\}_{i \in J}^{} $  will be used to describe the concatenation of scalar values to form a multidimensional feature vector $f=\{f_1,f_2,...,f_n\}$.}  that can be used as a template.
-These algorithms are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.
+%All these methods provide a feature $ f $~({\highlight{The notation $f=\{f_i\}_{1}^{n} $ or $f=\{f_i\}_{i \in J}^{} $}  \hl{will be used to describe the concatenation of scalar values to form a multidimensional feature vector} \highlight{$f=\{f_1,f_2,...,f_n\}$}.})
+
+\highlight{All these methods provide a feature $ f $ that can be used as a template}.
+The notation $f=\{f_i\}_{1}^{n} $ or $f=\{f_i\}_{i \in J}^{} $  is used to describe the concatenation of scalar values to form a multidimensional feature vector $f=\{f_1,f_2,...,f_n\}$.
+These algorithms are all based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length $N$ of the EEG segment in sample points.
 These features are used to create dictionaries or template databases.
 Finally, these templates provide the basis for the pattern matching algorithm and offline classification.
-These algorithm were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
-To maintain reproducibility, the dataset described in Section~\ref{EEGStream} and the source code has been made available in the online repository of the Code Ocean platform under the name \textit{EEGWave}.
+Algorithms were implemented on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).
+To maintain reproducibility, the dataset described in Section~\ref{EEGStream} and the source code has been made available in the online repository of the Code Ocean platform under the name \textit{EEGWave}.%%\footnote is here.
 
@@
 Given a signal, this optimization technique, tries to find the indexes of $m$ atoms and their weights (contributions) that minimize,
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent which is the error between the signal and its approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.
+which is the error between the signal and its approximation constructed by the weighted $w_i$ atoms $g_{i}$, and calculating the euclidean norm ${\left\lVert \cdot \right\rVert}_{2}$.
 The algorithm goes by first setting the approximating signal $\tilde{x}_{0}$  as the original signal itself,
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent and setting the iterative counter $k$ as $1$. Hence, it searches recurrently the best template out of the dictionary  that matches current approximation.
-
+and setting the iterative counter $k$ as $1$. Hence, it searches recurrently the best template out of the dictionary  that matches current approximation.  
 \begin{equation}
@@
 \end{equation}
-
-\noindent where $g_{i}$ are all the available scaled, translated and modulated atoms from the dictionary.
+ where $g_{i}$ are all the available scaled, translated and modulated atoms from the dictionary.
 The operation $\left\lvert \cdot \right\rvert$ corresponds to the absolute value of the inner product.
 This step determines the atom selection process, and their contribution is calculated based on
-
 %The operation $\left\lvert < \cdot,\cdot > \right\rvert$ corresponds to the absolute value of the inner product. 
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $k$ representing the index of the selected atom $g_{k}$ and ${\left\lVert \cdot \right\rVert}_{2}$ its euclidean norm.
+with $k$ representing the index of the selected atom $g_{k}$ and ${\left\lVert \cdot \right\rVert}_{2}$ its euclidean norm.
 Finally the contribution of each atom is subtracted from the next approximation~\citep{Cohen2014,Sanei2007, Mallat1993}
-
 \begin{equation}
@@
 
-The stopping criteria can be established based on a limiting threshold on Equation \ref{eq:mperror} or based on a predetermined number of steps and selected atoms.
+The stopping criteria can be established based on a limiting threshold on Equation (\ref{eq:mperror}) or based on a predetermined number of steps and selected atoms.
 Two variants of this algorithm are evaluated.
@@
 In \textit{MP 2} the coefficients of Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary~\citep{Vareka2012}.
-For the first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building, an eventually sparse, vector with them:
-
+For the first version, the matching against the template is evaluated according to Equation (\ref{eq:mperror}) directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building, an eventually sparse, vector with them:
 \begin{equation}
@@
 \end{equation}
-
-\noindent where $D$ is the size of the dictionary.
+where $D$ is the size of the dictionary.
 
@@
 
-\subsection{Permutation Entropy - PE}
-
-Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and increasingly for Epilepsy pre-ictal detection \citep{Bandt2002}.
+\subsection{Permutation Entropy---PE}
+
+Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with~applications on Anesthesia, Sleep Stage evaluation and increasingly for Epilepsy pre-ictal detection \citep{Bandt2002}.
 This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found along the signal.
@@
 Let's consider a signal on a window of length $W$ represented by the sample points
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent and resampled by $\tau$ intervals, starting from the sampling point $n$, doing
-
+and resampled by $\tau$ intervals, starting from the sampling point $n$, doing
 \begin{equation}
@@
 This sequence is of order $m$, which is the number of sample points used to derive the ordinal element called $\pi$.
-There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a decreasing order relationship~\citep{Keller2017}.
-For example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(1,3,2)$.
-Thus, along the signal window there can be at most $k$ different ordinal (and overlapping) elements $\pi_{s}$
-
+There are $m!$ ways in which this sequence can be orderly arranged, according to the position that each sample point holds within the sequence in a decreasing order \hl{relationship}~\citep{Keller2017}.
+For~example if $m=3$, and the first sample point is the bigger, the second is the smaller and the third one is in the middle, the ordinal element $\pi$ corresponds to $(1,3,2)$.
+Thus, along the signal window there can be at most $k$ different ordinal (and overlapping) elements $\pi_{s}$%%please add reference
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $k = W-(m-1) \tau$.
+with $k = W-(m-1) \tau$.
 The probability density function \textit{pdf} for all the available permutations of order $m$ should be $ \textbf{p} = (p_1,p_2,...,p_{m!}) $ with $ \sum_{i=1}^{m!} p_{i} = 1 $.
@@
 Hence, the time series window is mapped to a new set of $k$ ordinal elements, and the \textit{pdf} can be calculated by the empirical permutation entropy,
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $1 \leq i \leq m!$.
+with $1 \leq i \leq m!$.
 The Iverson Bracket $ \left[ \cdot \right] $ resolves to $1$ when their logical proposition argument is true, $0$ otherwise.
@@
 The empirical permutation entropy can be calculated from the histogram as,
-
 \begin{equation}
@@
 The implemented code was derived from~\citep{Unakafova2013}, and the model description from \citep{Berger2017}.
-This procedure produces a scalar number for the given signal window of size $W$.
+This~procedure produces a scalar number for the given signal window of size $W$.
 To derive a feature, a sliding window procedure must be implemented to cover an entire segment of length $N$.
-Thus, the length of the feature is $N - (W + \tau (m - 1))$.
-
+Thus, the~length of the feature is $N - (W + \tau (m - 1))$.
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $u$ varying on a sample by sample basis along the signal, starting from the specified index.
-
-\subsection{Slope Horizontal Chain Code - SHCC}
+with $u$ varying on a sample by sample basis along the signal, starting from the specified index.
+
+\subsection{Slope Horizontal Chain Code---SHCC}
 
@@
 A signal of length $N$, can be represented by a list of ordered-pairs $e$,
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent and it can be divided into $G$ different blocks.  These blocks are obtained by resampling the original signal from the index
-
+and it can be divided into $G$ different blocks.  These blocks are obtained by resampling the original signal from the index 
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $n$ being the original sampling index on $ 1 \leq n \leq N $ and $\lfloor \cdot \rfloor$ being the floor operation, i.e.
-rounding of the number argument to the closest smaller integer number.
-On the other hand, $\Delta$ can be obtained by
-
+with $n$ being the original sampling index on $ 1 \leq n \leq N $ and $\lfloor \cdot \rfloor$ being the floor operation, i.e., rounding of the number argument to the closest smaller integer number.
+On the other hand, $\Delta$ can be obtained~by
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $ G < N $ and using instead $\lceil \cdot \rceil$ as the ceil operation, the rounding to the closest bigger integer number. Lastly, the value $m$ can be derived from
-
+with $ G < N $ and using instead $\lceil \cdot \rceil$ as the ceil operation, the rounding to the closest bigger integer number. Lastly, the value $m$ can be derived from
 \begin{equation}
@@
 This resampling produces a new sequence of values,
-
 \begin{equation}
@@
 The next step is the normalization of each ordered-pair as vectors $\textbf{x'} = (x'_{1},...,x'_{G})$ and $\textbf{y'} = (y'_{1},...,y'_{G})$ according to
-
 \begin{equation}
@@
 \end{equation}
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $\textbf{1}$ being the vector of length $G$ with all their components equal to $1$.
+with $\textbf{1}$ being the vector of length $G$ with all their components equal to $1$.
 Hence, the scalar components $\hat{x}_{s}$ of $\hat{\textbf{x}}$, and $\hat{y}_{s}$ of $\hat{\textbf{y}}$, with $s$ varying between $1$ and $G$ are effectively normalized to $\hat{x}_{s},\hat{y}_{s} \in [0,1]$.
-\\
-
-Finally, the feature is constructed by calculating the point-to-point slope against the horizontal plane,
-
+
+Finally, the feature is constructed by calculating the point-to-point slope against the horizontal~plane,
 \begin{equation}
@@
 
-\subsection{Scale Invariant Feature Transform - SIFT}
-
-SIFT~\citep{Lowe2004} is a very successful feature extraction technique from Computer Vision .
-It has a biomimetic inspiration on how the visual cortex analyze images based on orientations~\citep{Edelman1997}.
-This method has been used in \citep{Ramele2016} to analyze EEG signals based on their plots on 2D images.
+\subsection{Scale Invariant Feature Transform---SIFT}
+
+SIFT~\citep{Lowe2004} is a very successful feature extraction technique from Computer Vision.
+It has a biomimetic inspiration on how the visual cortex analyze images based on \hl{orientations}~\citep{Edelman1997}.
+This~method has been used in \citep{Ramele2016} to analyze EEG signals based on their plots on 2D images.%%please add reference
 
 The first step of the algorithm is the plot generation based on single-channel EEG segments $ x(n) $.  Hence, this signal is normalized by the z-score~\citep{Zhang2013}: 
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $\delta$ being the signal magnification factor and $\bar{x}$ and $\sigma_{x}$, the mean and standard deviation of $x$ on the signal segment.
-The width of the image is determined based on the $1$-second length size of the segment in sample units.
+with $\delta$ being the signal magnification factor and $\bar{x}$ and $\sigma_{x}$, the mean and standard deviation of $x$ on the signal segment.
+The width of the image is determined based on the $1$-s length size of the segment in sample units.
 This corresponds to the sampling frequency $F_s$ of the EEG signal segment.
 The width is adjusted by multiplying by the magnification factor $\delta$,
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent whereas the height is calculated based on the peak-to-peak amplitude of the signal within the segment,
-
+whereas the height is calculated based on the peak-to-peak amplitude of the signal within the segment,
 \begin{equation}
@@
 
-Equation \ref{eq:sift5} determines the vertical position of the image where the signal's zero value will be located.
-
+Equation (\ref{eq:sift5}) determines the vertical position of the image where the signal's zero value will be~located.
 \begin{equation}
@@
 Finally, a binary, black-and-white image plot is generated based on
-
 \begin{equation}
 I(z_1,z_2) = \left\{ \begin{array}{rl}
-255 & \text{if} \,  z_1 = \delta \  n; \! z_2 = \tilde{x}(n) + z \\
+255 & \text{if} \,  z_1 = \delta \  n; \  z_2 = \tilde{x}(n) + z \\
 0   & \mbox{otherwise}
@@
 \end{equation}
-
-\noindent where $z_1$ and $z_2$ are the image coordinates values, $255$ represents white and $0$ is the background black color of the plot.
+where $z_1$ and $z_2$ are the image coordinates values, $255$ represents white and $0$ is the background black color of the plot.
 These points are interpolated using the Bresenham algorithm~\citep{Ramele2016}.
@@
 Therefore, for each block $(i,j)$ within the patch, a histogram $h(i,j,\theta)$ of the gradient orientations, for 8 circular orientations $\theta$, are calculated.
-This histogram is concatenated for all the 16 blocks and a feature is thus formed:
-
+This histogram is concatenated for all the 16 blocks and a feature is thus~formed:
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $i$ and $j$ belonging to $I = \{ 0,1,2,3 \} $ and localizing the 16 blocks within the grid.
+with $i$ and $j$ belonging to $I = \{ 0,1,2,3 \} $ and localizing the 16 blocks within the grid.
 The angles $\theta$ that belong to $\Theta$ are the eight possible equidistant values between $0$ and $315$.
@@
 The objective of the following experiments is to assess the performance of the algorithms that aim to recognize the shape of the P300 waveform, obtained after averaging signal segments.
-This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.
+This~performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.
 The experiments are performed by the offline evaluation of the character identification rate of a Visual P300-Based BCI Speller application.
 
-Farwell and Donchin P300 Speller~\citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
+Farwell and Donchin P300 \hl{Speller}~\citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.
 This procedure exploits a cognitive phenomena raised by the oddball paradigm~\citep{Luck2005}: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.
@@
 Therefore, P300 experiments data is often mined with \textit{null-signals}.
-These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the expected signal element is not generated.
+These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or~concentrated, the~expected signal element is not generated.
 This lack of certainty may be in detriment of any conducted analysis and can be misleading or difficult to deal with.
@@
 The participants are healthy and have normal or corrected-to-normal vision and no history of neurological disorders.
-These voluntary subjects are aged between 20-40 years old.
+These voluntary subjects are aged between 20--40 years old.
 EEG data is collected in a single recording session.
-Each subject is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.
+Each subject is seated in a comfortable chair, with~her/his vision aligned to a computer screen located one meter in front of her/him.
 The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.
-Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.
+Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10--20 international system.
 Reference is set to the right ear lobe and ground is preset as the AFz position.
@@
 Each trial is composed of $10$ intensification sequences of the $6$ columns and $6$ rows of the Speller Matrix.
-This yields exactly $120$ intensifications of rows and columns per trial.
-The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ seconds.
-This provides a $4$ Hz frequency of flashes on the screen.
-The initial pause and the inter-trial pauses are set to $20$ seconds.
-The whole experiment lasts for around $1400$ seconds.
+This~yields exactly $120$ intensifications of rows and columns per trial.
+The duration of each intensification as well as the Inter-Stimulus Interval, the pause between stimulations, are set to $0.125$ s.
+This~provides a $4$ Hz frequency of flashes on the screen.
+The initial pause and the inter-trial pauses are set to $20$ s.
+The whole experiment lasts for around $1400$ s.
 This produces an EEG stream which contains $4200$ marked sections where $3500$ of them are labeled as \textit{True} and the remaining $700$ as \textit{False}.
 The extracted EEG signals are band-pass filtered using a 4th order Butterworth digital filter between $0.1$ and $10$ Hz and a $50$ Hz notch filter is applied to remove line AC noise.
-The EEG trace is finally downsampled to $16$ Hz.
-Segments of $1$-second length are extracted according to the markers information and those with variations larger than $70 \mu V$ are identified as artifacts and eliminated.
+The EEG trace is finally downsampled to $16$~Hz.
+Segments of $1$-s length are extracted according to the markers information and those with variations larger than 70 $\upmu$V are identified as artifacts and eliminated.
 
@@
 They do not receive any extra information on the screen.
-None of them have any experience with a BCI device.
+None~of them have any experience with a BCI device.
 A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.
@@
 Everything is there, except the P300 ERP component.
-Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.
+Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and~those timing locations are used to superimpose the extracted ERP waveform.
 By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of the evoked potential.
@@
 A sample P300 ERP obtained from the trial number $2$ of Subject 8 can be seen in Figure  \ref{fig:erptemplate1}.
-These templates are selected due to their shapes more closely resembling the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.
-They are produced by extracting segments for this subject and by point-to-point coherently average them.
-
+These templates are selected due to their shapes more closely resembling the prototypical P300 \mbox{\hl{waveform}~\citep{Rao2013,Clerc2016}}.
+They are produced by extracting segments for this subject and by point-to-point coherently average~them.
+%%please add references
 %After the modification it was verified that the flickering effect of the 4 Hz signal is mantained and that alpha
@@
 \centering
-\includegraphics[width=12cm]{images/erptemplate1.eps}
+\includegraphics[width=11cm]{images/erptemplate1.eps}
 \caption{ERP Template obtained from the coherent point-to-point ensemble average from the signals of Subject Number Eight of the BNCI Horizon public dataset 008-2014.
-The template is $1s$ long which is 256 sample points, and the eight channels are superimposed with different colors.
+The template is 1 s long which is 256 sample points, and the eight channels are superimposed with different colors.
 The P3b component can be seen around the sample index $150$ and $200$.}
@@
 Once templates are procured, a random \textit{False} segment for the same subject is obtained.
-This is used as a baseline signal and is added to the template, conforming a new segment which has a superimposed P300 template.
-This procedure continues until the $700$ segments marked as \textit{True} are completed.
+This~is used as a baseline signal and is added to the template, conforming a new segment which has a superimposed P300 template.
+This procedure continues until the $700$ segments marked as \textit{True} are~completed.
 
@@
 
-Figure~\ref{fig:gains} shows a $5s$ sample of the EEG trace obtained with the MNE library~\citep{Gramfort2013}.
+Figure~\ref{fig:gains} shows a 5 s sample of the EEG trace obtained with the MNE library~\citep{Gramfort2013}.
 Channel $S$ represents the twelve different stimulus markers (columns or rows) while channel $L$ represent the label (\textit{True} vs \textit{False}).
@@
 Subfigures (c) and (d) show only one second of channels Cz and L from the same segment.
-The superimposed ERP can be devised enclosed by the vertical bars, around $31.5$s, where in (d) the peak is slightly bigger.
+The superimposed ERP can be devised enclosed by the vertical bars, around $31.5$ s, where in (d) the peak is slightly bigger.
 Figure~\ref{fig:gaincheck} shows the obtained ensemble average ERPs as result of superimposing the template signal into the EEG stream, time-locked to the stimulus onset.
@@
 
-\begin{figure}[htb]
+\begin{figure}[H]
 \centering
-\subfigure[EEG trace of the original signal.
-The horizontal axis represents $10$ seconds of the EEG stream, from the $28$th second up to the $38$th.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
-\subfigure[The same $10$-seconds eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
-\subfigure[EEG sample of Cz and L channel of the original EEG trace. Only $1$ second is shown here, at the $31$th second.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
-\subfigure[The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
-\caption{Eight-channel EEG signal for Subject Number 1 of the pseudo-real dataset without and with the superimposed ERP Template.
+\subfigure[~EEG trace of the original signal.
+The horizontal axis represents $10$ s of the EEG stream, from the $28$th second up to the $38$th.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
+\subfigure[~The same $10$-s eight-channel signal segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
+\subfigure[~EEG sample of Cz and L channel of the original EEG trace. Only $1$ second is shown here, at the $31$th second.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
+\subfigure[~The same segment with the superimposed template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
+\caption{~Eight-channel EEG signal for Subject Number 1 of the pseudo-real dataset without and with the superimposed ERP Template.
 The channel L, the mark which identifies where to superimpose the P300 ERP, is shown as well as the channel S which identifies the stimulus that was presented.
-On (c) and (d) the small variation that was introduced by the superimposition of the ERP can be seen enclosed by the vertical bars, where the slope of the bump on subfigure (d) is slightly bigger}
+On (\textbf{c},\textbf{d}) the small variation that was introduced by the superimposition of the ERP can be seen enclosed by the vertical bars, where the slope of the bump on subfigure (\textbf{d}) is slightly bigger}
 \label{fig:gains}
@@
 
-\begin{itemize}
-\item Experiment 1 - Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.
+\begin{itemize}[leftmargin=*,labelsep=5.8mm]
+\item Experiment 1---Letter Identification Performance: the letter identification performance of each one of these methods on the artificially generated pseudo-real dataset.
 The pool of $70$ P300 ERP waveforms, either obtained from the same subject in the passive-modality or from each subject in the active-modality are used to compose the artificial P300 wave in the pseudo-real dataset.
 Templates are randomly selected.
-\item Experiment 2 - Latency Noise:  Instead of superimposing the P300 ERPs over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
-The lagging value is picked from a uniform distribution $U(0,0.4)$ [s] ranging from $0$ to $0.4$ of the $1s$ segment size~\citep{DaPelo2018}.
-\item Experiment 3 - Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.
+\item Experiment 2---Latency Noise:  Instead of superimposing the P300 ERPs over the EEG trace at the exact locations where stimulus onsets are situated, an artificial latency lag is added.
+The~lagging value is picked from a uniform distribution $U(0,0.4)$ [s] ranging from $0$ to $0.4$ of the 1 s segment~size~\citep{DaPelo2018}.
+\item Experiment 3---Component Amplitude Noise: the amplitude of the main P3b component of the ERP template is randomly altered.
 This component is defined to be located from the stimulus onset between $148$ ms up to $996$ ms which is around $840$ ms long.
@@
 
-Described algorithms produce a feature $f$ for each EEG segment.
+Described algorithms produce a feature $f$ for each averaged EEG segment.
 The aim of the classification procedure is to identify for the remaining $20$ trials which of the 6 features $f$ that were obtained for row intensification, labeled by $\{ 1,...,6 \}$, and which of the 6 features for column intensification, named $\{ 7,...,12 \}$ are the ones that elicited the P300 response on the averaged EEG segment.
 The row number of the matrix can be obtained by doing
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent with $q_i$ being the set of k-nearest neighbors of the feature $f_u$ with $u$ varying from $1$ to $6$.
+with $q_i$ being the set of k-nearest neighbors of the feature $f_u$ with $u$ varying from $1$ to $6$.
 The parameter $k$ represents the number of neighbors chosen from the dictionary of templates.
 The column can be obtained in the same way,
-
 \begin{equation}
@@
 \centering
-\includegraphics[width=15cm]{images/dictionary.eps}
+\includegraphics[width=14cm]{images/dictionary.eps}
 \caption{Coherently averaged signals segments of $1$-second length containing the superimposed ERP.
-Vertical axis unit is $\mu V$.
+Vertical axis unit is $\upmu$V.
 Each one is extracted from the EEG signal of the Subject 1 of the pseudo-real dataset.
-These averaged signals correspond to the 15 first trials (2 averaged signals from each trial, one belonging to the column flashing and the other to row flashing).
-These are the templates used to build a dictionary per channel per subject and are used by the classification algorithm described in \ref{section:classification}.}
+These averaged signals correspond to the 15 first trials (2 averaged signals from each trial, one~belonging to the column flashing and the other to row flashing).
+These are the templates used to build a dictionary per channel per subject and are used by the classification algorithm described in~Section \ref{section:classification}.}
 \label{fig:dictionaryfig}
@@
 \includegraphics[width=15cm]{images/1-1.eps}
-\caption{\textbf{Passive Modality - Experiment 1}:Speller performance curves obtained for each method for the four subjects that performed the active modality protocol.
+\caption{{Passive Modality---Experiment 1}: Speller performance curves obtained for each method for the four subjects that performed the passive modality protocol.
 Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.
@@
 \centering
-\includegraphics[width=15cm]{images/2-1.eps}
-\caption{\textbf{Active Modality - Experiment 1}:Speller performance curves obtained for each method for the four subjects that performed the active modality protocol.
+\includegraphics[width=16.5cm]{images/2-1.eps}
+\caption{{Active Modality---Experiment 1}:Speller performance curves obtained for each method for the four subjects that performed the active modality protocol.
 Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.
@@
 \centering
-\includegraphics[width=15cm]{images/1-2.eps}
-\caption{\textbf{Passive Modality - Experiment 2}: Performance curves for four subjects for the five algorithms when a random latency is included when superimposing the P300 signal template.}
+\includegraphics[width=16cm]{images/1-2.eps}
+\caption{{Passive Modality---Experiment 2}: Performance curves for four subjects for the five algorithms when a random latency is included when superimposing the P300 signal template.}
 \label{fig:1-2}
@@
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=15cm]{images/2-2.eps}
-\caption{\textbf{Active Modality - Experiment 2}: Performance curves for the four subjects for the five algorithms.  A random latency is included while superimposing the P300 signal template.}
-\label{fig:2-2}
-\end{figure}
-
-Finally, Figure \ref{fig:1-3} and \ref{fig:2-3} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template is randomly attenuated.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=15cm]{images/1-3.eps}
-\caption{\textbf{Passive Modality - Experiment 3}:Performance curves for four subjects for the five algorithms when he amplitude of the P3b component of the template is randomly attenuated. }
-\label{fig:1-3}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=15cm]{images/2-3.eps}
-\caption{\textbf{Active Modality - Experiment 3}:Performance curves for four subjects for the five algorithms when the amplitude of the P3b component of the template is randomly attenuated. }
-\label{fig:2-3}
-\end{figure}
+Finally, Figures \ref{fig:1-3} and \ref{fig:2-3} represents the performance values obtained for the Experiment 3, when~the amplitude of the P3b component of the template is randomly attenuated.
 
@@
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=15cm]{images/PerformanceBCICompetition.eps}
-\caption{Speller performance obtained for the Dataset IIb of the BCI Competition II (2003) for each one of the algorithms.
-An offline BCI Simulation is performed using the first $42$ trials as training and the remaining $31$ as testing.
-The horizontal axis show the number of intensification sequences, from $0$ to $15$ for this dataset, while the vertical axis show the performance rate.}
-\label{fig:performancebcicompetition}
-\end{figure}
-
+ \vspace{-6pt}
 \begin{table}[H]
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=16cm]{images/2-2.eps}
+\caption{{Active Modality---Experiment 2}: Performance curves for the four subjects for the five algorithms.  A random latency is included while superimposing the P300 signal template.}
+\label{fig:2-2}
+\end{figure}
+
+ \vspace{-6pt}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=16cm]{images/1-3.eps}
+\caption{{Passive Modality---Experiment 3}:Performance curves for four subjects for the five algorithms when the amplitude of the P3b component of the template is randomly attenuated. }
+\label{fig:1-3}
+\end{figure}
+ \vspace{-6pt}
+\begin{figure}[H]
+\centering
+\includegraphics[width=16cm]{images/2-3.eps}
+\caption{{Active Modality---Experiment 3}:Performance curves for four subjects for the five algorithms when the amplitude of the P3b component of the template is randomly attenuated. }
+\label{fig:2-3}
+\end{figure}
+
+ \vspace{-6pt}
+\begin{figure}[H]
+\centering
+\includegraphics[width=16cm]{images/PerformanceBCICompetition.eps}
+\caption{Speller performance obtained for the Dataset IIb of the BCI Competition II (2003) for each one of the algorithms.
+An offline BCI Simulation is performed using the first $42$ trials as training and the remaining $31$ as testing.
+The horizontal axis show the number of intensification sequences, from $0$ to $15$ for this dataset, while the vertical axis show the performance rate.}
+\label{fig:performancebcicompetition}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %\centering
-%%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
+%%% \tablesize{} %% You can specify the fontsize here, e.g.,  \tablesize{\footnotesize}. If commented out \small will be used.
 %\begin{tabular}{ccc}
@@
 
-Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Neuper2003,Kathner2017}.
+Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and~for some algorithms close to the usable\hl{ threshold of} $70\%$~\citep{Neuper2003,Kathner2017}.
 When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.
-Results for this competition have shown perfect classification with tailored algorithms~\citep{Kundu2018}.
+Results for this competition have shown perfect classification with tailored \hl{algorithms}~\citep{Kundu2018}.
 This level is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.
-It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.
-
+It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel \hl{signal}~\citep{Rakotomamonjy2008,Gribonval2008}.
+%%please add reference.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+\section{Conclusions}
 \label{section:conclusion}
@@
 We verified that it is possible to obtain discriminating information from the underlying signal based exclusively on an automated method of processing the waveforms.
-This brings the possibility to use these techniques to implement intelligible~\citep{j2018challenge} automatic detection procedures, i.e.
-systems that are able to emphasize clearly and noticeable what are the factors that caused the system action, decision or classification.
+This brings the possibility to use these techniques to implement intelligible~\citep{j2018challenge} automatic detection procedures, i.e., systems that are able to emphasize clearly and noticeable what are the factors that caused the system action, decision or classification.
 This is due to the fact that they are based on metrics which can be visually verified.
@@
 Additionally, this may foster collaboration in a multidisciplinary environment and may ease the acceptance and translation of BCI technology~\citep{Chavarriaga2017}.
-The reason being, for caregivers and medical staff, particularly those with the expertise of the clinical EEG which is based on waveforms, they may feel a natural understanding of how the system is performing.
+The~reason being, for caregivers and medical staff, particularly those with the expertise of the clinical EEG which is based on waveforms, they may feel a natural understanding of how the system is~performing.
 
@@
 Another benefit of these methodologies is that they have a potential universal applicability.
-As they are only analyzing waveforms, they can be explored in other disciplines where the structure or shape of the waveform is of relevance.
+As~they are only analyzing waveforms, they can be explored in other disciplines where the structure or shape of the waveform is of relevance.
 Analyzing signals by their waveforms is relative common in chemical analysis~\citep{Skoog2000}, seismic analysis in Geology~\citep{Owens1984}, and quantitative financial analysis.
@@
 %\supplementary{The following are available online at www.mdpi.com/link, Figure S1: title, Table S1: title, Video S1: title.}
-
+\authorcontributions{This projects is part of a the first author's PhD Thesis which is directed by J.M.S. and codirected by A.J.V.}
+
+\funding{\hl{This project was supported by the ITBACyT-15 funding program issued by ITBA University. }}%%please add funding information
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{This project was supported by the ITBACyT-15 funding program issued by ITBA University.
-We would like to thank to Dr.
+\acknowledgments{We would like to thank to Dr.
 Valentina Unakafova for providing the Permutation Entropy algorithm and to Dr.
@@
 
-%We would like to thank to Dr. Jian Zhang for providing details of the MIDS algorithm, to Dr. Valentina Unakafova for providing the Permutation Entropy algorithm
-
+\conflictsofinterest{The authors declare no conflict of interest.} 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{This projects is part of a the first author's PhD Thesis which is directed by Juan Miguel Santos and codirected by Ana Julia Villar.}
 
@@
 %% optional
+
+\newpage
+
 \abbreviations{The following abbreviations are used in this manuscript:\\
 
-\noindent EEG: electroencephalography\\
-BCI: Brain Computer Interfaces\\
-BMI: Brain Machine Interfaces\\
-BNCI: Brain-Neural Computer Interfaces\\
-SNR: Signal to Noise Ratio\\
-CNS: Central Nervous System\\
-AC: Alternating Current\\
-DC: Direct Current\\
-ERP: Event-Related Potential\\
-P300: Positive deflection at 300 ms\\
-ITR: Information Transfer Rate\\
-BTR: Bit Transfer Rate\\
-SIFT: Scale Invariant Feature Transform\\
-SHCC: Slope Horizontal Chain Code\\
-PE: Permutation Entropy\\
-MP: Matching Pursuit\\
-ICU: Intensive Care Unit\\
-EKG: Electrocardiogram\\
-PAA:Period Amplitude Analysis\\
-SVM:Support Vector Machine
-}
+\noindent 
+\begin{tabular}{@{}ll}
+EEG& electroencephalography\\
+BCI&Brain Computer Interfaces\\
+BMI&Brain Machine Interfaces\\
+BNCI& Brain-Neural Computer Interfaces\\
+SNR& Signal to Noise Ratio\\
+CNS&Central Nervous System\\
+AC&Alternating Current\\
+DC&Direct Current\\
+ERP&Event-Related Potential\\
+P300&Positive deflection at 300 ms\\
+ITR&Information Transfer Rate\\
+BTR&Bit Transfer Rate\\
+SIFT&Scale Invariant Feature Transform\\
+SHCC&Slope Horizontal Chain Code\\
+PE&Permutation Entropy\\
+MP&Matching Pursuit\\
+ICU&Intensive Care Unit\\
+EKG&Electrocardiogram\\
+PAA&Period Amplitude Analysis\\
+SVM&Support Vector Machine
+\end{tabular}}
 
@@
 %\section{}
-%All appendix sections must be cited in the main text. In the appendixes, Figures, Tables, etc. should be labeled starting with `A', e.g., Figure A1, Figure A2, etc.
+%All appendix sections must be cited in the main text. In the appendixes, Figures, Tables, etc. should be labeled starting with `A', e.g.,, Figure A1, Figure A2, etc.
 
@@
 %=====================================
-\bibliography{brainscience}
+\reftitle{References}
+%=====================================
+% References, variant B: external bibliography
+%=====================================
+%\bibliography{brainscience}
+\begin{thebibliography}{-------}
+\providecommand{\natexlab}[1]{#1}
+
+\bibitem[Wolpaw \em{et~al.}(2002)Wolpaw, Birbaumer, McFarland, Pfurtscheller,
+  and Vaughan]{Wolpaw2002}
+Wolpaw, J.R.; Birbaumer, N.; McFarland, D.J.; Pfurtscheller, G.; Vaughan, T.M.
+\newblock {Brain-computer interfaces for communication and control.}
+\newblock {\em Clin. Neurophysiol.} {\bf 2002}, {\em 113},~767--91.
+
+\bibitem[Lutz \em{et~al.}(2008)Lutz, Sanderson, and Scherbov]{Lutz2008}
+Lutz, W.; Sanderson, W.; Scherbov, S.
+\newblock {The coming acceleration of global population ageing}.
+\newblock {\em Nature} {\bf 2008}, {\em 451},~716--719.
+
+\bibitem[Domingo(2012)]{Domingo2012}
+Domingo, M.C.
+\newblock {An overview of the Internet of Things for people with disabilities}.
+\newblock {\em J. Netw. Comput. Appl.} {\bf 2012}, {\em
+  35},~584--596.
+
+\bibitem[Guger \em{et~al.}(2017)Guger, Allison, and Lebedev]{Guger2017}
+Guger, C.; Allison, B.Z.; Lebedev, M.A.
+\newblock {Introduction}. In {\em Brain Computer Interface Research: A State of
+  the Art Summary 6}; Springer: Cham, Switzerland, 2017; pp. 1--8.
+
+\bibitem[Schomer and Silva(2010)]{Schomer2010}
+Schomer, D.L.; Silva, F.L.D.
+\newblock {\em Niedermeyer's Electroencephalography: Basic Principles, Clinical
+  Applications, and~Related Fields}; Oxford, UK; Oxford University Press:  \hl{2017}.%%please add location of the publisher
+
+\bibitem[Puce and H{\"{a}}m{\"{a}}l{\"{a}}inen(2017)]{Puce2017}
+Puce, A.; H{\"{a}}m{\"{a}}l{\"{a}}inen, M.S.
+\newblock {A review of issues related to data acquisition and analysis in
+  EEG/MEG studies}.
+\newblock {\em Brain Sci.} {\bf 2017}, {\em 7}, 58.
+
+\bibitem[{De Vos} and Debener(2014)]{DeVos2014}
+{De Vos}, M.; Debener, S.
+\newblock {Mobile eeg: Towards brain activity monitoring during natural action
+  and cognition}.
+\newblock {\em Int. J. Psychophysiol.} {\bf 2014}, {\em
+  91},~1--2.
+
+\bibitem[Hartman(2005)]{Hartman2005}
+Hartman, A.L.
+\newblock {\em {Atlas of EEG Patterns}}; Philadelphia, USA; Lippincott Williams \& Wilkins, \hl{2005};
+
+\bibitem[Yuste \em{et~al.}(2017)Yuste, Goering, {Ag{\"{u}}eray Arcas}, Bi,
+  Carmena, Carter, Fins, Friesen, Gallant, Huggins, Illes, Kellmeyer, Klein,
+  Marblestone, Mitchell, Parens, Pham, Rubel, Sadato, Sullivan, Teicher,
+  Wasserman, Wexler, Whittaker, and Wolpaw]{Yuste2017}
+Yuste, R.; Goering, S.; {Ag{\"{u}}eray Arcas}, B.; Bi, G.; Carmena, J.M.;
+  Carter, A.; Fins, J.J.; Friesen, P.; Gallant, J.; Huggins, J.E.; et al.
+\newblock {Four ethical priorities for neurotechnologies and AI}.
+\newblock {\em Nature} {\bf 2017}, {\em 551},~159--163.
+
+\bibitem[Jestico \em{et~al.}(1977)Jestico, Fitch, Gilliatt, and
+  Willison]{Jestico1977}
+Jestico, J.; Fitch, P.; Gilliatt, R.W.; Willison, R.G.
+\newblock {Automatic and rapid visual analysis of sleep stages and epileptic
+  activity. A preliminary report}.
+\newblock {\em Electroencephalogr. Clin. Neurophysiol.} {\bf 1977},
+  {\em 43},~438--441.
+
+\bibitem[Jansen(1991)]{Jansen1991}
+Jansen, B.H.
+\newblock {Quantitative analysis of electroencephalograms: is there chaos in
+  the future?}
+\newblock {\em Int. J. Bio-Med.~Comput.} {\bf 1991}, {\em
+  27},~95--123.
+
+\bibitem[Nijboer and Broermann(2009)]{Nijboer2009}
+Nijboer, F.; Broermann, U.
+\newblock {Brain–Computer Interfaces for Communication and Control in
+  Locked-in Patients}; Springer: Berlin/Heidelberg, Germany, 2009; pp. 185--201.
+
+\bibitem[Wei \em{et~al.}(2017)Wei, Zou, Zhang, and Chen]{Wei2017}
+Wei, Z.C.; Zou, J.Z.; Zhang, J.; Chen, L.L.
+\newblock {Automatic recognition of epileptic discharges based on shape
+  similarity in time-domain}.
+\newblock {\em Biomed. Signal Process. Control} {\bf 2017}, {\em
+  33},~236--244.
+
+\bibitem[Thakor and Tong(2004)]{Thakor2004}
+Thakor, N.V.; Tong, S.
+\newblock {Advances in Quantitative Electroencephalogram Analysis Methods}.
+\newblock {\em Annu. Rev. Biomed. Eng.} {\bf 2004}, {\em
+  6},~453--495.
+
+\bibitem[Jackson and Bolger(2014)]{Jackson2014}
+Jackson, A.F.; Bolger, D.J.
+\newblock The neurophysiological bases of EEG and EEG measurement: A review for
+  the rest of us.
+\newblock {\em Psychophysiology} {\bf 2014}, {\em 51},~1061--1071.
+
+\bibitem[Haberman and Spinelli(2012)]{Haberman2012}
+Haberman, M.A.; Spinelli, E.M.
+\newblock {A multichannel EEG acquisition scheme based on single ended
+  amplifiers and digital DRL}.
+\newblock {\em IEEE Trans. Biomed. Circuits Syst.} {\bf
+  2012}, {\em 6},~614--618.
+
+\bibitem[Weeda \em{et~al.}(2012)Weeda, Grasman, Waldorp, van~de Laar, van~der
+  Molen, and Huizenga]{Weeda2012}
+Weeda, W.D.; Grasman, R.P.P.P.; Waldorp, L.J.; van~de Laar, M.C.; van~der
+  Molen, M.W.; Huizenga, H.M.
+\newblock {A fast and reliable method for simultaneous waveform, amplitude and
+  latency estimation of single-trial EEG/MEG data}.
+\newblock {\em PLoS ONE} {\bf 2012}, {\em 7}, doi:10.1371/journal.pone.0038292
+.
+
+\bibitem[Farzan \em{et~al.}(2017)Farzan, Atluri, Frehlich, Dhami, Kleffner,
+  Price, Lam, Frey, Milev, Ravindran, McAndrews, Wong, Blumberger, Daskalakis,
+  Vila-Rodriguez, Alonso, Brenner, Liotti, Dharsee, Arnott, Evans, Rotzinger,
+  and Kennedy]{Farzan2017}
+Farzan, F.; Atluri, S.; Frehlich, M.; Dhami, P.; Kleffner, K.; Price, R.; Lam,
+  R.W.; Frey, B.N.; Milev, R.; Ravindran, A.; et al.
+\newblock {Standardization of electroencephalography for multi-site,
+  multi-platform and multi-investigator studies: Insights from the canadian
+  biomarker integration network in depression}.
+\newblock {\em Sci.~Rep.} {\bf 2017}, {\em 7},~7473.
+
+\bibitem[Cole and Voytek(2017)]{Cole2017}
+Cole, S.R.; Voytek, B.
+\newblock {Brain Oscillations and the Importance of Waveform Shape}.
+\newblock {\em Trends Cogn. Sci.} {\bf 2017}, {\em 21},~137--149.
+
+\bibitem[Buzs{\'{a}}ki \em{et~al.}(2012)Buzs{\'{a}}ki, Anastassiou, and
+  Koch]{Buzsaki2012}
+Buzs{\'{a}}ki, G.; Anastassiou, C.A.; Koch, C.
+\newblock {The origin of extracellular fields and currents-EEG, ECoG, LFP and
+  spikes}.
+\newblock {\em Nat. Rev. Neurosci.} {\bf 2012}, {\em 13},~407--420,
+ % \href{http://xxx.lanl.gov/abs/NIHMS150003}{{\normalfont [NIHMS150003]}}.
+
+\bibitem[Giagante \em{et~al.}(2003)Giagante, Oddo, Silva, Consalvo, Centurion,
+  D'Alessio, Solis, Salgado, Seoane, Saidon, and Kochen]{Giagante2003}
+Giagante, B.; Oddo, S.; Silva, W.; Consalvo, D.; Centurion, E.; D'Alessio, L.;
+  Solis, P.; Salgado, P.; Seoane, E.; Saidon, P.; Kochen, S.
+\newblock {Clinical-electroencephalogram patterns at seizure onset in patients
+  with hippocampal sclerosis}.
+\newblock {\em Clin. Neurophysiol.} {\bf 2003}, {\em 114},~2286--2293.
+
+\bibitem[Rodenbeck \em{et~al.}(2006)Rodenbeck, Binder, Geisler, Danker-Hopfe,
+  Lund, Raschke, Wee{\ss}, and Schulz]{Rodenbeck2006}
+Rodenbeck, A.; Binder, R.; Geisler, P.; Danker-Hopfe, H.; Lund, R.; Raschke,
+  F.; Wee{\ss}, H.G.; Schulz, H.
+\newblock {A~review of sleep EEG patterns. Part I: A compilation of amended
+  rules for their visual recognition according to Rechtschaffen and Kales}.
+\newblock {\em Somnologie} {\bf 2006}, {\em 10},~159--175.
+
+\bibitem[Uchida \em{et~al.}(1999)Uchida, Feinberg, March, Atsumi, and
+  Maloney]{Uchida1999}
+Uchida, S.; Feinberg, I.; March, J.D.; Atsumi, Y.; Maloney, T.
+\newblock {A comparison of period amplitude analysis and FFT power spectral
+  analysis of all-night human sleep EEG}.
+\newblock {\em Physiol. Behav.} {\bf 1999}, {\em 67},~121--131.
+
+\bibitem[Britton \em{et~al.}(2016)Britton, Frey, Hopp, Korb, Koubeissi,
+  Lievens, Pestana-Knight, and St.~Louis]{EEGIntro}
+Britton, J.W.; Frey, L.C.; Hopp, J.L.; Korb, P.; Koubeissi, M.Z.; Lievens,
+  W.E.; Pestana-Knight, E.M.; St.~Louis, E.K.
+\newblock \emph{Electroencephalography (EEG): An Introductory Text and Atlas of
+  Normal and Abnormal Findings in Adults, Children, and Infants}; Chicago, USA; American Epilepsy Society; \hl{2016}.%please add publisher and location
+
+\bibitem[Luck(2005)]{Luck2005}
+Luck, S.J.
+\newblock {\em {An Introduction to the Event-Related Potential Technique}};
+ USA; MIT Press;  \hl{2005}; Volume~78,  p. 388,%please add publisher and location
+ % \href{http://xxx.lanl.gov/abs/9780262621960}{{\normalfont [9780262621960]}}.
+
+\bibitem[Tatum \em{et~al.}(2008)Tatum, Husain, Benbadis, and Kaplan]{Tatum2008}
+Tatum, W.; Husain, A.; Benbadis, S.; Kaplan, P.
+\newblock {\em {Handbook of EEG Interpretation}}; Demos Medical Publishing: New York, NY, USA, 2008.
+
+\bibitem[Cacioppo \em{et~al.}(2007)Cacioppo, Tassinary, and
+  Berntson]{Cacioppo2007}
+Cacioppo, J.; Tassinary, L.G.; Berntson, G.G.
+\newblock {\em {The Handbook of Psychophysiology}}; USA ;Cambridge University Press; \hl{2007}.%please add publisher and location
+
+\bibitem[Kappenman and Luck(2012)]{Kappenman2012}
+Kappenman, E.S.; Luck, S.J.
+\newblock {\em {The Oxford Handbook of Event-Related Potential Components}};
+ New York, USA; Oxford University Press; \hl{2012}.%please add publisher and location
+
+\bibitem[Ouyang \em{et~al.}(2017)Ouyang, Hildebrandt, Sommer, and
+  Zhou]{Ouyang2017}
+Ouyang, G.; Hildebrandt, A.; Sommer, W.; Zhou, C.
+\newblock {Exploiting the intra-subject latency variability from single-trial
+  event-related potentials in the P3 time range: A review and comparative
+  evaluation of methods}.
+\newblock {\em Neurosci. Biobehav. Rev.} {\bf 2017}, {\em
+  75},~1--21.
+
+\bibitem[Sanei and Chambers(2007)]{Sanei2007}
+Sanei, S.; Chambers, J.
+\newblock {\em {EEG Signal Processing}};  Chichester, UK; Wiley; \hl{2007}.%please add publisher and location
+
+\bibitem[Wulsin \em{et~al.}(2011)Wulsin, Gupta, Mani, Blanco, and
+  Litt]{Wulsin2011}
+Wulsin, D.F.; Gupta, J.R.; Mani, R.; Blanco, J.A.; Litt, B.
+\newblock {Modeling electroencephalography waveforms with semi-supervised deep
+  belief nets: Fast classification and anomaly measurement}.
+\newblock {\em J. Neural Eng.} {\bf 2011}, {\em 8},~036015,
+ % \href{http://xxx.lanl.gov/abs/036015}{{\normalfont [036015]}}.
+
+\bibitem[Hirsch and Richard(2010)]{Lawrence2010}
+Hirsch, L.J.; Richard, B.P.
+\newblock {\em {Atlas of EEG in Critical Care}}; Wiley-Blackwell: Hoboken, NJ, USA, 2010; p.
+  348.
+
+\bibitem[Subha \em{et~al.}(2010)Subha, Joseph, {Acharya U}, and Lim]{Subha2010}
+Subha, D.P.; Joseph, P.K.; {Acharya U}, R.; Lim, C.M.
+\newblock {EEG signal analysis: a survey.}
+\newblock {\em J. Med. Syst.} {\bf 2010}, {\em 34},~195--212.
+
+\bibitem[Mak \em{et~al.}(2012)Mak, McFarland, Vaughan, McCane, Tsui, Zeitlin,
+  Sellers, and Wolpaw]{Mak2012}
+Mak, J.N.; McFarland, D.J.; Vaughan, T.M.; McCane, L.M.; Tsui, P.Z.; Zeitlin,
+  D.J.; Sellers, E.W.; Wolpaw, J.R.
+\newblock {EEG correlates of P300-based brain-computer interface (BCI)
+  performance in people with amyotrophic lateral sclerosis}.
+\newblock {\em J. Neural Eng.} {\bf 2012}, {\em 9}, 026014.
+
+\bibitem[M{\"{u}}ller-Putz \em{et~al.}(2015)M{\"{u}}ller-Putz, Riedl, and
+  Wriessnegger]{MullerPutz2015}
+M{\"{u}}ller-Putz, G.R.; Riedl, R.; Wriessnegger, S.C.
+\newblock {Electroencephalography (EEG) as a research tool in the information
+  systems discipline: Foundations, measurement, and applications}.
+\newblock {\em Commun. Assoc. Inf. Syst.} {\bf
+  2015}, {\em 37},~911--948.
+
+\bibitem[Shah and Wusthoff(2015)]{Shah2015}
+Shah, N.A.; Wusthoff, C.J.
+\newblock {How to use: Amplitude-integrated EEG (aEEG)}.
+\newblock {\em Arch. Dis. Child. Educ. Pract.~Ed.} {\bf 2015}, {\em 100},~75--81.
+
+\bibitem[Moher \em{et~al.}(2009)Moher, Liberati, Tetzlaff, Altman, Altman,
+  Antes, Atkins, Barbour, Barrowman, Berlin, Clark, Clarke, Cook, D'Amico,
+  Deeks, Devereaux, Dickersin, Egger, Ernst, G{\o}tzsche, Grimshaw, Guyatt,
+  Higgins, Ioannidis, Kleijnen, Lang, Magrini, McNamee, Moja, Mulrow, Napoli,
+  Oxman, Pham, Rennie, Sampson, Schulz, Shekelle, Tovey, and
+  Tugwell]{Moher2009}
+Moher, D.; Liberati, A.; Tetzlaff, J.; Altman, D.G.; Altman, D.; Antes, G.;
+  Atkins, D.; Barbour, V.; Barrowman, N.; Berlin, J.A.; et al.
+\newblock {Preferred reporting items for systematic reviews and meta-analyses:
+  The PRISMA statement}.
+\newblock {\em PLoS Med.} {\bf 2009}, {\em 6},~e1000097.
+
+\bibitem[Allen and Mills(2004)]{allen2004signal}
+Allen, R.L.; Mills, D.
+\newblock {\em Signal Analysis: Time, Frequency, Scale, and Structure}; John
+  Wiley \& Sons: Hoboken, NJ, USA, 2004.
+
+\bibitem[{Wolpaw, Jonathan R}(2012)]{WolpawJonathanR2012}
+{Wolpaw, Jonathan R}, W.E.
+\newblock {\em {Brain-Computer Interfaces: Principles and Practice}}; Oxford
+  University Press: Oxford, UK, 2012; p. 400.
+
+\bibitem[Thakor(2009)]{Thakor2009}
+Thakor, N.
+\newblock {\em {Quantitative EEG Analysis Methods and Clinical Applications}};
+  Norwood, MA, USA; Artech House Series Engineering in Medicine and Biology; \hl{2009}; p. 440.%please add publisher and location
+
+\bibitem[Ja{\v{s}}kowski and Verleger(2000)]{Jaskowski2000}
+Ja{\v{s}}kowski, P.; Verleger, R.
+\newblock {An evaluation of methods for single-trial estimation of P3 latency}.
+\newblock {\em Psychophysiology} {\bf 2000}, {\em 37},~153--162.
+
+\bibitem[Zhang and Luo(2011)]{Zhang2011}
+Zhang, D.; Luo, Y.
+\newblock {The P1 latency of single-trial ERPs estimated by two peak-picking
+  strategies}.
+\newblock {In~Proceedings of the 2011 4th International Conference on Biomedical
+  Engineering and Informatics (BMEI)}, Shanghai, China, 15--17 October 2011; Volume 2,~pp. 882--886.
+
+\bibitem[Alvarado-Gonz{\'{a}}lez \em{et~al.}(2016)Alvarado-Gonz{\'{a}}lez,
+  Gardu{\~{n}}o, Bribiesca, Y{\'{a}}{\~{n}}ez-Su{\'{a}}rez, and
+  Medina-Ba{\~{n}}uelos]{Alvarado-Gonzalez2016}
+Alvarado-Gonz{\'{a}}lez, M.; Gardu{\~{n}}o, E.; Bribiesca, E.;
+  Y{\'{a}}{\~{n}}ez-Su{\'{a}}rez, O.; Medina-Ba{\~{n}}uelos, V.
+\newblock {P300 Detection Based on EEG Shape Features}.
+\newblock {\em Comput. Math. Methods Med.} {\bf 2016},
+  {\em 2016},~1--14.
+
+\bibitem[Farwell and Donchin(1988)]{Farwell1988}
+Farwell, L.A.; Donchin, E.
+\newblock {Talking off the top of your head: toward a mental prosthesis
+  utilizing event-related brain potentials}.
+\newblock {\em Electroencephalogr. Clin. Neurophysiol.} {\bf 1988},
+  {\em 70},~510--523.
+
+\bibitem[F.F.(1976)]{Klein1976}
+Klein, F.F.
+\newblock {A waveform analyzer applied to the human EEG}.
+\newblock {\em IEEE Trans. Biomed. Eng.} {\bf 1976}, {\em
+  23},~246--252.
+
+\bibitem[Yamaguchi \em{et~al.}(2009)Yamaguchi, Fujio, Inoue, and
+  Pfurtscheller]{Yamaguchi2009}
+Yamaguchi, T.; Fujio, M.; Inoue, K.; Pfurtscheller, G.
+\newblock {Design Method of Morphological Structural Function for Pattern
+  Recognition of EEG Signals During Motor Imagery and Cognition}.
+\newblock  In Proceedings of the 2009 Fourth International Conference on Innovative Computing, Information and Control (ICICIC), Kaohsiung, Taiwan, 7--9 December 2009; pp. 1558--1561.
+
+\bibitem[Uchida \em{et~al.}(1996)Uchida, Matsuura, Ogata, Yamamoto, and
+  Aikawa]{Uchida1996}
+Uchida, S.; Matsuura, M.; Ogata, S.; Yamamoto, T.; Aikawa, N.
+\newblock {Computerization of Fujimori's method of waveform recognition a
+  review and methodological considerations for its application to all-night
+  sleep EEG}.
+\newblock {\em J. Neurosci. Methods} {\bf 1996}, {\em 64},~1--12.
+
+\bibitem[Vincent and Deville(2010)]{Vincent2010}
+Vincent, E.; Deville, Y.
+\newblock {\em {Handbook of Blind Source Separation---Independent Component
+  Analysis and Applications}}; Elsevier: Amsterdam, The Netherlands, 2010; p. 831.
+
+\bibitem[Mallat and Zhang(1993)]{Mallat1993}
+Mallat, S.G.; Zhang, Z.
+\newblock {Matching Pursuits With Time-Frequency Dictionaries}.
+\newblock {\em IEEE Trans. Signal Process.} {\bf 1993}, {\em
+  41},~3397--3415.
+
+\bibitem[{Chandran KS} \em{et~al.}(2016){Chandran KS}, Mishra, Shirhatti, and
+  Ray]{ChandranKS2016}
+{Chandran KS}, S.; Mishra, A.; Shirhatti, V.; Ray, S.
+\newblock {Comparison of Matching Pursuit Algorithm with Other Signal
+  Processing Techniques for Computation of the Time-Frequency Power Spectrum of
+  Brain Signals}.
+\newblock {\em J.~Neurosci.} {\bf 2016}, {\em 36},~3399--3408.
+
+\bibitem[Cohen(2014)]{Cohen2014}
+Cohen, M.X.
+\newblock {\em {Analyzing Neural Time Series Data: Theory and Practice}};
+  Cambridge MA, USA; MIT Press; \hl{2014}.%%%please add publisher and location
+
+\bibitem[Vařeka(2012)]{Vareka2012}
+Vařeka, L.
+\newblock {Matching pursuit for p300-based brain-computer interfaces}.
+\newblock {In Proceedings of the 2012 35th International Conference on Telecommunications and Signal Processing}, Prague, Czech Republic, 3--4 July 2012; pp. 513--516.
+
+\bibitem[Bandt and Pompe(2002)]{Bandt2002}
+Bandt, C.; Pompe, B.
+\newblock {Permutation Entropy: A Natural Complexity Measure for Time Series}.
+\newblock {\em \mbox{Phys. Rev. Lett.}} {\bf 2002}, {\em 88},~174102.
+
+\bibitem[Nicolaou and Georgiou(2010)]{Nicolaou2010}
+Nicolaou, N.; Georgiou, J.
+\newblock {Permutation entropy: A new feature for brain-computer interfaces}.
+\newblock  In~Proceedings of the 2010 IEEE Biomedical Circuits and Systems Conference, BioCAS 2010, Paphos, Cyprus, 3--5 November 2010; pp. 49--52.
+
+\bibitem[Unakafova and Keller(2013)]{Unakafova2013}
+Unakafova, V.; Keller, K.
+\newblock {Efficiently Measuring Complexity on the Basis of Real-World Data}.
+\newblock {\em Entropy} {\bf 2013}, {\em 15},~4392--4415.
+
+\bibitem[Berger \em{et~al.}(2017)Berger, Schneider, Kochs, and
+  Jordan]{Berger2017}
+Berger, S.; Schneider, G.; Kochs, E.F.; Jordan, D.
+\newblock {Permutation entropy: Too complex a measure for EEG time series?}
+\newblock {\em Entropy} {\bf 2017}, {\em 19},~692.
+
+\bibitem[Lowe(2004)]{Lowe2004}
+Lowe, G.
+\newblock {SIFT - The Scale Invariant Feature Transform}.
+\newblock {\em Int. J.} {\bf 2004}, {\em 2},~91--110.
+
+\bibitem[Edelman \em{et~al.}(1997)Edelman, Intrator, and Poggio]{Edelman1997}   %NOTE: This paper is unplublished.  We added the information to cite it as it is cited in many other works.
+Edelman, S.; Intrator, N.; Poggio, T.
+\newblock {Complex cells and object recognition}.
+\newblock {\em Unpublished manuscript} \hl{1997}, pp. 1--12.%%%%%please add publisher and location
+
+\bibitem[Ramele \em{et~al.}(2016)Ramele, Villar, and Santos]{Ramele2016}
+Ramele, R.; Villar, A.J.; Santos, J.M.
+\newblock {BCI classification based on signal plots and SIFT descriptors}.
+\newblock  In~Proceedings of the 4th International Winter Conference on Brain-Computer Interface (BCI), Yongpyong, South, 22--24 February 2016; pp. 1--4.
+
+\bibitem[Zhang \em{et~al.}(2013)Zhang, Xu, Guo, Zhang, Li, and Yao]{Zhang2013}
+Zhang, R.; Xu, P.; Guo, L.; Zhang, Y.; Li, P.; Yao, D.
+\newblock {Z-Score Linear Discriminant Analysis for EEG Based Brain-Computer
+  Interfaces}.
+\newblock {\em PLoS ONE} {\bf 2013}, {\em 8},~e74433.
+
+\bibitem[Vedaldi and Fulkerson(2010)]{Vedaldi2010}
+Vedaldi, A.; Fulkerson, B.
+\newblock {VLFeat - An open and portable library of computer vision
+  algorithms}.
+\newblock {\em Design} {\bf 2010}, {\em 3},~1--4.
+
+\bibitem[{Quian Quiroga} and Garcia(2003)]{QuianQuiroga2003}
+{Quian Quiroga}, R.; Garcia, H.
+\newblock {Single-trial event-related potentials with wavelet denoising}.
+\newblock {\em Clin.~Neurophysiol.} {\bf 2003}, {\em 114},~376--390.
+
+\bibitem[{Da Pelo} \em{et~al.}(2018){Da Pelo}, {De Tommaso}, Monaco,
+  Stramaglia, Bellotti, and Tangaro]{DaPelo2018}
+{Da Pelo}, P.; {De Tommaso}, M.; Monaco, A.; Stramaglia, S.; Bellotti, R.;
+  Tangaro, S.
+\newblock {Trial latencies estimation of event-related potentials in EEG by
+  means of genetic algorithms}.
+\newblock {\em J. Neural Eng.} {\bf 2018}, {\em 15},~026016.
+
+\bibitem[Harris(1978)]{Harris1978}
+Harris, F.J.
+\newblock {On the Use of Windows for Harmonic Analysis with the Discrete
+  Fourier Transform}.
+\newblock {\em Proc. IEEE} {\bf 1978}, {\em 66},~51--83.
+
+\bibitem[Krusienski \em{et~al.}(2006)Krusienski, Sellers, Cabestaing, Bayoudh,
+  McFarland, Vaughan, and Wolpaw]{Krusienski2006}
+Krusienski, D.J.; Sellers, E.W.; Cabestaing, F.; Bayoudh, S.; McFarland, D.J.;
+  Vaughan, T.M.; Wolpaw, J.R.
+\newblock {A~comparison of classification techniques for the P300 Speller}.
+\newblock {\em J. Neural Eng.} {\bf 2006}, {\em 3},~299--305.
+
+\bibitem[Kaper \em{et~al.}(2004)Kaper, Meinicke, Grossekathoefer, Lingner, and
+  Ritter]{Kaper2004}
+Kaper, M.; Meinicke, P.; Grossekathoefer, U.; Lingner, T.; Ritter, H.
+\newblock {BCI competition 2003 - Data set IIb: Support vector machines for the
+  P300 speller paradigm}.
+\newblock {\em IEEE Trans. Biomed. Eng.} {\bf 2004}, {\em
+  51},~1073--1076.
+
+\bibitem[Blankertz(2002)]{Blankertz2002}    %% This is unpublished document.  We copied how other works cited this paper.  There are not Publisher nor location information.  Remove if necessary.
+Blankertz, B.
+\newblock Documentation second wadsworth BCI dataset (P300 evoked potentials)
+  data acquired using BCI2000 P300 Speller Paradigm.
+\newblock In {\em BCI Classification Contest November}; {\hl{2002}}.%%%please add publisher and location
+
+\bibitem[Riccio \em{et~al.}(2013)Riccio, Simione, Schettini, Pizzimenti,
+  Inghilleri, Belardinelli, Mattia, and Cincotti]{Riccio2013}
+Riccio, A.; Simione, L.; Schettini, F.; Pizzimenti, A.; Inghilleri, M.;
+  Belardinelli, M.O.; Mattia, D.; Cincotti,~F.
+\newblock {Attention and P300-based BCI performance in people with amyotrophic
+  lateral sclerosis}.
+\newblock {\em \mbox{Front. Hum. Neurosci.}} {\bf 2013}, {\em 7},~732.
+
+\bibitem[Brunner \em{et~al.}(2014)Brunner, Blankertz, Cincotti, K{\"{u}}bler,
+  Mattia, Miralles, Nijholt, and Otal]{Brunner2014}
+Brunner, C.; Blankertz, B.; Cincotti, F.; K{\"{u}}bler, A.; Mattia, D.;
+  Miralles, F.; Nijholt, A.; Otal, B.
+\newblock {BNCI Horizon 2020---Towards a Roadmap for Brain/Neural Computer
+  Interaction}.
+\newblock {\em Lect. Notes Comput. Sci.} {\bf 2014}, {\em
+  8513},~475--486.
+
+\bibitem[Renard \em{et~al.}(2010)Renard, Lotte, Gibert, Congedo, Maby,
+  Delannoy, Bertrand, and L{\'{e}}cuyer]{Renard2010}
+Renard, Y.; Lotte, F.; Gibert, G.; Congedo, M.; Maby, E.; Delannoy, V.;
+  Bertrand, O.; L{\'{e}}cuyer, A.
+\newblock {OpenViBE: An Open-Source Software Platform to Design, Test, and Use
+  Brain Computer Interfaces in Real and Virtual Environments}.
+\newblock {\em Presence Teleoper. Virtual Environ.} {\bf 2010},
+  {\em 19},~35--53.
+
+\bibitem[Gramfort \em{et~al.}(2013)Gramfort, Luessi, Larson, Engemann,
+  Strohmeier, Brodbeck, Goj, Jas, Brooks, Parkkonen, and
+  H{\"{a}}m{\"{a}}l{\"{a}}inen]{Gramfort2013}
+Gramfort, A.; Luessi, M.; Larson, E.; Engemann, D.A.; Strohmeier, D.; Brodbeck,
+  C.; Goj, R.; Jas, M.; Brooks, T.; Parkkonen, L.;
+  H{\"{a}}m{\"{a}}l{\"{a}}inen, M.
+\newblock {MEG and EEG data analysis with MNE-Python}.
+\newblock {\em Front. Neurosci.} {\bf 2013}, {\em 7},~267.
+
+\bibitem[Boiman \em{et~al.}(2008)Boiman, Shechtman, and Irani]{Boiman2008}
+Boiman, O.; Shechtman, E.; Irani, M.
+\newblock {In defense of nearest-neighbor based image classification}.
+\newblock {In~Proceedings of the 26th IEEE Conference on Computer Vision and Pattern Recognition (CVPR)}, Anchorage, AK, USA, 23--28 June 2008.
+
+\bibitem[K{\"{a}}thner \em{et~al.}(2017)K{\"{a}}thner, Halder,
+  Hinterm{\"{u}}ller, Espinosa, Guger, Miralles, Vargiu, Dauwalder,
+  Rafael-Palou, Sol{\`{a}}, Daly, Armstrong, Martin, and
+  K{\"{u}}bler]{Kathner2017}
+K{\"{a}}thner, I.; Halder, S.; Hinterm{\"{u}}ller, C.; Espinosa, A.; Guger, C.;
+  Miralles, F.; Vargiu, E.; Dauwalder, S.; Rafael-Palou, X.; Sol{\`{a}}, M.;
+  Daly, J.M.; Armstrong, E.; Martin, S.; K{\"{u}}bler, A.
+\newblock {A multifunctional brain-computer interface intended for home use: An
+  evaluation with healthy participants and potential end users with dry and
+  gel-based electrodes}.
+\newblock {\em Front. Neurosci.} {\bf 2017}, {\em 11}, 286.
+
+\bibitem[Gribonval \em{et~al.}(2008)Gribonval, Rauhut, Schnass, and
+  Vandergheynst]{Gribonval2008}
+Gribonval, R.; Rauhut, H.; Schnass, K.; Vandergheynst, P.
+\newblock {Atoms of all channels, unite! Average case analysis of multi-channel
+  sparse recovery using greedy algorithms}.
+\newblock {\em J. Fourier Anal. Appl.} {\bf 2008}, {\em
+  14},~655--687.
+
+\bibitem[Chavarriaga \em{et~al.}(2017)Chavarriaga, Fried-Oken, Kleih, Lotte,
+  and Scherer]{Chavarriaga2017}
+Chavarriaga, R.; Fried-Oken, M.; Kleih, S.; Lotte, F.; Scherer, R.
+\newblock {Heading for new shores! Overcoming pitfalls in BCI design}.
+\newblock {\em Brain-Comput. Interfaces} {\bf 2017}, {\em 4},~60--73.
+
+\bibitem[Tjepkema-Cloostermans \em{et~al.}(2018)Tjepkema-Cloostermans,
+  de~Carvalho, and van Putten]{Tjepkema-Cloostermans2018}
+Tjepkema-Cloostermans, M.C.; de~Carvalho, R.C.; van Putten, M.J.
+\newblock {Deep learning for detection of focal epileptiform discharges from
+  scalp EEG recordings}.
+\newblock {\em Clin. Neurophysiol.} {\bf 2018}, {\em 129},~2191--2196.
+
+\bibitem[J~Bragg(2018)]{j2018challenge}
+Bragg, M.J.
+\newblock The Challenge of Crafting Intelligible Intelligence.
+\newblock  In Proceedings of the ACM Symposium on User Interface Software and Technology UIST 2018, Berlin, Germany, 14--17 October 2018.
+
+\bibitem[Criminisi and Shotton(2013)]{Criminisi2013}
+Criminisi, A.; Shotton, J.
+\newblock {\em Decision Forests for Computer Vision and Medical Image
+  Analysis}; Springer Science \& Business Media: Berlin, Germany, 2013.
+
+\bibitem[Gu and Jin(2012)]{Gu2012}
+Gu, S.; Jin, Y.
+\newblock {Heterogeneous classifier ensembles for EEG-based motor imaginary
+  detection}.
+\newblock  In Proceedings of the 2012 12th UK Workshop on Computational Intelligence (UKCI),  Edinburgh, UK, 5--7 September 2012; pp. 1--8.
+
+\bibitem[Skoog \em{et~al.}(2000)Skoog, West, Holler, and Crouch]{Skoog2000}
+Skoog, D.A.; West, D.M.; Holler, F.J.; Crouch, S.R.
+\newblock\emph{ Analytical Chemistry: An Introduction};
+\newblock {Saunders College Publishing: Philadelphia, PA, USA,} {2000}.
+
+\bibitem[Owens \em{et~al.}(1984)Owens, Zandt, and Taylor]{Owens1984}
+Owens, T.J.; Zandt, G.; Taylor, S.R.
+\newblock Seismic evidence for an ancient rift beneath the Cumberland Plateau,
+  Tennessee: A detailed analysis of broadband teleseismic P waveforms.
+\newblock {\em J. Geophys. Res. Solid Earth} {\bf 1984}, {\em
+  89},~7783--7795.
+
+\bibitem[Stockman \em{et~al.}(1976)Stockman, Kanal, and Kyle]{Stockman1976}
+Stockman, G.; Kanal, L.; Kyle, M.
+\newblock Structural pattern recognition of carotid pulse waves using a general
+  waveform parsing system.
+\newblock {\em Commun. ACM} {\bf 1976}, {\em 19},~688--695.
+
+% Missing references in the last version. Added from here !
+
+\bibitem[Kundu and Ari(2018)]{Kundu2018}
+Kundu, S.; Ari, S.
+\newblock P300 Detection with Brain–Computer Interface Application Using PCA
+  and Ensemble of Weighted SVMs.
+\newblock {\em IETE Journal of Research} {\bf 2018}, {\em 64},~406--414
+
+\bibitem[Tzimourta \em{et~al.}(2018)Tzimourta, Tsoulos, Bilero, Tzallas,
+  Tsipouras, and Giannakeas]{Tzimourta2018}
+Tzimourta, K.D.; Tsoulos, I.; Bilero, T.; Tzallas, A.T.; Tsipouras, M.G.;
+  Giannakeas, N.
+\newblock Direct Assessment of Alcohol Consumption in Mental State Using Brain
+  Computer Interfaces and Grammatical Evolution.
+\newblock {\em Inventions} {\bf 2018}, {\em 3}.
+
+\bibitem[{Kevric Jasmin} and Subasi(2016)]{KevricJasmin2016}
+{Kevric Jasmin}.; Subasi, A.
+\newblock {The Impact of Mspca Signal De-Noising In Real- Time Wireless Brain
+  Computer Interface System}.
+\newblock {\em Southeast Europe Journal of Soft Computing} {\bf 2016}, {\em
+  4},~1 -- 12.
+
+\bibitem[Cruz \em{et~al.}(2018)Cruz, Pires, and Nunes]{Cruz2018}
+Cruz, A.; Pires, G.; Nunes, U.J.
+\newblock Double ErrP Detection for Automatic Error Correction in an ERP-Based
+  BCI Speller.
+\newblock {\em IEEE Transactions on Neural Systems and Rehabilitation
+  Engineering} {\bf 2018}, {\em 26},~26--36.
+
+\bibitem[Boostani \em{et~al.}(2017)Boostani, Karimzadeh, and
+  Nami]{Boostani2017}
+Boostani, R.; Karimzadeh, F.; Nami, M.
+\newblock A comparative review on sleep stage classification methods in
+  patients and healthy individuals.
+\newblock {\em Computer Methods and Programs in Biomedicine} {\bf 2017}, {\em
+  140},~77 -- 91.
+
+\bibitem[Dimitriadis \em{et~al.}(2018)Dimitriadis, Salis, and
+  Linden]{Dimitriadis2018}
+Dimitriadis, S.I.; Salis, C.; Linden, D.
+\newblock A novel, fast and efficient single-sensor automatic sleep-stage
+  classification based on complementary cross-frequency coupling estimates.
+\newblock {\em Clinical Neurophysiology} {\bf 2018}, {\em 129},~815 -- 828.
+
+\bibitem[Keller \em{et~al.}(2017)Keller, Mangold, Stolz, and
+  Werner]{Keller2017}
+Keller, K.; Mangold, T.; Stolz, I.; Werner, J.
+\newblock Permutation Entropy: New Ideas and Challenges.
+\newblock {\em Entropy} {\bf 2017}, {\em 19}.
+
+\bibitem[Rakotomamonjy and Guigue(2008)]{Rakotomamonjy2008}
+Rakotomamonjy, A.; Guigue, V.
+\newblock {BCI Competition III: Dataset II- Ensemble of SVMs for BCI P300
+  Speller}.
+\newblock {\em IEEE Transactions on Biomedical Engineering} {\bf 2008}, {\em
+  55},~1147--1154.
+
+\bibitem[Neuper \em{et~al.}(2003)Neuper, Müller, Kübler, Birbaumer, and
+  Pfurtscheller]{Neuper2003}
+Neuper, C.; Müller, G.; Kübler, A.; Birbaumer, N.; Pfurtscheller, G.
+\newblock Clinical application of an EEG-based brain–computer interface: a
+  case study in a patient with severe motor impairment.
+\newblock {\em Clinical Neurophysiology} {\bf 2003}, {\em 114},~399 -- 409.
+
+\bibitem[Rao(2013)]{Rao2013}
+Rao, R.P.N.
+\newblock {\em {Brain-computer interfacing : an introduction}}; New York, USA; Cambridge
+  University Press,  2013; p. 319.
+
+\bibitem[Clerc \em{et~al.}(2016)Clerc, Bougrain, and Lotte]{Clerc2016}
+Clerc, M.; Bougrain, L.; Lotte, F.
+\newblock {\em {Brain-computer interfaces, Technology and applications
+  2(Cognitive Science)}}; London, UK; ISTE Ltd. and Wiley,  2016; p. 324.
+
+\end{thebibliography}
 
